### 2015-03-30 docs/parasite_mortatlity_manuscript.tex
Added convergence equivalent simulations

% Document
\documentclass[12pt, a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{authblk}
\usepackage{lipsum}

% Fig. and table formating
\usepackage{epsfig}
\usepackage{tabu}
\usepackage{rotating}
\usepackage{pbox}
\usepackage{framed, multicol}
\usepackage[framemethod=TikZ]{mdframed}
\usepackage{float}
\usepackage[left=1.25 in, right=1.25 in, top=1.25 in, bottom=1.25 in]{geometry}

\usepackage{caption}

% Fonts - Mathtime
%\usepackage{txfonts}
\usepackage{amsmath} % Add amssymb if not using Mathtime
\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

% Text
\setlength{\parindent}{0.5in}
\frenchspacing  \tolerance = 800  \hyphenpenalty = 800

\usepackage{lineno} % Line numbers
\def\linenumberfont{\normalfont\footnotesize\ttfamily}
\setlength\linenumbersep{0.2 in}

\usepackage{setspace}

% Format section and subsection headers
\makeatletter
\renewcommand{\section}{\@startsection
{section}%                   % the name
{1}%                         % the level
{0mm}%                       % the indent
{-\baselineskip}%            % the before skip
{0.5\baselineskip}%          % the after skip
{\normalfont\bf\large}} % the style

\renewcommand{\subsection}{\@startsection
{subsection}%                   % the name
{2}%                         % the level
{0mm}%                       % the indent
{-\baselineskip}%            % the before skip
{0.5\baselineskip}%          % the after skip
{\normalfont\bf}} % the style
\makeatother

% Other
\usepackage{graphicx}
\usepackage[singlelinecheck=false,font=small,labelfont=bf]{caption}
\usepackage[justification=centering]{caption}

% Bibliography
\usepackage[numbers, compress]{natbib} % Bibliography - APA
\bibpunct{(}{)}{;}{a}{}{,}

% Format the Bibliography appropriately
% increase \bibhang to take care of the numbers
\setlength{\bibhang}{0pc}
\makeatletter
%patch \@lbibitem to print the current number before the authors
\patchcmd{\@lbibitem}
  {]}
  {] \theNAT@ctr. \newline }
  {}{}
\makeatother

%%%%%% FRONT MATTER %%%%%%%%%

\begin{document}

\doublespacing

\section*{Methods}

\subsection*{Overview of PIHM approaches}

We propose and two alternative approaches for estimating LD50 and PD from
parasite intensity data.  Our two approaches, as well as the Adjei Method,
follow the same general steps.  First, all approaches assume that the
distribution of parasites across hosts in the pre-mortality population follows
a negative binomial distribution and, using the Crofton Method, each approach
starts by estimating the total number hosts  ($N_p$), the mean number of
parasites per host ($\mu_p$), and the aggregation of the parasites across hosts
before PIHM ($k_p$) (see Appendix X for a description of the Crofton Method).
Second, each method assumes that the probability of host survival with $x$ parasites is given by the logistic host-survival function
\begin{equation}
    h(x | a, b) = h_x = \dfrac{e^{a + b \log(x)}}{1 + e^{a + b \log(x)}}
    \label{eq:logistic}
\end{equation}
Using equation \ref{eq:logistic}, each method uses either a likelihood or chi-squared approach to estimate $a$ and $b$.  Finally, from the estimated parameters $a$ and $b$,
$LD_{50}$ and $PD$ can be calculated by the equations \citep{Adjei1986}

\begin{equation}
    LD_{50} = \exp(\frac{a}{b})
    \label{eq:ld50}
\end{equation}

and

\begin{equation}
    PD = 1 -  \dfrac{\sum_{x=0}^\infty h_x * f_x}{\sum_{x=0}^\infty f_x}
    \label{eq:pd}
\end{equation}
where $f_x$ is the probability mass function for the pre-mortality distribution of parasites per hosts which follows a negative binomial distribution with mean $\mu_p$ and aggregation parameter $k_p$. For
each method, the differences in predictions of
$LD_{50}$ and $PD$ are a result of how a given method each calculates $a$ and $b$.

\subsection*{Estimating $a$ and $b$ with the Adjei Method}

To estimate $a$ and $b$, the Adjei Method proceeds as follows \citep{Adjei1986}

\singlespacing
\begin{enumerate}
    \item Estimate $N_p$, $\mu_p$, and $k_p$ using the Crofton Method
    \item Given $N_p$, $\mu_p$, and $k_p$, estimate the expected number of hosts in category $i$ via equation $N_p f_i$.  $i$ specifies either a given parasite intensity or some range of parasites intensities (i.e. number of hosts with 100 parasites or number of hosts with 100-150 parasites) and there are $m$ categories.
    \item Calculate the observed number of hosts in each category $i$.  Assume that the observed number of hosts in each category $i$ is binomial distributed with the total number of ``trials'' equal to the expected number of hosts in $i$.
    \item If observed number of hosts in category $i$ is greater than the expected number of hosts in $i$, let observed in $i$ equal to expected in $i$.
    \item Run a generalized linear model in with a binomial random component and a logistic link to estimate $a$ and $b$.
\end{enumerate}

\doublespacing
See \cite{Adjei1986} and Appendix X for a more thorough description of this approach and some supporting examples.

\subsection*{Estimating $a$ and $b$ with chi-squared approach}

The first alternative approach that we propose relies on estimating $a$ and $b$ by minimizing the $\chi^2$ statistic.  The approach proceeds as follows

\singlespacing
\begin{enumerate}
    \item Estimate $N_p$, $\mu_p$, and $k_p$ using the Crofton Method.
    \item Specify $m$ categories of parasite intensities where each category $i$ either contains a single parasite intensity or a range of parasite intensities.   Calculate the expected number of hosts in each category $i $ by $N_p$ times the probability that a host is category $i$ and is alive.  This can be written as $N_p * f_i * h_i$ where the appropriate summation is taken if $i$ is a range of parasite intensities.
    \item Calculate the observed number of hosts in each category $i$.
    \item Calculate the $\chi^2$ statistic for the observed and expected $n$ categories using the standard notation $X^2 = \sum_{i=1}^m \frac{(observed_i - expected_i)^2}{expected_i}$
    \item Find $a$ and $b$ that minimize this $X^2$ statistic.
\end{enumerate}

\doublespacing

This approach does not require any alteration of the observed data as in the Adjei approach and uses the same iterative minimization technique as the Crofton Method \citep{Lester1977}.

\subsection*{Estimating $a$ and $b$ with the likelihood approach}

The second alternative method relies on maximizing the likelihood of the observed data and proceeds as follows.

\singlespacing
\begin{enumerate}
    \item Estimate $N_p$, $\mu_p$, and $k_p$ using the Crofton Method.
    \item Specify the probability distribution that a host has $x$ parasites and is alive as

    \begin{equation}
        p(x; a, b, \mu_p, k_p) = \phi h(x; a, b) f(x; \mu_p, k_p)
    \end{equation}

    where $\phi$ is a normalizing constant. % equal to $(\sum_{x=0}^\infty h(x; a, b) * f(x; \mu_p, k_p))^{-1}$.

    \item Assuming hosts are independent, the likelihood of a datasets with $n$ hosts can be written as $L(a, b | x, \mu_p, k_p) = \Pi_{i=0}^n p(x_i ; a, b, \mu_p, k_p)$.

    \item Use standard optimization techniques to estimate $a$ and $b$

\end{enumerate}

\doublespacing
The likelihood approach is advantageous because standard statistical techniques can then be used to assess whether parasite-induced host mortality is playing a significant role in a population [reword].

\subsection*{Dataset simulation and comparison of PIHM approaches}

To compare the ability of these methods to recover $LD_{50}$ and $PD$, we randomly generated data using the
following procedure.  First, we drew $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$.  Second, we chose values of $a$ and $b$ and calculated the probability of survival
for all $N_p$ hosts using equation \ref{eq:logistic}.  Third, we drew $N_p$ random numbers from a uniform distribution
between 0 and 1 and if host survival probability was less than this random
number, the host experienced parasite-induced mortality.  The surviving
hosts comprised the dataset that would be obtained in the field.

% For a simulated datasets, we estimated the parameters $a$ and $b$ using each of the three approached outlines above. For combination of $a$ and $b$ we then estimated $LD_{50}$ and $PD$. To
% determine the efficacy of each method, we estimated the bias and precision
% of the resulting $LD_{50}$ and $PD$ estimates \citep{Walther2005}.  To compare
% the bias and precision across parameter combinations we used the
% standardized bias and the standardized precision \citep{Walther2005}.

Using the simulated datasets, we devised two scenarios to test the above approaches. In the first scenario, we assumed
that the values of $N_p$, $\mu_p$, and $k_p$ were known and tested the ability of both
methods to recover the true values of $a$ and $b$ over increasing values of
$N_p$.  While this first scenario is unrealistic because the parameters $N_p$,
$\mu_p$, and $k_p$ are always unknown, we implemented this scenario as a baseline to
establish the efficacy of the methods independent of the Crofton Method used to estimate $N_p$, $\mu_p$ and $k_p$.   If a
method could not return unbiased [word choice] estimates of $LD_{50}$ and/or $PD$ under these
idealized conditions, we took this as evidence for unreliability of this method. The second scenario
proceeded exactly as the first, except that we estimated $N_p$, $mu_p$, and
$k_p$ using the Crofton Method.

For each scenario, we used two different $LD_{50}$
values: $\exp(2) = 7.39$ parasites with $\mu_p = 10$ and $\exp(3) = 20.08$ parasites with  $\mu_p = 30$. For each $LD_{50}$ value we choose three combinations of $a$ and $b$ that
represented three different patterns of how host survival decreased with
parasite intensity:  gradual, moderate and a sharp decrease in host survival
probability (Figure X).  Finally we examined three levels of parasite
aggregation, $k_p = 0.1, 0.5, 1$, which are realistic values of parasite aggregation in
natural populations \citep{Shaw1998}.  For each of these parameter
combinations we simulated 150 datasets, estimated $a$, $b$, $LD_{50}$, and $PD$ and calculated the standardized bias and
precision \citep{Walther2005} for these estimates over varying pre-mortality host population sizes  $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]. $N_p$ is not technically the sample size on which the methods are being
tested because parasite-induced mortality reduces $N_p$ for each simulated
dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.

% However, to facilitate comparison across parameter combinations, we
% used $N_p$ as a proxy for sample size.  Finally, for low values of $N_p$, there
% were some simulated datasets where the Chi-squared Method and the Likelihood
% Method failed to converge.  These trials were discarded \dots [or should we save them
% to see what is going on?]. The highest failure percentage we observed was X for
% the X method with $N_p$ = 300 [need to look into this further/use a more robust
% convergence method for the likelihood approach].

\section*{Results}

\subsection*{Scenario I: Known pre-mortality parameters}

\subsection*{Scenario II: Unknown pre-mortality parameters}

\section*{Discussion}

\singlespacing
\bibliographystyle{/Users/mqwilber/Dropbox/Documents/Bibformats/ecology_letters.bst}
\bibliography{/Users/mqwilber/Dropbox/Documents/Bibfiles/Projects_and_Permits-parasite_host_mortality}

\end{document}

### 2015-03-30 code/crofton_method.py

@@ -5,6 +5,7 @@ import scipy.optimize as opt
 import pandas as pd
 import statsmodels.api as sm
 import pymc as pm
+import matplotlib.pyplot as plt
 
 
 """
@@ -315,42 +316,6 @@ def fit_glm(all_data):
     return res.params
 
 
-def infected_lost(params, all_data, full_para=True):
-    """
-    Use formula from Adjei et al to calculate percent infected lost.
-
-    Parameters
-    ----------
-    params : array-like
-        Length 2. First value is a parameter of GLM and second parameter is
-        b parameter of the GLM
-    all_data : dataframe
-        DataFrame returned from adjei_fitting_method.
-        Columns 'emp', 'pred', 'para'
-    full_para : bool
-        If the column para includes the zero class
-    """
-
-    a = params[0]
-    b = params[1]
-    surv_prob = lambda x, a, b: np.exp(a + b * np.log(x)) / \
-                                    (1 + np.exp(a + b * np.log(x)))
-    pred = np.array(all_data['pred'])
-    para = np.array(all_data['para'])
-
-    # Proportion infected lost
-    if full_para:
-
-        surv_vals = surv_prob(para[1:], a, b)
-        inf_lost = 1 - np.sum(surv_vals * pred[1:]) / np.sum(pred[1:])
-
-        numer = pred[0] + np.sum(surv_vals * pred[1:])
-        denom = np.sum(pred)
-        tot_lost = 1 - numer / denom
-        return inf_lost, tot_lost
-
-    else:
-        return (1 - np.sum(surv_prob(para, a, b) * pred) / np.sum(pred), None)
 
 
 def surv_prob(x, a, b):
@@ -360,12 +325,9 @@ def surv_prob(x, a, b):
     x = np.atleast_1d(x)
 
     probs = np.empty(len(x))
-
-    for i, val in enumerate(x):
-        if val == 0:
-            probs[i] = 1
-        else:
-            probs[i] = np.exp(a + b * np.log(val)) / (1 + np.exp(a + b * np.log(val)))
+    probs[x == 0] = 1
+    ind = x != 0
+    probs[ind] = np.exp(a + b * np.log(x[ind])) / (1 + np.exp(a + b * np.log(x[ind])))
 
     return probs
 
@@ -423,6 +385,23 @@ def scaled_precision(data):
 
     return 100 * np.std(data, ddof=1) / np.mean(data)
 
[…]

### 2015-04-02 docs/parasite_mortatlity_manuscript.tex
Reran analyses with updated truncation a d made LD50 plots

@@
 \doublespacing
+\linenumbers
 
@@
 
-We propose and two alternative approaches for estimating LD50 and PD from
+We propose two alternative approaches for estimating LD50 and PD from
 parasite intensity data.  Our two approaches, as well as the Adjei Method,
@@
 before PIHM ($k_p$) (see Appendix X for a description of the Crofton Method).
+We discuss in the following sections that our alternative approaches do not have to start with the Crofton Method, but is a useful starting point when comparing our methods to the Adjei Method.
 Second, each method assumes that the probability of host survival with $x$ parasites is given by the logistic host-survival function
@@
 \begin{equation}
-    PD = 1 -  \dfrac{\sum_{x=0}^\infty h_x * f_x}{\sum_{x=0}^\infty f_x}
+    PD = 1 -  \sum_{x=0}^\infty h_x * f_x
     \label{eq:pd}
@@
 For
-each method, the differences in predictions of
+each method, any differences in predictions of
 $LD_{50}$ and $PD$ are a result of how a given method each calculates $a$ and $b$.
@@
     \item If observed number of hosts in category $i$ is greater than the expected number of hosts in $i$, let observed in $i$ equal to expected in $i$.
-    \item Run a generalized linear model in with a binomial random component and a logistic link to estimate $a$ and $b$.
+    \item Run a generalized linear model with a binomial random component and a logistic link to estimate $a$ and $b$.
 \end{enumerate}
@@
     \item Calculate the observed number of hosts in each category $i$.
-    \item Calculate the $\chi^2$ statistic for the observed and expected $n$ categories using the standard notation $X^2 = \sum_{i=1}^m \frac{(observed_i - expected_i)^2}{expected_i}$
-    \item Find $a$ and $b$ that minimize this $X^2$ statistic.
+    \item Calculate the $\chi^2$ statistic for the observed and expected $n$ categories using the standard notation $\chi^2 = \sum_{i=1}^m \frac{(observed_i - expected_i)^2}{expected_i}$
+    \item Find $a$ and $b$ that minimize this $\chi^2$ statistic.
 \end{enumerate}
@@
 \doublespacing
-
-This approach does not require any alteration of the observed data as in the Adjei approach and uses the same iterative minimization technique as the Crofton Method \citep{Lester1977}.
+The advantage of this approach over the Adjei approach is that is does not require any alteration of the observed data.
+Moreover, this approach can theoretically be extended such that $N_p$, $k_p$ and $\mu_p$ can be estimated directly from minimization of the $\chi^2$ statistic rather than via the Crofton Method.
+In practice, this minimization is difficult because host-parasite datasets are often not large enough and/or PIHM is not strong enough to uniquely estimate 5 parameters.
+Therefore, the Crofton Method provides an easy way to estimate $N_p$, $k_p$, and $\mu_p$ before minimizing the $\chi^2$ statistic to find $a$ and $b$.
 
@@
 
-The second alternative method relies on maximizing the likelihood of the observed data and proceeds as follows.
+The second alternative method relies on maximizing the likelihood of the observed data under a PIHM model.  This method proceeds as follows.
 
@@
     \begin{equation}
-        p(x; a, b, \mu_p, k_p) = \phi h(x; a, b) f(x; \mu_p, k_p)
+        p(x; a, b, \mu_p, k_p) = \phi h_x f_x
     \end{equation}
 
-    where $\phi$ is a normalizing constant. % equal to $(\sum_{x=0}^\infty h(x; a, b) * f(x; \mu_p, k_p))^{-1}$.
+    where $\phi$ is a normalizing constant. equal to $(\sum_{x=0}^\infty h(x; a, b) * f(x; \mu_p, k_p))^{-1}$.
 
@@
 \doublespacing
-The likelihood approach is advantageous because standard statistical techniques can then be used to assess whether parasite-induced host mortality is playing a significant role in a population [reword].
+The likelihood approach is advantageous over both of the aforementioned approaches because 1) it requires only estimating 4 parameters ($a$, $b$, $\mu_p$, and $k_p$) rather than 5 parameters 2) all 4 parameters can either be estimated jointly or $\mu_p$ and $k_p$ can first be estimated using the Crofton method and 3) standard statistical techniques, such as likelihood ratio tests or AIC, can be used to assess whether a model with PIHM is better than a model without PIHM.
 
@@
 
+Across all parameter combinations we examined, the two alternative approaches always provided less or equal bias when estimating $LD_{50}$ and $PD$ than the Adjei Method.
+For large sample sizes, all three methods were approximately equally effective at predicting $LD_{50}$ and $PD$ (Figure X).
+However, as sample size decreased the Adjei Method began to show significant bias in its predictions of $LD_{50}$ and $PD$.
+As sample sized decreased, the Adjei Method tended to first show bias in host-parasite systems when the host-survival decreased sharply with increasing parasite intensity (i.e.
+$a = 30, b = -15$).
+Approaching the smallest sample size, the Adjei Method generally showed the bias for all types of host-survival functions.
+
+For all three methods, the aggregation in the pre-mortality host population had strong effects on the bias of the $LD_{50}$ and $PD$ estimates.
+As $k_p$ decreased, the bias increased for all methods.
+This trend was most pronounced in the Adjei Method followed by the Chi-squared Method and was least pronounced in the Likelihood Method.
+These general trend held for different $LD_{50}$s (Appendix X).
+[Run same analyses with a different mean]
+
+% Precision, bias in $a$ and $b$
+
 \subsection*{Scenario II: Unknown pre-mortality parameters}
 
+When the Crofton Method was first used to estimate
+
+\subsection*{Scenario III: Efficacy of alternative methods under-realistic conditions}
+
+As both alternative methods were superior to the Adjei Method for estimating the host survival function, we focused whether either of these methods would be practical under realistic field conditions.
+
 \section*{Discussion}
+
+% The chi-squared method and the Adjei Method are more sensitive to the crofton method because the rely on the estimate of N_p which the Crofton Method does not estimate well.
+The likelihood approach doesn't rely on this estimate and is less sensitive to the crofton method.
 

### 2015-04-02 code/crofton_method.py

@@ -1,11 +1,13 @@
 from __future__ import division
 import numpy as np
 import macroeco.models as mod
+import macroeco.compare as comp
 import scipy.optimize as opt
 import pandas as pd
 import statsmodels.api as sm
 import pymc as pm
 import matplotlib.pyplot as plt
+from scipy.stats import chisqprob
 
 
 """
@@ -189,27 +191,27 @@ def likelihood_method(data, crof_params=None, max_sum=1000, guess=[10, -5]):
     -----
 
     """
-    kern = lambda x, mu, k, a, b: surv_prob(x, a, b) * \
-                                        mod.nbinom.pmf(x, mu, k)
-    pmf = lambda x, mu, k, a, b: kern(x, mu, k, a, b) / \
-                    sum(kern(np.arange(0, max_sum), mu, k, a, b))
+    # kern = lambda x, mu, k, a, b: surv_prob(x, a, b) * \
+    #                                     mod.nbinom.pmf(x, mu, k)
+    # pmf = lambda x, mu, k, a, b: kern(x, mu, k, a, b) / \
+    #                 sum(kern(np.arange(0, max_sum), mu, k, a, b))
 
-    def likefxn1(params, x):
-        """ Likelihood fxn for all parameters """
+    # def likefxn1(params, x):
+    #     """ Likelihood fxn for all parameters """
 
-        mu, k, a, b = params
-        return -np.sum(np.log(pmf(x, mu, k, a, b)))
+    #     mu, k, a, b = params
+    #     return -np.sum(np.log(pmf(x, mu, k, a, b)))
 
-    def likefxn2(params, x, mu, k):
-        """ Likelihood fxn for just a and b """
+    # def likefxn2(params, x, mu, k):
+    #     """ Likelihood fxn for just a and b """
 
-        a, b = params
-        return -np.sum(np.log(pmf(x, mu, k, a, b)))
+    #     a, b = params
+    #     return -np.sum(np.log(pmf(x, mu, k, a, b)))
 
     if crof_params:
         N, mu, k = crof_params
 
-        params = opt.fmin(likefxn2, guess, args=(data, mu, k), disp=0)
+        params = opt.fmin(likefxn2, guess, args=(data, mu, k), disp=1)
         out_params = [mu, k] + list(params)
 
     else:
@@ -220,6 +222,24 @@ def likelihood_method(data, crof_params=None, max_sum=1000, guess=[10, -5]):
                     args=(data,))
 
     return tuple(out_params)
+
+# PMF for host mortality
[…]

### 2015-04-02 code/old_R_code/analyses_pres_11_12_2014.R

@@ -1,0 +1,54 @@
+
+source("analysis_fxns.R")
+
+## 1. Run analysis on full raccon data
+
+# Load in data
+raccoon_data = read.csv("../data/all_raccoon_data.csv")
+pdf("full_raccoon_hist.pdf", width=7, height=5)
+hist(raccoon_data$worms, breaks=30, col='black', border='white',
+                xlab="Log Num. Parasites", ylab="Num. Raccoons",
+                xlim=c(0, 150), main="")
+dev.off()
+
+#raccoon_results = estimate_mortality(raccoon_data$worms, plot=T, weights=F,
+#            title="Raccoon Mortality, No Truncation", force_it=FALSE)
+
+# raccoon_results_force = estimate_mortality(raccoon_data$worms, plot=T, weights=T,
+#             title="Raccoon Mortality, No Truncation, with force", force_it=T)
+
+raccoon_results_trun = estimate_mortality(raccoon_data$worms, plot=T, weights=T,
+             title="Raccoon Mortality, Truncation", force_it=F, trun=120)
+
+# # # Plot observed vs predicted
+# # # How much death does this estimate estimate
+# # tcol = "#0000ff60"
+# # barplot(raccoon_results$obs, width=1, col='red', ylim=c(0, 50))
+# # points(1:length(raccoon_results$obs), raccoon_results$pred)
+
+# # hist(raccoon_results$observed, col=tcol, add=TRUE, freq=FALSE)
+
+# # # Truncated analysis. Truncated at 40
+# raccoon_results_trun = estimate_mortality(raccoon_data$worms, plot=T,
+#                 weights=T, trun=40, title="Raccoon Mortality, Truncation")
+
+# ## 2. Run analysis of chytrid data.  Try it with binned and unbinned
+
+# Load chytrid data; epizootic and enzootic
+# epi_data = read.csv("../data/ze_epizootic.csv")
+# en_data = read.csv("../data/ze_enzootic.csv")
+
+# en_results = estimate_mortality_bin(en_data$ZE1, plot=T, weights=T, trun=5000,
+#                 title="Enzootic Chtyrid Mortality")
+
+# en_results_force = estimate_mortality_bin(en_data$ZE1, plot=T, weights=T,
+#             trun=5000, title="Enzootic Chtyrid Mortality, Forced",
+#             force_it=TRUE)
+
+
+
+
+
+
+
+

### 2015-04-02 code/old_R_code/analysis_fxns.R

@@ -1,0 +1,220 @@
+library(MASS)
+
+## Defines analysis functions for analysing mortality data
+
+hx_func = function(x, a, b){
+    # Survival fxn
+    exp(a + b *log(x)) / (1 + exp(a + b *log( x)))
+}
+
+
+Fx_func = function(x, k, mu, num){
+    # NBD x number
+    dnbinom(x, size=k, mu=mu) * num
+}
+
+fit_model = function(obs, pred, para_number, num, k, mu,
+        plot=FALSE, weights=FALSE, title="Default", save=TRUE){
+
+    # obs : Observed hosts with i parasites
+    # pred : Predicted hosts with i parasites, should match obs
+    # num : Total number of host in the sample
+    # para_number : vector with the number of parasites in each bin
+    # k : k of the NBD fir to data
+    # mu : mu of the NBD fit to data
+    # plot : if True plot the survival fxn
+    # weights : if True, run glm with observations wiegthed by sample size
+    #
+    # Returns : list of Z0 (prop, of population dead), Z1 (prop, of infected
+    #                population dead), a (parameter of surv. fxn),
+    #                b (parameter of surv fxn)
+
+    if(weights){
+        fits_glm = glm(cbind(obs, pred - obs) ~
+                log(para_number), family=binomial, weights=pred)
+    }
+    else{
+        fits_glm = glm(cbind(obs, pred - obs) ~
+                log(para_number), family=binomial)
+    }
+
+    a = coef(fits_glm)[1]
+    b = coef(fits_glm)[2]
+
+    # # Calculating host mortality due to parasites
+    all_parasites = seq(1, max(para_number), 1)
+    numerator = c(Fx_func(0, k, mu, num), Fx_func(all_parasites, k, mu, num)) *
+                                c(1, hx_func(all_parasites, a, b))
+    sum_numerator = sum(numerator)
+
+    x_with_0 = 0:max(para_number)
+    denom = c(Fx_func(x_with_0, k, mu, num))
+    sum_denom = sum(denom)
+
+    Z0 = 1 - sum_numerator / sum_denom
+
+    Z1 = 1 - (sum(hx_func(all_parasites, a, b) * Fx_func(all_parasites, k, mu, num)) /
+                        sum(Fx_func(all_parasites, k, mu, num)))
+
+    if(plot){
+
+        if(save){
[…]

### 2015-04-02 code/old_R_code/log2bins.R

@@ -1,0 +1,112 @@
+library(MASS)
+
+#file = "../data/raccoon_data10_24.csv"
+#data=read.csv(file)
+#bp <- data$Weinstein
+#obs = bp[!is.na(bp)]
+# file = "../data/ze_adults_enzootic.csv"
+# data = read.csv(file)
+# obs = data$ZE1
+# obs = round(obs)
+file = "../data/all_raccoon_data.csv"
+data=read.csv(file)
+obs <- data$worms
+
+# Bin observe data
+log_base = 2
+start = 1
+max_num = ceiling(log(max(obs), base=log_base))
+bins = c(-0.1, log_base^(start:max_num))
+cuts = cut(obs, bins)
+df = data.frame(list("data"=obs, "bins"=cuts))
+observed = aggregate(obs~bins, df, length)
+
+# # Get predicted numbers from negative binomial
+trun_obs = obs
+fits = fitdistr(trun_obs, "negative binomial", list(size=.5, mu= 15), lower=.001)
+vec = seq(0, max(obs), 1)
+probs = dnbinom(vec, size=fits$estimate["size"], mu=fits$estimate["mu"])
+pred = probs * length(obs)
+
+tbins = 0:max(obs)
+new_cuts = cut(tbins, bins)
+pred_df = data.frame(list("data"=pred, "bins"=new_cuts))
+predicted = aggregate(pred~bins, pred_df, sum)
+
+# # Round these estimates?
+predicted$pred = round(predicted$pred)
+
+# Observed can't be bigger than predicted
+ind = observed$obs > predicted$pred
+observed$obs[ind] = predicted$pred[ind]
+
+# Set observed parasites to upper bound of the group
+group_upper = log_base^(start:max_num)
+
+# # # Maximum likelihood function
+
+# logistregfun = function(params, x, N, y){
+
+#     a = params[1]
+#     b = params[2]
+#     p.pred = exp(a + b *log(x)) / (1 + exp(a + b * log(x)))
+#     -sum(dbinom(y, size=N, prob=p.pred, log=TRUE))
+# }
+
+# opt1 = optim(fn = logistregfun, par = c(a=.5,b=.5),
+#                 x = group_upper, N=predicted$pred, y=observed$obs,
+#                 method="BFGS")
+
+# a = opt1$par[1]
+# b = opt1$par[2]
+
+hx = observed$obs / predicted$pred
+
[…]

### 2015-04-02 code/old_R_code/mortality_simulation.R

@@ -1,0 +1,85 @@
+# Code for simulating parasite mortality
+
+# Load fxns
+source("analysis_fxns.R")
+
+# Parameters
+num_hosts = 10000
+k = 1
+mu = 15
+a = 6 # LD50 a
+b = -2  # LD50 b
+LD50 = exp(a / abs(b))
+
+survival_prob = function(x, a, b){
+
+    res = exp(a + b * log(x)) / (1 + exp(a + b * log(x)))
+    return(res)
+
+}
+
+# 1. Equilibrium is established
+sim_data = rnbinom(num_hosts, size=k, mu=mu)
+sorted_sim = as.data.frame(table(sim_data))
+sim_para = as.numeric(levels(sorted_sim$sim_data)[as.integer(sorted_sim$sim_data)])
+sim_freq = sorted_sim$Freq
+
+# 2. Parasite induced mortality occurs
+surv_probs = survival_prob(sim_data, a, b)
+surv_probs[is.na(surv_probs)] = 1
+
+# Determine if an individual dies or not
+rands = runif(length(sim_data))
+surv = rands < surv_probs
+alive_hosts = sim_data[surv]
+
+true_death = 1 - sum(surv) / length(surv)
+
+par(mfrow=c(2,2))
+
+# Plot survival fxn
+max_para =
+vals = seq(1, max(sim_data), 1)
+plot(log(vals), survival_prob(vals, a, b), 'l', xlab="Log Num. Parasites",
+            ylab="Prob. Survival of Host", main="True Survival Curve")
+
+
+trun_val = .25 * LD50
+
+# How much death does this estimate estimate
+tcol = "#0000ff60"
+hist(log(sim_data), col='red', breaks=15, freq=T, xlab="Log Num. Parasites",
+        ylab="Num. Hosts", main="Pre vs Post Mortality", xlim=c(0,7))
+hist(log(alive_hosts), col=tcol, breaks=15, add=TRUE, freq=T)
+abline(v=log(LD50), col='grey', lwd=4, lty='dashed')
+abline(v=log(trun_val), col='black', lwd=4, lty='dashed')
+
+legend("topright", c("Pre-mortality", "Observed"), fill=c("red", tcol),
+    cex=0.6)
+
+ests1 = estimate_mortality(alive_hosts, plot=TRUE, weights=T,
+            trun=trun_val, force_it=T, save=FALSE,
+            title="Observed Survival Curve")
+
[…]

### 2015-04-02 code/old_R_code/parasite_morality_code.R

@@ -1,0 +1,117 @@
+library(MASS)
+
+# file = "../data/worm_data.csv"
+# data=read.csv(file)
+# bp <- data$adBPtot
+
+file = "../data/all_raccoon_data.csv"
+data=read.csv(file)
+bp <- data$worms
+
+# Count number of racs with a given parasite
+count_data = as.data.frame(table(bp))
+full_data = data.frame(list("bp" = seq(0, max(bp), 1), "Freq" = rep(0, max(bp) + 1)))
+full_data[as.numeric(as.vector(count_data$bp)) + 1, "Freq"] = count_data$Freq
+obs = full_data$Freq
+
+
+# Get predicted numbers from negative binomial
+fits = fitdistr(bp, "negative binomial", list(size=.5, mu= 15), lower=.001)
+vec = seq(0, max(bp), 1)
+probs = dnbinom(vec, size=fits$estimate["size"], mu=fits$estimate["mu"])
+pred = probs * length(bp)  # Maybe set fractional raccoons to zero?
+
+
+# # Observed can't be greater than predicted...seems like a hack-
+# #is binomial the best assumed distribution- cant have obs >pred
+
+parasite_nums = seq(0, max(bp), 1)
+new_pred = round(pred)
+new_obs = obs
+ind = obs > new_pred
+new_obs[ind] = new_pred[ind]
+
+# Drop where pred is zero- removes fraction individuals
+# start at i=2 because dont include 0 class as should have 100%surv
+ind_nozeros = new_pred != 0
+trun_obs = new_obs[ind_nozeros][2:sum(ind_nozeros)]
+trun_pred = new_pred[ind_nozeros][2:sum(ind_nozeros)]
+trun_x = parasite_nums[ind_nozeros][2:sum(ind_nozeros)]
+
+
+# # Maximum likelihood function
+
+logistregfun = function(params, x, N, y){
+
+	a = params[1]
+	b = params[2]
+	p.pred = exp(a + b *log(x)) / (1 + exp(a + b * log(x)))
+	-sum(dbinom(y, size=N, prob=p.pred, log=TRUE))
+}
+
+opt1 = optim(fn = logistregfun, par = c(a=.5,b=.5),
+				x = trun_x, N=trun_pred, y=trun_obs,
+				method="BFGS")
+
+#a=opt1$par[1]
+#b=opt1$par[2]
+
+### GLM in R: Should give the same results at the direct optimization
+fits_glm = glm(cbind(trun_obs, trun_pred - trun_obs) ~
[…]

### 2015-04-02 code/old_R_code/power_analysis.R

@@ -1,0 +1,77 @@
+# Code for simulating parasite mortality...should put these all into functions
+
+# Load fxns
+source("analysis_fxns.R")
+
+# Parameters
+num_hosts = 10000
+k = .5
+mu = 15
+a = 6 # LD50 a
+b = -2  # LD50 b
+LD50 = exp(a / abs(b))
+
+survival_prob = function(x, a, b){
+
+    res = exp(a + b * log(x)) / (1 + exp(a + b * log(x)))
+    return(res)
+
+}
+
+ks = seq(0.01, 1, length=50)
+trun_vals = seq(0.1, 1, length=50)
+a_vals = matrix(NA, nrow=length(ks), ncol=length(trun_vals))
+b_vals = matrix(NA, nrow=length(ks), ncol=length(trun_vals))
+mort_ratio = matrix(NA, nrow=length(ks), ncol=length(trun_vals))
+
+mort_diff = matrix(NA, nrow=length(ks), ncol=length(trun_vals))
+
+for(i in 1:length(ks)){
+    for(j in 1:length(trun_vals)){
+
+        # 1. Equilibrium is established
+        sim_data = rnbinom(num_hosts, size=ks[i], mu=mu)
+        sorted_sim = as.data.frame(table(sim_data))
+        sim_para = as.numeric(levels(sorted_sim$sim_data)[as.integer(sorted_sim$sim_data)])
+        sim_freq = sorted_sim$Freq
+
+        # 2. Parasite induced mortality occurs
+        surv_probs = survival_prob(sim_data, a, b)
+        surv_probs[is.na(surv_probs)] = 1
+
+        # Determine if an individual dies or not
+        rands = runif(length(sim_data))
+        surv = rands < surv_probs
+        alive_hosts = sim_data[surv]
+
+        true_death = 1 - sum(surv) / length(surv)
+
+        trun_val = trun_vals[j] * LD50
+
+        ests1 = estimate_mortality(alive_hosts, plot=F, weights=T,
+                    trun=trun_val, force_it=T, save=FALSE,
+                        title="Observed Survival Curve")
+
+        a = ests1$fits$a
+        b = ests1$fits$b
+
+        a_vals[i, j] = a
+        b_vals[i, j] = b
+        mort_ratio[i, j] = ests1$fits$Z0 / true_death
+        mort_diff[i, j] = ests1$fits$Z0 - true_death
+
[…]

### 2015-04-02 code/pihm_methods.py

@@ -1,0 +1,198 @@
+from __future__ import division
+import numpy as np
+import macroeco.models as mod
+import scipy.optimize as opt
+import pandas as pd
+import statsmodels.api as sm
+import pymc as pm
+import matplotlib.pyplot as plt
+
+"""
+Module contains the updated functions and classes for assessing
+parasite-induced host mortality (PIHM)
+
+Functions
+---------
+
+"""
+
+class PIHM:
+    """
+    Class PIHM (parasite-induced host morality) provides an interface for
+    analyzing the effect of parasite-induced host mortality on a given dataset
+
+    """
+
+    def __init__(self, data=None):
+        """
+        """
+
+        self.data = data
+        self.Np = None
+        self.mup = None
+        self.kp = None
+        self.a = None
+        self.b = None
+
+    def crofton_method(self, bin_edges, guess=None, full_output=False):
+        """
+        Crofton's method for estimating the pre-mortality parameters N_p
+        (population size before mortality),
+        mu_p (mean parasite load before mortality), and
+        k_p (parasite aggregation before mortality)
+
+        Parameters
+        ----------
+        data : array
+            Parasite data across hosts
+        bin_edges : list
+            List specifying upper and lower bin boundaries. Upper right most bin
+            boundary is inclusive; all other upper bounds are exclusive.  Example:
+            You want to group as follows [0, 1, 3], [4, 5, 6], [7, 8, 9] you would
+            specify [0, 4, 7, 9.1].  This can be a little weird, but if you want to
+            bin from [0], [1], [2], you would specify [0, 1, 2, 2.9].  You have to
+            be a bit careful with the upper bound because it is INCLUSIVE.
+        guess : None or tuple
+            If tuple, should be a guess for mu and k of the negative binomial
+        full_output : bool
+            If full_output is False, just returns N_p, mu_p, and k_p. If True,
+            return opt_params : (N_p, mu_p, k_p)
+            obs : observed number of hosts in groups
+            pred: predicted number of hosts in groups
+            splits: the group boundaries
+            np.sum((obs - pred)**2 / pred): minimized chi-squared statistic
+
+        Returns
+        -------
+        See full_output
+
+
+        """
+
+        data = self.data
+
+        obs = np.histogram(data, bins=bin_edges)[0]
+
+        splits = []
[…]

### 2015-04-20 docs/parasite_mortatlity_manuscript.tex
Updated parasite_mortality manuscript with new methods

--- 2015-03-30 docs/parasite_mortatlity_manuscript.tex
@@
 
+\linenumbers
+\section*{Introduction}
+
+Infectious agents can have major impacts on animal populations through changing
+population dynamics and stability (Dobson 1992), altering predator prey
+interactions (Joly 2004), and even causing species decline and extinction
+(Viana 2015, Daszak 1999).  Accurate estimates of parasite-induced host
+mortality (PIHM) in wild animals are important for understanding what regulates both
+host and parasite populations and to make predictions about disease
+transmission in natural systems. Although a negative impact on host fitness is
+a fundamental component of parasitism (Lafferty and Kuris 2002, Poulin book,
+other definition sources), it is notoriously difficult to quantify PIHM in wild animal populations (cite, McCallum 2008?, lester
+1984).
+
+To conclusively identify PIHM in wild animal populations, it is necessary to
+experimentally infect and track host populations in ways that are rarely
+possible in most host-parasite systems (citations).  Instead, parasitologists
+are often only able to collect a certain number of hosts and determine each
+host's parasite intensity.  This snapshot, distributional data is far from the
+ideal type of data for addressing questions regarding PIHM, but the reality is
+that this is the type of data on which most questions regarding PIHM are asked
+(citations). The two primary questions that one may wish to ask given a snapshot host-
+parasite dataset are: 1) Is PIHM occurring in this system? and 2) How does
+host survival change as parasite intensity increases?  MORE
+
+The first of these two questions was addressed by \cite{Crofton1971a} developed
+a method to test for PIHM using the truncation of the negative binomial
+distribution.  In short, the Crofton Method assumes that the distribution of
+parasites across hosts before mortality occurs follows a negative binomial
+distribution \citep{AndersonandMay1978,Shaw1998}.  As heavily infected hosts
+begin to die, the negative binomial distribution gets truncated because heavily
+infected hosts dies and are no longer observed in a sample. In other words, the
+observed host-parasite distribution and the pre-mortality host-parasite
+distribution will predict substantially different numbers of of highly infected
+hosts (because those have died due to infection) but similar number of hosts with low infection
+loads (because those have survived).  \citep{Crofton1971a} noted that by
+starting with all the observed data and iteratively fitting a negative binomial
+distribution to hosts with lower and lower parasite loads, one could determine
+whether or not PIHM was occurring in the system (Figure XA).  This was done
+graphically by determining whether parameters of the truncated negative binomial distributions showed a substantial change as hosts with smaller and smaller parasite intensities were fitted (Figure XB).
+The Crofton Method and this graphical technique for determining whether or not PIHM is occurring are both still used \citep{Ferguson2011}.
+A more thorough description and implementation of the Crofton Method is given in Appendix X.
+
+The second question regarding PIHM is how host survival changes as parasite
+intensity increases. The Crofton Method on its own cannot answer this question
+and \cite{Adjei1986} proposed a method to determine how host survival
+probability changes with increasing parasite-load.  The Adjei Method proceeds
+by first using the Crofton Method to estimate the pre-mortality parameters for
+a host-parasite distribution (describe what these are) and then, given these parameters, estimates a host-survival function that describes how the probability of host-survival changes with increasing parasite load (see Appendix X for a full description).
+From this host-survival function, the Adjei Method can estimate important host-parasite quantities such as the parasite intensity at which 50\% of hosts succumb to PIHM ($LD_{50}$), as well as the percent of hosts in a population succumbing to PIHM \citep{Adjei1986}.
+
+Asking questions regardin
+
 \section*{Methods}
 
-\subsection*{Overview of PIHM approaches}
+\subsection*{The likelihood method for estimating PIHM}
 
-We propose and two alternative approaches for estimating LD50 and PD from
-parasite intensity data.  Our two approaches, as well as the Adjei Method,
-follow the same general steps.  First, all approaches assume that the
-distribution of parasites across hosts in the pre-mortality population follows
-a negative binomial distribution and, using the Crofton Method, each approach
-starts by estimating the total number hosts  ($N_p$), the mean number of
-parasites per host ($\mu_p$), and the aggregation of the parasites across hosts
-before PIHM ($k_p$) (see Appendix X for a description of the Crofton Method).
-Second, each method assumes that the probability of host survival with $x$ parasites is given by the logistic host-survival function
+Given the potential deficiencies of the Adjei Method, we provide an alternative
+approach for estimating parasite-induced host mortality (PIHM) that makes less
+assumption than the previously proposed Adjei Method.  The likelihood method
+does not require any binning or alteration of the data, potentially reduces the
+number of parameters that need to be estimated, and allows for standard
+statistical techniques to be used to assess the significance of PIHM in a
+system.
+
+The likelihood method makes the following assumption about the host-parasite
+system. First, as with all previously proposed methods for estimating PIHM, the
+likelihood method assumes that the pre-mortality distribution follows a
+negative binomial distribution ($g(x)$) with parameter $\mu_p$ and $k_p$. The
+validity of the assumption is an inherent problem with all the PIHM methods
+proposed to date and we address this thoroughly in the Discussion. The second
+assumption that the likelihood method makes is that the host survival function
+takes the form of a logistic curve given by
+
 \begin{equation}
@@
 \end{equation}
-Using equation \ref{eq:logistic}, each method uses either a likelihood or chi-squared approach to estimate $a$ and $b$.  Finally, from the estimated parameters $a$ and $b$,
-$LD_{50}$ and $PD$ can be calculated by the equations \citep{Adjei1986}
+
+where $x$ is the parasite intensity in a given host and $a$ and $b$ are the two parameters of the function.
+Generally, a larger $a$ allows for hosts to tolerate larger parasite loads before experiencing parasite-induced mortality and a more negative $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases.
+The value $exp(a / |b|)$ (notation) is typically referred to as the $LD_50$, which gives the parasite intensity at which 50\% of host experience mortality.
+With these two explicit assumptions, the likelihood method tries to estimate the 4 parameters $\mu_p$, $k_p$, $a$, and $b$.
+
+To estimate these parameters, we need to define a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.
+Using the standard rules of conditional probability  This distribution can be written as
 
 \begin{equation}
-    LD_{50} = \exp(\frac{a}{b})
-    \label{eq:ld50}
+    P(x | \text{survival}) = \dfrac{P(\text{survival} | x) * P(x)}{P(\text{survival})}
+    \label{eq:concept}
 \end{equation}
 
-and
+One can now see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(survival) = \sum_0^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$.
+Therefore equation \ref{eq:concept} can be written as
 
 \begin{equation}
-    PD = 1 -  \dfrac{\sum_{x=0}^\infty h_x * f_x}{\sum_{x=0}^\infty f_x}
-    \label{eq:pd}
-\end{equation}
-where $f_x$ is the probability mass function for the pre-mortality distribution of parasites per hosts which follows a negative binomial distribution with mean $\mu_p$ and aggregation parameter $k_p$.
-For
-each method, the differences in predictions of
-$LD_{50}$ and $PD$ are a result of how a given method each calculates $a$ and $b$.
-
-\subsection*{Estimating $a$ and $b$ with the Adjei Method}
-
-To estimate $a$ and $b$, the Adjei Method proceeds as follows \citep{Adjei1986}
-
-\singlespacing
-\begin{enumerate}
-    \item Estimate $N_p$, $\mu_p$, and $k_p$ using the Crofton Method
-    \item Given $N_p$, $\mu_p$, and $k_p$, estimate the expected number of hosts in category $i$ via equation $N_p f_i$.
-$i$ specifies either a given parasite intensity or some range of parasites intensities (i.e.
-number of hosts with 100 parasites or number of hosts with 100-150 parasites) and there are $m$ categories.
-    \item Calculate the observed number of hosts in each category $i$.
-Assume that the observed number of hosts in each category $i$ is binomial distributed with the total number of ``trials'' equal to the expected number of hosts in $i$.
-    \item If observed number of hosts in category $i$ is greater than the expected number of hosts in $i$, let observed in $i$ equal to expected in $i$.
-    \item Run a generalized linear model in with a binomial random component and a logistic link to estimate $a$ and $b$.
-\end{enumerate}
-
-\doublespacing
-See \cite{Adjei1986} and Appendix X for a more thorough description of this approach and some supporting examples.
-
-\subsection*{Estimating $a$ and $b$ with chi-squared approach}
-
-The first alternative approach that we propose relies on estimating $a$ and $b$ by minimizing the $\chi^2$ statistic.  The approach proceeds as follows
-
-\singlespacing
-\begin{enumerate}
-    \item Estimate $N_p$, $\mu_p$, and $k_p$ using the Crofton Method.
-    \item Specify $m$ categories of parasite intensities where each category $i$ either contains a single parasite intensity or a range of parasite intensities.
-Calculate the expected number of hosts in each category $i $ by $N_p$ times the probability that a host is category $i$ and is alive.
-This can be written as $N_p * f_i * h_i$ where the appropriate summation is taken if $i$ is a range of parasite intensities.
-    \item Calculate the observed number of hosts in each category $i$.
-    \item Calculate the $\chi^2$ statistic for the observed and expected $n$ categories using the standard notation $X^2 = \sum_{i=1}^m \frac{(observed_i - expected_i)^2}{expected_i}$
-    \item Find $a$ and $b$ that minimize this $X^2$ statistic.
-\end{enumerate}
-
-\doublespacing
-
-This approach does not require any alteration of the observed data as in the Adjei approach and uses the same iterative minimization technique as the Crofton Method \citep{Lester1977}.
-
-\subsection*{Estimating $a$ and $b$ with the likelihood approach}
-
-The second alternative method relies on maximizing the likelihood of the observed data and proceeds as follows.
-
-\singlespacing
-\begin{enumerate}
-    \item Estimate $N_p$, $\mu_p$, and $k_p$ using the Crofton Method.
-    \item Specify the probability distribution that a host has $x$ parasites and is alive as
-
-    \begin{equation}
-        p(x; a, b, \mu_p, k_p) = \phi h(x; a, b) f(x; \mu_p, k_p)
+    P(x | survival) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
+    \label{eq:dist}
 \end{equation}
 
-    where $\phi$ is a normalizing constant. % equal to $(\sum_{x=0}^\infty h(x; a, b) * f(x; \mu_p, k_p))^{-1}$.
+Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, $b$ that maximize the likelihood of an observed host-parasite dataset $\mathbf{x}$.
 
-    \item Assuming hosts are independent, the likelihood of a datasets with $n$ hosts can be written as $L(a, b | x, \mu_p, k_p) = \Pi_{i=0}^n p(x_i ; a, b, \mu_p, k_p)$.
+Alternatively, one could apply the Crofton Method to estimate $\mu_p$ and $k_p$ and then find the maximum likelihood estimates of $a$ and $b$ and the corresponding $LD_{50}$.
+A final option would be to follow the example of \citep{Ferguson2011} and assume $k_p = 1$ and only estimate $a$, $b$ and $\mu_p$.
 
-    \item Use standard optimization techniques to estimate $a$ and $b$
+\subsection{Testing the ability of approaches to identify PIHM}
 
-\end{enumerate}
-
-\doublespacing
-The likelihood approach is advantageous because standard statistical techniques can then be used to assess whether parasite-induced host mortality is playing a significant role in a population [reword].
-
-\subsection*{Dataset simulation and comparison of PIHM approaches}
-
-To compare the ability of these methods to recover $LD_{50}$ and $PD$, we randomly generated data using the
-following procedure.  First, we drew $N_p$ randomly infected hosts from a
-negative binomial distribution with parameters $\mu_p$ and $k_p$.  Second, we chose values of $a$ and $b$ and calculated the probability of survival
+To test the ability of the Adjei Method and the Likelihood Method to identify whether or not PIHM was occurring in a system, we randomly generated data using the following procedure.
+First, we drew $N_p$ randomly infected hosts from a
+negative binomial distribution with parameters $\mu_p$ and $k_p$.
+This represented the dataset observed before mortality.
+Second, we chose values of $a$ and $b$ and calculated the probability of survival
 for all $N_p$ hosts using equation \ref{eq:logistic}.  Third, we drew $N_p$ random numbers from a uniform distribution
@@
 number, the host experienced parasite-induced mortality.  The surviving
-hosts comprised the dataset that would be obtained in the field.
+hosts comprised the dataset ($\mathbf{x}$) that would be obtained in the field, after PIHM.
 
-% For a simulated datasets, we estimated the parameters $a$ and $b$ using each of the three approached outlines above. For combination of $a$ and $b$ we then estimated $LD_{50}$ and $PD$. To
-% determine the efficacy of each method, we estimated the bias and precision
-% of the resulting $LD_{50}$ and $PD$ estimates \citep{Walther2005}.  To compare
-% the bias and precision across parameter combinations we used the
-% standardized bias and the standardized precision \citep{Walther2005}.
+Using both the pre-mortality and post-mortality simulated datasets,  we assumed
+that the values of $N_p$, $\mu_p$, and $k_p$ were known and tested the ability of both methods to correctly determine whether or not PIHM was occurring.
+While this scenario is unrealistic because the parameters $N_p$,
+$\mu_p$, and $k_p$ are always unknown, we implemented this scenario as a baseline to
+establish the efficacy of the methods independent of the estimates of $N_p$, $\mu_p$ and $k_p$.
+For the Adjei Method, $N_p$, $\mu_p$, and $k_p$ are estimated using the Crofton Method, while $\mu_p$ and $k_p$ in the likelihood method can be estimated jointly with $a$ and $b$ or via the Crofton Method.
+If a
+method could not correctly predict whether or not PIHM was occurring under these idealistic conditions, we considered this strong evidence of the unreliability of this method.
 
-Using the simulated datasets, we devised two scenarios to test the above approaches. In the first scenario, we assumed
-that the values of $N_p$, $\mu_p$, and $k_p$ were known and tested the ability of both
-methods to recover the true values of $a$ and $b$ over increasing values of
-$N_p$.  While this first scenario is unrealistic because the parameters $N_p$,
-$\mu_p$, and $k_p$ are always unknown, we implemented this scenario as a baseline to
-establish the efficacy of the methods independent of the Crofton Method used to estimate $N_p$, $\mu_p$ and $k_p$.   If a
-method could not return unbiased [word choice] estimates of $LD_{50}$ and/or $PD$ under these
-idealized conditions, we took this as evidence for unreliability of this method. The second scenario
-proceeded exactly as the first, except that we estimated $N_p$, $mu_p$, and
-$k_p$ using the Crofton Method.
+We used three different values of $\mu_p$ (10, 100, 500) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and sharp decreases in host survival with increasing parasite intensity.
+For a given $\mu_p$, each survival function had the same $LD_{50}$, but different values of $a$ and $b$ (Table X).
+We examined each $\mu_p$-survival function pair at  three levels of parasite
+aggregation, $k_p = 0.1, 0.5, 1$, which are realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
+combinations we simulated 150 datasets and tested the probability of each method incorrectly identifying PIHM in the pre-mortality dataset (Type I error) and incorrectly failing to identify PIHM in the post-mortality dataset (Type II error).
+For each method, we we used likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level 0.05 (Appendix X).
+We tested all each parameter combinations for pre-mortality population sizes of $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
+10000].
 
-For each scenario, we used two different $LD_{50}$
-values: $\exp(2) = 7.39$ parasites with $\mu_p = 10$ and $\exp(3) = 20.08$ parasites with  $\mu_p = 30$. For each $LD_{50}$ value we choose three combinations of $a$ and $b$ that
-represented three different patterns of how host survival decreased with
-parasite intensity:  gradual, moderate and a sharp decrease in host survival
-probability (Figure X).  Finally we examined three levels of parasite
-aggregation, $k_p = 0.1, 0.5, 1$, which are realistic values of parasite aggregation in
-natural populations \citep{Shaw1998}.  For each of these parameter
-combinations we simulated 150 datasets, estimated $a$, $b$, $LD_{50}$, and $PD$ and calculated the standardized bias and
+\subsection*{Testing ability of PIHM approaches to recover survival function}
+
+To compare the ability of the Adjei Method and the likelihood method to recover $LD_{50}$ and the parameters $a$ and $b$ or the survival function, we used the same simulation procedure and parameter combinations described above.For each parameter
+combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
 precision \citep{Walther2005} for these estimates over varying pre-mortality host population sizes  $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
@@
 
-% However, to facilitate comparison across parameter combinations, we
-% used $N_p$ as a proxy for sample size.  Finally, for low values of $N_p$, there
-% were some simulated datasets where the Chi-squared Method and the Likelihood
-% Method failed to converge.  These trials were discarded \dots [or should we save them
-% to see what is going on?]. The highest failure percentage we observed was X for
-% the X method with $N_p$ = 300 [need to look into this further/use a more robust
-% convergence method for the likelihood approach].
+\subsection*{Application to data}
+
+We tested the likelihood method on the datasets given in \citep{Crofton1971a} and \citep{Adjei1986} which both papers reported seeing PIHM in the respective populations.
 
@@
 
-\subsection*{Scenario I: Known pre-mortality parameters}
+\subsection*{Detecting PIHM}
 
-\subsection*{Scenario II: Unknown pre-mortality parameters}
+\subsection*{Recovery of the survival function}
+
+\subsection*{Application to data}
 
 \section*{Discussion}
+
+Parasite-induced host mortality is of substantial interest in many systems, but determining whether it is occurring given only observational data is notoriously difficult.
+Many of the previous methods derived to determine the effect of PIHM on a host-parasite system limit their inference to answering the yes or no question of whether or not PIHM is occurring.
+While a relevant question, it is often of interest to know something about the host survival function which can provide information regarding important properties of the host-parasite system, such as $LD_{50}$ and percent of the population suffering PIHM.
+
+We show that the Adjei Method, the only currently proposed method to estimate the host survival function and the $LD_{50}$ from observational PIHM, has some serious methodological problems that result in biased estimates of the host survival function even under the most idealistic conditions.
+Interestingly, despite these flaws, the Adjei Method can still produce unbiased and precise estimates of the $LD_{50}$ when host-parasite systems show aggregation close to $k = 1$
+
+To attempt to ameliorate the flaws in the Adjei Method, we proposed a more
+general method to determine both whether or not PIHM is occurring in a system
+and the shape of the survival function.  We show that this method is
+asymptotically unbiased when estimating the host-survival function for all of
+the parameter space that we explored, but can produce seriously biased
+estimates of the host survival function for sample sizes typically observed in
+many host-parasite studies.  However, we found that the likelihood method
+produces unbiased and precise estimates of the $LD_{50}$ for small, realistic
+sample sizes.
+
+Moreover, the likelihood method...
+
+% Discuss all of the problems with these methods
+
+What do these findings tell us about our ability to go beyond saying whether or
+not PIHM is occurring in a system?  While we have generalized and improved upon
+the previously existing methods for estimating PIHM, we cannot belie the fact
+that estimating the host survival function from observational data alone is
+ladened with assumptions and difficulties.
+
+The most fundamental
+assumption of all methods for estimating PIHM is that the shape of the pre-
+mortality host-parasite distribution is known. In the discrete case, this
+distribution is negative binomial, while in the continuous case the distribution can be gamma of exponential (\citep{Ferguson2011}).  While there is substantial empirical and
+theoretical evidence to justify the use of the negative binomial distribution
+as the pre-mortality distribution for macroparasites across hosts (Crofton, Shaw, Anderson and Gordon, etc), it is widely recognized that different processes can lead to a variety of distributions of parasites across hosts (Wilber, Duerr etc).
+However, the critical assumption of the pre-mortality distribution is not that the processes leading the pre-mortality distribution generate a negative binomial distribution, but rather that the pre-mortality distribution is well-fit by a negative binomial.
+The extreme flexibility of the negative binomial distribution in the discrete case of the gamma distribution in the continuous case make them reasonable candidate distributions for the pre-mortality distributions.
+Therefore, we do not see this assumption as central problem in any of the proposed methods.
+
+However, to use the pre-mortality distribution to infer whether or not the PIHM
+is occurring in a system requires an explicit assumption about the host
+survival function and the shape of the post-mortality distribution.  Regarding
+the host-survival function, all methods of PIHM assume that the host-survival
+function is such that uninfected individuals and individuals with low parasite
+intensity experience essentially no PIHM.  \cite{Lanciani1989} illustrated this
+by showing that when hosts experienced a linear decrease in survival
+probability the Crofton Method could not detect PIHM.  This result, relates to
+the shape of the post-mortality distribution.  Given that the pre-mortality
+distribution is well-fit by a negative binomial, a linear host-survival
+function will result in a post-mortality distribution that is also well-fit by
+a negative binomial distribution (why?).  In this case, one would be unable to
+identify PIHM because the pre-mortality distribution
+
+As parasitologist, we often want to get the most out of our
 

### 2015-04-20 code/analysis_for_manuscript.py
Added code for simulating parameters

@@ -1,0 +1,85 @@
+import pihm_methods as pm
+reload(pm)
+import numpy as np
+import configparser
+import pandas as pd
+
+"""
+Description
+-----------
+
+Full simulation for the manuscript
+
+MORE
+
+"""
+
+
+# Parameters
+ks = [1, 0.5, .1]
+# mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], []),
+#            100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
+mu_ld50 =  {500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
+Nps = [300, 500, 1000, 2000, 5000, 7500, 10000]
+N_samp = 150
+
+# ks = [1]
+# mu_ld50 = {10: ([(20, -10)], [0, 1, 2, 3, 4, 4.9])}
+# Nps = [5000]
+# N_samp = 50
+
+
+scenario_1_results = {}
+
+pihm = pm.PIHM()
+
+for mup in mu_ld50.iterkeys():
+
+  scenario_1_results[mup] = {}
+
+  for ld50 in mu_ld50[mup][0]:
+
+    scenario_1_results[mup][ld50] = {}
+
+    for kp in ks:
+
+      scenario_1_results[mup][ld50][kp] = {}
+
+      for Np in Nps:
+
+        # Print to see progress
+        print((mup, ld50, kp, Np))
+
+        results_like = []
+        results_adjei = []
+        samp_sizes = []
+
+        # Set the parameters
+
+        for i in xrange(N_samp):
+
+          pihm.set_all_params(Np, mup, kp, ld50[0], ld50[1])
+          pihm.data = pihm.get_pihm_samples()[0]
+
+          samp_sizes.append(len(pihm.data))
+
+          try:
+
+            # Likelihood results
+            like_out = pihm.likelihood_method(full_fit=False, disp=False)[-2:]
+
+            # Adjei results
+            adjei_out = pihm.adjei_method([], [], no_bins=True,
+                                                run_crof=False)[-2:]
+
+            results_like.append(like_out)
+            results_adjei.append(adjei_out)
+
+          except:
+            pass
+
+        scenario_1_results[mup][ld50][kp][Np] = \
+                        (results_like, results_adjei, samp_sizes)
+
+
+pd.to_pickle(scenario_1_results, "../results/scenario1_analysis_results2.pkl")

### 2015-04-20 code/analyses_pres_11_12_2014.R
Remove excess files

@@ -1,54 +1,0 @@
-
-source("analysis_fxns.R")
-
-## 1. Run analysis on full raccon data
-
-# Load in data
-raccoon_data = read.csv("../data/all_raccoon_data.csv")
-pdf("full_raccoon_hist.pdf", width=7, height=5)
-hist(raccoon_data$worms, breaks=30, col='black', border='white',
-                xlab="Log Num. Parasites", ylab="Num. Raccoons",
-                xlim=c(0, 150), main="")
-dev.off()
-
-#raccoon_results = estimate_mortality(raccoon_data$worms, plot=T, weights=F,
-#            title="Raccoon Mortality, No Truncation", force_it=FALSE)
-
-# raccoon_results_force = estimate_mortality(raccoon_data$worms, plot=T, weights=T,
-#             title="Raccoon Mortality, No Truncation, with force", force_it=T)
-
-raccoon_results_trun = estimate_mortality(raccoon_data$worms, plot=T, weights=T,
-             title="Raccoon Mortality, Truncation", force_it=F, trun=120)
-
-# # # Plot observed vs predicted
-# # # How much death does this estimate estimate
-# # tcol = "#0000ff60"
-# # barplot(raccoon_results$obs, width=1, col='red', ylim=c(0, 50))
-# # points(1:length(raccoon_results$obs), raccoon_results$pred)
-
-# # hist(raccoon_results$observed, col=tcol, add=TRUE, freq=FALSE)
-
-# # # Truncated analysis. Truncated at 40
-# raccoon_results_trun = estimate_mortality(raccoon_data$worms, plot=T,
-#                 weights=T, trun=40, title="Raccoon Mortality, Truncation")
-
-# ## 2. Run analysis of chytrid data.  Try it with binned and unbinned
-
-# Load chytrid data; epizootic and enzootic
-# epi_data = read.csv("../data/ze_epizootic.csv")
-# en_data = read.csv("../data/ze_enzootic.csv")
-
-# en_results = estimate_mortality_bin(en_data$ZE1, plot=T, weights=T, trun=5000,
-#                 title="Enzootic Chtyrid Mortality")
-
-# en_results_force = estimate_mortality_bin(en_data$ZE1, plot=T, weights=T,
-#             trun=5000, title="Enzootic Chtyrid Mortality, Forced",
-#             force_it=TRUE)
-
-
-
-
-
-
-
-

### 2015-04-20 code/analysis_fxns.R

@@ -1,220 +1,0 @@
-library(MASS)
-
-## Defines analysis functions for analysing mortality data
-
-hx_func = function(x, a, b){
-    # Survival fxn
-    exp(a + b *log(x)) / (1 + exp(a + b *log( x)))
-}
-
-
-Fx_func = function(x, k, mu, num){
-    # NBD x number
-    dnbinom(x, size=k, mu=mu) * num
-}
-
-fit_model = function(obs, pred, para_number, num, k, mu,
-        plot=FALSE, weights=FALSE, title="Default", save=TRUE){
-
-    # obs : Observed hosts with i parasites
-    # pred : Predicted hosts with i parasites, should match obs
-    # num : Total number of host in the sample
-    # para_number : vector with the number of parasites in each bin
-    # k : k of the NBD fir to data
-    # mu : mu of the NBD fit to data
-    # plot : if True plot the survival fxn
-    # weights : if True, run glm with observations wiegthed by sample size
-    #
-    # Returns : list of Z0 (prop, of population dead), Z1 (prop, of infected
-    #                population dead), a (parameter of surv. fxn),
-    #                b (parameter of surv fxn)
-
-    if(weights){
-        fits_glm = glm(cbind(obs, pred - obs) ~
-                log(para_number), family=binomial, weights=pred)
-    }
-    else{
-        fits_glm = glm(cbind(obs, pred - obs) ~
-                log(para_number), family=binomial)
-    }
-
-    a = coef(fits_glm)[1]
-    b = coef(fits_glm)[2]
-
-    # # Calculating host mortality due to parasites
-    all_parasites = seq(1, max(para_number), 1)
-    numerator = c(Fx_func(0, k, mu, num), Fx_func(all_parasites, k, mu, num)) *
-                                c(1, hx_func(all_parasites, a, b))
-    sum_numerator = sum(numerator)
-
-    x_with_0 = 0:max(para_number)
-    denom = c(Fx_func(x_with_0, k, mu, num))
-    sum_denom = sum(denom)
-
-    Z0 = 1 - sum_numerator / sum_denom
-
-    Z1 = 1 - (sum(hx_func(all_parasites, a, b) * Fx_func(all_parasites, k, mu, num)) /
-                        sum(Fx_func(all_parasites, k, mu, num)))
-
-    if(plot){
-
-        if(save){
[…]

### 2015-04-20 code/log2bins.R

@@ -1,112 +1,0 @@
-library(MASS)
-
-#file = "../data/raccoon_data10_24.csv"
-#data=read.csv(file)
-#bp <- data$Weinstein
-#obs = bp[!is.na(bp)]
-# file = "../data/ze_adults_enzootic.csv"
-# data = read.csv(file)
-# obs = data$ZE1
-# obs = round(obs)
-file = "../data/all_raccoon_data.csv"
-data=read.csv(file)
-obs <- data$worms
-
-# Bin observe data
-log_base = 2
-start = 1
-max_num = ceiling(log(max(obs), base=log_base))
-bins = c(-0.1, log_base^(start:max_num))
-cuts = cut(obs, bins)
-df = data.frame(list("data"=obs, "bins"=cuts))
-observed = aggregate(obs~bins, df, length)
-
-# # Get predicted numbers from negative binomial
-trun_obs = obs
-fits = fitdistr(trun_obs, "negative binomial", list(size=.5, mu= 15), lower=.001)
-vec = seq(0, max(obs), 1)
-probs = dnbinom(vec, size=fits$estimate["size"], mu=fits$estimate["mu"])
-pred = probs * length(obs)
-
-tbins = 0:max(obs)
-new_cuts = cut(tbins, bins)
-pred_df = data.frame(list("data"=pred, "bins"=new_cuts))
-predicted = aggregate(pred~bins, pred_df, sum)
-
-# # Round these estimates?
-predicted$pred = round(predicted$pred)
-
-# Observed can't be bigger than predicted
-ind = observed$obs > predicted$pred
-observed$obs[ind] = predicted$pred[ind]
-
-# Set observed parasites to upper bound of the group
-group_upper = log_base^(start:max_num)
-
-# # # Maximum likelihood function
-
-# logistregfun = function(params, x, N, y){
-
-#     a = params[1]
-#     b = params[2]
-#     p.pred = exp(a + b *log(x)) / (1 + exp(a + b * log(x)))
-#     -sum(dbinom(y, size=N, prob=p.pred, log=TRUE))
-# }
-
-# opt1 = optim(fn = logistregfun, par = c(a=.5,b=.5),
-#                 x = group_upper, N=predicted$pred, y=observed$obs,
-#                 method="BFGS")
-
-# a = opt1$par[1]
-# b = opt1$par[2]
-
-hx = observed$obs / predicted$pred
-
[…]

### 2015-04-20 code/mortality_simulation.R

@@ -1,85 +1,0 @@
-# Code for simulating parasite mortality
-
-# Load fxns
-source("analysis_fxns.R")
-
-# Parameters
-num_hosts = 10000
-k = 1
-mu = 15
-a = 6 # LD50 a
-b = -2  # LD50 b
-LD50 = exp(a / abs(b))
-
-survival_prob = function(x, a, b){
-
-    res = exp(a + b * log(x)) / (1 + exp(a + b * log(x)))
-    return(res)
-
-}
-
-# 1. Equilibrium is established
-sim_data = rnbinom(num_hosts, size=k, mu=mu)
-sorted_sim = as.data.frame(table(sim_data))
-sim_para = as.numeric(levels(sorted_sim$sim_data)[as.integer(sorted_sim$sim_data)])
-sim_freq = sorted_sim$Freq
-
-# 2. Parasite induced mortality occurs
-surv_probs = survival_prob(sim_data, a, b)
-surv_probs[is.na(surv_probs)] = 1
-
-# Determine if an individual dies or not
-rands = runif(length(sim_data))
-surv = rands < surv_probs
-alive_hosts = sim_data[surv]
-
-true_death = 1 - sum(surv) / length(surv)
-
-par(mfrow=c(2,2))
-
-# Plot survival fxn
-max_para =
-vals = seq(1, max(sim_data), 1)
-plot(log(vals), survival_prob(vals, a, b), 'l', xlab="Log Num. Parasites",
-            ylab="Prob. Survival of Host", main="True Survival Curve")
-
-
-trun_val = .25 * LD50
-
-# How much death does this estimate estimate
-tcol = "#0000ff60"
-hist(log(sim_data), col='red', breaks=15, freq=T, xlab="Log Num. Parasites",
-        ylab="Num. Hosts", main="Pre vs Post Mortality", xlim=c(0,7))
-hist(log(alive_hosts), col=tcol, breaks=15, add=TRUE, freq=T)
-abline(v=log(LD50), col='grey', lwd=4, lty='dashed')
-abline(v=log(trun_val), col='black', lwd=4, lty='dashed')
-
-legend("topright", c("Pre-mortality", "Observed"), fill=c("red", tcol),
-    cex=0.6)
-
-ests1 = estimate_mortality(alive_hosts, plot=TRUE, weights=T,
-            trun=trun_val, force_it=T, save=FALSE,
-            title="Observed Survival Curve")
-
[…]

### 2015-04-20 code/parasite_morality_code.R

@@ -1,117 +1,0 @@
-library(MASS)
-
-# file = "../data/worm_data.csv"
-# data=read.csv(file)
-# bp <- data$adBPtot
-
-file = "../data/all_raccoon_data.csv"
-data=read.csv(file)
-bp <- data$worms
-
-# Count number of racs with a given parasite
-count_data = as.data.frame(table(bp))
-full_data = data.frame(list("bp" = seq(0, max(bp), 1), "Freq" = rep(0, max(bp) + 1)))
-full_data[as.numeric(as.vector(count_data$bp)) + 1, "Freq"] = count_data$Freq
-obs = full_data$Freq
-
-
-# Get predicted numbers from negative binomial
-fits = fitdistr(bp, "negative binomial", list(size=.5, mu= 15), lower=.001)
-vec = seq(0, max(bp), 1)
-probs = dnbinom(vec, size=fits$estimate["size"], mu=fits$estimate["mu"])
-pred = probs * length(bp)  # Maybe set fractional raccoons to zero?
-
-
-# # Observed can't be greater than predicted...seems like a hack-
-# #is binomial the best assumed distribution- cant have obs >pred
-
-parasite_nums = seq(0, max(bp), 1)
-new_pred = round(pred)
-new_obs = obs
-ind = obs > new_pred
-new_obs[ind] = new_pred[ind]
-
-# Drop where pred is zero- removes fraction individuals
-# start at i=2 because dont include 0 class as should have 100%surv
-ind_nozeros = new_pred != 0
-trun_obs = new_obs[ind_nozeros][2:sum(ind_nozeros)]
-trun_pred = new_pred[ind_nozeros][2:sum(ind_nozeros)]
-trun_x = parasite_nums[ind_nozeros][2:sum(ind_nozeros)]
-
-
-# # Maximum likelihood function
-
-logistregfun = function(params, x, N, y){
-
-	a = params[1]
-	b = params[2]
-	p.pred = exp(a + b *log(x)) / (1 + exp(a + b * log(x)))
-	-sum(dbinom(y, size=N, prob=p.pred, log=TRUE))
-}
-
-opt1 = optim(fn = logistregfun, par = c(a=.5,b=.5),
-				x = trun_x, N=trun_pred, y=trun_obs,
-				method="BFGS")
-
-#a=opt1$par[1]
-#b=opt1$par[2]
-
-### GLM in R: Should give the same results at the direct optimization
-fits_glm = glm(cbind(trun_obs, trun_pred - trun_obs) ~
[…]

### 2015-04-20 code/power_analysis.R

@@ -1,77 +1,0 @@
-# Code for simulating parasite mortality...should put these all into functions
-
-# Load fxns
-source("analysis_fxns.R")
-
-# Parameters
-num_hosts = 10000
-k = .5
-mu = 15
-a = 6 # LD50 a
-b = -2  # LD50 b
-LD50 = exp(a / abs(b))
-
-survival_prob = function(x, a, b){
-
-    res = exp(a + b * log(x)) / (1 + exp(a + b * log(x)))
-    return(res)
-
-}
-
-ks = seq(0.01, 1, length=50)
-trun_vals = seq(0.1, 1, length=50)
-a_vals = matrix(NA, nrow=length(ks), ncol=length(trun_vals))
-b_vals = matrix(NA, nrow=length(ks), ncol=length(trun_vals))
-mort_ratio = matrix(NA, nrow=length(ks), ncol=length(trun_vals))
-
-mort_diff = matrix(NA, nrow=length(ks), ncol=length(trun_vals))
-
-for(i in 1:length(ks)){
-    for(j in 1:length(trun_vals)){
-
-        # 1. Equilibrium is established
-        sim_data = rnbinom(num_hosts, size=ks[i], mu=mu)
-        sorted_sim = as.data.frame(table(sim_data))
-        sim_para = as.numeric(levels(sorted_sim$sim_data)[as.integer(sorted_sim$sim_data)])
-        sim_freq = sorted_sim$Freq
-
-        # 2. Parasite induced mortality occurs
-        surv_probs = survival_prob(sim_data, a, b)
-        surv_probs[is.na(surv_probs)] = 1
-
-        # Determine if an individual dies or not
-        rands = runif(length(sim_data))
-        surv = rands < surv_probs
-        alive_hosts = sim_data[surv]
-
-        true_death = 1 - sum(surv) / length(surv)
-
-        trun_val = trun_vals[j] * LD50
-
-        ests1 = estimate_mortality(alive_hosts, plot=F, weights=T,
-                    trun=trun_val, force_it=T, save=FALSE,
-                        title="Observed Survival Curve")
-
-        a = ests1$fits$a
-        b = ests1$fits$b
-
-        a_vals[i, j] = a
-        b_vals[i, j] = b
-        mort_ratio[i, j] = ests1$fits$Z0 / true_death
-        mort_diff[i, j] = ests1$fits$Z0 - true_death
-
[…]

### 2015-04-20 code/pihm_methods.py
Added function for extracting simulation parameters

@@ -436,6 +436,83 @@ def likefxn2(params, x, mu, k):
     return -np.sum(np.log(pihm_pmf(x, mu, k, a, b)))
 
 
+def extract_simulation_results(sim_results, keys, method_name, param):
+    """
+
+    Method to extract parameters from a simulation dictionary as generated
+    by analysis scripts
+
+    Parameters
+    ----------
+    sim_results : dict
+        A dictionary with analysis results
+
+    keys : list
+        First item is a value for mup, second item is a tuple specify the
+        (a, b) pair, third iterm is kp.
+
+    method_name: str
+        Either likelihood or adjei
+
+    param: str
+        either "a", "b", or "ld50"
+
+    Returns
+    : tuple
+        samp_sizes, biases, precisions
+
+
+    """
+
+    sims = sim_results[keys[0]][keys[1]][keys[2]]
+
+    Np_vals = list(sims.viewkeys())
+
+    if method_name == "likelihood":
+        index = 0
+    else:
+        index = 1
+
+    samp_sizes = []
+    biases = []
+    precisions = []
+
+    for Np in Np_vals:
+
+        res = sims[Np]
+
+        try:
+            a_vals, b_vals = zip(*res[index])
+        except:
+            continue
+
+        N_vals = res[2]
+
+        samp_sizes.append(np.mean(N_vals))
+
+        if param == "a":
+
+            biases.append(scaled_bias(np.array(a_vals), keys[1][0]))
+            precisions.append(scaled_precision(np.array(a_vals)))
+
+        elif param == "b":
+
+            biases.append(scaled_bias(np.array(b_vals), keys[1][1]))
+            precisions.append(scaled_precision(np.array(b_vals)))
+
+        elif param == "ld50":
+
+            ld50_vals = np.exp(np.array(a_vals) / np.abs(b_vals))
+            truth = np.exp(keys[1][0] / np.abs(keys[1][1]))
+
+            biases.append(scaled_bias(ld50_vals, truth))
+            precisions.append(scaled_precision(ld50_vals))
+        else:
+            raise KeyError("Don't recognize parameter: should be a, b, or ld50")
+
+    return samp_sizes, biases, precisions
+
+
 def split_data(bin_edges):
     """
     Given bin edges, split data in fully enumerate bins. Example: [1, 4, 8]

### 2015-04-20 code/pihm_methods.py
Added PIHM tests for likelihood and adjei method

@@ -1,11 +1,13 @@
 from __future__ import division
 import numpy as np
 import macroeco.models as mod
+import macroeco.compare as comp
 import scipy.optimize as opt
 import pandas as pd
 import statsmodels.api as sm
 import pymc as pm
 import matplotlib.pyplot as plt
+from scipy.stats import chisqprob
 
 """
 Module contains the updated functions and classes for assessing
@@ -65,6 +67,13 @@ class PIHM:
         """
 
         return self.Np, self.mup, self.kp, self.a, self.b
+
+    def fit_nbd(self, k_array=np.linspace(0.01, 20, 100)):
+        """
+        Fit a NBD to the parasite data
+        """
+
+        return mod.nbinom.fit_mle(self.data, k_array=k_array)
 
 
     def crofton_method(self, crof_bin_edges, guess=None):
@@ -178,7 +187,7 @@ class PIHM:
 
 
     def adjei_method(self, full_bin_edges, crof_bin_edges, no_bins=True,
-                    run_crof=False):
+                    run_crof=False, test_sig=False):
         """
         Implements adjei method.  Truncated negative binomial method is fit using
         Crofton Method and Adjei method is used to estimate morality function.
@@ -199,6 +208,11 @@ class PIHM:
             be a bit careful with the upper bound because it is INCLUSIVE.
         run_crof : bool
             If True, runs the crofton method. Otherwise does not.
+        test_sig : bool
+            If True, returns an additional tuple with the null_deviance, the
+            deviance, the chi-squared value,  and the p-value of full model.
+            A p-value of less than 0.05 (or alpha) indicates that PIHM is
+            significant.
 
         Returns
         -------
@@ -224,7 +238,6 @@ class PIHM:
         full_theor_data = np.repeat(np.arange(0, max_worms + 1),
                 theor_data.astype(int))
         theor_binned = np.histogram(full_theor_data, bins=full_bin_edges)[0]
-
 
         # Put everything into self.dataframe
         if no_bins:
@@ -253,15 +266,23 @@ class PIHM:
         self.adjei_all_data = all_data
 
         # Fit the glm model
-        params = fit_glm(all_data)
+        params, null_deviance, deviance = fit_glm(all_data)
         self.a, self.b = params
         a = params[0]
         b = params[1]
 
-        return self.get_all_params()
+        if test_sig:
+
+            chi = null_deviance - deviance
+            p = chisqprob(chi, 1)
[…]

### 2015-04-21 docs/parasite_mortatlity_manuscript.tex
Fleshed out discussion, introduction, and results

@@
 experimentally infect and track host populations in ways that are rarely
-possible in most host-parasite systems (citations).  Instead, parasitologists
-are often only able to collect a certain number of hosts and determine each
-host's parasite intensity.  This snapshot, distributional data is far from the
+possible in most host-parasite systems \citep{McCallum2000a}.  Instead, parasitologists
+are often only able to determine the parasite intensity on some number of sampled hosts.  This snapshot, distributional data is far from the
 ideal type of data for addressing questions regarding PIHM, but the reality is
@@
 
-The first of these two questions was addressed by \cite{Crofton1971a} developed
+The first of these two questions was addressed by \cite{Crofton1971a} who developed
 a method to test for PIHM using the truncation of the negative binomial
@@
 distribution \citep{AndersonandMay1978,Shaw1998}.  As heavily infected hosts
-begin to die, the negative binomial distribution gets truncated because heavily
-infected hosts dies and are no longer observed in a sample. In other words, the
+begin to die, the negative binomial distribution is truncated because heavily
+infected hosts die and are no longer observed in a sample. In other words, the
 observed host-parasite distribution and the pre-mortality host-parasite
 distribution will predict substantially different numbers of of highly infected
-hosts (because those have died due to infection) but similar number of hosts with low infection
+hosts (because those have died due to infection), but similar number of hosts with low infection
 loads (because those have survived).  \citep{Crofton1971a} noted that by
@@
 and \cite{Adjei1986} proposed a method to determine how host survival
-probability changes with increasing parasite-load.  The Adjei Method proceeds
+probability changes with increasing parasite load.  The Adjei Method proceeds
 by first using the Crofton Method to estimate the pre-mortality parameters for
-a host-parasite distribution (describe what these are) and then, given these parameters, estimates a host-survival function that describes how the probability of host-survival changes with increasing parasite load (see Appendix X for a full description).
-From this host-survival function, the Adjei Method can estimate important host-parasite quantities such as the parasite intensity at which 50\% of hosts succumb to PIHM ($LD_{50}$), as well as the percent of hosts in a population succumbing to PIHM \citep{Adjei1986}.
+a host-parasite distribution (describe what these are) and then, given these
+parameters, estimates a host-survival function that describes how the
+probability of host-survival changes with increasing parasite load (see
+Appendix X for a full description).  From this host-survival function, the
+Adjei Method can estimate important host-parasite quantities such as the
+parasite intensity at which 50\% of hosts succumb to PIHM ($LD_{50}$), as well
+as the percent of hosts in a population succumbing to PIHM \citep{Adjei1986}. A more thorough description of the Adjei Method is given in Appendix X.
 
-Asking questions regardin
+Despite these methods both being over two decades old, they are still the
+primary means of answering questions about PIHM given distributional data
+\citep[\cite{Royce1990} but see][for an alternative to the Crofton
+Method]{Ferguson2011}. However, both methods have a few important limitations.
+The Crofton Method (and the method recently proposed by \cite{Ferguson2011})
+relies on a visual test to determine whether or not PIHM is occurring in a
+system.  With no clear decision rule, it can be difficult to determine the
+significance of PIHM in a given host-parasite system.  Moreover, previous
+studies have shown that the Crofton Method can fail to detect PIHM depending on
+the shape of the host survival function \citep{Lanciani1989}.
+
+In theory, the Adjei Method can provide a means to assess the significance of
+PIHM in a system as well as determine the shape of the host survival function.
+In practice, however, the Adjei Method has never been thoroughly tested and
+relies on a number of questionable data manipulations.  For example, in order
+for the Adjei Method to be used to estimate PIHM, the observed data must be
+altered; a less than ideal property for any statistical method.
+
+In this study, we propose a simple, novel method for estimating both whether or
+not PIHM is occurring in a host-parasite system and the shape of the survival
+function.  We compare the ability of our method to answer the two PIHM
+questions given above to the results given by the Adjei Method.  We show that
+our method is superior to the Adjei Method and that the Adjei Method has some
+serious deficiencies when trying to answer either of the above questions.
 
 \section*{Methods}
+
+\subsection*{The Adjei method for estimating PIHM}
 
@@
 To estimate these parameters, we need to define a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.
-Using the standard rules of conditional probability  This distribution can be written as
+Using the standard rules of conditional probability this distribution can be written as
 
@@
 
-Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, $b$ that maximize the likelihood of an observed host-parasite dataset $\mathbf{x}$.
+Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset $\mathbf{x}$.
 
@@
 
-We tested the likelihood method on the datasets given in \citep{Crofton1971a} and \citep{Adjei1986} which both papers reported seeing PIHM in the respective populations.
+We tested the likelihood method on the 6 datasets given in \cite{Crofton1971a}
+and the 4 datasets given in \citep{Adjei1986} in which both papers reported
+seeing PIHM in the respective populations.  We examined both whether or not the
+Likelihood Method or the Adjei Method detected significant PIHM in the given
+datasets and the resulting $LD_{50}$ predictions from both methods.  We then
+compared these results to whether the authors stated that PIHM was a
+significant in the given system.
 
@@
 
+The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$ for all combinations of parameters examined in this study (Figure X).
+Even small sample sizes (< 500 hosts), the Likelihood Methods estimate of $LD_{50}$ was largely unbiased, with small biases occurring for host survival functions that were gradual (i.e.
+$\mu_p = 10, a = 5, b = -2.5; \mu_p = 100, a = 5, b= -1.5; \mu_p = 500, a = 10, b = -1.8$)   On the other hand, the Adjei Method always produced more biased estimates of the $LD_{50}$ than the Likelihood Method across all parameter combinations.
+For $\mu_p = 10$, the $LD_{50}$ estimates from the Adjei Method were largely unbiased for large samples sizes, but showed increasing bias as sample size decreased, particularly for steep survival functions ($a = 20, b = -10$).
+As $\mu_p$ increased, the Adjei method produced biased estimates of $LD_{50}$ across all sample sizes, with bias increasing as sample size decreased.
+
+The precision of the $LD_{50}$ estimates for the Likelihood method decreased (increasing Coefficient of Variation) as sample size decreased for all parameter combinations we examined.
+The $LD_{50}$ estimates from the Adjei Method showed a similar pattern, with large decreases in precision occurring for the steepest survival function across all values of $\mu_p$ (Figure X).
+
+In terms of the host survival function, the Likelihood Method gave
+asymptotically unbiased estimates of $a$ and $b$ as sample size increased for
+all parameter combinations considered.  However, as sample size decreased, the
+likelihood method tended to produce severely biased estimates of $a$ and $b$.
+This was generally more pronounced for steeper survival functions and more
+aggregated pre-mortality distributions (Figure X).  The Adjei Method produced
+biased estimates of $a$ and $b$ across all sample sizes, with the bias
+consistently being larger when the survival function was steeper. The bias of
+the Adjei Method also increased as $\mu_p$ increased.
+
 \subsection*{Application to data}
@@
 Parasite-induced host mortality is of substantial interest in many systems, but determining whether it is occurring given only observational data is notoriously difficult.
-Many of the previous methods derived to determine the effect of PIHM on a host-parasite system limit their inference to answering the yes or no question of whether or not PIHM is occurring.
-While a relevant question, it is often of interest to know something about the host survival function which can provide information regarding important properties of the host-parasite system, such as $LD_{50}$ and percent of the population suffering PIHM.
-
-We show that the Adjei Method, the only currently proposed method to estimate the host survival function and the $LD_{50}$ from observational PIHM, has some serious methodological problems that result in biased estimates of the host survival function even under the most idealistic conditions.
-Interestingly, despite these flaws, the Adjei Method can still produce unbiased and precise estimates of the $LD_{50}$ when host-parasite systems show aggregation close to $k = 1$
+We show that the Adjei Method, the only currently proposed method to estimate the host survival function and the $LD_{50}$ from observational PIHM, has some serious methodological problems that result in biased estimates of the host survival function and the $LD_{50}$ even under the most idealistic conditions.
 
@@
 
-What do these findings tell us about our ability to go beyond saying whether or
-not PIHM is occurring in a system?  While we have generalized and improved upon
-the previously existing methods for estimating PIHM, we cannot belie the fact
-that estimating the host survival function from observational data alone is
-ladened with assumptions and difficulties.
-
-The most fundamental
+While we have generalized and improved upon
+the previously existing methods for answering questions about PIHM, we cannot belie the fact
+that estimating PIHM from observational data alone is
+ladened with assumptions and difficulties \citep{McCallum2000a}. The most fundamental
 assumption of all methods for estimating PIHM is that the shape of the pre-
 mortality host-parasite distribution is known. In the discrete case, this
-distribution is negative binomial, while in the continuous case the distribution can be gamma of exponential (\citep{Ferguson2011}).  While there is substantial empirical and
+distribution is negative binomial, while in the continuous case the distribution can be gamma or exponential (\citep{Ferguson2011}).  While there is substantial empirical and
 theoretical evidence to justify the use of the negative binomial distribution
-as the pre-mortality distribution for macroparasites across hosts (Crofton, Shaw, Anderson and Gordon, etc), it is widely recognized that different processes can lead to a variety of distributions of parasites across hosts (Wilber, Duerr etc).
+as the pre-mortality distribution for macroparasites across hosts \citep{Calabrese2011,Anderson1982a,Shaw1998}, it is widely recognized that different processes can lead to a variety of distributions of parasites across hosts (Wilber, Duerr etc).
 However, the critical assumption of the pre-mortality distribution is not that the processes leading the pre-mortality distribution generate a negative binomial distribution, but rather that the pre-mortality distribution is well-fit by a negative binomial.
-The extreme flexibility of the negative binomial distribution in the discrete case of the gamma distribution in the continuous case make them reasonable candidate distributions for the pre-mortality distributions.
+The extreme flexibility of the negative binomial distribution makes it a reasonable candidate distribution for the pre-mortality distributions.
 Therefore, we do not see this assumption as central problem in any of the proposed methods.
@@
 survival function and the shape of the post-mortality distribution.  Regarding
-the host-survival function, all methods of PIHM assume that the host-survival
+the host-survival function, all currently proposed methods of PIHM assume that the host-survival
 function is such that uninfected individuals and individuals with low parasite
@@
 by showing that when hosts experienced a linear decrease in survival
-probability the Crofton Method could not detect PIHM.  This result, relates to
-the shape of the post-mortality distribution.  Given that the pre-mortality
-distribution is well-fit by a negative binomial, a linear host-survival
-function will result in a post-mortality distribution that is also well-fit by
-a negative binomial distribution (why?).  In this case, one would be unable to
-identify PIHM because the pre-mortality distribution
+probability the Crofton Method could not detect PIHM.
+As the most fundamental models of host-parasite dynamics predict a linear decrease in host survival probability with increasing host load \citep{AndersonandMay1978}, the failure of these methods to detect this relationship is a significant disconnect between empirical and theoretical disease ecology.
 
-As parasitologist, we often want to get the most out of our
+Related to the above result, all of these methods require that the post-mortality distribution be significantly different from a negative binomial distribution.
+This is necessary because none of the above methods will be able to detect PIHM if a negative binomial distribution is an adequate fit to the post-mortality distribution (describe why?).
+As many observed host-parasite distributions are not significantly different from a negative binomial distribution, there are limited cases where these methods can even be considered.
+
+Finally, all of these methods assume that the truncation of a negative binomial
+distribution is due to PIHM, but simulation studies by \citep{Anderson1982a}
+show that a variety of other processes can lead to the truncation of a negative
+binomial distribution such as within host parasite density-dependence, age-
+dependent variation in host resistance and rate of infection
+\citep{McCallum2000a}.  Therefore, even detecting ``significant'' PIHM in a
+dataset does not mean that PIHM is cause of the truncation.
+
+Given these numerous caveats, is there a place in parasitology for methods that
+estimate PIHM from distributional data?  We are in agreement with
+\cite{Lester1984} that, at the very least, methods for estimating PIHM can at
+least provide some preliminary insight into whether or not PIHM is worth
+looking into.  However, we stress that these methods should only be used as an
+exploratory tool when assessing the role of PIHM in a system and potential
+users should critically evaluate whether they think they have a large enough
+sample sizes and appropriate host survival functions for the methods proposed
+in this paper to be applicable.  Even if they are applicable, inferring PIHM
+from distributional substitute for field or laboratory experiments and an in
+depth understanding of the natural history of the host-parasite system under
+question.
 

### 2015-04-21 code/analysis_for_manuscript_typeIandII.py
Added and updated script for the manuscript analysis

@@ -7,7 +7,7 @@ import pandas as pd
 Description
 -----------
 
-Full simulation for the bias and precision section of the parasite-induced
+Full simulation for type I and type II error section of the parasite-induced
 host mortality manuscript.
 
 This script looks at all combinations of parameters that I am using in the
@@ -35,6 +35,16 @@ mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], []),
 
 Nps = [300, 500, 1000, 2000, 5000, 7500, 10000]
 N_samp = 150
+
+# ks = [1]
+
+# mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], [])}
+
+#             # 100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
+#             # 500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
+
+# Nps = [300, 500, 1000, 2000, 5000, 7500, 10000]
+# N_samp = 1
 
 scenario_1_results = {}
 
@@ -66,21 +76,35 @@ for mup in mu_ld50.iterkeys():
         for i in xrange(N_samp):
 
           pihm.set_all_params(Np, mup, kp, ld50[0], ld50[1])
-          pihm.data = pihm.get_pihm_samples()[0]
 
-          samp_sizes.append(len(pihm.data))
+          data_post, data_pre = pihm.get_pihm_samples()
+
+          samp_sizes.append(len(data_post))
 
           try:
 
-            # Likelihood results
-            like_out = pihm.likelihood_method(full_fit=False, disp=False)[-2:]
+            # Test on pre-mort
+            pihm.data = data_pre
+
+            # Likelihood method
+            like_p_typeI = pihm.test_for_pihm_w_likelihood(fixed_pre=True,
+                                  disp=False)[1]
 
             # Adjei results
-            adjei_out = pihm.adjei_method([], [], no_bins=True,
-                                                run_crof=False)[-2:]
+            adjei_p_typeI = pihm.test_for_pihm_w_adjei([], [], no_bins=True)[1]
 
-            results_like.append(like_out)
-            results_adjei.append(adjei_out)
+            # Test on post-mort
+            pihm.data = data_post
+
+            # Likelihood method
+            like_p_typeII = pihm.test_for_pihm_w_likelihood(fixed_pre=True,
+                                disp=False)[1]
+
+            # Adjei results
[…]

### 2015-04-21 code/pihm_methods.py
Updated PIHM testing for adjei and likelihood method

@@ -275,7 +275,7 @@ class PIHM:
 
             chi = null_deviance - deviance
             p = chisqprob(chi, 1)
-            return (self.get_all_params(), (chi, p))
+            return (self.get_all_params(), (null_deviance, deviance))
 
         else:
             return self.get_all_params()
@@ -380,16 +380,32 @@ class PIHM:
     def test_for_pihm_w_adjei(self, full_bin_edges, crof_bin_edges,
                         no_bins=True, run_crof=False):
         """
+
+        Test a dataset for parasite induced host mortality using the likelihood
+        method
+
+        This method compares a reduced model (negative binomial distribution)
+        to a full model (Negative binomail with PIHM).  The two models are
+        nested and differ by two parameters: a and b.  This amounts to fitting
+        the model a negative binomial distribution to the data, then fitting
+        the full model to the data and comparing likelihoods using a likelihood
+        ratio test.  The likelihood ratio should be approximately chi-squared
+        with the dof equal to the difference in the parameters.
+
         """
 
-        param, hypoth = self.adjei_method(full_bin_edges, crof_bin_edges,
+        params, (null_deviance, deviance) = self.adjei_method(full_bin_edges, crof_bin_edges,
                             no_bins=no_bins, run_crof=run_crof, test_sig=True)
 
-        return hypoth
+        chi_sq = null_deviance - deviance
+        prob = chisqprob(chi_sq, 1)
+
+        return chi_sq, prob, null_deviance, deviance
 
 
     def test_for_pihm_w_likelihood(self, guess=[10, -5],
-                        k_array=np.linspace(0.05, 2, 100), fixed_pre=False):
+                        k_array=np.linspace(0.05, 2, 100),
+                        fixed_pre=False, disp=True):
         """
         Test a dataset for parasite induced host mortality using the likelihood
         method
@@ -408,7 +424,12 @@ class PIHM:
             Guesses for a and b
         k_array : array
             Array of values over which to search to best fit k
-        fixed_pre
+        fixed_pre : bool
+            If True, the premortality parameters are fixed (mup and kp). Else,
+            they are jointly estimated from the data
+        disp : bool
+            If True, convergence message is printed.  If False, no convergence
+            method is printed
 
 
         Returns
@@ -420,7 +441,8 @@ class PIHM:
         if not fixed_pre:
 
             # Get full nll
-            params = self.likelihood_method(full_fit=True, guess=guess)
[…]

### 2015-04-22 code/analysis_for_manuscript_test_data.py
Analyzed crofton and adjei data

@@ -1,0 +1,191 @@
+import pihm_methods as pm
+reload(pm)
+import numpy as np
+import pandas as pd
+
+"""
+Description
+-----------
+
+Testing the Adjei Method and the Likelihood Method against known data
+
+A couple different tests
+
+1. Run Crofton Method and then test for PIHM (Adjei and Likelihood)
+2. For Likelihood, fit with all parameters
+
+"""
+
+### Crofton Data for 6 different stations ###
+
+para_crof = np.arange(0, 9)
+
+# Data from Crofton 1971
+st1_obs = np.array([161, 111, 67, 65, 50, 30, 33, 13, 8])
+st2_obs = np.array([189, 129, 86, 51, 27, 14, 8, 1, 2])
+st3_obs = np.array([458, 81, 40, 22, 19, 4, 6, 3, 0])
+st4_obs = np.array([164, 147, 92, 43, 25, 11, 3, 0, 1])
+st5_obs = np.array([140, 77, 30, 14, 10, 3, 2, 0, 0])
+st6_obs = np.array([153, 29, 6, 2, 1, 0, 0, 0, 0])
+
+# Raw data
+st1_raw = np.repeat(para_crof, st1_obs)
+st2_raw = np.repeat(para_crof, st2_obs)
+st3_raw = np.repeat(para_crof, st3_obs)
+st4_raw = np.repeat(para_crof, st4_obs)
+st5_raw = np.repeat(para_crof, st5_obs)
+st6_raw = np.repeat(para_crof, st6_obs)
+
+all_crof_data = [st1_raw, st2_raw, st3_raw, st4_raw, st5_raw, st6_raw]
+
+bins = [0, 1, 2, 3, 4, 4.9]
+
+crof = pm.PIHM()
+
+test_like_full_crof = []
+test_like_full_ld50_crof = []
+test_like_crof = []
+test_like_ld50_crof = []
+test_adjei_crof = []
+test_adjei_ld50_crof = []
+
+for raw in all_crof_data:
+
+    crof.data = raw
+    crof.crofton_method(bins)
+    test_like_crof.append(crof.test_for_pihm_w_likelihood(fixed_pre=True)[1])
+
+    # Calculate the LD50 for the like
+    res1 = crof.likelihood_method(full_fit=False)
+    test_like_ld50_crof.append(np.exp(res1[-2] / np.abs(res1[-1])))
+
+    test_adjei_crof.append(crof.test_for_pihm_w_adjei([], [])[1])
+
[…]

### 2015-04-22 code/test_pihm_methods.py
Added unit test of Crofton method and likelihood pihm

@@ -14,7 +14,130 @@ import macroeco.models as mod
 class TestPIHM(TestCase):
 
     def test_crofton_method(self):
-        pass
+        """
+        Testing Crofton Method on data from Crofton 1971
+
+        """
+
+        chi_sq = lambda pred, obs: np.sum((obs - pred)**2 / pred)
+
+        para_crof = np.arange(0, 9)
+
+        crof = PIHM()
+
+        # Data from Crofton 1971
+        st1_obs = np.array([161, 111, 67, 65, 50, 30, 33, 13, 8])
+        st2_obs = np.array([189, 129, 86, 51, 27, 14, 8, 1, 2])
+        st3_obs = np.array([458, 81, 40, 22, 19, 4, 6, 3, 0])
+        st4_obs = np.array([164, 147, 92, 43, 25, 11, 3, 0, 1])
+        st5_obs = np.array([140, 77, 30, 14, 10, 3, 2, 0, 0])
+        st6_obs = np.array([153, 29, 6, 2, 1, 0, 0, 0, 0])
+
+        # Raw data
+        st1_raw = np.repeat(para_crof, st1_obs)
+        st2_raw = np.repeat(para_crof, st2_obs)
+        st3_raw = np.repeat(para_crof, st3_obs)
+        st4_raw = np.repeat(para_crof, st4_obs)
+        st5_raw = np.repeat(para_crof, st5_obs)
+        st6_raw = np.repeat(para_crof, st6_obs)
+
+        # Crofton predicted with truncations!
+        st1_pred = {
+            3: np.array([162.2, 103.9, 77.4, 60.4, 48.2, 38.9, 31.7, 26.0, 21.4]),
+            4: np.array([162.4, 103.4, 77.6, 61.2, 49.4, 40.5, 33.5, 27.9, 23.3]),
+            5: np.array([161.2, 107.0, 78.0, 58.7, 44.7, 34.4, 26.6, 20.7, 16.1]),
+            6: np.array([162.1, 104.6, 77.3, 59.5, 46.7, 37.1, 29.7, 23.9, 19.3]),
+            7: np.array([160.2, 108.8, 79.2, 58.9, 44.3, 33.6, 25.5, 19.5, 14.9]),
+            8: np.array([158.4, 111.7, 80.8, 59.0, 43.2, 31.7, 23.3, 17.2, 12.7])}
+
+        st2_pred = {
+            3: np.array([188.6, 130.7, 83.6 , 52.0, 31.9, 19.4, 11.7, 7.0, 4.2]),
[…]

### 2015-04-22 docs/parasite_mortatlity_manuscript.tex
Filled in results and added to discussion with overview of results

@@
 
-In this study, we propose a simple, novel method for estimating both whether or
+In this study, we propose a novel method for estimating both whether or
 not PIHM is occurring in a host-parasite system and the shape of the survival
@@
 questions given above to the results given by the Adjei Method.  We show that
-our method is superior to the Adjei Method and that the Adjei Method has some
+our method is superior to the Adjei Method and that the Adjei Method has
 serious deficiencies when trying to answer either of the above questions.
@@
 
-\subsection*{The Adjei method for estimating PIHM}
+\subsection*{The Adjei Method for estimating PIHM}
 
-\subsection*{The likelihood method for estimating PIHM}
+\subsection*{The Likelihood Method for estimating PIHM}
 
@@
 
-We tested the likelihood method on the 6 datasets given in \cite{Crofton1971a}
-and the 4 datasets given in \citep{Adjei1986} in which both papers reported
-seeing PIHM in the respective populations.  We examined both whether or not the
-Likelihood Method or the Adjei Method detected significant PIHM in the given
-datasets and the resulting $LD_{50}$ predictions from both methods.  We then
-compared these results to whether the authors stated that PIHM was a
-significant in the given system.
+We tested the ability of the Adjei Method and the Likelihood Method to identify
+PIHM on 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
+given in \cite{Adjei1986}. In the \cite{Crofton1971a} datasets, the host was
+the snail \emph{Gammarus pulex} which acts as the intermediate host for the
+acanthocephalan \emph{Polmorphus minutus}. In the \cite{Adjei1986} datasets,
+the hosts were two species of lizard fish \emph{Saurida tumbil} and
+\emph{Saurida undosquamis} that were infected by the cestode parasite
+\emph{Callitetrarhynchus gracilis}.  Males and females of both fish species
+were considered separately.
+
+In both these studies, the authors reported PIHM and we test whether the Adjei
+Method and or the Likelihood Method also predict PIHM. For the 6 datasets from
+\cite{Crofton1971a}, we truncated the data at 4 parasites, applied the Crofton
+Method to estimate the pre-mortality distribution, and then ran the Likelihood
+Method and Adjei Method using these pre-mortality parameters.  For the
+\cite{Adjei1986} datasets, we followed the same procedure as the authors and
+first truncated the data at 2 parasites and then fit the Crofton Method for the
+female fish of both species.  We then parameterized the male pre-mortality
+distributions for each species with the results from the females.  Finally, we
+applied the Adjei Method and the Likelihood Method to determine whether or not
+PIHM was significant for these species and compared our results to those given by the authors.
 
@@
 
-The precision of the $LD_{50}$ estimates for the Likelihood method decreased (increasing Coefficient of Variation) as sample size decreased for all parameter combinations we examined.
+The precision of the $LD_{50}$ estimates for the Likelihood Method decreased (increasing coefficient of variation) as sample size decreased for all parameter combinations we examined.
 The $LD_{50}$ estimates from the Adjei Method showed a similar pattern, with large decreases in precision occurring for the steepest survival function across all values of $\mu_p$ (Figure X).
@@
 
+Of the 10 datasets we considered, the previous authors claimed to detect PIHM
+in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
+from the pre-mortality parameters of the Crofton Method detected significant
+PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
+dataset in which the the Likelihood Method parameterized from the Crofton
+Method did not detect a significant effect of PIHM was the the Adjei dataset
+for \emph{S. tumbil}.  For this case there was a marginally significant effect
+of PIHM the likelihood ratio test giving $\chi^2_{df=2} = 5.34; p = 0.069$.
+
+The Adjei Method detected significant PIHM in 9 of the 10 datasets given (Table \ref{table:pihm}).
+This is consistent with the the previous results which show that the Adjei Method has a very high Type I error rate.
+The Likelihood Method in which all parameters were jointly estimated only predicted significant PIHM in 4 of the 10 datasets.
+
+Discuss LD50 values?
+
 \section*{Discussion}
 
-Parasite-induced host mortality is of substantial interest in many systems, but determining whether it is occurring given only observational data is notoriously difficult.
-We show that the Adjei Method, the only currently proposed method to estimate the host survival function and the $LD_{50}$ from observational PIHM, has some serious methodological problems that result in biased estimates of the host survival function and the $LD_{50}$ even under the most idealistic conditions.
+Parasite-induced host mortality is of substantial interest in many systems, but
+determining whether it is occurring given only observational data is
+notoriously difficult.  We show that the Adjei Method, the only currently
+proposed method to estimate the host survival function and the $LD_{50}$ from
+observational PIHM, has some serious methodological problems that result in
+biased estimates of the host survival function and the $LD_{50}$ even under the most idealistic conditions.
+Moreover, we show that the Adjei Method has a seriously inflated Type I error rate, meaning it will often detect PIHM even when it is not present.
 
@@
 asymptotically unbiased when estimating the host-survival function for all of
-the parameter space that we explored, but can produce seriously biased
-estimates of the host survival function for sample sizes typically observed in
-many host-parasite studies.  However, we found that the likelihood method
+the parameter space that we explored and we found that it
 produces unbiased and precise estimates of the $LD_{50}$ for small, realistic
 sample sizes.
+Moreover, this novel method has reasonable type I and type II error rates when attempting to predict whether or not PIHM is occurring in a system.
+However, we note that the Likelihood Method produces seriously biased
+estimates of the host survival function ($a$ and $b$) for sample sizes typically observed in many host-parasite studies and should not be used to estimate the exact shape of the host survival function.
 
-Moreover, the likelihood method...
+We fit both the Likelihood Method and the Adjei Method to empirical data to
+determine whether they could detect PIHM that had been previously reported
+based on visual assessments.  Consistent with our simulation results, we found
+that the Adjei Method tended to detect PIHM where it had not been previously
+reported, while the Likelihood Method's detection of PIHM was consistent with
+previously PIHM in a given dataset. Taken together, these results suggest that the Adjei Method is a fundamentally
+flawed method for detecting PIHM and describing features of the host survival
+function and we recommend using the Likelihood Method for detecting PIHM and describing attributes of the host survival function.
 
@@
 
-While we have generalized and improved upon
+While we improved upon
 the previously existing methods for answering questions about PIHM, we cannot belie the fact
@@
 Related to the above result, all of these methods require that the post-mortality distribution be significantly different from a negative binomial distribution.
-This is necessary because none of the above methods will be able to detect PIHM if a negative binomial distribution is an adequate fit to the post-mortality distribution (describe why?).
+This is necessary because none of the above methods will be able to detect PIHM if a negative binomial distribution is an adequate fit to the post-mortality distribution [describe why?].
 As many observed host-parasite distributions are not significantly different from a negative binomial distribution, there are limited cases where these methods can even be considered.
@@
 Finally, all of these methods assume that the truncation of a negative binomial
-distribution is due to PIHM, but simulation studies by \citep{Anderson1982a}
-show that a variety of other processes can lead to the truncation of a negative
+distribution is due to PIHM, but previous studies have shownthat a variety of other processes can lead to the truncation of a negative
 binomial distribution such as within host parasite density-dependence, age-
 dependent variation in host resistance and rate of infection
-\citep{McCallum2000a}.  Therefore, even detecting ``significant'' PIHM in a
+\citep{McCallum2000a,Anderson1982a,Rousset1996}.  Therefore, even detecting ``significant'' PIHM in a
 dataset does not mean that PIHM is cause of the truncation.
@@
 estimate PIHM from distributional data?  We are in agreement with
-\cite{Lester1984} that, at the very least, methods for estimating PIHM can at
-least provide some preliminary insight into whether or not PIHM is worth
-looking into.  However, we stress that these methods should only be used as an
+\cite{Lester1984} that, at the very least, methods for estimating PIHM can provide preliminary insight into whether or not PIHM is worth further exploration.
+However, we stress that these methods should only be used as an
 exploratory tool when assessing the role of PIHM in a system and potential
@@
 in this paper to be applicable.  Even if they are applicable, inferring PIHM
-from distributional substitute for field or laboratory experiments and an in
-depth understanding of the natural history of the host-parasite system under
-question.
+from distributional data is no substitute for field or laboratory experiments and an in
+depth understanding of the natural history of the host-parasite system under consideration.
 
@@
 
+\renewcommand{\arraystretch}{1.2}
+\begin{sidewaystable}
+
+    \caption{Comparison of the PIHM predictions of previously used host-parasite datasets to those given by the Adjei Method and the Likelihood Method.
+The first column specifies the identity of the dataset, the second column specifies whether or not the author indicated that PIHM was occurring in the system, the third column indicates whether or not the Likelihood Method with parameters from the Crofton Method detects significant PIHM, the third column indicates whether the Likelihood Method with jointly estimated pre-mortality parameters detects PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
+If a method detects significant PIHM the $LD_{50}$ is given is parentheses}
+
+    \centering
+    \begin{tabular}{l l p{3cm} p{3cm} l}
+
+    \hline\hline
+    Data Set & Author detected PIHM? & Likelihood Method w/ Crofton Parameters & Likelihood Method w/out Crofton Parameters & Adjei Method \\
+
+    \hline
+    Crofton, Station 1 & Yes & Yes (7.27) & Yes (8.00) & Yes (9.33) \\
+    Crofton, Station 2 & Yes & Yes (6.92) & No & Yes (14.95)\\
+    Crofton, Station 3 & Yes & Yes (5.93) & Yes (7.04) & Yes (5.98) \\
+    Crofton, Station 4 & No & No & No & Yes (7.99) \\
+    Crofton, Station 5 & No & No & No & Yes (10.58) \\
+    Crofton, Station 6 & No & No & No & No \\
+    Adjei, \emph{S. tumbil} female & Yes (5.7) & No  & No   & Yes (6.37) \\
+    Adjei, \emph{S. tumbil} male & Yes (3.4) & Yes (3.42) & Yes (3.28) & Yes (3.66)  \\
+    Adjei, \emph{S. undosquamis} female & Yes (3.2) & Yes (3.04) & No  & Yes (3.11) \\
+    Adjei, \emph{S. undosquamis} male & Yes (1.8) & Yes (1.83) & Yes (2.34) & Yes (1.78) \\
+
+    \end{tabular}
+    \label{table:pihm}
+
+\end{sidewaystable}
+
 \end{document}

### 2015-04-22 code/pihm_methods.py
Extracting type I and power simulation results

@@ -518,7 +518,8 @@ def likefxn2(params, x, mu, k):
     return -np.sum(np.log(pihm_pmf(x, mu, k, a, b)))
 
 
-def extract_simulation_results(sim_results, keys, method_name, param):
+def extract_simulation_results(sim_results, keys, method_name, param,
+                                alpha=0.05):
     """
 
     Method to extract parameters from a simulation dictionary as generated
@@ -537,7 +538,12 @@ def extract_simulation_results(sim_results, keys, method_name, param):
         Either likelihood or adjei
 
     param: str
-        either "a", "b", or "ld50"
+        either "a", "b", "ld50" or "p".  If the param is p, then the type I
+        and type II errors are extracted.  This is only applicable for
+        results from the type I and type II error simulations
+
+    alpha : float
+        The significance level.  Only used if param == "p"
 
     Returns
     : tuple
@@ -564,6 +570,8 @@ def extract_simulation_results(sim_results, keys, method_name, param):
         res = sims[Np]
 
         try:
+            # If the param is p, a_vals are type I p_vals
+            # and b_vals are type II p_vals
             a_vals, b_vals = zip(*res[index])
         except:
             continue
@@ -589,6 +597,15 @@ def extract_simulation_results(sim_results, keys, method_name, param):
 
             biases.append(scaled_bias(ld50_vals, truth))
             precisions.append(scaled_precision(ld50_vals))
+
+        elif param == "p":
+
+            # Type I error
+            biases.append(np.sum(np.array(a_vals) < alpha) / len(a_vals))
+
+            # Power
+            precisions.append(np.sum(np.array(b_vals) < alpha) / len(b_vals))
+
         else:
             raise KeyError("Don't recognize parameter: should be a, b, or ld50")
 

### 2015-04-22 code/pihm_methods.py
Manually added GLM constant

@@ -400,7 +400,7 @@ class PIHM:
         chi_sq = null_deviance - deviance
         prob = chisqprob(chi_sq, 1)
 
-        return chi_sq, prob, null_deviance, deviance
+        return chi_sq, prob, null_deviance, deviance, params
 
 
     def test_for_pihm_w_likelihood(self, guess=[10, -5],
@@ -461,7 +461,7 @@ class PIHM:
         chi_sq = 2 * (-full_nll - (-red_nll))
         prob = chisqprob(chi_sq, 2)
 
-        return chi_sq, prob, full_nll, red_nll
+        return chi_sq, prob, full_nll, red_nll, params
 
 
 def pihm_pmf(x, mup, kp, a, b, max_sum=5000):
@@ -652,9 +652,12 @@ def fit_glm(all_data):
     all_data = all_data[['emp', 'pred', 'para']].copy()
     all_data['diff'] = np.array(all_data.pred) - np.array(all_data.emp)
     all_data['log_para'] = np.log(all_data['para'])
-    new_data = sm.add_constant(all_data)
-    glm_fit = sm.GLM(new_data[['emp', 'diff']],
-                    new_data[['const', 'log_para']], family=sm.families.Binomial())
+
+    all_data['const'] = 1
+
+    glm_fit = sm.GLM(all_data[['emp', 'diff']],
+                    all_data[['const', 'log_para']],
+                    family=sm.families.Binomial())
     res = glm_fit.fit()
 
     return res.params, res.null_deviance, res.deviance

### 2015-04-22 code/analysis_for_manuscript_typeIandII.py
Testing type I and power analysis for why type I is high for large Np

@@ -26,25 +26,27 @@ Author : Mark Wilber
 """
 
 
-# Parameters
-ks = [1, 0.5, .1]
+# # Parameters
+# ks = [1, 0.5, .1]
 
-mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], []),
-            100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
-            500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
-
-Nps = [300, 500, 1000, 2000, 5000, 7500, 10000]
-N_samp = 150
-
-# ks = [1]
-
-# mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], [])}
-
-#             # 100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
-#             # 500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
+# mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], []),
+#             100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
+#             500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
 
 # Nps = [300, 500, 1000, 2000, 5000, 7500, 10000]
-# N_samp = 1
+# N_samp = 150
+
+ks = [1]
+
+mu_ld50 = {500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
+
+#{10: ([(5, -2.5), (10, -5), (20, -10)], [])}
+
+            # 100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
+            # 500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
+
+Nps = [300] #[300, 500, 1000, 2000, 5000, 7500, 10000]
+N_samp = 5
 
 scenario_1_results = {}
 
@@ -81,36 +83,44 @@ for mup in mu_ld50.iterkeys():
 
           samp_sizes.append(len(data_post))
 
-          try:
+          #try:
 
-            # Test on pre-mort
-            pihm.data = data_pre
+          # Test on pre-mort
+          pihm.data = data_pre
 
-            # Likelihood method
-            like_p_typeI = pihm.test_for_pihm_w_likelihood(fixed_pre=True,
-                                  disp=False)[1]
+          # Likelihood method
+          like_res_typeI = pihm.test_for_pihm_w_likelihood(fixed_pre=True,
[…]

### 2015-04-23 code/analysis_for_manuscript.py
Ran simulations with mu at 50 rather than 500

@@ -31,10 +31,25 @@ ks = [1, 0.5, .1]
 
 mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], []),
             100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
-            500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
+            50: ([(5, -1.4), (10, -2.8), (20, -5.6)], [])}
 
 Nps = [300, 500, 1000, 2000, 5000, 7500, 10000]
 N_samp = 150
+
+# ks = [1] # 0.5, .1]
+
+# mu_ld50 = {50: ([(5, -1.4), (10, -2.8), (20, -5.6)], [])}
+
+# # {100: ([(5, -1.15), (10, -2.3), (20, -4.6)], [])}
+
+# # {10: ([(5, -2.5), (10, -5), (20, -10)], []),
+# #             100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
+# #             500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
+
+# Nps = [300] # 500, 1000, 2000, 5000, 7500, 10000]
+# N_samp = 10
+
+
 
 scenario_1_results = {}
 

### 2015-04-23 code/analysis_for_manuscript_typeIandII.py

@@ -27,26 +27,15 @@ Author : Mark Wilber
 
 
 # # Parameters
-# ks = [1, 0.5, .1]
+# Parameters
+ks = [1, 0.5, .1]
 
-# mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], []),
-#             100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
-#             500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
+mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], []),
+            100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
+            50: ([(5, -1.4), (10, -2.8), (20, -5.6)], [])}
 
-# Nps = [300, 500, 1000, 2000, 5000, 7500, 10000]
-# N_samp = 150
-
-ks = [1]
-
-mu_ld50 = {500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
-
-#{10: ([(5, -2.5), (10, -5), (20, -10)], [])}
-
-            # 100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
-            # 500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
-
-Nps = [300] #[300, 500, 1000, 2000, 5000, 7500, 10000]
-N_samp = 5
+Nps = [50, 100, 200, 300, 400, 500] #[300, 500, 1000, 2000, 5000, 7500, 10000]
+N_samp = 150
 
 scenario_1_results = {}
 
@@ -83,44 +72,44 @@ for mup in mu_ld50.iterkeys():
 
           samp_sizes.append(len(data_post))
 
-          #try:
+          try:
 
-          # Test on pre-mort
-          pihm.data = data_pre
+            # Test on pre-mort
+            pihm.data = data_pre
 
-          # Likelihood method
-          like_res_typeI = pihm.test_for_pihm_w_likelihood(fixed_pre=True,
-                                disp=True, k_array=np.linspace(0.01, 2, 200))
+            # Likelihood method
+            like_res_typeI = pihm.test_for_pihm_w_likelihood(fixed_pre=True,
+                                  disp=False, k_array=np.linspace(0.01, 2, 200))
 
-          like_p_typeI = like_res_typeI[1] #(like_res_typeI[1], like_res_typeI[-1][-1])
[…]

### 2015-04-23 code/crofton_method.py
Added option fitting for crofton methods

@@ -840,8 +840,14 @@ def test_for_pihm(data, guess=[10, -5], crof_params=None):
     params = likelihood_method(data, crof_params=crof_params, guess=guess)
     full_nll = likefxn1(params, data)
 
-    mle_fit = mod.nbinom.fit_mle(data, k_array=np.linspace(.1, 2, 100))
-    red_nll = comp.nll(data, mod.nbinom(*mle_fit))
+
+    if crof_params:
+        mu, k = np.array(crof_params)[1:]
+        red_nll = comp.nll(data, mod.nbinom(mu, k))
+
+    else:
+        mle_fit = mod.nbinom.fit_mle(data, k_array=np.linspace(.1, 2, 100))
+        red_nll = comp.nll(data, mod.nbinom(*mle_fit))
 
     chi_sq = 2 * (-full_nll - (-red_nll))
     prob = chisqprob(chi_sq, 2)

### 2015-04-23 docs/parasite_mortatlity_manuscript.tex
Added table for empirical work, title, introduction, and most of results
Still need Figures

@@
 
+\title{Comparing methods for estimating parasite-induced host mortality from distributional data: implementation and limitations}
+\author{Mark Wilber, Sara Weinstein, and Cherie Briggs}
+
 \begin{document}
+
+\maketitle
 
@@
 
-Infectious agents can have major impacts on animal populations through changing
-population dynamics and stability (Dobson 1992), altering predator prey
-interactions (Joly 2004), and even causing species decline and extinction
-(Viana 2015, Daszak 1999).  Accurate estimates of parasite-induced host
-mortality (PIHM) in wild animals are important for understanding what regulates both
-host and parasite populations and to make predictions about disease
-transmission in natural systems. Although a negative impact on host fitness is
-a fundamental component of parasitism (Lafferty and Kuris 2002, Poulin book,
-other definition sources), it is notoriously difficult to quantify PIHM in wild animal populations (cite, McCallum 2008?, lester
-1984).
+Infectious agents [parasites?] can have major impacts on animal populations
+through changing population dynamics and stability \citep{Dobson1992}, altering
+predator prey interactions \citep{Joly2004}, and even causing species decline
+and extinction \citep{DeCastro2005a,McCallum2012b}. Accurate estimates of
+parasite-induced host mortality (PIHM) in wild animals are important for
+understanding what regulates both host and parasite populations and to make
+predictions about disease transmission in natural systems. Although a negative
+impact on host fitness is a fundamental component of parasitism
+\citep{Lafferty2002} [more], it is notoriously difficult to quantify PIHM in
+wild animal populations \citep{Lester1984,McCallum2000a}.  While both macro and
+microparasites have detrimental and possibly fatal effects on hosts, for the
+remainder of this paper we limit our discussion to macroparasites that can be
+discretely counted within a host.
+
+[More discussion of PIHM before the nitty-gritty?]
 
 To conclusively identify PIHM in wild animal populations, it is necessary to
-experimentally infect and track host populations in ways that are rarely
-possible in most host-parasite systems \citep{McCallum2000a}.  Instead, parasitologists
+experimentally infect and track host populations; a method that is rarely
+possible in most wild animal systems \citep{McCallum2000a}.  Instead, parasitologists
 are often only able to determine the parasite intensity on some number of sampled hosts.  This snapshot, distributional data is far from the
-ideal type of data for addressing questions regarding PIHM, but the reality is
-that this is the type of data on which most questions regarding PIHM are asked
-(citations). The two primary questions that one may wish to ask given a snapshot host-
-parasite dataset are: 1) Is PIHM occurring in this system? and 2) How does
-host survival change as parasite intensity increases?  MORE
+ideal type of data for addressing questions regarding PIHM, but this is the type of data on which most questions regarding PIHM are asked \citep[e.g.][]{Ferguson2011,Royce1990,Lanciani1989,Lester1984,Lester1977}.
+In particular, parasitologists are often interested in asking two questions from this data: 1) Is PIHM occurring in this system?
+and if PIHM is occurring 2) Can we quantify the way in which parasite intensity is affecting host survival?
 
-The first of these two questions was addressed by \cite{Crofton1971a} who developed
-a method to test for PIHM using the truncation of the negative binomial
+The first of these PIHM questions was addressed by \cite{Crofton1971a} who developed
+a method to test for the presence of PIHM using the truncation of the negative binomial
 distribution.  In short, the Crofton Method assumes that the distribution of
@@
 distribution \citep{AndersonandMay1978,Shaw1998}.  As heavily infected hosts
-begin to die, the negative binomial distribution is truncated because heavily
-infected hosts die and are no longer observed in a sample. In other words, the
+begin to die, the negative binomial distribution is truncated and these hosts are no longer observed in a sample. In other words, the
 observed host-parasite distribution and the pre-mortality host-parasite
@@
 hosts (because those have died due to infection), but similar number of hosts with low infection
-loads (because those have survived).  \citep{Crofton1971a} noted that by
+loads (because those have survived).  \cite{Crofton1971a} noted that by
 starting with all the observed data and iteratively fitting a negative binomial
 distribution to hosts with lower and lower parasite loads, one could determine
-whether or not PIHM was occurring in the system (Figure XA).  This was done
-graphically by determining whether parameters of the truncated negative binomial distributions showed a substantial change as hosts with smaller and smaller parasite intensities were fitted (Figure XB).
-The Crofton Method and this graphical technique for determining whether or not PIHM is occurring are both still used \citep{Ferguson2011}.
-A more thorough description and implementation of the Crofton Method is given in Appendix X.
+whether or not PIHM was occurring in the system and estimate the parameters of the pre-mortality distribution.  This was done
+graphically by determining whether parameters of the truncated negative binomial distributions showed a substantial change as hosts with smaller and smaller parasite intensities were fitted (Figure X).
+The Crofton Method and this graphical technique are both still currently used \citep{Ferguson2011}.
+A more thorough description and implementation of the Crofton Method is given in Appendix X and we discuss the validity of its assumptions in the \emph{Discussion}.
 
-The second question regarding PIHM is how host survival changes as parasite
-intensity increases. The Crofton Method on its own cannot answer this question
-and \cite{Adjei1986} proposed a method to determine how host survival
-probability changes with increasing parasite load.  The Adjei Method proceeds
+The second question regarding PIHM attempts to quantify how parasite intensity affects host survival.
+This question asks substantially more from the data and \cite{Adjei1986} proposed a method to attempt to answer it.
+The Adjei Method proceeds
 by first using the Crofton Method to estimate the pre-mortality parameters for
-a host-parasite distribution (describe what these are) and then, given these
-parameters, estimates a host-survival function that describes how the
+a host-parasite distribution and then, given these
+parameters, estimates a host survival function that describes how the
 probability of host-survival changes with increasing parasite load (see
-Appendix X for a full description).  From this host-survival function, the
+\emph{Methods}).  From this host survival function, the
 Adjei Method can estimate important host-parasite quantities such as the
@@
 
-Despite these methods both being over two decades old, they are still the
+Despite these methods both being over three decades old, they are still the
 primary means of answering questions about PIHM given distributional data
-\citep[\cite{Royce1990} but see][for an alternative to the Crofton
+\citep[but see][for an alternative to the Crofton
 Method]{Ferguson2011}. However, both methods have a few important limitations.
-The Crofton Method (and the method recently proposed by \cite{Ferguson2011})
-relies on a visual test to determine whether or not PIHM is occurring in a
+The Crofton Method, and more recent methods \citep{Ferguson2011},
+rely on a visual test to determine whether or not PIHM is occurring in a
 system.  With no clear decision rule, it can be difficult to determine the
-significance of PIHM in a given host-parasite system.  Moreover, previous
-studies have shown that the Crofton Method can fail to detect PIHM depending on
-the shape of the host survival function \citep{Lanciani1989}.
+significance of PIHM in a given host-parasite system and use consistent criteria for identifying PIHM across systems.
+In theory, the Adjei Method can ameliorate this problem.
+In addition to providing information on the host-survival function, this method can also be used to assess the significance of
+PIHM in a system.
+In practice, however, the Adjei Method has never been thoroughly tested and
+relies on a number of questionable data manipulations.
 
-In theory, the Adjei Method can provide a means to assess the significance of
-PIHM in a system as well as determine the shape of the host survival function.
-In practice, however, the Adjei Method has never been thoroughly tested and
-relies on a number of questionable data manipulations.  For example, in order
-for the Adjei Method to be used to estimate PIHM, the observed data must be
-altered; a less than ideal property for any statistical method.
-
-In this study, we propose a novel method for estimating both whether or
-not PIHM is occurring in a host-parasite system and the shape of the survival
-function.  We compare the ability of our method to answer the two PIHM
-questions given above to the results given by the Adjei Method.  We show that
-our method is superior to the Adjei Method and that the Adjei Method has
-serious deficiencies when trying to answer either of the above questions.
+In this study, we have
+three primary goals. First we wish to test reliability of the Adjei Method for
+answering the aforementioned questions regarding PIHM.  Second, we propose a
+novel method for answering these questions and compares its efficacy against the Adjei Method.  Third, we
+discuss the limitations of inferring PIHM from distributional data alone and
+whether any method for inferring PIHM has a place in the future of
+parasitology.
 
@@
 
+The Adjei Method for estimating PIHM has two steps.  The first
+step is to estimate the parameters of the pre-mortality host-parasite
+distribution using the Crofton Method.  The three parameters estimated are the
+total number of hosts before mortality $N_p$,  the mean number of parasites per
+host before mortality $\mu_p$, and the aggregation of parasites before
+mortality given by the parameter $k_p$ from a negative binomials distribution.
+When $k_p$ is small, parasites are highly aggregated within hosts and when
+$k_p$ is large parasites are more evenly spread among hosts \citep{Wilson2002}.
+The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][]{Royce1990,Lester1984} and we provide a tested implementation of the method in Appendix X.
+
+The second step of the Adjei Method is to estimate the shape of the host survival function. \cite{Adjei1986} assume that the host survival function follows the logistic form
+
+\begin{equation}
+    h(x | a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+    \label{eq:logistic}
+\end{equation}
+
+where $x$ is the parasite intensity in a given host and $a$ and $b$ are the two
+parameters of the function. Generally, a larger $a$ allows for hosts to
+tolerate larger parasite loads before experiencing parasite-induced mortality
+and a larger $b$ leads to a more rapid decline in the probability of host
+survival as parasite intensity increases.  The value $\exp(a / b)$ is referred
+to as the $LD_{50}$, which gives the parasite intensity at which 50\% of hosts
+experience mortality.
+
+To estimate this function the Adjei Method proceeds as follows.  First, it
+calculates the expected number of hosts with a given parasite load $x$ by using
+the equation $g(x ; \mu_p, k_p) * N_p$ where $g(x ; \cdot)$ is the negative binomial pre-
+mortality distribution.  Second, the observed and predicted number of hosts
+with $x$ parasites are paired as a single data point and the method then assumes that
+this data point follows a binomial distribution with the total number of
+``trials'' equal to the predicted number of hosts and the total number of
+``successes'' equal to the observed number of hosts. In some cases, the
+observed number of hosts is greater than the expected number of hosts and the
+Adjei Method alters the data so that the observed is equal to the predicted
+\citep{Adjei1986}.
+After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation \ref{eq:logistic}.
+This model provides estimates for parameters $a$, $b$ and $LD_{50}$.
+
+While not included in the original implementation of the Adjei Method, a
+$\chi^2$ test with a degrees of freedom of 1 can be used to assess whether a GLM model that includes parasite
+intensity as a predictor of host survival probability is a ``better'' model than a
+GLM without this predictor.  This allows the Adjei Method to determine whether
+PIHM is a significant factor in a host-parasite system.
+
+The Adjei Method's most glaring deficiency is the need to alter the observed
+data in order to fit the model into the binomial GLM framework.  A second more
+subtle problem with the Adjei Method is the potential need to bin data in order
+to predict greater than one host in a given parasite intensity class.  For
+example, if the total number of hosts pre-mortality was 50, the mean number of
+parasites per host pre-mortality was 100 and the aggregation parameter was
+$k_p$, applying the equation $g(x ; \mu_p=100, k_p=1) * 50$ would result in
+less than 1 individual in all parasite intensities $x$. In other words, the
+Adjei Method cannot be applied to samples with either very high mean parasite
+loads, small sample sizes, or both without some sort of binning of the data.
+While this is not a flaw ``per se'', it does add a certain level of
+subjectivity (i.e. which bins should you use?) to a method that already has
+serious potential issues.  In this analysis, we always assume the Adjei Method is not binning the data, though we provide code for applying the binning method in Appendix X.
+
 \subsection*{The Likelihood Method for estimating PIHM}
@@
 approach for estimating parasite-induced host mortality (PIHM) that makes less
-assumption than the previously proposed Adjei Method.  The likelihood method
+assumptions and provides more reliable answers to the PIHM questions outlined above.  The Likelihood Method we propose
 does not require any binning or alteration of the data, potentially reduces the
@@
 
-The likelihood method makes the following assumption about the host-parasite
-system. First, as with all previously proposed methods for estimating PIHM, the
-likelihood method assumes that the pre-mortality distribution follows a
-negative binomial distribution ($g(x)$) with parameter $\mu_p$ and $k_p$. The
-validity of the assumption is an inherent problem with all the PIHM methods
-proposed to date and we address this thoroughly in the Discussion. The second
+As with all previously proposed methods for estimating PIHM, the
+Likelihood Method first assumes that the pre-mortality distribution follows a
+negative binomial distribution $g(x; \mu_p, k_p)$.  The second
 assumption that the likelihood method makes is that the host survival function
-takes the form of a logistic curve given by
-
-\begin{equation}
-    h(x | a, b) = h_x = \dfrac{e^{a + b \log(x)}}{1 + e^{a + b \log(x)}}
-    \label{eq:logistic}
-\end{equation}
-
-where $x$ is the parasite intensity in a given host and $a$ and $b$ are the two parameters of the function.
-Generally, a larger $a$ allows for hosts to tolerate larger parasite loads before experiencing parasite-induced mortality and a more negative $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases.
-The value $exp(a / |b|)$ (notation) is typically referred to as the $LD_50$, which gives the parasite intensity at which 50\% of host experience mortality.
-With these two explicit assumptions, the likelihood method tries to estimate the 4 parameters $\mu_p$, $k_p$, $a$, and $b$.
+takes the form of a logistic curve given by equation \ref{eq:logistic}.  With these two explicit assumptions, the Likelihood Method estimates the 4 parameters $\mu_p$, $k_p$, $a$, and $b$.
 
 To estimate these parameters, we need to define a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.
-Using the standard rules of conditional probability this distribution can be written as
+Using standard rules of conditional probability this distribution can be written as
 
@@
 
-One can now see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(survival) = \sum_0^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$.
+One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_0^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$.
 Therefore equation \ref{eq:concept} can be written as
@@
 \begin{equation}
-    P(x | survival) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
+    P(x | \text{survival}) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
     \label{eq:dist}
@@
 
-Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset $\mathbf{x}$.
+Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset.
+Alternatively, one could apply the Crofton Method to estimate $\mu_p$ and $k_p$ and then find the maximum likelihood estimates of $a$ and $b$ and the corresponding $LD_{50}$.
 
-Alternatively, one could apply the Crofton Method to estimate $\mu_p$ and $k_p$ and then find the maximum likelihood estimates of $a$ and $b$ and the corresponding $LD_{50}$.
-A final option would be to follow the example of \citep{Ferguson2011} and assume $k_p = 1$ and only estimate $a$, $b$ and $\mu_p$.
+To estimate that significance of PIHM in a host-parasite system, can use a
+likelihood ratio test in which the full model is given by equation
+\ref{eq:dist} and the reduced model is given by a negative binomial
+distribution.  If PIHM is significant in the system, the resulting likelihood
+ratio statistic should be greater $\chi^2_{2,\text{ } \alpha=0.05}$.
 
-\subsection{Testing the ability of approaches to identify PIHM}
+\subsection*{Testing the ability of approaches to identify PIHM}
 
@@
 
-We used three different values of $\mu_p$ (10, 100, 500) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and sharp decreases in host survival with increasing parasite intensity.
+We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and sharp decreases in host survival with increasing parasite intensity.
 For a given $\mu_p$, each survival function had the same $LD_{50}$, but different values of $a$ and $b$ (Table X).
 We examined each $\mu_p$-survival function pair at  three levels of parasite
-aggregation, $k_p = 0.1, 0.5, 1$, which are realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
-combinations we simulated 150 datasets and tested the probability of each method incorrectly identifying PIHM in the pre-mortality dataset (Type I error) and incorrectly failing to identify PIHM in the post-mortality dataset (Type II error).
-For each method, we we used likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level 0.05 (Appendix X).
-We tested all each parameter combinations for pre-mortality population sizes of $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
-10000].
+aggregation, $k_p = 0.1, 0.5, 1$ - realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
+combinations we simulated 150 datasets and tested the probability of each method incorrectly identifying PIHM in the pre-mortality dataset (type I error) and correctly identifying PIHM in the post-mortality dataset (power).
+For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level $\alpha = 0.05$.
+We tested all each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500].
+$N_p$ is not technically the sample size on which the methods are being
+tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
+dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.
 
@@
 
-To compare the ability of the Adjei Method and the likelihood method to recover $LD_{50}$ and the parameters $a$ and $b$ or the survival function, we used the same simulation procedure and parameter combinations described above.For each parameter
+To compare the ability of the Adjei Method and the Likelihood Method to recover $LD_{50}$ and the parameters $a$ and $b$ or the survival function, we used the same simulation procedure and parameter combinations described above.For each parameter
 combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
 precision \citep{Walther2005} for these estimates over varying pre-mortality host population sizes  $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
-10000]. $N_p$ is not technically the sample size on which the methods are being
-tested because parasite-induced mortality reduces $N_p$ for each simulated
-dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.
+10000]. We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.
 
@@
 PIHM on 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
-given in \cite{Adjei1986}. In the \cite{Crofton1971a} datasets, the host was
+given in \cite{Adjei1986} (Table X). In the \cite{Crofton1971a} datasets, the host was
 the snail \emph{Gammarus pulex} which acts as the intermediate host for the
@@
 In both these studies, the authors reported PIHM and we test whether the Adjei
-Method and or the Likelihood Method also predict PIHM. For the 6 datasets from
+Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
 \cite{Crofton1971a}, we truncated the data at 4 parasites, applied the Crofton
@@
 
+The Adjei Method showed highly inflated Type I error (i.e. falsely detected
+PIHM where there was no PIHM) for all parameter combinations that we
+considered.  This method also showed the unintuitive pattern of Type I error
+rate decreasing as sample size decreased.  This pattern is due to the issue of
+binning discussed in the \emph{Methods}. For small samples sizes, the
+applicability of the Adjei Method is compromised without binning the observed
+data in some way.  On the other hand, the Likelihood Method showed a type I
+error rate at or near the pre-set level of 0.05 for all parameter combinations
+and sample sizes considered.
+
+The ability of the Adjei Method to detect PIHM given that it was occurring in a
+system (i.e. the power) was close to 1 for larger sample sizes and tended to
+decrease as sample size decreased.  The Likelihood Method was a power close to
+one for all parameter combinations and sample sizes considered.  With gradual
+survival functions (black lines; Figure X), the Likelihood Method showed a
+decreased ability to detect PIHM for small sample sizes.
+
 \subsection*{Recovery of the survival function}
 
-The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$ for all combinations of parameters examined in this study (Figure X).
-Even small sample sizes (< 500 hosts), the Likelihood Methods estimate of $LD_{50}$ was largely unbiased, with small biases occurring for host survival functions that were gradual (i.e.
-$\mu_p = 10, a = 5, b = -2.5; \mu_p = 100, a = 5, b= -1.5; \mu_p = 500, a = 10, b = -1.8$)   On the other hand, the Adjei Method always produced more biased estimates of the $LD_{50}$ than the Likelihood Method across all parameter combinations.
-For $\mu_p = 10$, the $LD_{50}$ estimates from the Adjei Method were largely unbiased for large samples sizes, but showed increasing bias as sample size decreased, particularly for steep survival functions ($a = 20, b = -10$).
-As $\mu_p$ increased, the Adjei method produced biased estimates of $LD_{50}$ across all sample sizes, with bias increasing as sample size decreased.
+The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$
+for all combinations of parameters examined in this study (Figure X).  Even
+small sample sizes (< 500 hosts), the Likelihood Methods estimate of $LD_{50}$
+was largely unbiased, with small biases occurring for host survival functions
+that were gradual (i.e. $\mu_p = 10, a = 5, b = -2.5; \mu_p = 100, a = 5, b=
+-1.5; \mu_p = 50, a = 5, b = -1.4$)   On the other hand, the Adjei Method
+always produced more biased estimates of the $LD_{50}$ than the Likelihood
+Method across all parameter combinations.  For $\mu_p = 10$, the $LD_{50}$
+estimates from the Adjei Method were largely unbiased for large samples sizes,
+but showed increasing bias as sample size decreased, particularly for steep
+survival functions ($a = 20, b = -10$).  As $\mu_p$ increased, the Adjei method
+produced biased estimates of $LD_{50}$ across all sample sizes, with bias
+increasing as sample size decreased.
 
-The precision of the $LD_{50}$ estimates for the Likelihood Method decreased (increasing coefficient of variation) as sample size decreased for all parameter combinations we examined.
-The $LD_{50}$ estimates from the Adjei Method showed a similar pattern, with large decreases in precision occurring for the steepest survival function across all values of $\mu_p$ (Figure X).
+The precision of the $LD_{50}$ estimates for the Likelihood Method decreased
+(increasing coefficient of variation) as sample size decreased for all
+parameter combinations we examined.  The $LD_{50}$ estimates from the Adjei
+Method showed a similar pattern, with large decreases in precision occurring
+for the steepest survival function across all values of $\mu_p$ (Figure X).
 
@@
 all parameter combinations considered.  However, as sample size decreased, the
-likelihood method tended to produce severely biased estimates of $a$ and $b$.
+Likelihood Method tended to produce severely biased estimates of $a$ and $b$.
 This was generally more pronounced for steeper survival functions and more
@@
 
-Of the 10 datasets we considered, the previous authors claimed to detect PIHM
+Of the 10 datasets we considered, the previous authors visually detect PIHM
 in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
@@
 Method did not detect a significant effect of PIHM was the the Adjei dataset
-for \emph{S. tumbil}.  For this case there was a marginally significant effect
-of PIHM the likelihood ratio test giving $\chi^2_{df=2} = 5.34; p = 0.069$.
+for \emph{S. tumbil}.  For this dataset there was a marginally significant effect
+of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$).
 
@@
 This is consistent with the the previous results which show that the Adjei Method has a very high Type I error rate.
-The Likelihood Method in which all parameters were jointly estimated only predicted significant PIHM in 4 of the 10 datasets.
-
-Discuss LD50 values?
+The Likelihood Method in which all parameters were jointly estimated predicted significant PIHM in 4 of the 10 datasets.
 
@@
 
-Parasite-induced host mortality is of substantial interest in many systems, but
-determining whether it is occurring given only observational data is
+Determining whether PIHM is a significant factor for a host populations is critically important in many systems, but detecting and describing PIHM given only observational data is
 notoriously difficult.  We show that the Adjei Method, the only currently
@@
 Moreover, we show that the Adjei Method has a seriously inflated Type I error rate, meaning it will often detect PIHM even when it is not present.
+Moreover, for small, realistic sample sizes the Adjei Method behaves erratically; a consequence of the need to subjectively bin the data in order to predict parasite intensity classes with at least one host.
 
@@
 asymptotically unbiased when estimating the host-survival function for all of
-the parameter space that we explored and we found that it
-produces unbiased and precise estimates of the $LD_{50}$ for small, realistic
-sample sizes.
-Moreover, this novel method has reasonable type I and type II error rates when attempting to predict whether or not PIHM is occurring in a system.
-However, we note that the Likelihood Method produces seriously biased
-estimates of the host survival function ($a$ and $b$) for sample sizes typically observed in many host-parasite studies and should not be used to estimate the exact shape of the host survival function.
+the parameter space that we explored and we found that it produces unbiased and
+precise estimates of the $LD_{50}$ for small, realistic sample sizes. Moreover,
+this novel method has a type I error rate close to the pre-set level of $\alpha = 0.05$ and high power for detecting PIHM, given it is occurring, for realistic samples sizes.
+  However, we note that
+the Likelihood Method produces seriously biased estimates of the host survival
+function ($a$ and $b$) for sample sizes typically observed in many host-
+parasite studies.  The bias was most severe for steep host survival functions,
+due to large changes in the values of $a$ and $b$ only slightly changing an
+already steep survival function.  Given these results, neither the
+Likelihood Method or the Adjei Method could confidently recover the exact shape of the host survival function for small, realistic sample sizes.
 
-We fit both the Likelihood Method and the Adjei Method to empirical data to
+We also fit both the Likelihood Method and the Adjei Method to empirical data to
 determine whether they could detect PIHM that had been previously reported
@@
 reported, while the Likelihood Method's detection of PIHM was consistent with
-previously PIHM in a given dataset. Taken together, these results suggest that the Adjei Method is a fundamentally
+previously reported PIHM in a given dataset.  Taken together, these results suggest that the Adjei Method is a fundamentally
 flawed method for detecting PIHM and describing features of the host survival
@@
 
-% Discuss all of the problems with these methods
-
-While we improved upon
-the previously existing methods for answering questions about PIHM, we cannot belie the fact
-that estimating PIHM from observational data alone is
-ladened with assumptions and difficulties \citep{McCallum2000a}. The most fundamental
-assumption of all methods for estimating PIHM is that the shape of the pre-
-mortality host-parasite distribution is known. In the discrete case, this
-distribution is negative binomial, while in the continuous case the distribution can be gamma or exponential (\citep{Ferguson2011}).  While there is substantial empirical and
-theoretical evidence to justify the use of the negative binomial distribution
-as the pre-mortality distribution for macroparasites across hosts \citep{Calabrese2011,Anderson1982a,Shaw1998}, it is widely recognized that different processes can lead to a variety of distributions of parasites across hosts (Wilber, Duerr etc).
-However, the critical assumption of the pre-mortality distribution is not that the processes leading the pre-mortality distribution generate a negative binomial distribution, but rather that the pre-mortality distribution is well-fit by a negative binomial.
-The extreme flexibility of the negative binomial distribution makes it a reasonable candidate distribution for the pre-mortality distributions.
-Therefore, we do not see this assumption as central problem in any of the proposed methods.
+While we improved upon the previously existing methods for answering questions
+about PIHM, we cannot belie the fact that estimating PIHM from observational
+data alone is ladened with assumptions and difficulties \citep{McCallum2000a}.
+The most fundamental assumption of all methods for estimating PIHM is that the
+shape of the pre-mortality host-parasite distribution is known. In the
+discrete case, this distribution is negative binomial, while in the continuous
+case the distribution can be gamma or exponential \citep{Ferguson2011}.
+While there is substantial empirical and theoretical evidence to justify the
+use of the negative binomial distribution as the pre-mortality distribution for
+macroparasites across hosts \citep{Calabrese2011,Anderson1982a,Shaw1998}, it is
+widely recognized that different processes can lead to a variety of
+distributions of parasites across hosts \citep{Isham1995,Grenfell1995b,Wilson2002, Duerr2003}.  However, the
+critical assumption of the pre-mortality distribution is not that the processes
+leading to the pre-mortality distribution generate a negative binomial
+distribution, but rather that the pre-mortality distribution is well-fit by a
+negative binomial. The extreme flexibility of the negative binomial
+distribution makes it a reasonable candidate distribution for the pre-mortality
+distributions.  Therefore, we do not see this assumption as central problem in
+any of the proposed methods.
 
@@
 survival function and the shape of the post-mortality distribution.  Regarding
-the host-survival function, all currently proposed methods of PIHM assume that the host-survival
-function is such that uninfected individuals and individuals with low parasite
-intensity experience essentially no PIHM.  \cite{Lanciani1989} illustrated this
-by showing that when hosts experienced a linear decrease in survival
-probability the Crofton Method could not detect PIHM.
-As the most fundamental models of host-parasite dynamics predict a linear decrease in host survival probability with increasing host load \citep{AndersonandMay1978}, the failure of these methods to detect this relationship is a significant disconnect between empirical and theoretical disease ecology.
+the host-survival function, all currently proposed methods of PIHM assume that
+the host-survival function is such that uninfected individuals and individuals
+with low parasite intensity experience essentially no PIHM.
+\cite{Lanciani1989} illustrated the importance of this assumption by showing
+that when hosts experienced a linear decrease in survival probability the
+Crofton Method could not detect PIHM.   As the most fundamental models of host-
+parasite dynamics assume a linear decrease in host survival probability with
+increasing host load \citep{AndersonandMay1978}, the failure of these methods
+to detect this relationship is a significant disconnect between empirical and
+theoretical disease ecology [though I haven't explicitly tested this with the likelihood method].  However, empirical work has shown that non-linear
+functions of host survival are not uncommon in host-parasite systems
+\citep{Benesh2011}, so this assumption alone does not preclude the use of PIHM
+methods on empirical data.
 
-Related to the above result, all of these methods require that the post-mortality distribution be significantly different from a negative binomial distribution.
-This is necessary because none of the above methods will be able to detect PIHM if a negative binomial distribution is an adequate fit to the post-mortality distribution [describe why?].
-As many observed host-parasite distributions are not significantly different from a negative binomial distribution, there are limited cases where these methods can even be considered.
+Regarding the shape of the post-mortality distribution, all of these methods
+require that the post-mortality distribution be significantly different from a
+negative binomial distribution.  This is necessary because none of the above
+methods will be able to detect PIHM if a negative binomial distribution is an
+adequate fit to the post-mortality distribution.
+This is simply because there will be no need for a more complex model (either truncation of the negative binomial or the model given in equation \ref{eq:dist}) if a negative binomial distribution already fits the data.
+As many
+observed host-parasite distributions are not significantly different from a
+negative binomial distribution, there may be limited cases where these PIHM methods
+can even be considered.
 
 Finally, all of these methods assume that the truncation of a negative binomial
-distribution is due to PIHM, but previous studies have shownthat a variety of other processes can lead to the truncation of a negative
+distribution is due to PIHM, but previous studies have shown that a variety of other processes can lead to the truncation of a negative
 binomial distribution such as within host parasite density-dependence, age-
-dependent variation in host resistance and rate of infection
+dependent variation in host resistance and heterogeneous rate of infection
 \citep{McCallum2000a,Anderson1982a,Rousset1996}.  Therefore, even detecting ``significant'' PIHM in a
@@
 estimate PIHM from distributional data?  We are in agreement with
-\cite{Lester1984} that, at the very least, methods for estimating PIHM can provide preliminary insight into whether or not PIHM is worth further exploration.
-However, we stress that these methods should only be used as an
+\cite{Lester1984} that, at the very least, methods for estimating PIHM can
+provide preliminary insight into whether or not PIHM is worth further
+exploration.  However, we stress that these methods should only be used as an
 exploratory tool when assessing the role of PIHM in a system and potential
 users should critically evaluate whether they think they have a large enough
-sample sizes and appropriate host survival functions for the methods proposed
+sample sizes and appropriate host survival functions/post-mortality distributions for the methods proposed
 in this paper to be applicable.  Even if they are applicable, inferring PIHM
-from distributional data is no substitute for field or laboratory experiments and an in
-depth understanding of the natural history of the host-parasite system under consideration.
+from distributional data is no substitute for field or laboratory experiments
+and/or in depth understanding of the natural history of the host-parasite
+system under consideration.
 
@@
     \hline\hline
-    Data Set & Author detected PIHM? & Likelihood Method w/ Crofton Parameters & Likelihood Method w/out Crofton Parameters & Adjei Method \\
+    Data Set (sample size) & Author detected PIHM? & Likelihood Method w/ Crofton Parameters & Likelihood Method w/out Crofton Parameters & Adjei Method \\
 
     \hline
-    Crofton, Station 1 & Yes & Yes (7.27) & Yes (8.00) & Yes (9.33) \\
-    Crofton, Station 2 & Yes & Yes (6.92) & No & Yes (14.95)\\
-    Crofton, Station 3 & Yes & Yes (5.93) & Yes (7.04) & Yes (5.98) \\
-    Crofton, Station 4 & No & No & No & Yes (7.99) \\
-    Crofton, Station 5 & No & No & No & Yes (10.58) \\
-    Crofton, Station 6 & No & No & No & No \\
-    Adjei, \emph{S. tumbil} female & Yes (5.7) & No  & No   & Yes (6.37) \\
-    Adjei, \emph{S. tumbil} male & Yes (3.4) & Yes (3.42) & Yes (3.28) & Yes (3.66)  \\
-    Adjei, \emph{S. undosquamis} female & Yes (3.2) & Yes (3.04) & No  & Yes (3.11) \\
-    Adjei, \emph{S. undosquamis} male & Yes (1.8) & Yes (1.83) & Yes (2.34) & Yes (1.78) \\
+    Crofton, Station 1 ($n=538$) & Yes & Yes (7.27) & Yes (8.00) & Yes (9.33) \\
+    Crofton, Station 2 ($n=507$) & Yes & Yes (6.92) & No & Yes (14.95)\\
+    Crofton, Station 3 ($n=633$) & Yes & Yes (5.93) & Yes (7.04) & Yes (5.98) \\
+    Crofton, Station 4 ($n=486$) & No & No & No & Yes (7.99) \\
+    Crofton, Station 5 ($n=276$) & No & No & No & Yes (10.58) \\
+    Crofton, Station 6 ($n=191$) & No & No & No & No \\
+    Adjei, \emph{S. tumbil} female ($n=446$) & Yes (5.7) & No  & No   & Yes (6.37) \\
+    Adjei, \emph{S. tumbil} male ($n=452$) & Yes (3.4) & Yes (3.42) & Yes (3.28) & Yes (3.66)  \\
+    Adjei, \emph{S. undosquamis} female ($n=2573$) & Yes (3.2) & Yes (3.04) & No  & Yes (3.11) \\
+    Adjei, \emph{S. undosquamis} male ($n=2440$) & Yes (1.8) & Yes (1.83) & Yes (2.34) & Yes (1.78) \\
 
@@
 
+\begin{figure}
+    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/conceptual_fig.pdf}
+
+    \caption{Five potential shapes for a host-survival functions.
+PIHM should be easier to detect for steeper host survival functions \citep{Lanciani1989}, but we may expect the bias in the parameter estimates to increase as it becomes increasingly difficult to distinguish between steep survival functions.}
+\end{figure}
+
+\begin{figure}
+
+\end{figure}
+
 \end{document}

### 2015-04-24 docs/parasite_mortatlity_manuscript.tex
Sent draft off to Sara

@@
 
+\section*{Abstract}
+
+TODO
+
 \doublespacing
@@
 
-Infectious agents [parasites?] can have major impacts on animal populations
+Infectious agents can have major impacts on animal populations
 through changing population dynamics and stability \citep{Dobson1992}, altering
-predator prey interactions \citep{Joly2004}, and even causing species decline
+predator-prey interactions \citep{Joly2004}, and even causing species decline
 and extinction \citep{DeCastro2005a,McCallum2012b}. Accurate estimates of
@@
 \citep{Lafferty2002} [more], it is notoriously difficult to quantify PIHM in
-wild animal populations \citep{Lester1984,McCallum2000a}.  While both macro and
-microparasites have detrimental and possibly fatal effects on hosts, for the
-remainder of this paper we limit our discussion to macroparasites that can be
-discretely counted within a host.
-
-[More discussion of PIHM before the nitty-gritty?]
+wild animal populations \citep{Lester1984,McCallum2000a}.
 
@@
 are often only able to determine the parasite intensity on some number of sampled hosts.  This snapshot, distributional data is far from the
-ideal type of data for addressing questions regarding PIHM, but this is the type of data on which most questions regarding PIHM are asked \citep[e.g.][]{Ferguson2011,Royce1990,Lanciani1989,Lester1984,Lester1977}.
-In particular, parasitologists are often interested in asking two questions from this data: 1) Is PIHM occurring in this system?
-and if PIHM is occurring 2) Can we quantify the way in which parasite intensity is affecting host survival?
+ideal type of data for addressing questions regarding PIHM, but this is the type of data on which most questions regarding PIHM are asked \citep{Ferguson2011,Royce1990,Lanciani1989,Lester1984,Lester1977}.
+In particular, parasitologists are often interested in asking two questions from this data: 1) Is PIHM occurring in a host-parasite system?
+and if PIHM is occurring 2) Can the effect of parasite intensity on host survival be quantified?
+While both macro and
+microparasites have detrimental and possibly fatal effects on hosts, for the
+remainder of this paper we limit our discussion to macroparasites that can be
+discretely counted within a host.
 
@@
 observed host-parasite distribution and the pre-mortality host-parasite
-distribution will predict substantially different numbers of of highly infected
-hosts (because those have died due to infection), but similar number of hosts with low infection
-loads (because those have survived).  \cite{Crofton1971a} noted that by
-starting with all the observed data and iteratively fitting a negative binomial
+distribution will predict substantially different numbers of hosts with high infect intensity (because those have died due to infection), but a similar number of hosts with low infection
+intensity (because those have survived).  \cite{Crofton1971a} noted that by
+starting with all observed hosts and iteratively fitting a negative binomial
 distribution to hosts with lower and lower parasite loads, one could determine
-whether or not PIHM was occurring in the system and estimate the parameters of the pre-mortality distribution.  This was done
-graphically by determining whether parameters of the truncated negative binomial distributions showed a substantial change as hosts with smaller and smaller parasite intensities were fitted (Figure X).
-The Crofton Method and this graphical technique are both still currently used \citep{Ferguson2011}.
-A more thorough description and implementation of the Crofton Method is given in Appendix X and we discuss the validity of its assumptions in the \emph{Discussion}.
+whether or not PIHM was occurring and estimate the parameters of the host distribution before parasite-induced mortality.  This was done
+graphically by determining whether the parameters of the negative binomial distributions fit to different truncations of the data showed a substantial change as the truncation point moved from heavily infected hosts to lightly infected hosts.
+The Crofton Method and this graphical technique are both still currently used to assess whether PIHM is a occurring in a system \citep{Ferguson2011}.
+We give a thorough description and implementation of the Crofton Method in \emph{Supplementary Information} (SI) 1 and discuss the validity of its assumptions in the \emph{Discussion}.
 
-The second question regarding PIHM attempts to quantify how parasite intensity affects host survival.
-This question asks substantially more from the data and \cite{Adjei1986} proposed a method to attempt to answer it.
-The Adjei Method proceeds
+The second question regarding PIHM moves beyond a simple yes or no answer and attempts to quantify how parasite intensity affects host survival.
+\cite{Adjei1986} proposed a method to answer this question
 by first using the Crofton Method to estimate the pre-mortality parameters for
 a host-parasite distribution and then, given these
-parameters, estimates a host survival function that describes how the
-probability of host-survival changes with increasing parasite load (see
-\emph{Methods}).  From this host survival function, the
-Adjei Method can estimate important host-parasite quantities such as the
+parameters, estimating a host survival function that described how the
+probability of host-survival changed with increasing parasite load (see
+\emph{Methods}).  With this host survival function, one can estimate important host-parasite quantities such as the
 parasite intensity at which 50\% of hosts succumb to PIHM ($LD_{50}$), as well
-as the percent of hosts in a population succumbing to PIHM \citep{Adjei1986}. A more thorough description of the Adjei Method is given in Appendix X.
+as the percent of hosts in a population succumbing to PIHM \citep{Adjei1986}.
 
@@
 rely on a visual test to determine whether or not PIHM is occurring in a
-system.  With no clear decision rule, it can be difficult to determine the
-significance of PIHM in a given host-parasite system and use consistent criteria for identifying PIHM across systems.
+system.  With no clear decision rule, it can be difficult to consistently determine the
+significance of PIHM across different host-parasite systems.
 In theory, the Adjei Method can ameliorate this problem.
@@
 answering the aforementioned questions regarding PIHM.  Second, we propose a
-novel method for answering these questions and compares its efficacy against the Adjei Method.  Third, we
+novel method for answering these questions, compare its efficacy against the Adjei Method, and test both methods ability to detect PIHM on empirical data.  Third, we
 discuss the limitations of inferring PIHM from distributional data alone and
@@
 host before mortality $\mu_p$, and the aggregation of parasites before
-mortality given by the parameter $k_p$ from a negative binomials distribution.
-When $k_p$ is small, parasites are highly aggregated within hosts and when
-$k_p$ is large parasites are more evenly spread among hosts \citep{Wilson2002}.
-The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][]{Royce1990,Lester1984} and we provide a tested implementation of the method in Appendix X.
+mortality given by the parameter $k_p$ from a negative binomial distribution.
+When $k_p$ is small, parasites are highly aggregated among hosts and when
+$k_p$ is large parasites are more evenly spread out \citep{Wilson2002}.
+The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][]{Royce1990,Lester1984} and we provide a tested implementation of the method in SI 3.
 
@@
 \end{equation}
-
 where $x$ is the parasite intensity in a given host and $a$ and $b$ are the two
-parameters of the function. Generally, a larger $a$ allows for hosts to
-tolerate larger parasite loads before experiencing parasite-induced mortality
+parameters of the logistic function. Generally, a larger $a$ allows for hosts to
+tolerate larger parasite intensities before experiencing parasite-induced mortality
 and a larger $b$ leads to a more rapid decline in the probability of host
@@
 calculates the expected number of hosts with a given parasite load $x$ by using
-the equation $g(x ; \mu_p, k_p) * N_p$ where $g(x ; \cdot)$ is the negative binomial pre-
-mortality distribution.  Second, the observed and predicted number of hosts
+the equation $g(x ; \mu_p, k_p) * N_p$ where $g(x ; \cdot)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
 with $x$ parasites are paired as a single data point and the method then assumes that
@@
 example, if the total number of hosts pre-mortality was 50, the mean number of
-parasites per host pre-mortality was 100 and the aggregation parameter was
-$k_p$, applying the equation $g(x ; \mu_p=100, k_p=1) * 50$ would result in
+parasites per host pre-mortality was 100 and the aggregation parameter was 1, applying the equation $g(x ; \mu_p=100, k_p=1) * 50$ would result in
 less than 1 individual in all parasite intensities $x$. In other words, the
@@
 loads, small sample sizes, or both without some sort of binning of the data.
-While this is not a flaw ``per se'', it does add a certain level of
+While this is not a flaw \emph{per se}, it does add a certain level of
 subjectivity (i.e. which bins should you use?) to a method that already has
-serious potential issues.  In this analysis, we always assume the Adjei Method is not binning the data, though we provide code for applying the binning method in Appendix X.
+serious potential issues.  In this analysis, we always assume the Adjei Method is not binning the data, though we provide code for applying the binning method in SI 3.
 
@@
 negative binomial distribution $g(x; \mu_p, k_p)$.  The second
-assumption that the likelihood method makes is that the host survival function
+assumption is that the host survival function
 takes the form of a logistic curve given by equation \ref{eq:logistic}.  With these two explicit assumptions, the Likelihood Method estimates the 4 parameters $\mu_p$, $k_p$, $a$, and $b$.
@@
 
-To estimate that significance of PIHM in a host-parasite system, can use a
-likelihood ratio test in which the full model is given by equation
+To estimate that significance of PIHM in a host-parasite system, a
+likelihood ratio test can be used in which the full model is given by equation
 \ref{eq:dist} and the reduced model is given by a negative binomial
-distribution.  If PIHM is significant in the system, the resulting likelihood
-ratio statistic should be greater $\chi^2_{2,\text{ } \alpha=0.05}$.
+distribution.  If PIHM is not significant in the system, the resulting likelihood
+ratio statistic should approximately follow a $\chi^2$ distribution with degrees of freedom equal to 2.  We provide the code for implementing this Likelihood Method in SI 3.
 
-\subsection*{Testing the ability of approaches to identify PIHM}
+\subsection*{Question 1: Is PIHM occurring?}
 
@@
 number, the host experienced parasite-induced mortality.  The surviving
-hosts comprised the dataset ($\mathbf{x}$) that would be obtained in the field, after PIHM.
+hosts comprised the dataset that would be obtained in the field, after PIHM.
 
@@
 establish the efficacy of the methods independent of the estimates of $N_p$, $\mu_p$ and $k_p$.
-For the Adjei Method, $N_p$, $\mu_p$, and $k_p$ are estimated using the Crofton Method, while $\mu_p$ and $k_p$ in the likelihood method can be estimated jointly with $a$ and $b$ or via the Crofton Method.
+For the Adjei Method, $N_p$, $\mu_p$, and $k_p$ are estimated using the Crofton Method, while $\mu_p$ and $k_p$ in the Likelihood Method can be estimated jointly with $a$ and $b$ or via the Crofton Method.
 If a
@@
 
-We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and sharp decreases in host survival with increasing parasite intensity.
-For a given $\mu_p$, each survival function had the same $LD_{50}$, but different values of $a$ and $b$ (Table X).
+We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in host survival with increasing parasite intensity (Figure \ref{fig:surv}).
+For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.
 We examined each $\mu_p$-survival function pair at  three levels of parasite
-aggregation, $k_p = 0.1, 0.5, 1$ - realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
-combinations we simulated 150 datasets and tested the probability of each method incorrectly identifying PIHM in the pre-mortality dataset (type I error) and correctly identifying PIHM in the post-mortality dataset (power).
+aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
+combinations we simulated 150 datasets and tested the probability of each method incorrectly identifying PIHM in the pre-mortality dataset (Type I error) and correctly identifying PIHM in the post-mortality dataset (power).
 For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level $\alpha = 0.05$.
-We tested all each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500].
+We tested each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500].
 $N_p$ is not technically the sample size on which the methods are being
@@
 
-\subsection*{Testing ability of PIHM approaches to recover survival function}
+\subsection*{Question 2: Can the effect of parasite intensity on host survival be quantified?}
 
-To compare the ability of the Adjei Method and the Likelihood Method to recover $LD_{50}$ and the parameters $a$ and $b$ or the survival function, we used the same simulation procedure and parameter combinations described above.For each parameter
+To compare the ability of the Adjei Method and the Likelihood Method to recover $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above.For each parameter
 combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
@@
 We tested the ability of the Adjei Method and the Likelihood Method to identify
-PIHM on 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
-given in \cite{Adjei1986} (Table X). In the \cite{Crofton1971a} datasets, the host was
+PIHM in 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
+given in \cite{Adjei1986} (Table \ref{table:pihm}). In the \cite{Crofton1971a} datasets, the host was
 the snail \emph{Gammarus pulex} which acts as the intermediate host for the
@@
 
-In both these studies, the authors reported PIHM and we test whether the Adjei
+In both these studies, the authors reported PIHM in some of the datasets and we test whether the Adjei
 Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
@@
 applied the Adjei Method and the Likelihood Method to determine whether or not
-PIHM was significant for these species and compared our results to those given by the authors.
+PIHM was significant for these species and compared our results to those given by the authors.  All fitting to data was done with the Python code provided in SI 3.
 
@@
 
-\subsection*{Detecting PIHM}
+\subsection*{Question 1: Is PIHM occurring?}
 
-The Adjei Method showed highly inflated Type I error (i.e. falsely detected
-PIHM where there was no PIHM) for all parameter combinations that we
-considered.  This method also showed the unintuitive pattern of Type I error
+The Adjei Method showed highly inflated Type I error rates (i.e. falsely detected
+PIHM) for all parameter combinations that we
+considered (Figure \ref{fig:typeI10}; SI2 Figs 1, 2).  This method also showed the unintuitive pattern of Type I error
 rate decreasing as sample size decreased.  This pattern is due to the issue of
@@
 applicability of the Adjei Method is compromised without binning the observed
-data in some way.  On the other hand, the Likelihood Method showed a type I
+data in some way.  In contrast, the Likelihood Method showed a Type I
 error rate at or near the pre-set level of 0.05 for all parameter combinations
@@
 system (i.e. the power) was close to 1 for larger sample sizes and tended to
-decrease as sample size decreased.  The Likelihood Method was a power close to
+decrease as sample size decreased (Figure \ref{fig:typeI10}; SI2 Figs 1, 2).  The Likelihood Method had a power close to
 one for all parameter combinations and sample sizes considered.  With gradual
-survival functions (black lines; Figure X), the Likelihood Method showed a
-decreased ability to detect PIHM for small sample sizes.
+survival functions (solid black lines; Figure \ref{fig:typeI10}, SI2 Figs 1, 2), the Likelihood Method showed a
+decreased power to detect PIHM for small sample sizes.
 
-\subsection*{Recovery of the survival function}
+% TODO: Need to figure out what to do with the mu = 100 that are being dropped
+% because the sample size is too small!
+
+\subsection*{Question 2: Can the effect of parasite intensity on host survival be quantified?}
 
 The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$
-for all combinations of parameters examined in this study (Figure X).  Even
-small sample sizes (< 500 hosts), the Likelihood Methods estimate of $LD_{50}$
+for all combinations of parameters examined in this study (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  Even for
+small sample sizes (< 500 hosts), the Likelihood Method's estimate of $LD_{50}$
 was largely unbiased, with small biases occurring for host survival functions
-that were gradual (i.e. $\mu_p = 10, a = 5, b = -2.5; \mu_p = 100, a = 5, b=
--1.5; \mu_p = 50, a = 5, b = -1.4$)   On the other hand, the Adjei Method
+that were gradual.  On the other hand, the Adjei Method
 always produced more biased estimates of the $LD_{50}$ than the Likelihood
-Method across all parameter combinations.  For $\mu_p = 10$, the $LD_{50}$
+Method across all parameter combinations (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  For $\mu_p = 10$, the $LD_{50}$
 estimates from the Adjei Method were largely unbiased for large samples sizes,
 but showed increasing bias as sample size decreased, particularly for steep
-survival functions ($a = 20, b = -10$).  As $\mu_p$ increased, the Adjei method
+survival functions.  As $\mu_p$ increased, the Adjei method
 produced biased estimates of $LD_{50}$ across all sample sizes, with bias
-increasing as sample size decreased.
+increasing as sample size decreased (SI2 Fig 3, 4).
 
@@
 (increasing coefficient of variation) as sample size decreased for all
-parameter combinations we examined.  The $LD_{50}$ estimates from the Adjei
+parameter combinations we examined (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  The $LD_{50}$ estimates from the Adjei
 Method showed a similar pattern, with large decreases in precision occurring
-for the steepest survival function across all values of $\mu_p$ (Figure X).
+for the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).
 
@@
 asymptotically unbiased estimates of $a$ and $b$ as sample size increased for
-all parameter combinations considered.  However, as sample size decreased, the
+all parameter combinations considered (Figure \ref{fig:biasa}, SI2 Fig 5, 6).  However, as sample size decreased, the
 Likelihood Method tended to produce severely biased estimates of $a$ and $b$.
 This was generally more pronounced for steeper survival functions and more
-aggregated pre-mortality distributions (Figure X).  The Adjei Method produced
+aggregated pre-mortality distributions (e.g. Figure \ref{fig:biasa}).  The Adjei Method produced
 biased estimates of $a$ and $b$ across all sample sizes, with the bias
 consistently being larger when the survival function was steeper. The bias of
-the Adjei Method also increased as $\mu_p$ increased.
+the Adjei Method's estimate of $a$ also increased as $\mu_p$ increased (Figure \ref{fig:biasa}, SI2 Fig 5, 6).
 
@@
 
-Of the 10 datasets we considered, the previous authors visually detect PIHM
+Of the 10 datasets we considered, the previous authors visually detected PIHM
 in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
@@
 PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
-dataset in which the the Likelihood Method parameterized from the Crofton
-Method did not detect a significant effect of PIHM was the the Adjei dataset
-for \emph{S. tumbil}.  For this dataset there was a marginally significant effect
+dataset in which the the Likelihood Method did not detect a significant effect of PIHM was the Adjei dataset
+for female \emph{S. tumbil}.  For this dataset there was a marginally significant effect
 of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$).
@@
 The Adjei Method detected significant PIHM in 9 of the 10 datasets given (Table \ref{table:pihm}).
-This is consistent with the the previous results which show that the Adjei Method has a very high Type I error rate.
-The Likelihood Method in which all parameters were jointly estimated predicted significant PIHM in 4 of the 10 datasets.
+This is consistent with the previous results which show that the Adjei Method has a very high Type I error rate.
 
@@
 
-Determining whether PIHM is a significant factor for a host populations is critically important in many systems, but detecting and describing PIHM given only observational data is
-notoriously difficult.  We show that the Adjei Method, the only currently
-proposed method to estimate the host survival function and the $LD_{50}$ from
-observational PIHM, has some serious methodological problems that result in
-biased estimates of the host survival function and the $LD_{50}$ even under the most idealistic conditions.
-Moreover, we show that the Adjei Method has a seriously inflated Type I error rate, meaning it will often detect PIHM even when it is not present.
-Moreover, for small, realistic sample sizes the Adjei Method behaves erratically; a consequence of the need to subjectively bin the data in order to predict parasite intensity classes with at least one host.
+Determining whether PIHM is a significant factor for a host populations is
+critically important in many systems, but detecting and describing PIHM given
+only observational data is notoriously difficult.  We show that the Adjei
+Method, the only currently proposed method to estimate the host survival
+function and the $LD_{50}$ from observational data, has some serious
+methodological problems that result in biased estimates even under the most idealistic conditions.
+Moreover, we show that the Adjei Method has a seriously inflated Type I error
+rate, meaning it will often detect PIHM even when it is not present. Moreover,
+for small, realistic sample sizes the Adjei Method behaves erratically; a
+consequence of the need to subjectively bin the data in order to predict
+parasite intensity classes with at least one host.
 
@@
 general method to determine both whether or not PIHM is occurring in a system
-and the shape of the survival function.  We show that this method is
+and to quantify the survival function.  We show that this method is
 asymptotically unbiased when estimating the host-survival function for all of
@@
 precise estimates of the $LD_{50}$ for small, realistic sample sizes. Moreover,
-this novel method has a type I error rate close to the pre-set level of $\alpha = 0.05$ and high power for detecting PIHM, given it is occurring, for realistic samples sizes.
+this novel method has a Type I error rate close to the pre-set level of $\alpha = 0.05$ and high power for detecting PIHM for realistic samples sizes.
   However, we note that
@@
 reported, while the Likelihood Method's detection of PIHM was consistent with
-previously reported PIHM in a given dataset.  Taken together, these results suggest that the Adjei Method is a fundamentally
-flawed method for detecting PIHM and describing features of the host survival
-function and we recommend using the Likelihood Method for detecting PIHM and describing attributes of the host survival function.
+previously reported PIHM in a given dataset.  Taken together, these results suggest that the Adjei Method is fundamentally
+flawed.  We recommend using the Likelihood Method for detecting PIHM and describing attributes of the host survival function.
 
-While we improved upon the previously existing methods for answering questions
+While we have improved upon the previously existing methods for answering questions
 about PIHM, we cannot belie the fact that estimating PIHM from observational
@@
 The most fundamental assumption of all methods for estimating PIHM is that the
-shape of the pre-mortality host-parasite distribution is known. In the
-discrete case, this distribution is negative binomial, while in the continuous
-case the distribution can be gamma or exponential \citep{Ferguson2011}.
+shape of the pre-mortality host-parasite distribution is known and follows a negative binomial distribution.
 While there is substantial empirical and theoretical evidence to justify the
@@
 parasite dynamics assume a linear decrease in host survival probability with
-increasing host load \citep{AndersonandMay1978}, the failure of these methods
+increasing parasite intensity \citep{AndersonandMay1978}, the failure of these methods
 to detect this relationship is a significant disconnect between empirical and
@@
 users should critically evaluate whether they think they have a large enough
-sample sizes and appropriate host survival functions/post-mortality distributions for the methods proposed
+sample sizes and an appropriate host survival function/post-mortality distribution for the methods developed
 in this paper to be applicable.  Even if they are applicable, inferring PIHM
@@
 system under consideration.
+
+\section*{Acknowledgments}
+
+TODO
 
@@
     \caption{Comparison of the PIHM predictions of previously used host-parasite datasets to those given by the Adjei Method and the Likelihood Method.
-The first column specifies the identity of the dataset, the second column specifies whether or not the author indicated that PIHM was occurring in the system, the third column indicates whether or not the Likelihood Method with parameters from the Crofton Method detects significant PIHM, the third column indicates whether the Likelihood Method with jointly estimated pre-mortality parameters detects PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
-If a method detects significant PIHM the $LD_{50}$ is given is parentheses}
+The first column specifies the identity of the dataset, the second column specifies whether or not the author indicated that PIHM was occurring in the system, the third column indicates whether or not the Likelihood Method with parameters from the Crofton Method detects significant PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
+If a method detected significant PIHM the predicted $LD_{50}$ is given in parentheses}
 
     \centering
-    \begin{tabular}{l l p{3cm} p{3cm} l}
+    \begin{tabular}{l  p{3cm} p{3cm} l}
 
     \hline\hline
-    Data Set (sample size) & Author detected PIHM? & Likelihood Method w/ Crofton Parameters & Likelihood Method w/out Crofton Parameters & Adjei Method \\
+    Data Set (sample size) & Author detected PIHM? & Likelihood Method?  & Adjei Method? \\
 
     \hline
-    Crofton, Station 1 ($n=538$) & Yes & Yes (7.27) & Yes (8.00) & Yes (9.33) \\
-    Crofton, Station 2 ($n=507$) & Yes & Yes (6.92) & No & Yes (14.95)\\
-    Crofton, Station 3 ($n=633$) & Yes & Yes (5.93) & Yes (7.04) & Yes (5.98) \\
-    Crofton, Station 4 ($n=486$) & No & No & No & Yes (7.99) \\
-    Crofton, Station 5 ($n=276$) & No & No & No & Yes (10.58) \\
-    Crofton, Station 6 ($n=191$) & No & No & No & No \\
-    Adjei, \emph{S. tumbil} female ($n=446$) & Yes (5.7) & No  & No   & Yes (6.37) \\
-    Adjei, \emph{S. tumbil} male ($n=452$) & Yes (3.4) & Yes (3.42) & Yes (3.28) & Yes (3.66)  \\
-    Adjei, \emph{S. undosquamis} female ($n=2573$) & Yes (3.2) & Yes (3.04) & No  & Yes (3.11) \\
-    Adjei, \emph{S. undosquamis} male ($n=2440$) & Yes (1.8) & Yes (1.83) & Yes (2.34) & Yes (1.78) \\
+    Crofton, Station 1 ($n=538$) & Yes & Yes (7.27) & Yes (9.33) \\
+    Crofton, Station 2 ($n=507$) & Yes & Yes (6.92) &  Yes (14.95)\\
+    Crofton, Station 3 ($n=633$) & Yes & Yes (5.93) &  Yes (5.98) \\
+    Crofton, Station 4 ($n=486$) & No & No &  Yes (7.99) \\
+    Crofton, Station 5 ($n=276$) & No & No & Yes (10.58) \\
+    Crofton, Station 6 ($n=191$) & No & No & No \\
+    Adjei, \emph{S. tumbil} female ($n=446$) & Yes (5.7) & No & Yes (6.37) \\
+    Adjei, \emph{S. tumbil} male ($n=452$) & Yes (3.4) & Yes (3.42) & Yes (3.66)  \\
+    Adjei, \emph{S. undosquamis} female ($n=2573$) & Yes (3.2) & Yes (3.04) & Yes (3.11) \\
+    Adjei, \emph{S. undosquamis} male ($n=2440$) & Yes (1.8) & Yes (1.83) & Yes (1.78) \\
 
@@
 PIHM should be easier to detect for steeper host survival functions \citep{Lanciani1989}, but we may expect the bias in the parameter estimates to increase as it becomes increasingly difficult to distinguish between steep survival functions.}
+    \label{fig:surv}
 \end{figure}
@@
 
+    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu10}
+
+    \caption{The Type I error rate and the power of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$.
+The first column gives the Type I error rate of each method for falsely detecting PIHM when none is present.
+The red line gives the the pre-set Type I error rate of $\alpha = 0.05$.
+The second column gives the power of a given method to detect PIHM when it is actually occurring.
+}
+    \label{fig:typeI10}
+
+\end{figure}
+
+\begin{figure}
+
+    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu10}
+
+    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+The first column gives the bias of each methods $LD_{50}$ estimate over 150 simulations.
+The second column gives the precision of each methods $LD_{50}$ estimate over 150 simulations.}
+
+    \label{fig:biasld50}
+
+\end{figure}
+
+\begin{figure}
+
+    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu10}
+
+    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+The first column gives the bias of each methods $a$ estimate over 150 simulations.
+The second column gives the precision of each methods $a$ estimate over 150 simulations.}
+
+    \label{fig:biasa}
+
 \end{figure}

### 2015-04-24 docs/supplementary_material.tex
Sent draft off to Sara

% Document
\documentclass[12pt, a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{authblk}
\usepackage{lipsum}

% Figure and table formating
\usepackage{epsfig}
\usepackage{tabu}
\usepackage{rotating}
\usepackage{pbox}
\usepackage{framed, multicol}
\usepackage[framemethod=TikZ]{mdframed}
\usepackage{float}
\usepackage[left=1.25 in, right=1.25 in, top=1.25 in, bottom=1.25 in]{geometry}
\usepackage{csvsimple}

% Fonts - Mathtime
%\usepackage{txfonts}
\usepackage{amsmath} % Add amssymb if not using Mathtime

% Text
\setlength{\parindent}{0.5in}
\frenchspacing  \tolerance = 800  \hyphenpenalty = 800

\usepackage{lineno} % Line numbers
\def\linenumberfont{\normalfont\footnotesize\ttfamily}
\setlength\linenumbersep{0.2 in}

\usepackage{setspace}

% Format section and subsection headers
\makeatletter
\renewcommand{\section}{\@startsection
{section}%                   % the name
{1}%                         % the level
{0mm}%                       % the indent
{-\baselineskip}%            % the before skip
{0.5\baselineskip}%          % the after skip
{\normalfont\bf\large}} % the style

\renewcommand{\subsection}{\@startsection
{subsection}%                   % the name
{2}%                         % the level
{0mm}%                       % the indent
{-\baselineskip}%            % the before skip
{0.5\baselineskip}%          % the after skip
{\normalfont\bf}} % the style
\makeatother

% Other
\usepackage{graphicx}
\usepackage[singlelinecheck=false,font=small,labelfont=bf]{caption}

% Bibliography
\usepackage[numbers, compress]{natbib} % Bibliography - APA
\bibpunct{(}{)}{;}{a}{}{,}

% Format the Bibliography appropriately
% increase \bibhang to take care of the numbers
\setlength{\bibhang}{0pc}
\makeatletter
% patch \@lbibitem to print the current number before the authors
\patchcmd{\@lbibitem}
  {]}
  {] \theNAT@ctr. \newline }
  {}{}
\makeatother

%%%%%% FRONT MATTER %%%%%%%%%

\begin{document}

\linenumbers

\doublespacing

\noindent
\Large{\textbf{Supporting Information}}

\normalsize

\section*{Appendix 1: Implementation of the Crofton Method}

The algorithm for fitting the Crofton Method \citep{Crofton1971a} proceeds as follows. First, obtain a dataset
with $n$ hosts where each host has some parasite intensity 0 to $p_{max}$. Starting
with the full dataset, guess a vector of pre-mortality parameters $(N_p, \mu_p, k_p)$ and given
these parameters calculate the predicted number of hosts with $0, 1, 2, \dots
p_{max}$.  Compare the expected number of hosts with $0, 1, 2, \dots  p_{max}$ parasites
to the observed number hosts with $0, 1, 2, \dots p_{max}$ parasites and calculate
the $\chi^2$-squared statistic associated with your observed and predicted vectors. In reality, one often has to bin the parasite intensity data because all
parasite intensities are not represented in the dataset. Continue to guess $(N_p,
mu_p, k_p)$ vectors until a set of parameters is found that minimizes the $\chi^2$-
squared statistic.

Second, choose a truncation value ($t_1$) such that $t_1 <
p_{max}$. Truncate the data such that $\text{data}_\text{truncated} <= t$ and repeat the above
iterative procedure to calculate another set of parameters $(N_{p2}, \mu_{p2}, k_{p2})$
that minimizes the $\chi^2$-squared statistic on the truncated data. Choose a new
truncated value $t_2 < t_1$ and repeat the first two steps. Continue to truncated
the dataset until it only contains hosts with 0, 1, and 2 parasites (or 3 bins).
As the method attempts to estimate three parameters, at least 3 classes are
needed if all 3 parameters are to be identifiable \citep{Royce1990}.

We provide an implementation and unit tests of the Crofton Method Supplementary Information 3.  Figure \ref{fig:crof_test} visually shows the results of these unit tests.

\section*{Appendix 2: Additional Figure}

See Figures 1 - 6.

\section*{Appendix 3: Code and unit tests for estimating parasite-induced host mortality}

Python code, unit tests, and a help file for the Crofton Method, the Adjei Method and the Likelihood Method can be found in \verb*#pihm_methods.py#, \verb*#test_pihm_methods.py#, and \verb*#help_file.txt#, respectively.

% These can be in the supplementary in later drafts!

\begin{figure}

    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu50}

    \caption{The type I error rate and the power of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$.  The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.  The red line gives the the pre-set type I error rate of $\alpha = 0.05$.  The second column gives the power of a given method to detect PIHM when it is actually occurring. }
    \label{fig:typeI50}

\end{figure}

\begin{figure}

    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu100}

    \caption{The type I error rate and the power of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$.  The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.  The red line gives the the pre-set type I error rate of $\alpha = 0.05$.  The second column gives the power of a given method to detect PIHM when it is actually occurring. }
    \label{fig:typeI100}

\end{figure}

\begin{figure}

    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu50}

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.  The first column gives the bias of each methods $LD_{50}$ estimate over 150 simulations. The second column gives the precision of each methods $LD_{50}$ estimate over 150 simulations.}

    \label{fig:biasld50_50}

\end{figure}

\begin{figure}

    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu100}

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.  The first column gives the bias of each methods $LD_{50}$ estimate over 150 simulations. The second column gives the precision of each methods $LD_{50}$ estimate over 150 simulations.}

    \label{fig:biasld50_100}

\end{figure}

\begin{figure}

    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu50}

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.  The first column gives the bias of each methods $a$ estimate over 150 simulations. The second column gives the precision of each methods $a$ estimate over 150 simulations.}

    \label{fig:biasa_50}

\end{figure}

\begin{figure}

    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu100}

    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.  The first column gives the bias of each methods $a$ estimate over 150 simulations. The second column gives the precision of each methods $a$ estimate over 150 simulations.}

    \label{fig:biasa_100}

\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/compare_crofton.pdf}
    \caption{A comparison of this papers implementation of the Crofton Method with the results given in \cite{Crofton1971a}.  Three of the 6 stations fit by in \citep{Crofton1971a} are shown here and all three show that the two implementations give very similar results.}
    \label{fig:crof_test}

\end{figure}

\singlespacing
\bibliographystyle{/Users/mqwilber/Dropbox/Documents/Bibformats/ecology_letters.bst}
\bibliography{/Users/mqwilber/Dropbox/Documents/Bibfiles/Projects_and_Permits-parasite_host_mortality}

\end{document}

### 2015-04-24 code/pihm_methods.py

@@ -146,6 +146,7 @@ class PIHM:
         Returns the observed and predicted vectors for the preset values of
         Np, mup, and kp.  Will throw an error in Np, mup and kp are not defined.
         You can either define these manually or run the crofton_method first.
+        Note that the predicted values are the predicted values WITHOUT mortality
 
         Parameters
         ----------
@@ -233,11 +234,24 @@ class PIHM:
 
         # 2. Bin empirical self.data and theoretical self.data
         empirical_binned = np.histogram(self.data, bins=full_bin_edges)[0]
-        theor_data = np.round(mod.nbinom.pmf(np.arange(0, max_worms + 1),
-                        self.mup, self.kp) * self.Np, decimals=0)
-        full_theor_data = np.repeat(np.arange(0, max_worms + 1),
-                theor_data.astype(int))
-        theor_binned = np.histogram(full_theor_data, bins=full_bin_edges)[0]
+
+        # Note that if Np is too low and mup is too high the Adjei method
+        # Won't work because the probability is too disperesed!  There won't be
+        # a complete individual in any of the classes so you can't fit the
+        # binomial. Need to try different bin sizes
+
+        split_edges = split_data(full_bin_edges)
+
+
+        theor_binned = np.array([np.round(np.sum(mod.nbinom.pmf(s, self.mup, self.kp)
+                        * self.Np), decimals=0) for s in split_edges])
+
+        # Group the theoretical data by bin edges
+
+        if np.all(theor_binned == 0):
+            raise AttributeError("With Np = %.2f and mup = %.2f,the are"  % (self.Np, self.mup) +
+                                  " no\ncomplete individuals in the given " +
+                                  "bins. Try different bin sizes")
 
         # Put everything into self.dataframe
         if no_bins:
@@ -272,9 +286,6 @@ class PIHM:
         b = params[1]
 
         if test_sig:
-
-            chi = null_deviance - deviance
-            p = chisqprob(chi, 1)
             return (self.get_all_params(), (null_deviance, deviance))
 
         else:
@@ -601,10 +612,13 @@ def extract_simulation_results(sim_results, keys, method_name, param,
         elif param == "p":
 
             # Type I error
-            biases.append(np.sum(np.array(a_vals) < alpha) / len(a_vals))
+            typeIp, typeIb = zip(*a_vals)
+            powerp, powerb = zip(*b_vals)
+
[…]

### 2015-04-24 code/test_pihm_methods.py

@@ -140,6 +140,7 @@ class TestPIHM(TestCase):
 
 
     def test_obs_pred(self):
+
         pass
 
     def test_adjei_method(self):
@@ -191,9 +192,11 @@ class TestPIHM(TestCase):
 
         pred_ld50 = np.round([ld50(f) for f in fits], decimals=1)
         adjei_ld50 = [5.7, 3.4, 3.2, 1.8]
+        exp_ld50 = [6.4, 3.7, 3.1, 1.8]
+
+        assert_array_equal(exp_ld50, pred_ld50)
 
         print(zip(pred_ld50, adjei_ld50))
-        1/0
 
     def test_likelihood_method(self):
 
@@ -260,6 +263,29 @@ class TestPIHM(TestCase):
         # Varying pre-mort values...shouldn't detect PIHM
         p = tobj.test_for_pihm_w_likelihood(fixed_pre=False)[1]
         assert_equal((p > 0.05), True)
+
+
+class TestAux(TestCase):
+
+    def test_split_data(self):
+
+        # Test boundary condition. Shouldn't return 8
+        bin_edges = [0, 3, 8]
+        pred = [np.arange(0, 3), np.arange(3, 8)]
+        obs = split_data(bin_edges)
+        [assert_array_equal(p, o) for p, o in zip(pred, obs)]
+
+        # Test upper boundary condition. Should return 8
+        bin_edges = [0, 3, 8.1]
+        pred = [np.arange(0, 3), np.arange(3, 9)]
+        obs = split_data(bin_edges)
+        [assert_array_equal(p, o) for p, o in zip(pred, obs)]
+
+        # Test full edges.  Should return all the edges
+        bin_edges = np.arange(0, 10.1)
+        pred = np.arange(0, 11)
+        obs = split_data(bin_edges)
+        [assert_array_equal(p, o) for p, o in zip(pred, obs)]
 
 
 

### 2015-04-27 docs/parasite_mortatlity_manuscript.tex
Made edits before sending to Cherie

@@
 answering the aforementioned questions regarding PIHM.  Second, we propose a
-novel method for answering these questions, compare its efficacy against the Adjei Method, and test both methods ability to detect PIHM on empirical data.  Third, we
+novel method for answering these questions, compare its efficacy against the Adjei Method, and test both methods' ability to detect PIHM on empirical data.  Third, we
 discuss the limitations of inferring PIHM from distributional data alone and
@@
 calculates the expected number of hosts with a given parasite load $x$ by using
-the equation $g(x ; \mu_p, k_p) * N_p$ where $g(x ; \cdot)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
+the equation $g(x ; \mu_p, k_p) * N_p$, where $g(x ; \cdot)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
 with $x$ parasites are paired as a single data point and the method then assumes that
@@
 
-One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_0^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$.
+One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$.
 Therefore equation \ref{eq:concept} can be written as
@@
 
-To compare the ability of the Adjei Method and the Likelihood Method to recover $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above.For each parameter
+To compare the ability of the Adjei Method and the Likelihood Method to recover $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above.
+For each parameter
 combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
 precision \citep{Walther2005} for these estimates over varying pre-mortality host population sizes  $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
-10000]. We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.
+10000].
+We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.
+Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.
 
@@
 one for all parameter combinations and sample sizes considered.  With gradual
-survival functions (solid black lines; Figure \ref{fig:typeI10}, SI2 Figs 1, 2), the Likelihood Method showed a
-decreased power to detect PIHM for small sample sizes.
+survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (solid black lines; Figure \ref{fig:typeI10}, SI2 Figs 1, 2).
 
@@
 In terms of the host survival function, the Likelihood Method gave
-asymptotically unbiased estimates of $a$ and $b$ as sample size increased for
+asymptotically unbiased estimates of $a$ as sample size increased for
 all parameter combinations considered (Figure \ref{fig:biasa}, SI2 Fig 5, 6).  However, as sample size decreased, the
-Likelihood Method tended to produce severely biased estimates of $a$ and $b$.
+Likelihood Method tended to produce severely biased estimates of $a$.
 This was generally more pronounced for steeper survival functions and more
 aggregated pre-mortality distributions (e.g. Figure \ref{fig:biasa}).  The Adjei Method produced
-biased estimates of $a$ and $b$ across all sample sizes, with the bias
+biased estimates of $a$ across all sample sizes, with the bias
 consistently being larger when the survival function was steeper. The bias of
@@
 the Likelihood Method produces seriously biased estimates of the host survival
-function ($a$ and $b$) for sample sizes typically observed in many host-
-parasite studies.  The bias was most severe for steep host survival functions,
+function ($a$ and $b$) for sample sizes typically observed in many host-parasite studies.  The bias was most severe for steep host survival functions,
 due to large changes in the values of $a$ and $b$ only slightly changing an
 already steep survival function.  Given these results, neither the
-Likelihood Method or the Adjei Method could confidently recover the exact shape of the host survival function for small, realistic sample sizes.
+Likelihood Method nor the Adjei Method could confidently recover the exact shape of the host survival function for small, realistic sample sizes.
 
@@
 previously reported PIHM in a given dataset.  Taken together, these results suggest that the Adjei Method is fundamentally
-flawed.  We recommend using the Likelihood Method for detecting PIHM and describing attributes of the host survival function.
+flawed and we recommend using the Likelihood Method for detecting PIHM and describing attributes of the host survival function.
 
@@
 distribution, but rather that the pre-mortality distribution is well-fit by a
-negative binomial. The extreme flexibility of the negative binomial
+negative binomial distribution. The extreme flexibility of the negative binomial
 distribution makes it a reasonable candidate distribution for the pre-mortality
@@
 that when hosts experienced a linear decrease in survival probability the
-Crofton Method could not detect PIHM.   As the most fundamental models of host-
-parasite dynamics assume a linear decrease in host survival probability with
+Crofton Method could not detect PIHM.   As the most fundamental models of host-parasite dynamics assume a linear decrease in host survival probability with
 increasing parasite intensity \citep{AndersonandMay1978}, the failure of these methods
 to detect this relationship is a significant disconnect between empirical and
-theoretical disease ecology [though I haven't explicitly tested this with the likelihood method].  However, empirical work has shown that non-linear
+theoretical disease ecology.  However, empirical work has shown that non-linear
 functions of host survival are not uncommon in host-parasite systems
@@
     \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
-The first column gives the bias of each methods $LD_{50}$ estimate over 150 simulations.
-The second column gives the precision of each methods $LD_{50}$ estimate over 150 simulations.}
+The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
+The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}
 
@@
     \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
-The first column gives the bias of each methods $a$ estimate over 150 simulations.
-The second column gives the precision of each methods $a$ estimate over 150 simulations.}
+The first column gives the bias of each method's $a$ estimate over 150 simulations.
+The second column gives the precision of each method's $a$ estimate over 150 simulations.}
 

### 2015-04-27 docs/supplementary_material.tex
Made edits before sending to Cherie

@@
     \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
-The first column gives the bias of each methods $LD_{50}$ estimate over 150 simulations.
-The second column gives the precision of each methods $LD_{50}$ estimate over 150 simulations.}
+The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
+The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}
 
@@
     \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
-The first column gives the bias of each methods $LD_{50}$ estimate over 150 simulations.
-The second column gives the precision of each methods $LD_{50}$ estimate over 150 simulations.}
+The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
+The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}
 
@@
     \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
-The first column gives the bias of each methods $a$ estimate over 150 simulations.
-The second column gives the precision of each methods $a$ estimate over 150 simulations.}
+The first column gives the bias of each method's $a$ estimate over 150 simulations.
+The second column gives the precision of each method's $a$ estimate over 150 simulations.}
 
@@
     \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
-The first column gives the bias of each methods $a$ estimate over 150 simulations.
-The second column gives the precision of each methods $a$ estimate over 150 simulations.}
+The first column gives the bias of each method's $a$ estimate over 150 simulations.
+The second column gives the precision of each method's $a$ estimate over 150 simulations.}
 
@@
     \caption{A comparison of this papers implementation of the Crofton Method with the results given in \cite{Crofton1971a}.
-Three of the 6 stations fit by in \citep{Crofton1971a} are shown here and all three show that the two implementations give very similar results.}
+Three of the 6 stations fit by \cite{Crofton1971a} are shown here and all three show that the two implementations give very similar results.}
     \label{fig:crof_test}

### 2015-05-04 docs/parasite_mortatlity_manuscript.tex
Incorporated most of Sara's comments. Still need citations

@@
 
-Infectious agents can have major impacts on animal populations
-through changing population dynamics and stability \citep{Dobson1992}, altering
-predator-prey interactions \citep{Joly2004}, and even causing species decline
-and extinction \citep{DeCastro2005a,McCallum2012b}. Accurate estimates of
-parasite-induced host mortality (PIHM) in wild animals are important for
-understanding what regulates both host and parasite populations and to make
-predictions about disease transmission in natural systems. Although a negative
-impact on host fitness is a fundamental component of parasitism
-\citep{Lafferty2002} [more], it is notoriously difficult to quantify PIHM in
-wild animal populations \citep{Lester1984,McCallum2000a}.
+Infectious agents can have major impacts on animal populations through changing
+population dynamics and stability, altering predator-prey interactions, and
+even causing species’ decline and extinction. Accurately estimating the impact
+of these infectious agents in wildlife is critical to both understanding what
+regulates host and parasite populations and making predictions about disease
+transmission. The impact of pathogens, such as rabies(), bovine TB(), and
+rinderpest(),  is typically quantified based on the presence or absence of
+disease, and does not account for the number of infectious agents present.
+Although sufficient for many bacterial and viral agents that reproduce within a
+host, for macroparasites, hosts cannot be simply categorized as infected and
+uninfected because pathology is linked to the intensity of infection (Anderson
+and May 1978).  Helminths exhibiting this intensity dependent pathology have
+significant impacts on human health (), domestic livestock economics (),
+wildlife survival ().  And, while it is generally assumed that some fraction of
+wild host populations must succumb to parasitic infections, it is notoriously
+difficult to actually quantify parasite-induced host mortality (PIHM) in wild
+animal populations. [more]
 
-To conclusively identify PIHM in wild animal populations, it is necessary to
-experimentally infect and track host populations; a method that is rarely
-possible in most wild animal systems \citep{McCallum2000a}.  Instead, parasitologists
-are often only able to determine the parasite intensity on some number of sampled hosts.  This snapshot, distributional data is far from the
-ideal type of data for addressing questions regarding PIHM, but this is the type of data on which most questions regarding PIHM are asked \citep{Ferguson2011,Royce1990,Lanciani1989,Lester1984,Lester1977}.
+Ideally, parasite-induced host mortality would be
+quantified by experimentally infecting and tracking individual hosts in the
+wild population; however, for logistical and ethical reasons this method is
+rarely feasible. Data on parasite intensity is much easier to collect and has
+often been used to identify the presence PIHM (cite studies) and quantify the
+relationship between infection intensity and host mortality (cite studies).
+
+Crofton (1971) first proposed that PIHM could be identified by comparing the
+observed parasite distribution in the host population to the distribution
+predicted in the absence of parasite-induced host mortality. This method
+(“Crofton Method”) assumes that, prior to host mortality, parasites are
+distributed in the host population following a negative binomial distribution;
+however, as intensity dependent pathology removes heavily infected hosts from
+the population, the tail of the distribution is truncated. Mortality is assumed
+to not occur in hosts with low intensity infections, thus by iteratively
+fitting a negative binomial distribution to hosts with lower and lower parasite
+loads, and comparing the tail end of this predicted distribution to the
+observed parasite data, one could determine both whether PIHM was occurring and
+the parasite distribution in the host population prior to parasite induced
+mortality. We give a thorough description and implementation of the Crofton
+Method in \emph{Supplementary Information (SI)} 1 and discuss the validity of
+its assumptions in the \emph{Discussion}.
+
+The Crofton Method may be able to detect the presence of PIHM however,
+quantifying the relationship between infection intensity and survival
+probability is more complicated. \cite{Adjei1986} suggested that these values
+could be quantified by using the Crofton Method to first estimate the pre-
+mortality parasite distribution and then, using those parameters, calculate the
+probability of host survival with increasing parasite load. To do this,
+\cite{Adjei1986} modeled host survival as a logistic function and then used a
+binomial generalized linear model (GLM) to estimate the logistic parameters.
+These methods appeared to provide an estimate the parasite intensity at which
+50\% of hosts exhibit PIHM (), as well as the unmeasurable fraction of the
+population that was lost (SI for adjei method details and extended discussion
+of issues?). However, to implement this method the observed data must be
+modified to fit the binomial GLM framework and subjectively binned when mean
+infection intensity is high or sample sizes are small.
+
+After 30 years, and despite clear limitations \citep{McCallum2000a}, these
+methods (particularly the Crofton Method) are still discussed among
+parasitologists and are the primary techniques for examining population level
+impacts of parasitism using parasite intensity data. In these methods, PIHM can
+only be identified by visually examining plots and, with no clear decision
+rule; it can be difficult to determine the significance of PIHM across
+different host-parasite systems. The survival function produced by the Adjei
+Method offers one solution; however, this method requires manipulating the
+original data and has never been thoroughly tested.
+
+Estimating parasite impacts on host population using intensity data should only
+be done if methods exist to permit unbiased and accurate estimates of PIHM. In
+this study, we first propose a novel likelihood method for calculating PIHM. We
+next use simulations to compare this likelihood method with previous Adjei
+method to (1) test the ability to detect occurrence of PIHM and (2) estimate
+lethal parasite load ($LD_{50}$) and associated survival function.  We then
+apply both methods to real datasets previously used in PIHM analyses and
+compare results. And, finally, we discuss the limitations of inferring PIHM
+from distributional data and whether any method for inferring PIHM has a place
+in quantitative parasitology.
+
+% Infectious agents can have major impacts on animal populations
+% through changing population dynamics and stability \citep{Dobson1992}, altering
+% predator-prey interactions \citep{Joly2004}, and even causing species decline
+% and extinction \citep{DeCastro2005a,McCallum2012b}. Accurate estimates of
+% parasite-induced host mortality (PIHM) in wild animals are important for
+% understanding what regulates both host and parasite populations and to make
+% predictions about disease transmission in natural systems. Although a negative
+% impact on host fitness is a fundamental component of parasitism
+% \citep{Lafferty2002} [more], it is notoriously difficult to quantify PIHM in
+% wild animal populations \citep{Lester1984,McCallum2000a}.
+
+% To conclusively identify PIHM in wild animal populations, it is necessary to
+% experimentally infect and track host populations; a method that is rarely
+% possible in most wild animal systems \citep{McCallum2000a}.  Instead, parasitologists
+% are often only able to determine the parasite intensity on some number of sampled hosts.  This snapshot, distributional data is far from the
+% ideal type of data for addressing questions regarding PIHM, but this is the type of data on which most questions regarding PIHM are asked \citep{Ferguson2011,Royce1990,Lanciani1989,Lester1984,Lester1977}.
 In particular, parasitologists are often interested in asking two questions from this data: 1) Is PIHM occurring in a host-parasite system?
@@
 While both macro and
-microparasites have detrimental and possibly fatal effects on hosts, for the
-remainder of this paper we limit our discussion to macroparasites that can be
-discretely counted within a host.
+% microparasites have detrimental and possibly fatal effects on hosts, for the
+% remainder of this paper we limit our discussion to macroparasites that can be
+% discretely counted within a host.
 
-The first of these PIHM questions was addressed by \cite{Crofton1971a} who developed
-a method to test for the presence of PIHM using the truncation of the negative binomial
-distribution.  In short, the Crofton Method assumes that the distribution of
-parasites across hosts before mortality occurs follows a negative binomial
-distribution \citep{AndersonandMay1978,Shaw1998}.  As heavily infected hosts
-begin to die, the negative binomial distribution is truncated and these hosts are no longer observed in a sample. In other words, the
-observed host-parasite distribution and the pre-mortality host-parasite
-distribution will predict substantially different numbers of hosts with high infect intensity (because those have died due to infection), but a similar number of hosts with low infection
-intensity (because those have survived).  \cite{Crofton1971a} noted that by
-starting with all observed hosts and iteratively fitting a negative binomial
-distribution to hosts with lower and lower parasite loads, one could determine
-whether or not PIHM was occurring and estimate the parameters of the host distribution before parasite-induced mortality.  This was done
-graphically by determining whether the parameters of the negative binomial distributions fit to different truncations of the data showed a substantial change as the truncation point moved from heavily infected hosts to lightly infected hosts.
+% The first of these PIHM questions was addressed by \cite{Crofton1971a} who developed
+% a method to test for the presence of PIHM using the truncation of the negative binomial
+% distribution.  In short, the Crofton Method assumes that the distribution of
+% parasites across hosts before mortality occurs follows a negative binomial
+% distribution \citep{AndersonandMay1978,Shaw1998}.  As heavily infected hosts
+% begin to die, the negative binomial distribution is truncated and these hosts are no longer observed in a sample. In other words, the
+% observed host-parasite distribution and the pre-mortality host-parasite
+% distribution will predict substantially different numbers of hosts with high infect intensity (because those have died due to infection), but a similar number of hosts with low infection
+% intensity (because those have survived).  \cite{Crofton1971a} noted that by
+% starting with all observed hosts and iteratively fitting a negative binomial
+% distribution to hosts with lower and lower parasite loads, one could determine
+% whether or not PIHM was occurring and estimate the parameters of the host distribution before parasite-induced mortality.  This was done
+% graphically by determining whether the parameters of the negative binomial distributions fit to different truncations of the data showed a substantial change as the truncation point moved from heavily infected hosts to lightly infected hosts.
 The Crofton Method and this graphical technique are both still currently used to assess whether PIHM is a occurring in a system \citep{Ferguson2011}.
@@
 
-The second question regarding PIHM moves beyond a simple yes or no answer and attempts to quantify how parasite intensity affects host survival.
+% The second question regarding PIHM moves beyond a simple yes or no answer and attempts to quantify how parasite intensity affects host survival.
 \cite{Adjei1986} proposed a method to answer this question
-by first using the Crofton Method to estimate the pre-mortality parameters for
-a host-parasite distribution and then, given these
-parameters, estimating a host survival function that described how the
-probability of host-survival changed with increasing parasite load (see
-\emph{Methods}).  With this host survival function, one can estimate important host-parasite quantities such as the
-parasite intensity at which 50\% of hosts succumb to PIHM ($LD_{50}$), as well
-as the percent of hosts in a population succumbing to PIHM \citep{Adjei1986}.
+% by first using the Crofton Method to estimate the pre-mortality parameters for
+% a host-parasite distribution and then, given these
+% parameters, estimating a host survival function that described how the
+% probability of host-survival changed with increasing parasite load (see
+% \emph{Methods}).  With this host survival function, one can estimate important host-parasite quantities such as the
+% parasite intensity at which 50\% of hosts succumb to PIHM ($LD_{50}$), as well
+% as the percent of hosts in a population succumbing to PIHM \citep{Adjei1986}.
 
-Despite these methods both being over three decades old, they are still the
-primary means of answering questions about PIHM given distributional data
-\citep[but see][for an alternative to the Crofton
-Method]{Ferguson2011}. However, both methods have a few important limitations.
-The Crofton Method, and more recent methods \citep{Ferguson2011},
-rely on a visual test to determine whether or not PIHM is occurring in a
-system.  With no clear decision rule, it can be difficult to consistently determine the
-significance of PIHM across different host-parasite systems.
+% Despite these methods both being over three decades old, they are still the
+% primary means of answering questions about PIHM given distributional data
+% \citep[but see][for an alternative to the Crofton
+% Method]{Ferguson2011}. However, both methods have a few important limitations.
+% The Crofton Method, and more recent methods \citep{Ferguson2011},
+% rely on a visual test to determine whether or not PIHM is occurring in a
+% system.  With no clear decision rule, it can be difficult to consistently determine the
+% significance of PIHM across different host-parasite systems.
 In theory, the Adjei Method can ameliorate this problem.
 In addition to providing information on the host-survival function, this method can also be used to assess the significance of
-PIHM in a system.
-In practice, however, the Adjei Method has never been thoroughly tested and
-relies on a number of questionable data manipulations.
+% PIHM in a system.
+% In practice, however, the Adjei Method has never been thoroughly tested and
+% relies on a number of questionable data manipulations.
 
-In this study, we have
-three primary goals. First we wish to test reliability of the Adjei Method for
-answering the aforementioned questions regarding PIHM.  Second, we propose a
-novel method for answering these questions, compare its efficacy against the Adjei Method, and test both methods' ability to detect PIHM on empirical data.  Third, we
-discuss the limitations of inferring PIHM from distributional data alone and
-whether any method for inferring PIHM has a place in the future of
-parasitology.
+% In this study, we have
+% three primary goals. First we wish to test reliability of the Adjei Method for
+% answering the aforementioned questions regarding PIHM.  Second, we propose a
+% novel method for answering these questions, compare its efficacy against the Adjei Method, and test both methods' ability to detect PIHM on empirical data.  Third, we
+% discuss the limitations of inferring PIHM from distributional data alone and
+% whether any method for inferring PIHM has a place in the future of
+% parasitology.
 
@@
 
-\subsection*{The Adjei Method for estimating PIHM}
+% \subsection*{The Adjei Method for estimating PIHM}
 
-The Adjei Method for estimating PIHM has two steps.  The first
-step is to estimate the parameters of the pre-mortality host-parasite
-distribution using the Crofton Method.  The three parameters estimated are the
-total number of hosts before mortality $N_p$,  the mean number of parasites per
-host before mortality $\mu_p$, and the aggregation of parasites before
-mortality given by the parameter $k_p$ from a negative binomial distribution.
-When $k_p$ is small, parasites are highly aggregated among hosts and when
-$k_p$ is large parasites are more evenly spread out \citep{Wilson2002}.
+% The Adjei Method for estimating PIHM has two steps.  The first
+% step is to estimate the parameters of the pre-mortality host-parasite
+% distribution using the Crofton Method.  The three parameters estimated are the
+% total number of hosts before mortality $N_p$,  the mean number of parasites per
+% host before mortality $\mu_p$, and the aggregation of parasites before
+% mortality given by the parameter $k_p$ from a negative binomial distribution.
+% When $k_p$ is small, parasites are highly aggregated among hosts and when
+% $k_p$ is large parasites are more evenly spread out \citep{Wilson2002}.
 The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][]{Royce1990,Lester1984} and we provide a tested implementation of the method in SI 3.
 
-The second step of the Adjei Method is to estimate the shape of the host survival function. \cite{Adjei1986} assume that the host survival function follows the logistic form
+% The second step of the Adjei Method is to estimate the shape of the host survival function. \cite{Adjei1986} assume that the host survival function follows the logistic form
 
-\begin{equation}
-    h(x | a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
-    \label{eq:logistic}
-\end{equation}
-where $x$ is the parasite intensity in a given host and $a$ and $b$ are the two
-parameters of the logistic function. Generally, a larger $a$ allows for hosts to
-tolerate larger parasite intensities before experiencing parasite-induced mortality
-and a larger $b$ leads to a more rapid decline in the probability of host
-survival as parasite intensity increases.  The value $\exp(a / b)$ is referred
-to as the $LD_{50}$, which gives the parasite intensity at which 50\% of hosts
-experience mortality.
+% \begin{equation}
+%     h(x | a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+%     \label{eq:logistic}
+% \end{equation}
+% where $x$ is the parasite intensity in a given host and $a$ and $b$ are the two
+% parameters of the logistic function. Generally, a larger $a$ allows for hosts to
+% tolerate larger parasite intensities before experiencing parasite-induced mortality
+% and a larger $b$ leads to a more rapid decline in the probability of host
+% survival as parasite intensity increases.  The value $\exp(a / b)$ is referred
+% to as the $LD_{50}$, which gives the parasite intensity at which 50\% of hosts
+% experience mortality.
 
-To estimate this function the Adjei Method proceeds as follows.  First, it
-calculates the expected number of hosts with a given parasite load $x$ by using
-the equation $g(x ; \mu_p, k_p) * N_p$, where $g(x ; \cdot)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
-with $x$ parasites are paired as a single data point and the method then assumes that
-this data point follows a binomial distribution with the total number of
-``trials'' equal to the predicted number of hosts and the total number of
-``successes'' equal to the observed number of hosts. In some cases, the
-observed number of hosts is greater than the expected number of hosts and the
-Adjei Method alters the data so that the observed is equal to the predicted
-\citep{Adjei1986}.
+% To estimate this function the Adjei Method proceeds as follows.  First, it
+% calculates the expected number of hosts with a given parasite load $x$ by using
+% the equation $g(x ; \mu_p, k_p) * N_p$, where $g(x ; \cdot)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
+% with $x$ parasites are paired as a single data point and the method then assumes that
+% this data point follows a binomial distribution with the total number of
+% ``trials'' equal to the predicted number of hosts and the total number of
+% ``successes'' equal to the observed number of hosts. In some cases, the
+% observed number of hosts is greater than the expected number of hosts and the
+% Adjei Method alters the data so that the observed is equal to the predicted
+% \citep{Adjei1986}.
 After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation \ref{eq:logistic}.
@@
 
-While not included in the original implementation of the Adjei Method, a
-$\chi^2$ test with a degrees of freedom of 1 can be used to assess whether a GLM model that includes parasite
-intensity as a predictor of host survival probability is a ``better'' model than a
-GLM without this predictor.  This allows the Adjei Method to determine whether
-PIHM is a significant factor in a host-parasite system.
+% While not included in the original implementation of the Adjei Method, a
+% $\chi^2$ test with a degrees of freedom of 1 can be used to assess whether a GLM model that includes parasite
+% intensity as a predictor of host survival probability is a ``better'' model than a
+% GLM without this predictor.  This allows the Adjei Method to determine whether
+% PIHM is a significant factor in a host-parasite system.
 
-The Adjei Method's most glaring deficiency is the need to alter the observed
-data in order to fit the model into the binomial GLM framework.  A second more
-subtle problem with the Adjei Method is the potential need to bin data in order
-to predict greater than one host in a given parasite intensity class.  For
-example, if the total number of hosts pre-mortality was 50, the mean number of
-parasites per host pre-mortality was 100 and the aggregation parameter was 1, applying the equation $g(x ; \mu_p=100, k_p=1) * 50$ would result in
-less than 1 individual in all parasite intensities $x$. In other words, the
-Adjei Method cannot be applied to samples with either very high mean parasite
-loads, small sample sizes, or both without some sort of binning of the data.
-While this is not a flaw \emph{per se}, it does add a certain level of
-subjectivity (i.e. which bins should you use?) to a method that already has
-serious potential issues.  In this analysis, we always assume the Adjei Method is not binning the data, though we provide code for applying the binning method in SI 3.
+% The Adjei Method's most glaring deficiency is the need to alter the observed
+% data in order to fit the model into the binomial GLM framework.  A second more
+% subtle problem with the Adjei Method is the potential need to bin data in order
+% to predict greater than one host in a given parasite intensity class.  For
+% example, if the total number of hosts pre-mortality was 50, the mean number of
+% parasites per host pre-mortality was 100 and the aggregation parameter was 1, applying the equation $g(x ; \mu_p=100, k_p=1) * 50$ would result in
+% less than 1 individual in all parasite intensities $x$. In other words, the
+% Adjei Method cannot be applied to samples with either very high mean parasite
+% loads, small sample sizes, or both without some sort of binning of the data.
+% While this is not a flaw \emph{per se}, it does add a certain level of
+% subjectivity (i.e. which bins should you use?) to a method that already has
+% serious potential issues.  In this analysis, we always assume the Adjei Method is not binning the data, though we provide code for applying the binning method in SI 3.
 
-\subsection*{The Likelihood Method for estimating PIHM}
+\subsection*{A novel, likelihood method for estimating PIHM}
 
-Given the potential deficiencies of the Adjei Method, we provide an alternative
-approach for estimating parasite-induced host mortality (PIHM) that makes less
-assumptions and provides more reliable answers to the PIHM questions outlined above.  The Likelihood Method we propose
-does not require any binning or alteration of the data, potentially reduces the
-number of parameters that need to be estimated, and allows for standard
-statistical techniques to be used to assess the significance of PIHM in a
-system.
+Here we propose an alternative approach using a likelihood method that does not
+require binning or data alteration, potentially reduces the number of
+parameters to be estimated, and allows the significance of PIHM to be estimated
+using standard statistical techniques.
 
@@
 assumption is that the host survival function
-takes the form of a logistic curve given by equation \ref{eq:logistic}.  With these two explicit assumptions, the Likelihood Method estimates the 4 parameters $\mu_p$, $k_p$, $a$, and $b$.
+takes the form of a logistic curve given by
+\begin{equation}
+    h(x | a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+    \label{eq:logistic}
+\end{equation}
+With these two explicit assumptions, the Likelihood Method estimates the 4 parameters $\mu_p$, $k_p$, $a$, and $b$.
 
-To estimate these parameters, we need to define a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.
+To estimate these parameters, we first define a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.
 Using standard rules of conditional probability this distribution can be written as
@@
 
-\subsection*{Question 1: Is PIHM occurring?}
+\subsection{Evaluating the likelihood and Adjei Methods}
+
+\emph{Question 1: Can we detect PIHM?}
 
@@
 
-\subsection*{Question 2: Can the effect of parasite intensity on host survival be quantified?}
+\subsection*{Question 2: Can we estimate fatal parasite intensity and the host survival function?}
 
-To compare the ability of the Adjei Method and the Likelihood Method to recover $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above.
+To compare the ability of the Adjei Method and the Likelihood Method to recover the $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above.
 For each parameter
@@
 
-\subsection*{Application to data}
+\subsection*{Application to real data}
 
@@
 the hosts were two species of lizard fish \emph{Saurida tumbil} and
-\emph{Saurida undosquamis} that were infected by the cestode parasite
+\emph{Saurida undosquamis} that were infected by the cestode
 \emph{Callitetrarhynchus gracilis}.  Males and females of both fish species
@@
 
-In both these studies, the authors reported PIHM in some of the datasets and we test whether the Adjei
+In both studies, the authors reported PIHM in some of the datasets and we test whether the Adjei
 Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
@@
 first truncated the data at 2 parasites and then fit the Crofton Method for the
-female fish of both species.  We then parameterized the male pre-mortality
+female fish of both species.  Then, following the original authors' methods, we parameterized the male pre-mortality
 distributions for each species with the results from the females.  Finally, we
@@
 
-\subsection*{Question 1: Is PIHM occurring?}
+\subsection*{Question 1: Detecting presence of PIHM}
 
@@
 rate decreasing as sample size decreased.  This pattern is due to the issue of
-binning discussed in the \emph{Methods}. For small samples sizes, the
+binning discussed in the \emph{Introduction} and \emph{SI X}. For small samples sizes, the
 applicability of the Adjei Method is compromised without binning the observed
@@
 
-\subsection*{Question 2: Can the effect of parasite intensity on host survival be quantified?}
+\subsection*{Question 2: Estimating the $LD_{50}$ and survival function}
 
@@
 was largely unbiased, with small biases occurring for host survival functions
-that were gradual.  On the other hand, the Adjei Method
-always produced more biased estimates of the $LD_{50}$ than the Likelihood
-Method across all parameter combinations (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  For $\mu_p = 10$, the $LD_{50}$
+that were gradual. The precision of the Likelihood Method's $LD_{50}$ estimates decreased
+(increasing coefficient of variation) as sample size decreased for all
+parameter combinations we examined (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).
+
+The Adjei Method
+always produced biased estimates of the $LD_{50}$ across all parameter combinations (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  For $\mu_p = 10$, the $LD_{50}$
 estimates from the Adjei Method were largely unbiased for large samples sizes,
@@
 produced biased estimates of $LD_{50}$ across all sample sizes, with bias
-increasing as sample size decreased (SI2 Fig 3, 4).
-
-The precision of the $LD_{50}$ estimates for the Likelihood Method decreased
-(increasing coefficient of variation) as sample size decreased for all
-parameter combinations we examined (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  The $LD_{50}$ estimates from the Adjei
-Method showed a similar pattern, with large decreases in precision occurring
+increasing as sample size decreased (SI2 Fig 3, 4). The $LD_{50}$ estimates from the Adjei
+Method showed large decreases in precision occurring
 for the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).
@@
 PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
-dataset in which the the Likelihood Method did not detect a significant effect of PIHM was the Adjei dataset
+dataset in which the Likelihood Method did not detect a significant effect of PIHM was the Adjei dataset
 for female \emph{S. tumbil}.  For this dataset there was a marginally significant effect
@@
 
-The Adjei Method detected significant PIHM in 9 of the 10 datasets given (Table \ref{table:pihm}).
-This is consistent with the previous results which show that the Adjei Method has a very high Type I error rate.
+The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}),  showing that the Adjei Method has a very high Type I error rate.
 
@@
 
-Determining whether PIHM is a significant factor for a host populations is
-critically important in many systems, but detecting and describing PIHM given
-only observational data is notoriously difficult.  We show that the Adjei
-Method, the only currently proposed method to estimate the host survival
-function and the $LD_{50}$ from observational data, has some serious
-methodological problems that result in biased estimates even under the most idealistic conditions.
-Moreover, we show that the Adjei Method has a seriously inflated Type I error
-rate, meaning it will often detect PIHM even when it is not present. Moreover,
-for small, realistic sample sizes the Adjei Method behaves erratically; a
-consequence of the need to subjectively bin the data in order to predict
-parasite intensity classes with at least one host.
+% Sara W Outline
+% 5 paragraph disc Outline:
+%  Describe benefits of new method (LM)
+% No data manipulation
+% Estimates fewer parameter
+% Statistical output
+%  A new method to quantitatively explore mortality using intensity data
 
-To attempt to ameliorate the flaws in the Adjei Method, we proposed a more
-general method to determine both whether or not PIHM is occurring in a system
-and to quantify the survival function.  We show that this method is
-asymptotically unbiased when estimating the host-survival function for all of
-the parameter space that we explored and we found that it produces unbiased and
-precise estimates of the $LD_{50}$ for small, realistic sample sizes. Moreover,
-this novel method has a Type I error rate close to the pre-set level of $\alpha = 0.05$ and high power for detecting PIHM for realistic samples sizes.
-  However, we note that
-the Likelihood Method produces seriously biased estimates of the host survival
-function ($a$ and $b$) for sample sizes typically observed in many host-parasite studies.  The bias was most severe for steep host survival functions,
-due to large changes in the values of $a$ and $b$ only slightly changing an
-already steep survival function.  Given these results, neither the
-Likelihood Method nor the Adjei Method could confidently recover the exact shape of the host survival function for small, realistic sample sizes.
+% Compare  LM with AM
+% LM always does better,
+% Detecting  PIHM: more power, less false detection
+% Ld50/survival curve: LM more precise/less biased. But these do get worse with decreasing sample size.
+% real data , consistent with original author’s results, supports AM type 1 error issues
+% AM is fundamentally flawed, LM is clearly better under ideal simulations, but can it be implemented in the real world?
 
-We also fit both the Likelihood Method and the Adjei Method to empirical data to
-determine whether they could detect PIHM that had been previously reported
-based on visual assessments.  Consistent with our simulation results, we found
-that the Adjei Method tended to detect PIHM where it had not been previously
-reported, while the Likelihood Method's detection of PIHM was consistent with
-previously reported PIHM in a given dataset.  Taken together, these results suggest that the Adjei Method is fundamentally
-flawed and we recommend using the Likelihood Method for detecting PIHM and describing attributes of the host survival function.
+% 3. Data set limitations (for the LM, because no one should even try the AM):
+% a.
+requires large sample sizes (n>X) which may be impossible to collect from the species we would be most interested in getting this data for, eg, large mammals of conservation concern.
+While 50-100 is very realistic for invertebrates or fish, not so for larger verts
+% b. highly aggregated or high mean intensity increase needed sample size (is this right for LM?) which may make this difficult for many helminths with worm burdens in the 100s-1000s
+% c. low parasite induced mortality may be hard to detect- particularly an issue for worms in definitive hosts
+% d.
+shape of survival curve matters, although often assumed to be linear (eg Anderson may model) can’t know apriori, and method fails  for both perfectly linear and exceptionally steep survival functions
+% e. math works, but not for all parasite-host systems
 
-While we have improved upon the previously existing methods for answering questions
-about PIHM, we cannot belie the fact that estimating PIHM from observational
-data alone is ladened with assumptions and difficulties \citep{McCallum2000a}.
-The most fundamental assumption of all methods for estimating PIHM is that the
-shape of the pre-mortality host-parasite distribution is known and follows a negative binomial distribution.
-While there is substantial empirical and theoretical evidence to justify the
-use of the negative binomial distribution as the pre-mortality distribution for
+% 4.even when the math works, should we believe it?
+% a. All methods rely on same fundamental assumptions of the crofton method.
+% b. assume that pre mort psite dist follows a negative binomial. And this is valid because the NB is extremely flexible and lots of empirical and theoretical work says so.
+% c. but, detecting mortality requires that the observed data be significantly different than the NB but because NB is so flexible it can be fit to almost anything, making it impossible to detect pihm
+% d.assume truncation from pihm, but could be caused by lots of other things, so detecting significant pihm does not mean it occurred.
+% f. serious fundamental issues that need to be recognized and acknowledge with any use of this technique
+
+% 5.(conclusion)Does any of this have a place in quantitative parasitology?
+
+In this study we propose a novel method, the Likelihood Method, for detecting
+and describing PIHM from parasite intensity data.  Unlike currently proposed
+methods for estimating PIHM, the Likelihood Method does not require any
+unnecessary data manipulation, uses fewer parameters [it can, but not always],
+and provides a quantitative decision rule for determining whether or not PIHM
+is occurring in a host-parasite system.
+
+Using simulated datasets, we show that the Likelihood Method correctly detected
+PIHM when it was occurring and correctly failed to detect PIHM when it was not
+occurring for sample sizes below 100 hosts. When applied to real data, the
+Likelihood Method gave results consistent with what was previously reported. In
+contrast, the Adjei Method, the only currently proposed method to estimate the
+host survival function and the $LD_{50}$ from parasite intensity data, consistently
+detected PIHM even when it was not present and behaved erratically for small sample sizes (< 200
+hosts); a consequence of the need to
+subjectively bin the data to predict parasite intensity classes with
+at least one host.  Finally, when applied to real data, the Adjei Method overestimated the occurrence of PIHM relative to what was previously reported.
+
+In addition to accurately detecting PIHM, the Likelihood Method also provided
+unbiased and precise estimates of a host survival function's $LD_{50}$ across
+all sample sizes we considered.  While the Likelihood Method did produce
+unbiased estimates of the host survival function ($a$ and $b$) for
+unrealistically large sample sizes, it gave severely biased estimates of the
+host survival function for realistic samples sizes. The bias was most severe
+for steep host survival functions due to large changes in the values of $a$ and
+$b$ only slightly changing an already steep survival function. The Adjei
+Method, on the other hand, gave biased predictions of the $LD_{50}$ and host
+survival function across all parameter combinations, with bias increasing with
+increasing parasite aggregation and increasing mean parasite intensity.  Taken
+together, these results show that Adjei Method is fundamentally flawed and the
+Likelihood Method may offer a potential alternative for estimating PIHM.
+
+While we have improved upon the previously existing methods for answering
+questions about PIHM, we cannot belie the fact that estimating PIHM from
+observational data alone is ladened with assumptions and difficulties
+\citep{McCallum2000a}. The most fundamental assumption of all methods for
+estimating PIHM is that the shape of the pre-mortality host-parasite
+distribution is known and follows a negative binomial distribution. While there
+is substantial empirical and theoretical evidence to justify the use of the
+negative binomial distribution as the pre-mortality distribution for
 macroparasites across hosts \citep{Calabrese2011,Anderson1982a,Shaw1998}, it is
 widely recognized that different processes can lead to a variety of
-distributions of parasites across hosts \citep{Isham1995,Grenfell1995b,Wilson2002, Duerr2003}.  However, the
-critical assumption of the pre-mortality distribution is not that the processes
-leading to the pre-mortality distribution generate a negative binomial
-distribution, but rather that the pre-mortality distribution is well-fit by a
-negative binomial distribution. The extreme flexibility of the negative binomial
+distributions of parasites across hosts
+\citep{Isham1995,Grenfell1995b,Wilson2002, Duerr2003}.  However, the critical
+assumption of the pre-mortality distribution is not that the processes leading
+to the pre-mortality distribution generate a negative binomial distribution,
+but rather that the pre-mortality distribution is well-fit by a negative
+binomial distribution. The extreme flexibility of the negative binomial
 distribution makes it a reasonable candidate distribution for the pre-mortality
@@
 the host-survival function is such that uninfected individuals and individuals
-with low parasite intensity experience essentially no PIHM.
-\cite{Lanciani1989} illustrated the importance of this assumption by showing
-that when hosts experienced a linear decrease in survival probability the
-Crofton Method could not detect PIHM.   As the most fundamental models of host-parasite dynamics assume a linear decrease in host survival probability with
-increasing parasite intensity \citep{AndersonandMay1978}, the failure of these methods
-to detect this relationship is a significant disconnect between empirical and
-theoretical disease ecology.  However, empirical work has shown that non-linear
-functions of host survival are not uncommon in host-parasite systems
-\citep{Benesh2011}, so this assumption alone does not preclude the use of PIHM
-methods on empirical data.
+with low parasite intensity experience essentially no PIHM. \cite{Lanciani1989}
+illustrated the importance of this assumption by showing that when hosts
+experienced a linear decrease in survival probability the Crofton Method could
+not detect PIHM.   As the most fundamental models of host-parasite dynamics
+assume a linear decrease in host survival probability with increasing parasite
+intensity \citep{AndersonandMay1978}, the failure of these methods to detect
+this relationship is a significant disconnect between empirical and theoretical
+disease ecology.  However, empirical work has shown that non-linear functions
+of host survival are not uncommon in host-parasite systems \citep{Benesh2011},
+so this assumption alone does not preclude the use of PIHM methods on empirical data.
 
@@
 methods will be able to detect PIHM if a negative binomial distribution is an
-adequate fit to the post-mortality distribution.
-This is simply because there will be no need for a more complex model (either truncation of the negative binomial or the model given in equation \ref{eq:dist}) if a negative binomial distribution already fits the data.
-As many
-observed host-parasite distributions are not significantly different from a
-negative binomial distribution, there may be limited cases where these PIHM methods
-can even be considered.
+adequate fit to the post-mortality distribution.  This is simply because there
+will be no need for a more complex model (either truncation of the negative
+binomial or the model given in equation \ref{eq:dist}) if a negative binomial
+distribution already fits the data. As many observed host-parasite
+distributions are not significantly different from a negative binomial
+distribution, there may be limited cases where these PIHM methods can even be
+considered.
 
 Finally, all of these methods assume that the truncation of a negative binomial
-distribution is due to PIHM, but previous studies have shown that a variety of other processes can lead to the truncation of a negative
-binomial distribution such as within host parasite density-dependence, age-
-dependent variation in host resistance and heterogeneous rate of infection
-\citep{McCallum2000a,Anderson1982a,Rousset1996}.  Therefore, even detecting ``significant'' PIHM in a
-dataset does not mean that PIHM is cause of the truncation.
+distribution is due to PIHM, but previous studies have shown that a variety of
+other processes can lead to the truncation of a negative binomial distribution
+such as within host parasite density-dependence, age- dependent variation in
+host resistance and heterogeneous rate of infection
+\citep{McCallum2000a,Anderson1982a,Rousset1996}.  Therefore, even detecting
+``significant'' PIHM in a dataset does not mean that PIHM is cause of the
+truncation.
 
-Given these numerous caveats, is there a place in parasitology for methods that
+Given these numerous caveats, is the is there a place in parasitology for methods that
 estimate PIHM from distributional data?  We are in agreement with

### 2015-05-04 docs/parasite_mortatlity_manuscript.tex
Replaced discussion with new version. Moved Adjei descrip to supplementary materials

@@
 \cite{Adjei1986} modeled host survival as a logistic function and then used a
-binomial generalized linear model (GLM) to estimate the logistic parameters.
+binomial generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI X for a technical description of the Adjei Method}).
 These methods appeared to provide an estimate the parasite intensity at which
@@
 
-% Infectious agents can have major impacts on animal populations
-% through changing population dynamics and stability \citep{Dobson1992}, altering
-% predator-prey interactions \citep{Joly2004}, and even causing species decline
-% and extinction \citep{DeCastro2005a,McCallum2012b}. Accurate estimates of
-% parasite-induced host mortality (PIHM) in wild animals are important for
-% understanding what regulates both host and parasite populations and to make
-% predictions about disease transmission in natural systems. Although a negative
-% impact on host fitness is a fundamental component of parasitism
-% \citep{Lafferty2002} [more], it is notoriously difficult to quantify PIHM in
-% wild animal populations \citep{Lester1984,McCallum2000a}.
-
-% To conclusively identify PIHM in wild animal populations, it is necessary to
-% experimentally infect and track host populations; a method that is rarely
-% possible in most wild animal systems \citep{McCallum2000a}.  Instead, parasitologists
-% are often only able to determine the parasite intensity on some number of sampled hosts.  This snapshot, distributional data is far from the
-% ideal type of data for addressing questions regarding PIHM, but this is the type of data on which most questions regarding PIHM are asked \citep{Ferguson2011,Royce1990,Lanciani1989,Lester1984,Lester1977}.
-In particular, parasitologists are often interested in asking two questions from this data: 1) Is PIHM occurring in a host-parasite system?
-and if PIHM is occurring 2) Can the effect of parasite intensity on host survival be quantified?
-While both macro and
-% microparasites have detrimental and possibly fatal effects on hosts, for the
-% remainder of this paper we limit our discussion to macroparasites that can be
-% discretely counted within a host.
-
-% The first of these PIHM questions was addressed by \cite{Crofton1971a} who developed
-% a method to test for the presence of PIHM using the truncation of the negative binomial
-% distribution.  In short, the Crofton Method assumes that the distribution of
-% parasites across hosts before mortality occurs follows a negative binomial
-% distribution \citep{AndersonandMay1978,Shaw1998}.  As heavily infected hosts
-% begin to die, the negative binomial distribution is truncated and these hosts are no longer observed in a sample. In other words, the
-% observed host-parasite distribution and the pre-mortality host-parasite
-% distribution will predict substantially different numbers of hosts with high infect intensity (because those have died due to infection), but a similar number of hosts with low infection
-% intensity (because those have survived).  \cite{Crofton1971a} noted that by
-% starting with all observed hosts and iteratively fitting a negative binomial
-% distribution to hosts with lower and lower parasite loads, one could determine
-% whether or not PIHM was occurring and estimate the parameters of the host distribution before parasite-induced mortality.  This was done
-% graphically by determining whether the parameters of the negative binomial distributions fit to different truncations of the data showed a substantial change as the truncation point moved from heavily infected hosts to lightly infected hosts.
-The Crofton Method and this graphical technique are both still currently used to assess whether PIHM is a occurring in a system \citep{Ferguson2011}.
-We give a thorough description and implementation of the Crofton Method in \emph{Supplementary Information} (SI) 1 and discuss the validity of its assumptions in the \emph{Discussion}.
-
-% The second question regarding PIHM moves beyond a simple yes or no answer and attempts to quantify how parasite intensity affects host survival.
-\cite{Adjei1986} proposed a method to answer this question
-% by first using the Crofton Method to estimate the pre-mortality parameters for
-% a host-parasite distribution and then, given these
-% parameters, estimating a host survival function that described how the
-% probability of host-survival changed with increasing parasite load (see
-% \emph{Methods}).  With this host survival function, one can estimate important host-parasite quantities such as the
-% parasite intensity at which 50\% of hosts succumb to PIHM ($LD_{50}$), as well
-% as the percent of hosts in a population succumbing to PIHM \citep{Adjei1986}.
-
-% Despite these methods both being over three decades old, they are still the
-% primary means of answering questions about PIHM given distributional data
-% \citep[but see][for an alternative to the Crofton
-% Method]{Ferguson2011}. However, both methods have a few important limitations.
-% The Crofton Method, and more recent methods \citep{Ferguson2011},
-% rely on a visual test to determine whether or not PIHM is occurring in a
-% system.  With no clear decision rule, it can be difficult to consistently determine the
-% significance of PIHM across different host-parasite systems.
-In theory, the Adjei Method can ameliorate this problem.
-In addition to providing information on the host-survival function, this method can also be used to assess the significance of
-% PIHM in a system.
-% In practice, however, the Adjei Method has never been thoroughly tested and
-% relies on a number of questionable data manipulations.
-
-% In this study, we have
-% three primary goals. First we wish to test reliability of the Adjei Method for
-% answering the aforementioned questions regarding PIHM.  Second, we propose a
-% novel method for answering these questions, compare its efficacy against the Adjei Method, and test both methods' ability to detect PIHM on empirical data.  Third, we
-% discuss the limitations of inferring PIHM from distributional data alone and
-% whether any method for inferring PIHM has a place in the future of
-% parasitology.
-
 \section*{Methods}
-
-% \subsection*{The Adjei Method for estimating PIHM}
-
-% The Adjei Method for estimating PIHM has two steps.  The first
-% step is to estimate the parameters of the pre-mortality host-parasite
-% distribution using the Crofton Method.  The three parameters estimated are the
-% total number of hosts before mortality $N_p$,  the mean number of parasites per
-% host before mortality $\mu_p$, and the aggregation of parasites before
-% mortality given by the parameter $k_p$ from a negative binomial distribution.
-% When $k_p$ is small, parasites are highly aggregated among hosts and when
-% $k_p$ is large parasites are more evenly spread out \citep{Wilson2002}.
-The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][]{Royce1990,Lester1984} and we provide a tested implementation of the method in SI 3.
-
-% The second step of the Adjei Method is to estimate the shape of the host survival function. \cite{Adjei1986} assume that the host survival function follows the logistic form
-
-% \begin{equation}
-%     h(x | a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
-%     \label{eq:logistic}
-% \end{equation}
-% where $x$ is the parasite intensity in a given host and $a$ and $b$ are the two
-% parameters of the logistic function. Generally, a larger $a$ allows for hosts to
-% tolerate larger parasite intensities before experiencing parasite-induced mortality
-% and a larger $b$ leads to a more rapid decline in the probability of host
-% survival as parasite intensity increases.  The value $\exp(a / b)$ is referred
-% to as the $LD_{50}$, which gives the parasite intensity at which 50\% of hosts
-% experience mortality.
-
-% To estimate this function the Adjei Method proceeds as follows.  First, it
-% calculates the expected number of hosts with a given parasite load $x$ by using
-% the equation $g(x ; \mu_p, k_p) * N_p$, where $g(x ; \cdot)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
-% with $x$ parasites are paired as a single data point and the method then assumes that
-% this data point follows a binomial distribution with the total number of
-% ``trials'' equal to the predicted number of hosts and the total number of
-% ``successes'' equal to the observed number of hosts. In some cases, the
-% observed number of hosts is greater than the expected number of hosts and the
-% Adjei Method alters the data so that the observed is equal to the predicted
-% \citep{Adjei1986}.
-After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation \ref{eq:logistic}.
-This model provides estimates for parameters $a$, $b$ and $LD_{50}$.
-
-% While not included in the original implementation of the Adjei Method, a
-% $\chi^2$ test with a degrees of freedom of 1 can be used to assess whether a GLM model that includes parasite
-% intensity as a predictor of host survival probability is a ``better'' model than a
-% GLM without this predictor.  This allows the Adjei Method to determine whether
-% PIHM is a significant factor in a host-parasite system.
-
-% The Adjei Method's most glaring deficiency is the need to alter the observed
-% data in order to fit the model into the binomial GLM framework.  A second more
-% subtle problem with the Adjei Method is the potential need to bin data in order
-% to predict greater than one host in a given parasite intensity class.  For
-% example, if the total number of hosts pre-mortality was 50, the mean number of
-% parasites per host pre-mortality was 100 and the aggregation parameter was 1, applying the equation $g(x ; \mu_p=100, k_p=1) * 50$ would result in
-% less than 1 individual in all parasite intensities $x$. In other words, the
-% Adjei Method cannot be applied to samples with either very high mean parasite
-% loads, small sample sizes, or both without some sort of binning of the data.
-% While this is not a flaw \emph{per se}, it does add a certain level of
-% subjectivity (i.e. which bins should you use?) to a method that already has
-% serious potential issues.  In this analysis, we always assume the Adjei Method is not binning the data, though we provide code for applying the binning method in SI 3.
 
@@
 
-In this study we propose a novel method, the Likelihood Method, for detecting
-and describing PIHM from parasite intensity data.  Unlike currently proposed
-methods for estimating PIHM, the Likelihood Method does not require any
-unnecessary data manipulation, uses fewer parameters [it can, but not always],
-and provides a quantitative decision rule for determining whether or not PIHM
-is occurring in a host-parasite system.
+Quantifying the impact of parasitism on wild host populations is critical in
+both disease modeling and wildlife management. Ideally the relationship between
+infection intensity and host survival would be measured experimentally, but for
+logistical and ethical reasons, this is often impossible \citep{McCallum2000a}.
+Looking for evidence of mortality in parasite distribution data requires the
+least amount of information, but is notoriously difficult to implement. The
+methodological flaws in the Adjei Method limit its utility, so here we propose
+an alternative, likelihood-based, method to estimated host survival and the
+$LD_{50}$ from observed parasite intensity data.  At a theoretical level this
+method is a significant improvement because it can require fewer parameters,
+provides a statistical decision rule for identifying PIHM and does not require
+any data manipulation.
 
-Using simulated datasets, we show that the Likelihood Method correctly detected
-PIHM when it was occurring and correctly failed to detect PIHM when it was not
-occurring for sample sizes below 100 hosts. When applied to real data, the
-Likelihood Method gave results consistent with what was previously reported. In
-contrast, the Adjei Method, the only currently proposed method to estimate the
-host survival function and the $LD_{50}$ from parasite intensity data, consistently
-detected PIHM even when it was not present and behaved erratically for small sample sizes (< 200
-hosts); a consequence of the need to
-subjectively bin the data to predict parasite intensity classes with
-at least one host.  Finally, when applied to real data, the Adjei Method overestimated the occurrence of PIHM relative to what was previously reported.
+Under simulated conditions, the likelihood method always out performs the Adjei Method. For simply detecting the presence of PIHM, the Likelihood
+Method is both more powerful and has fewer false detection events (Type I
+errors).  When both methods were applied to published datasets previously used
+in PIHM analyses, the Adjei Method tended to detect PIHM where it had not previously been
+reported, consistent with the high Type I error rate observed in our
+simulations. The likelihood method is also universally more precise and less
+biased in calculations of both the parasite $LD_{50}$ and host survival curve.
+However, while only the Likelihood Method produces precise and unbiased $LD_{50}$
+estimates, neither method can provide unbiased estimates of the host survival
+function at realistic sample sizes.  These simulations demonstrate that
+the novel Likelihood Method is more powerful and precise than the previously propose Adjei Method.
 
-In addition to accurately detecting PIHM, the Likelihood Method also provided
-unbiased and precise estimates of a host survival function's $LD_{50}$ across
-all sample sizes we considered.  While the Likelihood Method did produce
-unbiased estimates of the host survival function ($a$ and $b$) for
-unrealistically large sample sizes, it gave severely biased estimates of the
-host survival function for realistic samples sizes. The bias was most severe
-for steep host survival functions due to large changes in the values of $a$ and
-$b$ only slightly changing an already steep survival function. The Adjei
-Method, on the other hand, gave biased predictions of the $LD_{50}$ and host
-survival function across all parameter combinations, with bias increasing with
-increasing parasite aggregation and increasing mean parasite intensity.  Taken
-together, these results show that Adjei Method is fundamentally flawed and the
-Likelihood Method may offer a potential alternative for estimating PIHM.
+Although superior to the Adjei Method, the LM may still not always be
+applicable to real data.  The Likelihood Method requires relatively large
+sample sizes (n>50-100) [Are these relatively large? They seem small to me], that although reasonable
+to obtain for invertebrates or small fish may be completely unfeasible for many
+vertebrates, particularly the species of conservation concern where addressing
+the impact of parasitism would be most important. An even larger sample size is
+required to capture the full parasite distribution when parasites are highly
+aggregated, mean infection intensity is high, or parasite prevalence is low [same as aggregation],
+all of which are common in many parasitic helminths.  Low parasite-induced host
+mortality, as might be predicted in many definitive hosts, may also be very
+difficult to detect and require impossibly large sample sizes. And, even
+when sample size is sufficient, these methods can only detect PIHM is the host
+survival curve is non-linear \citep{Lanciani1989}. Most host-parasite models assume a linear
+relationship between survival and infection intensity \citep{AndersonandMay1978,McCallum2000a}, however
+nonlinear survival functions are not uncommon in empirical host-parasite
+systems \citep{Benesh2011}.
+And, while linear functions make PIHM undetectable, at the other extreme, steep, non-linear survival curves produce severely biased estimates of the survival function.
+Even the Likelihood Method is probably limited to
+detecting PIHM and estimating $LD_{50}$ in systems where greater than 50 hosts can be
+collected, parasites are common and only moderately aggregated, and
+substantial host mortality occurs at relatively low parasite intensity.
 
-While we have improved upon the previously existing methods for answering
-questions about PIHM, we cannot belie the fact that estimating PIHM from
-observational data alone is ladened with assumptions and difficulties
-\citep{McCallum2000a}. The most fundamental assumption of all methods for
-estimating PIHM is that the shape of the pre-mortality host-parasite
-distribution is known and follows a negative binomial distribution. While there
-is substantial empirical and theoretical evidence to justify the use of the
-negative binomial distribution as the pre-mortality distribution for
-macroparasites across hosts \citep{Calabrese2011,Anderson1982a,Shaw1998}, it is
-widely recognized that different processes can lead to a variety of
-distributions of parasites across hosts
-\citep{Isham1995,Grenfell1995b,Wilson2002, Duerr2003}.  However, the critical
-assumption of the pre-mortality distribution is not that the processes leading
-to the pre-mortality distribution generate a negative binomial distribution,
-but rather that the pre-mortality distribution is well-fit by a negative
-binomial distribution. The extreme flexibility of the negative binomial
-distribution makes it a reasonable candidate distribution for the pre-mortality
-distributions.  Therefore, we do not see this assumption as central problem in
-any of the proposed methods.
-
-However, to use the pre-mortality distribution to infer whether or not the PIHM
-is occurring in a system requires an explicit assumption about the host
-survival function and the shape of the post-mortality distribution.  Regarding
-the host-survival function, all currently proposed methods of PIHM assume that
-the host-survival function is such that uninfected individuals and individuals
-with low parasite intensity experience essentially no PIHM. \cite{Lanciani1989}
-illustrated the importance of this assumption by showing that when hosts
-experienced a linear decrease in survival probability the Crofton Method could
-not detect PIHM.   As the most fundamental models of host-parasite dynamics
-assume a linear decrease in host survival probability with increasing parasite
-intensity \citep{AndersonandMay1978}, the failure of these methods to detect
-this relationship is a significant disconnect between empirical and theoretical
-disease ecology.  However, empirical work has shown that non-linear functions
-of host survival are not uncommon in host-parasite systems \citep{Benesh2011},
-so this assumption alone does not preclude the use of PIHM methods on empirical data.
-
-Regarding the shape of the post-mortality distribution, all of these methods
-require that the post-mortality distribution be significantly different from a
-negative binomial distribution.  This is necessary because none of the above
-methods will be able to detect PIHM if a negative binomial distribution is an
-adequate fit to the post-mortality distribution.  This is simply because there
-will be no need for a more complex model (either truncation of the negative
-binomial or the model given in equation \ref{eq:dist}) if a negative binomial
-distribution already fits the data. As many observed host-parasite
-distributions are not significantly different from a negative binomial
-distribution, there may be limited cases where these PIHM methods can even be
-considered.
-
-Finally, all of these methods assume that the truncation of a negative binomial
-distribution is due to PIHM, but previous studies have shown that a variety of
-other processes can lead to the truncation of a negative binomial distribution
-such as within host parasite density-dependence, age- dependent variation in
-host resistance and heterogeneous rate of infection
-\citep{McCallum2000a,Anderson1982a,Rousset1996}.  Therefore, even detecting
-``significant'' PIHM in a dataset does not mean that PIHM is cause of the
-truncation.
+While we have improved on the existing methods for quantifying
+PIHM from parasite intensity data, all such methods require several
+fundamental, and potentially problematic assumptions.
+Nearly all methods to date derive from \cite[][but see \cite{Ferguson2011}]{Crofton1971a} and assume that, prior to any PIHM, parasites are distributed in the host population following a
+negative binomial distribution. But, it is fundamentally impossible to know
+what the pre-mortality parasite distribution was in a wild host population and
+it is widely recognized that different processes can lead to a variety of
+parasite distributions in hosts \citep{Anderson1982a, Duerr2003}. However, the negative binomial is extremely
+flexible and there is substantial empirical and theoretical evidence to support
+the assumption that, prior to any PIHM, parasite distributions can be fit by a negative binomial distribution \citep{Shaw1995,Shaw1998,Wilson2002}. Unfortunately, this
+flexibility in the distribution may also reduce our ability to detect PIHM. If
+a negative binomial can be fit to the observed post-mortality parasite
+distribution then, regardless of how lethal the parasite was, it will be
+impossible to detect PIHM because there is no need for a more complex model.
+Most observed parasite distributions are well fit by the negative binomial distribution \citep{Shaw1998}, suggesting
+that systems where these methods are applicable may be more the exception than
+the rule.  Finally, even when truncation of the negative binomial distribution is detected, it may be caused by other
+processes such as density dependence, age dependent variation in host
+resistance and heterogeneous infection rates \citep{McCallum2000a,Anderson1982a,Rousset1996}.  This means that in the rare event
+that PIHM is detected, it may actually not be the result of PIHM.
 
@@
 system under consideration.
+
+% In this study we propose a novel method, the Likelihood Method, for detecting
+% and describing PIHM from parasite intensity data.  Unlike currently proposed
+% methods for estimating PIHM, the Likelihood Method does not require any
+% unnecessary data manipulation, uses fewer parameters [it can, but not always],
+% and provides a quantitative decision rule for determining whether or not PIHM
+% is occurring in a host-parasite system.
+
+% Using simulated datasets, we show that the Likelihood Method correctly detected
+% PIHM when it was occurring and correctly failed to detect PIHM when it was not
+% occurring for sample sizes below 100 hosts. When applied to real data, the
+% Likelihood Method gave results consistent with what was previously reported. In
+% contrast, the Adjei Method, the only currently proposed method to estimate the
+% host survival function and the $LD_{50}$ from parasite intensity data, consistently
+% detected PIHM even when it was not present and behaved erratically for small sample sizes (< 200
+% hosts); a consequence of the need to
+% subjectively bin the data to predict parasite intensity classes with
+% at least one host.  Finally, when applied to real data, the Adjei Method overestimated the occurrence of PIHM relative to what was previously reported.
+
+% In addition to accurately detecting PIHM, the Likelihood Method also provided
+% unbiased and precise estimates of a host survival function's $LD_{50}$ across
+% all sample sizes we considered.  While the Likelihood Method did produce
+% unbiased estimates of the host survival function ($a$ and $b$) for
+% unrealistically large sample sizes, it gave severely biased estimates of the
+% host survival function for realistic samples sizes. The bias was most severe
+% for steep host survival functions due to large changes in the values of $a$ and
+% $b$ only slightly changing an already steep survival function. The Adjei
+% Method, on the other hand, gave biased predictions of the $LD_{50}$ and host
+% survival function across all parameter combinations, with bias increasing with
+% increasing parasite aggregation and increasing mean parasite intensity.  Taken
+% together, these results show that Adjei Method is fundamentally flawed and the
+% Likelihood Method may offer a potential alternative for estimating PIHM.
+
+% While we have improved upon the previously existing methods for answering
+% questions about PIHM, we cannot belie the fact that estimating PIHM from
+% observational data alone is ladened with assumptions and difficulties
+% \citep{McCallum2000a}. The most fundamental assumption of all methods for
+% estimating PIHM is that the shape of the pre-mortality host-parasite
+% distribution is known and follows a negative binomial distribution. While there
+% is substantial empirical and theoretical evidence to justify the use of the
+% negative binomial distribution as the pre-mortality distribution for
+% macroparasites across hosts \citep{Calabrese2011,Anderson1982a,Shaw1998}, it is
+% widely recognized that different processes can lead to a variety of
+% distributions of parasites across hosts
+% \citep{Isham1995,Grenfell1995b,Wilson2002, Duerr2003}.  However, the critical
+% assumption of the pre-mortality distribution is not that the processes leading
+% to the pre-mortality distribution generate a negative binomial distribution,
+% but rather that the pre-mortality distribution is well-fit by a negative
+% binomial distribution. The extreme flexibility of the negative binomial
+% distribution makes it a reasonable candidate distribution for the pre-mortality
+% distributions.  Therefore, we do not see this assumption as central problem in
+% any of the proposed methods.
+
+% However, to use the pre-mortality distribution to infer whether or not the PIHM
+% is occurring in a system requires an explicit assumption about the host
+% survival function and the shape of the post-mortality distribution.  Regarding
+% the host-survival function, all currently proposed methods of PIHM assume that
+% the host-survival function is such that uninfected individuals and individuals
+% with low parasite intensity experience essentially no PIHM. \cite{Lanciani1989}
+% illustrated the importance of this assumption by showing that when hosts
+% experienced a linear decrease in survival probability the Crofton Method could
+% not detect PIHM.   As the most fundamental models of host-parasite dynamics
+% assume a linear decrease in host survival probability with increasing parasite
+% intensity \citep{AndersonandMay1978}, the failure of these methods to detect
+% this relationship is a significant disconnect between empirical and theoretical
+% disease ecology.  However, empirical work has shown that non-linear functions
+% of host survival are not uncommon in host-parasite systems \citep{Benesh2011},
+% so this assumption alone does not preclude the use of PIHM methods on empirical data.
+
+% Regarding the shape of the post-mortality distribution, all of these methods
+% require that the post-mortality distribution be significantly different from a
+% negative binomial distribution.  This is necessary because none of the above
+% methods will be able to detect PIHM if a negative binomial distribution is an
+% adequate fit to the post-mortality distribution.  This is simply because there
+% will be no need for a more complex model (either truncation of the negative
+% binomial or the model given in equation \ref{eq:dist}) if a negative binomial
+% distribution already fits the data. As many observed host-parasite
+% distributions are not significantly different from a negative binomial
+% distribution, there may be limited cases where these PIHM methods can even be
+% considered.
+
+% Finally, all of these methods assume that the truncation of a negative binomial
+% distribution is due to PIHM, but previous studies have shown that a variety of
+% other processes can lead to the truncation of a negative binomial distribution
+% such as within host parasite density-dependence, age- dependent variation in
+% host resistance and heterogeneous rate of infection
+% \citep{McCallum2000a,Anderson1982a,Rousset1996}.  Therefore, even detecting
+% ``significant'' PIHM in a dataset does not mean that PIHM is cause of the
+% truncation.
+
+% Given these numerous caveats, is the is there a place in parasitology for methods that
+% estimate PIHM from distributional data?  We are in agreement with
+% \cite{Lester1984} that, at the very least, methods for estimating PIHM can
+% provide preliminary insight into whether or not PIHM is worth further
+% exploration.  However, we stress that these methods should only be used as an
+% exploratory tool when assessing the role of PIHM in a system and potential
+% users should critically evaluate whether they think they have a large enough
+% sample sizes and an appropriate host survival function/post-mortality distribution for the methods developed
+% in this paper to be applicable.  Even if they are applicable, inferring PIHM
+% from distributional data is no substitute for field or laboratory experiments
+% and/or in depth understanding of the natural history of the host-parasite
+% system under consideration.
 

### 2015-05-04 docs/supplementary_material.tex
Replaced discussion with new version. Moved Adjei descrip to supplementary materials

@@
 
-\section*{Appendix 2: Additional Figure}
+\section*{Appendix 2: Implementation of the Adjei Method}
+
+The Adjei Method for estimating PIHM has two steps.  The first
+step is to estimate the parameters of the pre-mortality host-parasite
+distribution using the Crofton Method.  The three parameters estimated are the
+total number of hosts before mortality $N_p$,  the mean number of parasites per
+host before mortality $\mu_p$, and the aggregation of parasites before
+mortality given by the parameter $k_p$ from a negative binomial distribution.
+When $k_p$ is small, parasites are highly aggregated among hosts and when
+$k_p$ is large parasites are more evenly spread out \citep{Wilson2002}.
+The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][]{Royce1990,Lester1984} and we provide a tested implementation of the method in SI 3.
+
+The second step of the Adjei Method is to estimate the shape of the host survival function. \cite{Adjei1986} assume that the host survival function follows the logistic form
+
+\begin{equation}
+    h(x | a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+    \label{eq:logistic}
+\end{equation}
+where $x$ is the parasite intensity in a given host and $a$ and $b$ are the two
+parameters of the logistic function. Generally, a larger $a$ allows for hosts to
+tolerate larger parasite intensities before experiencing parasite-induced mortality
+and a larger $b$ leads to a more rapid decline in the probability of host
+survival as parasite intensity increases.  The value $\exp(a / b)$ is referred
+to as the $LD_{50}$, which gives the parasite intensity at which 50\% of hosts
+experience mortality.
+
+To estimate this function the Adjei Method proceeds as follows.  First, it
+calculates the expected number of hosts with a given parasite load $x$ by using
+the equation $g(x ; \mu_p, k_p) * N_p$, where $g(x ; \cdot)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
+with $x$ parasites are paired as a single data point and the method then assumes that
+this data point follows a binomial distribution with the total number of
+``trials'' equal to the predicted number of hosts and the total number of
+``successes'' equal to the observed number of hosts. In some cases, the
+observed number of hosts is greater than the expected number of hosts and the
+Adjei Method alters the data so that the observed is equal to the predicted
+\citep{Adjei1986}.
+After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation \ref{eq:logistic}.
+This model provides estimates for parameters $a$, $b$ and $LD_{50}$.
+
+While not included in the original implementation of the Adjei Method, a
+$\chi^2$ test with a degrees of freedom of 1 can be used to assess whether a GLM model that includes parasite
+intensity as a predictor of host survival probability is a ``better'' model than a
+GLM without this predictor.  This allows the Adjei Method to determine whether
+PIHM is a significant factor in a host-parasite system.
+
+The Adjei Method's most glaring deficiency is the need to alter the observed
+data in order to fit the model into the binomial GLM framework.  A second more
+subtle problem with the Adjei Method is the potential need to bin data in order
+to predict greater than one host in a given parasite intensity class.  For
+example, if the total number of hosts pre-mortality was 50, the mean number of
+parasites per host pre-mortality was 100 and the aggregation parameter was 1, applying the equation $g(x ; \mu_p=100, k_p=1) * 50$ would result in
+less than 1 individual in all parasite intensities $x$. In other words, the
+Adjei Method cannot be applied to samples with either very high mean parasite
+loads, small sample sizes, or both without some sort of binning of the data.
+While this is not a flaw \emph{per se}, it does add a certain level of
+subjectivity (i.e. which bins should you use?) to a method that already has
+serious potential issues.  In this analysis, we always assume the Adjei Method is not binning the data, though we provide code for applying the binning method in SI 4.
+
+\section*{Appendix 3: Additional Figure}
 
@@
 
-\section*{Appendix 3: Code and unit tests for estimating parasite-induced host mortality}
+\section*{Appendix 4: Code and unit tests for estimating parasite-induced host mortality}
 

### 2015-05-04 docs/parasite_mortatlity_manuscript.tex
Added some citations in the intro

@@
 Infectious agents can have major impacts on animal populations through changing
-population dynamics and stability, altering predator-prey interactions, and
-even causing species’ decline and extinction. Accurately estimating the impact
-of these infectious agents in wildlife is critical to both understanding what
-regulates host and parasite populations and making predictions about disease
-transmission. The impact of pathogens, such as rabies(), bovine TB(), and
+population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
+even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
+of these infectious agents in wildlife is critical to understanding what
+regulates host and parasite populations, making predictions about disease
+transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies(), bovine TB(), and
 rinderpest(),  is typically quantified based on the presence or absence of
@@
 host, for macroparasites, hosts cannot be simply categorized as infected and
-uninfected because pathology is linked to the intensity of infection (Anderson
-and May 1978).  Helminths exhibiting this intensity dependent pathology have
+uninfected because pathology is linked to the intensity of infection \citep{AndersonandMay1979}.  Helminths exhibiting this intensity dependent pathology have
 significant impacts on human health (), domestic livestock economics (),
@@
 difficult to actually quantify parasite-induced host mortality (PIHM) in wild
-animal populations. [more]
+animal populations \citep{Lafferty2002, McCallum2000a}. [more]
 
@@
 wild population; however, for logistical and ethical reasons this method is
-rarely feasible. Data on parasite intensity is much easier to collect and has
-often been used to identify the presence PIHM (cite studies) and quantify the
-relationship between infection intensity and host mortality (cite studies).
+rarely feasible \citep{McCallum2000a}. Data on parasite intensity is much easier to collect and has
+often been used to identify the presence PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and quantify the
+relationship between infection intensity and host mortality \citep{Adjei1986} [others?].
 
-Crofton (1971) first proposed that PIHM could be identified by comparing the
+\cite{Crofton1971a} first proposed that PIHM could be identified by comparing the
 observed parasite distribution in the host population to the distribution
@@
 in quantitative parasitology.
+
+% Infectious agents can have major impacts on animal populations
+% through changing population dynamics and stability \citep{Dobson1992}, altering
+% predator-prey interactions \citep{Joly2004}, and even causing species decline
+% and extinction \citep{DeCastro2005a,McCallum2012b}. Accurate estimates of
+% parasite-induced host mortality (PIHM) in wild animals are important for
+% understanding what regulates both host and parasite populations and to make
+% predictions about disease transmission in natural systems. Although a negative
+% impact on host fitness is a fundamental component of parasitism
+% \citep{Lafferty2002} [more], it is notoriously difficult to quantify PIHM in
+% wild animal populations \citep{Lester1984,McCallum2000a}.
+
+% To conclusively identify PIHM in wild animal populations, it is necessary to
+% experimentally infect and track host populations; a method that is rarely
+% possible in most wild animal systems \citep{McCallum2000a}.  Instead, parasitologists
+% are often only able to determine the parasite intensity on some number of sampled hosts.  This snapshot, distributional data is far from the
+% ideal type of data for addressing questions regarding PIHM, but this is the type of data on which most questions regarding PIHM are asked \citep{Ferguson2011,Royce1990,Lanciani1989,Lester1984,Lester1977}.
+In particular, parasitologists are often interested in asking two questions from this data: 1) Is PIHM occurring in a host-parasite system?
+and if PIHM is occurring 2) Can the effect of parasite intensity on host survival be quantified?
+While both macro and
+% microparasites have detrimental and possibly fatal effects on hosts, for the
+% remainder of this paper we limit our discussion to macroparasites that can be
+% discretely counted within a host.
+
+% The first of these PIHM questions was addressed by \cite{Crofton1971a} who developed
+% a method to test for the presence of PIHM using the truncation of the negative binomial
+% distribution.  In short, the Crofton Method assumes that the distribution of
+% parasites across hosts before mortality occurs follows a negative binomial
+% distribution \citep{AndersonandMay1978,Shaw1998}.  As heavily infected hosts
+% begin to die, the negative binomial distribution is truncated and these hosts are no longer observed in a sample. In other words, the
+% observed host-parasite distribution and the pre-mortality host-parasite
+% distribution will predict substantially different numbers of hosts with high infect intensity (because those have died due to infection), but a similar number of hosts with low infection
+% intensity (because those have survived).  \cite{Crofton1971a} noted that by
+% starting with all observed hosts and iteratively fitting a negative binomial
+% distribution to hosts with lower and lower parasite loads, one could determine
+% whether or not PIHM was occurring and estimate the parameters of the host distribution before parasite-induced mortality.  This was done
+% graphically by determining whether the parameters of the negative binomial distributions fit to different truncations of the data showed a substantial change as the truncation point moved from heavily infected hosts to lightly infected hosts.
+The Crofton Method and this graphical technique are both still currently used to assess whether PIHM is a occurring in a system \citep{Ferguson2011}.
+We give a thorough description and implementation of the Crofton Method in \emph{Supplementary Information} (SI) 1 and discuss the validity of its assumptions in the \emph{Discussion}.
+
+% The second question regarding PIHM moves beyond a simple yes or no answer and attempts to quantify how parasite intensity affects host survival.
+\cite{Adjei1986} proposed a method to answer this question
+% by first using the Crofton Method to estimate the pre-mortality parameters for
+% a host-parasite distribution and then, given these
+% parameters, estimating a host survival function that described how the
+% probability of host-survival changed with increasing parasite load (see
+% \emph{Methods}).  With this host survival function, one can estimate important host-parasite quantities such as the
+% parasite intensity at which 50\% of hosts succumb to PIHM ($LD_{50}$), as well
+% as the percent of hosts in a population succumbing to PIHM \citep{Adjei1986}.
+
+% Despite these methods both being over three decades old, they are still the
+% primary means of answering questions about PIHM given distributional data
+% \citep[but see][for an alternative to the Crofton
+% Method]{Ferguson2011}. However, both methods have a few important limitations.
+% The Crofton Method, and more recent methods \citep{Ferguson2011},
+% rely on a visual test to determine whether or not PIHM is occurring in a
+% system.  With no clear decision rule, it can be difficult to consistently determine the
+% significance of PIHM across different host-parasite systems.
+In theory, the Adjei Method can ameliorate this problem.
+In addition to providing information on the host-survival function, this method can also be used to assess the significance of
+% PIHM in a system.
+% In practice, however, the Adjei Method has never been thoroughly tested and
+% relies on a number of questionable data manipulations.
+
+% In this study, we have
+% three primary goals. First we wish to test reliability of the Adjei Method for
+% answering the aforementioned questions regarding PIHM.  Second, we propose a
+% novel method for answering these questions, compare its efficacy against the Adjei Method, and test both methods' ability to detect PIHM on empirical data.  Third, we
+% discuss the limitations of inferring PIHM from distributional data alone and
+% whether any method for inferring PIHM has a place in the future of
+% parasitology.
 

### 2015-05-04 docs/parasite_mortatlity_manuscript.tex
Made some general edits on the second draft before sending to Sara

--- 2015-05-04 docs/parasite_mortatlity_manuscript.tex
@@
 host, for macroparasites, hosts cannot be simply categorized as infected and
-uninfected because pathology is linked to the intensity of infection (Anderson
-and May 1978).  Helminths exhibiting this intensity dependent pathology have
+uninfected because pathology is linked to the intensity of infection \citep{AndersonandMay1979}.  Helminths exhibiting this intensity dependent pathology have
 significant impacts on human health (), domestic livestock economics (),
-wildlife survival ().  And, while it is generally assumed that some fraction of
+wildlife survival ().  While it is generally assumed that some fraction of
 wild host populations must succumb to parasitic infections, it is notoriously
 difficult to actually quantify parasite-induced host mortality (PIHM) in wild
-animal populations. [more]
+animal populations \citep{Lafferty2002, McCallum2000a}. [more]
 
@@
 The Crofton Method may be able to detect the presence of PIHM however,
-quantifying the relationship between infection intensity and survival
+quantifying the relationship between infection intensity and host survival
 probability is more complicated. \cite{Adjei1986} suggested that these values
@@
 probability of host survival with increasing parasite load. To do this,
-\cite{Adjei1986} modeled host survival as a logistic function and then used a
-binomial generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI X for a technical description of the Adjei Method}).
+\cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI 2} for a technical description of the Adjei Method).
 These methods appeared to provide an estimate the parasite intensity at which
 50\% of hosts exhibit PIHM (), as well as the unmeasurable fraction of the
-population that was lost (SI for adjei method details and extended discussion
-of issues?). However, to implement this method the observed data must be
-modified to fit the binomial GLM framework and subjectively binned when mean
+population that was lost (SI 2). However, to implement this method the observed data must be
+modified to fit the GLM framework and subjectively binned when mean
 infection intensity is high or sample sizes are small.
@@
 be done if methods exist to permit unbiased and accurate estimates of PIHM. In
-this study, we first propose a novel likelihood method for calculating PIHM. We
-next use simulations to compare this likelihood method with previous Adjei
-method to (1) test the ability to detect occurrence of PIHM and (2) estimate
-lethal parasite load ($LD_{50}$) and associated survival function.  We then
+this study, we first propose a novel method for calculating PIHM. We
+next use simulations to compare our method with the previous Adjei
+Method to test the ability of both methods to (1) detect occurrence of PIHM and (2) estimate the
+lethal parasite load ($LD_{50}$) [see Table?] and the associated survival function.  We then
 apply both methods to real datasets previously used in PIHM analyses and
-compare results. And, finally, we discuss the limitations of inferring PIHM
+compare results. Finally, we discuss the limitations of inferring PIHM
 from distributional data and whether any method for inferring PIHM has a place
@@
 
-\subsection*{A novel, likelihood method for estimating PIHM}
+\subsection*{A novel, likelihood-based method for estimating PIHM}
 
-Here we propose an alternative approach using a likelihood method that does not
-require binning or data alteration, potentially reduces the number of
-parameters to be estimated, and allows the significance of PIHM to be estimated
+Here we propose a novel, likelihood-based method (henceforth Likelihood Method) that does not
+require binning or data alteration, reduces the number of
+parameters to be estimated, and allows the significance of PIHM to be determined
 using standard statistical techniques.
@@
 As with all previously proposed methods for estimating PIHM, the
-Likelihood Method first assumes that the pre-mortality distribution follows a
-negative binomial distribution $g(x; \mu_p, k_p)$.  The second
-assumption is that the host survival function
-takes the form of a logistic curve given by
+Likelihood Method first assumes that the pre-mortality distribution of parasites across hosts follows a
+negative binomial distribution $g(x; \mu_p, k_p)$, where $\mu_p$ is the mean parasite intensity in hosts before mortality and $k_p$ is the parasite aggregation before mortality (smaller $k_p$ leads to more aggregation).
+Previous methods have also required a parameter $N_p$ specifying the total number of hosts before mortality \citep{Crofton1971a,Adjei1986}, but this is not a necessary parameter in the Likelihood Method.
+
+The second assumption of the Likelihood Method is that the host survival function, the function specifying the probability of a host surviving with $x$ parasites, takes the form of a logistic curve given by
 \begin{equation}
-    h(x | a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+    h(x ; a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
     \label{eq:logistic}
 \end{equation}
-With these two explicit assumptions, the Likelihood Method estimates the 4 parameters $\mu_p$, $k_p$, $a$, and $b$.
+For this equation, the parasite intensity at which 50\% of host experience PIHM ($LD_{50}$) can be calculated by $\exp{a / b}$.
+With these two explicit assumptions, the Likelihood Method estimates four parameters: $\mu_p$, $k_p$, $a$, and $b$.
 
@@
 
-To estimate that significance of PIHM in a host-parasite system, a
+To estimate the significance of PIHM in a host-parasite system, a
 likelihood ratio test can be used in which the full model is given by equation
@@
 
-\subsection{Evaluating the likelihood and Adjei Methods}
+\subsection*{Evaluating the Adjei and Likelihood Methods}
 
@@
 
-To test the ability of the Adjei Method and the Likelihood Method to identify whether or not PIHM was occurring in a system, we randomly generated data using the following procedure.
+To test the ability of the Adjei and the Likelihood Methods to identify whether or not PIHM was occurring in a system, we randomly generated data using the following procedure.
 First, we drew $N_p$ randomly infected hosts from a
@@
 
-We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in host survival with increasing parasite intensity (Figure \ref{fig:surv}).
+We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:surv}).
 For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.
@@
 tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
-dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.
+dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.\\
 
-\subsection*{Question 2: Can we estimate fatal parasite intensity and the host survival function?}
+\noindent
+\emph{Question 2: Can we estimate fatal parasite intensity and the host survival function?}
 
@@
 applied the Adjei Method and the Likelihood Method to determine whether or not
-PIHM was significant for these species and compared our results to those given by the authors.  All fitting to data was done with the Python code provided in SI 3.
+PIHM was significant for these species and compared our results to those given by the authors.  All fitting to data was done with the code provided in SI 3.
 
@@
 considered (Figure \ref{fig:typeI10}; SI2 Figs 1, 2).  This method also showed the unintuitive pattern of Type I error
-rate decreasing as sample size decreased.  This pattern is due to the issue of
+rate decreasing as sample size decreased.  This pattern was due to the issue of
 binning discussed in the \emph{Introduction} and \emph{SI X}. For small samples sizes, the
@@
 
-The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}),  showing that the Adjei Method has a very high Type I error rate.
+The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results that the Adjei Method has a high Type I error rate [additional data?
+to emphasize this point.
+Raccoon data?].
 
 \section*{Discussion}
-
-% Sara W Outline
-% 5 paragraph disc Outline:
-%  Describe benefits of new method (LM)
-% No data manipulation
-% Estimates fewer parameter
-% Statistical output
-%  A new method to quantitatively explore mortality using intensity data
-
-% Compare  LM with AM
-% LM always does better,
-% Detecting  PIHM: more power, less false detection
-% Ld50/survival curve: LM more precise/less biased. But these do get worse with decreasing sample size.
-% real data , consistent with original author’s results, supports AM type 1 error issues
-% AM is fundamentally flawed, LM is clearly better under ideal simulations, but can it be implemented in the real world?
-
-% 3. Data set limitations (for the LM, because no one should even try the AM):
-% a.
-requires large sample sizes (n>X) which may be impossible to collect from the species we would be most interested in getting this data for, eg, large mammals of conservation concern.
-While 50-100 is very realistic for invertebrates or fish, not so for larger verts
-% b. highly aggregated or high mean intensity increase needed sample size (is this right for LM?) which may make this difficult for many helminths with worm burdens in the 100s-1000s
-% c. low parasite induced mortality may be hard to detect- particularly an issue for worms in definitive hosts
-% d.
-shape of survival curve matters, although often assumed to be linear (eg Anderson may model) can’t know apriori, and method fails  for both perfectly linear and exceptionally steep survival functions
-% e. math works, but not for all parasite-host systems
-
-% 4.even when the math works, should we believe it?
-% a. All methods rely on same fundamental assumptions of the crofton method.
-% b. assume that pre mort psite dist follows a negative binomial. And this is valid because the NB is extremely flexible and lots of empirical and theoretical work says so.
-% c. but, detecting mortality requires that the observed data be significantly different than the NB but because NB is so flexible it can be fit to almost anything, making it impossible to detect pihm
-% d.assume truncation from pihm, but could be caused by lots of other things, so detecting significant pihm does not mean it occurred.
-% f. serious fundamental issues that need to be recognized and acknowledge with any use of this technique
-
-% 5.(conclusion)Does any of this have a place in quantitative parasitology?
 
@@
 methodological flaws in the Adjei Method limit its utility, so here we propose
-an alternative, likelihood-based, method to estimated host survival and the
+an alternative, likelihood-based, method to estimate host survival and the
 $LD_{50}$ from observed parasite intensity data.  At a theoretical level this
-method is a significant improvement because it can require fewer parameters,
+method is a significant improvement over the previous methods because it requires fewer parameters,
 provides a statistical decision rule for identifying PIHM and does not require
@@
 
-Under simulated conditions, the likelihood method always out performs the Adjei Method. For simply detecting the presence of PIHM, the Likelihood
+Using simulated data, we found that the Likelihood Method always out performs the Adjei Method. For simply detecting the presence of PIHM, the Likelihood
 Method is both more powerful and has fewer false detection events (Type I
-errors).  When both methods were applied to published datasets previously used
+errors) [tense?].  When both methods were applied to published datasets previously used
 in PIHM analyses, the Adjei Method tended to detect PIHM where it had not previously been
 reported, consistent with the high Type I error rate observed in our
-simulations. The likelihood method is also universally more precise and less
-biased in calculations of both the parasite $LD_{50}$ and host survival curve.
+simulations. The Likelihood Method is also more precise and less
+biased in calculations of both the parasite $LD_{50}$ and host survival curve over the parameter values we considered.
 However, while only the Likelihood Method produces precise and unbiased $LD_{50}$
@@
 function at realistic sample sizes.  These simulations demonstrate that
-the novel Likelihood Method is more powerful and precise than the previously propose Adjei Method.
+the Likelihood Method is more powerful and precise than the previously propose Adjei Method.
 
-Although superior to the Adjei Method, the LM may still not always be
+Although superior to the Adjei Method, the Likelihood Method may still not always be
 applicable to real data.  The Likelihood Method requires relatively large
@@
 vertebrates, particularly the species of conservation concern where addressing
-the impact of parasitism would be most important. An even larger sample size is
+the impact of parasitism would be most important [citation?]. An even larger sample size is
 required to capture the full parasite distribution when parasites are highly
@@
 fundamental, and potentially problematic assumptions.
-Nearly all methods to date derive from \cite[][but see \cite{Ferguson2011}]{Crofton1971a} and assume that, prior to any PIHM, parasites are distributed in the host population following a
+Nearly all current methods derive from \cite{Crofton1971a} \citep[but see][]{Ferguson2011} and assume that, prior to any PIHM, parasites are distributed in the host population following a
 negative binomial distribution. But, it is fundamentally impossible to know
@@
 the rule.  Finally, even when truncation of the negative binomial distribution is detected, it may be caused by other
-processes such as density dependence, age dependent variation in host
-resistance and heterogeneous infection rates \citep{McCallum2000a,Anderson1982a,Rousset1996}.  This means that in the rare event
+processes such as within host density dependence, age dependent variation in host
+resistance and/or heterogeneous infection rates \citep{McCallum2000a,Anderson1982a,Rousset1996}.  This means that in the event
 that PIHM is detected, it may actually not be the result of PIHM.
@@
 
-% In this study we propose a novel method, the Likelihood Method, for detecting
-% and describing PIHM from parasite intensity data.  Unlike currently proposed
-% methods for estimating PIHM, the Likelihood Method does not require any
-% unnecessary data manipulation, uses fewer parameters [it can, but not always],
-% and provides a quantitative decision rule for determining whether or not PIHM
-% is occurring in a host-parasite system.
-
-% Using simulated datasets, we show that the Likelihood Method correctly detected
-% PIHM when it was occurring and correctly failed to detect PIHM when it was not
-% occurring for sample sizes below 100 hosts. When applied to real data, the
-% Likelihood Method gave results consistent with what was previously reported. In
-% contrast, the Adjei Method, the only currently proposed method to estimate the
-% host survival function and the $LD_{50}$ from parasite intensity data, consistently
-% detected PIHM even when it was not present and behaved erratically for small sample sizes (< 200
-% hosts); a consequence of the need to
-% subjectively bin the data to predict parasite intensity classes with
-% at least one host.  Finally, when applied to real data, the Adjei Method overestimated the occurrence of PIHM relative to what was previously reported.
-
-% In addition to accurately detecting PIHM, the Likelihood Method also provided
-% unbiased and precise estimates of a host survival function's $LD_{50}$ across
-% all sample sizes we considered.  While the Likelihood Method did produce
-% unbiased estimates of the host survival function ($a$ and $b$) for
-% unrealistically large sample sizes, it gave severely biased estimates of the
-% host survival function for realistic samples sizes. The bias was most severe
-% for steep host survival functions due to large changes in the values of $a$ and
-% $b$ only slightly changing an already steep survival function. The Adjei
-% Method, on the other hand, gave biased predictions of the $LD_{50}$ and host
-% survival function across all parameter combinations, with bias increasing with
-% increasing parasite aggregation and increasing mean parasite intensity.  Taken
-% together, these results show that Adjei Method is fundamentally flawed and the
-% Likelihood Method may offer a potential alternative for estimating PIHM.
-
-% While we have improved upon the previously existing methods for answering
-% questions about PIHM, we cannot belie the fact that estimating PIHM from
-% observational data alone is ladened with assumptions and difficulties
-% \citep{McCallum2000a}. The most fundamental assumption of all methods for
-% estimating PIHM is that the shape of the pre-mortality host-parasite
-% distribution is known and follows a negative binomial distribution. While there
-% is substantial empirical and theoretical evidence to justify the use of the
-% negative binomial distribution as the pre-mortality distribution for
-% macroparasites across hosts \citep{Calabrese2011,Anderson1982a,Shaw1998}, it is
-% widely recognized that different processes can lead to a variety of
-% distributions of parasites across hosts
-% \citep{Isham1995,Grenfell1995b,Wilson2002, Duerr2003}.  However, the critical
-% assumption of the pre-mortality distribution is not that the processes leading
-% to the pre-mortality distribution generate a negative binomial distribution,
-% but rather that the pre-mortality distribution is well-fit by a negative
-% binomial distribution. The extreme flexibility of the negative binomial
-% distribution makes it a reasonable candidate distribution for the pre-mortality
-% distributions.  Therefore, we do not see this assumption as central problem in
-% any of the proposed methods.
-
-% However, to use the pre-mortality distribution to infer whether or not the PIHM
-% is occurring in a system requires an explicit assumption about the host
-% survival function and the shape of the post-mortality distribution.  Regarding
-% the host-survival function, all currently proposed methods of PIHM assume that
-% the host-survival function is such that uninfected individuals and individuals
-% with low parasite intensity experience essentially no PIHM. \cite{Lanciani1989}
-% illustrated the importance of this assumption by showing that when hosts
-% experienced a linear decrease in survival probability the Crofton Method could
-% not detect PIHM.   As the most fundamental models of host-parasite dynamics
-% assume a linear decrease in host survival probability with increasing parasite
-% intensity \citep{AndersonandMay1978}, the failure of these methods to detect
-% this relationship is a significant disconnect between empirical and theoretical
-% disease ecology.  However, empirical work has shown that non-linear functions
-% of host survival are not uncommon in host-parasite systems \citep{Benesh2011},
-% so this assumption alone does not preclude the use of PIHM methods on empirical data.
-
-% Regarding the shape of the post-mortality distribution, all of these methods
-% require that the post-mortality distribution be significantly different from a
-% negative binomial distribution.  This is necessary because none of the above
-% methods will be able to detect PIHM if a negative binomial distribution is an
-% adequate fit to the post-mortality distribution.  This is simply because there
-% will be no need for a more complex model (either truncation of the negative
-% binomial or the model given in equation \ref{eq:dist}) if a negative binomial
-% distribution already fits the data. As many observed host-parasite
-% distributions are not significantly different from a negative binomial
-% distribution, there may be limited cases where these PIHM methods can even be
-% considered.
-
-% Finally, all of these methods assume that the truncation of a negative binomial
-% distribution is due to PIHM, but previous studies have shown that a variety of
-% other processes can lead to the truncation of a negative binomial distribution
-% such as within host parasite density-dependence, age- dependent variation in
-% host resistance and heterogeneous rate of infection
-% \citep{McCallum2000a,Anderson1982a,Rousset1996}.  Therefore, even detecting
-% ``significant'' PIHM in a dataset does not mean that PIHM is cause of the
-% truncation.
-
-% Given these numerous caveats, is the is there a place in parasitology for methods that
-% estimate PIHM from distributional data?  We are in agreement with
-% \cite{Lester1984} that, at the very least, methods for estimating PIHM can
-% provide preliminary insight into whether or not PIHM is worth further
-% exploration.  However, we stress that these methods should only be used as an
-% exploratory tool when assessing the role of PIHM in a system and potential
-% users should critically evaluate whether they think they have a large enough
-% sample sizes and an appropriate host survival function/post-mortality distribution for the methods developed
-% in this paper to be applicable.  Even if they are applicable, inferring PIHM
-% from distributional data is no substitute for field or laboratory experiments
-% and/or in depth understanding of the natural history of the host-parasite
-% system under consideration.
-
 \section*{Acknowledgments}
 
-TODO
+We thank X for their useful comments. This work was supported by X.
 
@@
 
+\newpage
+
+\renewcommand{\arraystretch}{1.2}
+
+\begin{table}
+
+    \caption{Definition of parameters and functions used in the main text}
+    \begin{tabular}{c c}
+    \hline
+    Parameter & Definition \\
+    \hline\hline
+    $\mu_p$ & Pre-mortality mean parasite intensity \\
+    $k_p$   & Pre-mortality parasite aggregation \\
+    $N_p$   & Pre-mortality host population size \\
+    $x$     & Number of parasites in a given host \\
+    $g(x; \mu_p, k_p)$ & Pre-mortality negative binomial distribution parasites distribution \\
+    $a$ & Parameter of the logistic host survival function \\
+    $b$ & Parameter of the logistic host survival function \\
+    $h(x; a, b)$ & The host survival function \\
+    $LD_{50}$ & $\exp(a / b)$, Parasite intensity at which 50\% of hosts die \\
+
+    \end{tabular}
+    \label{tab:params}
+\end{table}
+
 \renewcommand{\arraystretch}{1.2}
@@
     \caption{Comparison of the PIHM predictions of previously used host-parasite datasets to those given by the Adjei Method and the Likelihood Method.
-The first column specifies the identity of the dataset, the second column specifies whether or not the author indicated that PIHM was occurring in the system, the third column indicates whether or not the Likelihood Method with parameters from the Crofton Method detects significant PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
+The first column specifies the identity of the dataset, the second column specifies whether or not the author indicated that PIHM was occurring in the system, the third column indicates whether or not the Likelihood Method with pre-mortality parameters estimated from the Crofton Method detects significant PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
 If a method detected significant PIHM the predicted $LD_{50}$ is given in parentheses}

### 2015-05-06 docs/parasite_mortatlity_manuscript.tex
Made new figures for manuscript

@@
+% Document
 % Document
@@
 \usepackage{lipsum}
+\usepackage{tikz}
 
-% Fig. and table formating
+% Figure and table formating
 \usepackage{epsfig}
@@
 \usepackage[framemethod=TikZ]{mdframed}
+
+% Set up Frame
+\mdfdefinestyle{MyFrame}{%
+    linecolor=black,
+    skipabove=10pt,
+    outerlinewidth=2pt,
+    roundcorner=20pt,
+    innertopmargin=10pt,
+    innerbottommargin=10pt,%\baselineskip,
+    innerrightmargin=20pt,
+    innerleftmargin=20pt,
+    backgroundcolor=gray!50!white}
+
 \usepackage{float}
-\usepackage[left=1.25 in, right=1.25 in, top=1.25 in, bottom=1.25 in]{geometry}
-
-\usepackage{caption}
+\usepackage[left=1 in, right=1 in, top=1.25 in, bottom=1.25 in]{geometry}
 
@@
 \usepackage{amsmath} % Add amssymb if not using Mathtime
-\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
 
@@
 \def\linenumberfont{\normalfont\footnotesize\ttfamily}
-\setlength\linenumbersep{0.2 in}
+\setlength\linenumbersep{0.1 in}
 
@@
 \usepackage[singlelinecheck=false,font=small,labelfont=bf]{caption}
-\usepackage[justification=centering]{caption}
 
@@
 \makeatletter
-%patch \@lbibitem to print the current number before the authors
+% patch \@lbibitem to print the current number before the authors
 \patchcmd{\@lbibitem}
@@
 difficult to actually quantify parasite-induced host mortality (PIHM) in wild
-animal populations \citep{Lafferty2002, McCallum2000a}. [more]
+animal populations \citep{Lafferty2002, McCallum2000a}.
 
@@
 These methods appeared to provide an estimate the parasite intensity at which
-50\% of hosts exhibit PIHM (), as well as the unmeasurable fraction of the
+50\% of hosts exhibit PIHM ($LD_{50}$), as well as the unmeasurable fraction of the
 population that was lost (SI 2). However, to implement this method the observed data must be
@@
 Method to test the ability of both methods to (1) detect occurrence of PIHM and (2) estimate the
-lethal parasite load ($LD_{50}$) [see Table?] and the associated survival function.  We then
+lethal parasite load ($LD_{50}$) and the associated survival function.  We then
 apply both methods to real datasets previously used in PIHM analyses and
@@
 for the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).
+
+% Change statistical alpha to something else so I don't confuse it with pathogenecity alpha.
+
+% Set sample size to be small. (50 or 100) and then vary k and mu, Make some sort of heat map to show this?
+
+% What figures do we want to show?  Represent figures in a single frame? Pick one that shows the general pattern? Maybe drop the a figure to the Supplement?
+
+% Crop Figure 2 and merge Fig 3-4 to have 4 panels.
+
+% Fiddle with Figure 1 to make it a plate in The new figure 2? Inset it in some way.
+
+% Figure 3: Just focus on small (100) size and vary mu and k and see what happens. (Chose steep survival function for LD50)
 
@@
 \begin{figure}
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/conceptual_fig.pdf}
 
-    \caption{Five potential shapes for a host-survival functions.
-PIHM should be easier to detect for steeper host survival functions \citep{Lanciani1989}, but we may expect the bias in the parameter estimates to increase as it becomes increasingly difficult to distinguish between steep survival functions.}
-    \label{fig:surv}
+\begin{tikzpicture}
+
+    \node at (0, 0) (pic) {\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure1_partII_for_manuscript50}};
+
+    \node[above=0.1cm] at (pic.north) (concept) {\includegraphics[width=0.6\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure1_partI_for_manuscript}};
+
+    \draw[->] (0, 4.2)--(-1.5, 3.1);
+    \draw[->] (0, 4.2)--(1.5, 3.1);
+
+    \node at (-3.2, 4.9) {A.};
+    \node at (-6.5, 2.7) {B.};
+    \node at (1.4, 2.7) {C.};
+
+\end{tikzpicture}
+
+\caption{A) Five potential shapes for a host-survival functions.
+In the simulations we used a gradual survival function (dotted line), and moderate survival function (dashed line), and a steep survival function (solid line).
+The linear and immediate survival functions represent two potential extremes.
+For each of these survival functions and the parameter combinations described in the main text, we tested the Type I Error and Power of the Likelihood Method and Adjei Method.
+B) gives the Type I Error of each method over a range of pre-mortality sample sizes with $\mu_p = 50$ and $k_p = 0.5$.
+The red line shows the pre-set significance level of 0.05.
+C) gives the Power of each method for detecting PIHM over a range of post-mortality sample sizes for $\mu_p = 50$ and $k_p = 0.5$.
+In general, the Likelihood Method has higher power and lower Type I Error than the Adjei Method.
+See the SI X Fig X - X for Type I Error and Power results for all parameter combinations.}
+
 \end{figure}
@@
 \begin{figure}
+    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure2_for_manuscript.pdf}
 
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu10}
+    \caption{Bias and precision (coefficient of variation) for the Likelihood Method and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
+The pre-mortality parameters for this simulation were $\mu_p = 50$ and $k_p = 0.5$.
+The figure shows the simulations for three different host survival functions (gradual, moderate, and steep) with the same $LD_{50}$.
+Bias and precision results for all other parameter combinations can be found in Fig X - X in SI X.}
 
-    \caption{The Type I error rate and the power of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$.
+    \label{fig:question2}
+\end{figure}
+
+% \begin{figure}
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/conceptual_fig.pdf}
+
+%     \caption{Five potential shapes for a host-survival functions.
+PIHM should be easier to detect for steeper host survival functions \citep{Lanciani1989}, but we may expect the bias in the parameter estimates to increase as it becomes increasingly difficult to distinguish between steep survival functions.}
+%     \label{fig:surv}
+% \end{figure}
+
+% \begin{figure}
+
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu10}
+
+%     \caption{The Type I error rate and the power of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$.
 The first column gives the Type I error rate of each method for falsely detecting PIHM when none is present.
@@
 }
-    \label{fig:typeI10}
+%     \label{fig:typeI10}
 
-\end{figure}
+% \end{figure}
 
-\begin{figure}
+% \begin{figure}
 
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu10}
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu10}
 
-    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+%     \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
 The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
@@
 
-    \label{fig:biasld50}
+%     \label{fig:biasld50}
 
-\end{figure}
+% \end{figure}
 
-\begin{figure}
+% \begin{figure}
 
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu10}
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu10}
 
-    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+%     \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
 The first column gives the bias of each method's $a$ estimate over 150 simulations.
@@
 
-    \label{fig:biasa}
+%     \label{fig:biasa}
 
-\end{figure}
+% \end{figure}
 

### 2015-05-06 code/pihm_methods.py

@@ -13,9 +13,6 @@ from scipy.stats import chisqprob
 Module contains the updated functions and classes for assessing
 parasite-induced host mortality (PIHM)
 
-Functions
----------
-
 """
 
 class PIHM:

### 2015-05-07 code/analysis_for_manuscript_full_likelihood.py
Added final full likelihood analysis

@@ -1,0 +1,71 @@
+import pihm_methods as pm
+reload(pm)
+import numpy as np
+import pandas as pd
+
+"""
+Description
+-----------
+
+
+
+Author : Mark Wilber
+
+
+"""
+
+
+# Parameters
+ks = [1]
+
+mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], [])}
+
+Nps = np.round(np.logspace(1.8, 3.5, 10, base=10)).astype(np.int)#[50, 100, 200, 300, 400, 500] #[300, 500, 1000, 2000, 5000, 7500, 10000]
+N_samp = 150
+
+scenario_1_results = {}
+
+pihm = pm.PIHM()
+
+for mup in mu_ld50.iterkeys():
+
+  scenario_1_results[mup] = {}
+
+  for ld50 in mu_ld50[mup][0]:
+
+    scenario_1_results[mup][ld50] = {}
+
+    for kp in ks:
+
+      scenario_1_results[mup][ld50][kp] = {}
+
+      for Np in Nps:
+
+        # Print to see progress
+        print((mup, ld50, kp, Np))
+
+        results_like = []
+        samp_sizes = []
+
+        # Set the parameters
+
+        for i in xrange(N_samp):
+
+          pihm.set_all_params(Np, mup, kp, ld50[0], ld50[1])
+
+          data_post, data_pre = pihm.get_pihm_samples()
+
+          samp_sizes.append(len(data_post))
+
+          # Test on post_mort
+          pihm.data = data_post
+
+          # Likelihood method. Don't fix the premortatliy
+          like_res = pihm.test_for_pihm_w_likelihood(fixed_pre=False, disp=False)
+
+          # Pull out p-val and a and b
+          results_like.append((like_res[1], like_res[-1][-2:]))
+
+        scenario_1_results[mup][ld50][kp][Np] = (results_like, samp_sizes)
+
+pd.to_pickle(scenario_1_results, "../results/full_likelihood_analysis_results.pkl")

### 2015-05-07 docs/parasite_mortatlity_manuscript.tex
Edited manuscript supplementary material and made a few changes to figures

@@
 Using both the pre-mortality and post-mortality simulated datasets,  we assumed
-that the values of $N_p$, $\mu_p$, and $k_p$ were known and tested the ability of both methods to correctly determine whether or not PIHM was occurring.
-While this scenario is unrealistic because the parameters $N_p$,
-$\mu_p$, and $k_p$ are always unknown, we implemented this scenario as a baseline to
-establish the efficacy of the methods independent of the estimates of $N_p$, $\mu_p$ and $k_p$.
-For the Adjei Method, $N_p$, $\mu_p$, and $k_p$ are estimated using the Crofton Method, while $\mu_p$ and $k_p$ in the Likelihood Method can be estimated jointly with $a$ and $b$ or via the Crofton Method.
-If a
-method could not correctly predict whether or not PIHM was occurring under these idealistic conditions, we considered this strong evidence of the unreliability of this method.
+that the values of $N_p$, $\mu_p$, and $k_p$ were known and tested the ability
+of both methods to correctly determine whether or not PIHM was occurring.
+While this scenario is unrealistic because the parameters $N_p$, $\mu_p$, and
+$k_p$ are always unknown, we implemented this scenario as a baseline to
+establish the efficacy of the methods independent of the estimates of $N_p$,
+$\mu_p$ and $k_p$.  If a method could not correctly predict whether or not PIHM
+was occurring under these idealistic conditions, we considered this strong
+evidence of the unreliability of this method.  For the Adjei Method, $N_p$,
+$\mu_p$, and $k_p$ are estimated using the Crofton Method, while $\mu_p$ and
+$k_p$ in the Likelihood Method can be estimated jointly with $a$ and $b$ or via
+the Crofton Method (SI X).
 
@@
 combinations we simulated 150 datasets and tested the probability of each method incorrectly identifying PIHM in the pre-mortality dataset (Type I error) and correctly identifying PIHM in the post-mortality dataset (power).
-For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level $\alpha = 0.05$.
+For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.
 We tested each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500].
@@
 Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.
+
+\subsection*{Efficacy of the Likelihood Method with unknown pre-mortality parameters}
+
+In the above simulations, we tested the power of the Likelihood Method in the ideal scenario when the pre-mortality parameters were known (or estimated independently via the Crofton Method).
+In this simulation, test the ability of the Likelihood Method to correctly identify PIHM and estimate $LD_50$ when the pre-mortality parameters are not known.
+The simulations in the previous sections showed that the Likelihood method was highly effective at identifying PIHM and the $LD_{50}$ when $\mu_p = 10$ and $k_p = 1$, and as an illustrative example of the limitations of the Likelihood Method, we only use these pre-mortality values when simulating the dataset.
 
@@
 
-% Change statistical alpha to something else so I don't confuse it with pathogenecity alpha.
-
 % Set sample size to be small. (50 or 100) and then vary k and mu, Make some sort of heat map to show this?
-
-% What figures do we want to show?  Represent figures in a single frame? Pick one that shows the general pattern? Maybe drop the a figure to the Supplement?
-
-% Crop Figure 2 and merge Fig 3-4 to have 4 panels.
-
-% Fiddle with Figure 1 to make it a plate in The new figure 2? Inset it in some way.
-
-% Figure 3: Just focus on small (100) size and vary mu and k and see what happens. (Chose steep survival function for LD50)
 
@@
 
-The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results that the Adjei Method has a high Type I error rate [additional data?
-to emphasize this point.
-Raccoon data?].
+The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results that the Adjei Method has a high Type I error rate.
 
@@
 
-Using simulated data, we found that the Likelihood Method always out performs the Adjei Method. For simply detecting the presence of PIHM, the Likelihood
-Method is both more powerful and has fewer false detection events (Type I
-errors) [tense?].  When both methods were applied to published datasets previously used
+Using simulated data, we found that the Likelihood Method always out performed the Adjei Method. For simply detecting the presence of PIHM, the Likelihood
+Method was both more powerful and had fewer false detection events (Type I
+errors).  When both methods were applied to published datasets previously used
 in PIHM analyses, the Adjei Method tended to detect PIHM where it had not previously been
 reported, consistent with the high Type I error rate observed in our
-simulations. The Likelihood Method is also more precise and less
+simulations. The Likelihood Method was also more precise and less
 biased in calculations of both the parasite $LD_{50}$ and host survival curve over the parameter values we considered.
-However, while only the Likelihood Method produces precise and unbiased $LD_{50}$
-estimates, neither method can provide unbiased estimates of the host survival
+However, while only the Likelihood Method produced precise and unbiased $LD_{50}$
+estimates, neither method could provide unbiased estimates of the host survival
 function at realistic sample sizes.  These simulations demonstrate that
@@
 
-Although superior to the Adjei Method, the Likelihood Method may still not always be
-applicable to real data.  The Likelihood Method requires relatively large
-sample sizes (n>50-100) [Are these relatively large? They seem small to me], that although reasonable
-to obtain for invertebrates or small fish may be completely unfeasible for many
-vertebrates, particularly the species of conservation concern where addressing
-the impact of parasitism would be most important [citation?]. An even larger sample size is
-required to capture the full parasite distribution when parasites are highly
-aggregated, mean infection intensity is high, or parasite prevalence is low [same as aggregation],
-all of which are common in many parasitic helminths.  Low parasite-induced host
-mortality, as might be predicted in many definitive hosts, may also be very
-difficult to detect and require impossibly large sample sizes. And, even
-when sample size is sufficient, these methods can only detect PIHM is the host
-survival curve is non-linear \citep{Lanciani1989}. Most host-parasite models assume a linear
-relationship between survival and infection intensity \citep{AndersonandMay1978,McCallum2000a}, however
-nonlinear survival functions are not uncommon in empirical host-parasite
-systems \citep{Benesh2011}.
-And, while linear functions make PIHM undetectable, at the other extreme, steep, non-linear survival curves produce severely biased estimates of the survival function.
-Even the Likelihood Method is probably limited to
-detecting PIHM and estimating $LD_{50}$ in systems where greater than 50 hosts can be
-collected, parasites are common and only moderately aggregated, and
-substantial host mortality occurs at relatively low parasite intensity.
+Although superior to the Adjei Method, the Likelihood Method may still not
+always be applicable to real data.  Our simulations showed when the when pre-
+mortality parameters were not set a priori, the Likelihood Method needed at
+least X samples to have 80\% power and for steep survival functions and even
+more as the survival function became more gradual. While these sample sizes are
+reasonable for hosts such as invertebrates or small fish, they are completely
+unfeasible for many vertebrates, particularly the species of conservation
+concern where addressing the impact of parasitism would be most important. An
+even larger sample size would be required to capture the full parasite
+distribution when parasites are highly aggregated, mean infection intensity is
+high, or parasite prevalence is low [same as aggregation], all of which are
+common in many parasitic helminths.  Low parasite-induced host mortality, as
+might be predicted in many definitive hosts, may also be very difficult to
+detect and require impossibly large sample sizes. And, even when sample size is
+sufficient, these methods can only detect PIHM when the host survival curve is
+non-linear \citep{Lanciani1989}. Most host-parasite models assume a linear
+relationship between survival and infection intensity
+\citep{AndersonandMay1978,McCallum2000a}, however non-linear survival functions
+are not uncommon in empirical host-parasite systems \citep{Benesh2011}.  And,
+while linear functions make PIHM undetectable, at the other extreme, steep,
+non-linear survival curves produce severely biased estimates of the survival
+function. Give the interaction between all of these different factors, the
+Likelihood Method is probably limited to detecting PIHM and estimating
+$LD_{50}$ in systems where greater 100 hosts can be collected, parasites are
+common and only moderately aggregated, and substantial host mortality occurs at
+relatively low parasite intensity.
 
@@
 In the simulations we used a gradual survival function (dotted line), and moderate survival function (dashed line), and a steep survival function (solid line).
-The linear and immediate survival functions represent two potential extremes.
+The linear and immediate survival functions represent two potential extremes that we do not include in the simulations.
 For each of these survival functions and the parameter combinations described in the main text, we tested the Type I Error and Power of the Likelihood Method and Adjei Method.
-B) gives the Type I Error of each method over a range of pre-mortality sample sizes with $\mu_p = 50$ and $k_p = 0.5$.
+B) Gives the Type I Error of each method over a range of pre-mortality sample sizes with a pre-mortality mean parasite intensity ($\mu_p$) of 50 and pre-mortality parasite aggregation ($k_p$) at 0.5.
 The red line shows the pre-set significance level of 0.05.
-C) gives the Power of each method for detecting PIHM over a range of post-mortality sample sizes for $\mu_p = 50$ and $k_p = 0.5$.
+C) Gives the Power of each method for detecting PIHM over a range of post-mortality sample sizes for $\mu_p = 50$ and $k_p = 0.5$.
 In general, the Likelihood Method has higher power and lower Type I Error than the Adjei Method.
-See the SI X Fig X - X for Type I Error and Power results for all parameter combinations.}
+See the SI 3 Fig 1 - 3 for Type I Error and Power results for all parameter combinations.}
 
@@
 
-    \caption{Bias and precision (coefficient of variation) for the Likelihood Method and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
+    \caption{Bias (\textbf{A}) and precision (\textbf{B}) (coefficient of variation) for the Likelihood Method and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
 The pre-mortality parameters for this simulation were $\mu_p = 50$ and $k_p = 0.5$.
-The figure shows the simulations for three different host survival functions (gradual, moderate, and steep) with the same $LD_{50}$.
-Bias and precision results for all other parameter combinations can be found in Fig X - X in SI X.}
+The figure shows the simulations for three different host survival functions (gradual, moderate, and steep), each with the same $LD_{50}$.
+Bias and precision results of $LD_{50}$ and $a$ for all other parameter combinations can be found in SI 3 Fig 4 - 9.}
 
     \label{fig:question2}
+
 \end{figure}
 
-% \begin{figure}
-%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/conceptual_fig.pdf}
-
-%     \caption{Five potential shapes for a host-survival functions.
-PIHM should be easier to detect for steeper host survival functions \citep{Lanciani1989}, but we may expect the bias in the parameter estimates to increase as it becomes increasingly difficult to distinguish between steep survival functions.}
-%     \label{fig:surv}
-% \end{figure}
-
-% \begin{figure}
-
-%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu10}
-
-%     \caption{The Type I error rate and the power of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$.
-The first column gives the Type I error rate of each method for falsely detecting PIHM when none is present.
-The red line gives the the pre-set Type I error rate of $\alpha = 0.05$.
-The second column gives the power of a given method to detect PIHM when it is actually occurring.
-}
-%     \label{fig:typeI10}
-
-% \end{figure}
-
-% \begin{figure}
-
-%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu10}
-
-%     \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
-The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
-The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}
-
-%     \label{fig:biasld50}
-
-% \end{figure}
-
-% \begin{figure}
-
-%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu10}
-
-%     \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
-The first column gives the bias of each method's $a$ estimate over 150 simulations.
-The second column gives the precision of each method's $a$ estimate over 150 simulations.}
-
-%     \label{fig:biasa}
-
-% \end{figure}
-
 \end{document}

### 2015-05-07 docs/supplementary_material.tex
Edited manuscript supplementary material and made a few changes to figures

@@
 \usepackage[left=1.25 in, right=1.25 in, top=1.25 in, bottom=1.25 in]{geometry}
-\usepackage{csvsimple}
 
@@
 
-We provide an implementation and unit tests of the Crofton Method Supplementary Information 3.  Figure \ref{fig:crof_test} visually shows the results of these unit tests.
+We provide an implementation and unit tests of the Crofton Method Supplementary Information 4.
+Figure \ref{fig:crof_test} visually shows that our implementation of the Crofton Method agrees with results previously published by \cite{Crofton1971a}.
 
@@
 
-The Adjei Method for estimating PIHM has two steps.  The first
+The Adjei Method for estimating PIHM has two steps \citep{Adjei1986}.  The first
 step is to estimate the parameters of the pre-mortality host-parasite
@@
 $k_p$ is large parasites are more evenly spread out \citep{Wilson2002}.
-The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][]{Royce1990,Lester1984} and we provide a tested implementation of the method in SI 3.
+The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][ and above]{Royce1990,Lester1984} and we provide a tested implementation of the method in SI 4.
 
@@
 
-To estimate this function the Adjei Method proceeds as follows.  First, it
+To estimate this function, the Adjei Method first
 calculates the expected number of hosts with a given parasite load $x$ by using
@@
 
+% \section*{Appendix 3: Implementation of the Likelihood Method}
+
+% As described in the main text, the probability mass function for the likelihood method is given by
+
+% \begin{equation}
+%     P(x | \text{survival}) = \dfrac{P(\text{survival} | x) * P(x)}{P(\text{survival})}
+%     \label{eq:concept}
+% \end{equation}
+
+% One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$.
+Therefore equation \ref{eq:concept} can be written as
+
+% \begin{equation}
+%     P(x | \text{survival}) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
+%     \label{eq:dist}
+% \end{equation}
+
+% Using this equation, we seek to find estimates for $\mu_p$, $k_p$, $a$, and $b$
+% that maximize the likelihood of equation \ref{eq:dist} given a host-parasite
+% dataset. The first approach is use a standard maximization algorithm to try to
+% find all four parameters.  However, for small number of hosts and/or a gradual
+% survival function, the gradient of the likelihood will not contain much
+% information (check this) and the maximization will likely fail or give
+% unrealistic results. An alternative approach would be to parameterize $\mu_p$
+% and $k_p$ using the Crofton Method and then use equation \ref{eq:dist} to find the values for $a$ and $b$ that maximize the likelihood of a host-parasite dataset.
+
 \section*{Appendix 3: Additional Figure}
 
-See Figures 1 - 6.
+See Figures 1 - 9.
 
@@
 % These can be in the supplementary in later drafts!
+
+\begin{figure}
+
+    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu10}
+
+    \caption{The type I error rate and the power of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$.
+The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
+The red line gives the the pre-set type I error rate of $\alpha = 0.05$.
+The second column gives the power of a given method to detect PIHM when it is actually occurring.
+}
+    \label{fig:typeI10}
+
+\end{figure}
 
@@
 
+    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu10}
+
+    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
+The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}
+
+    \label{fig:biasld50_10}
+
+\end{figure}
+
+\begin{figure}
+
     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu50}
@@
     \label{fig:biasld50_100}
+
+\end{figure}
+
+\begin{figure}
+
+    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu10}
+
+    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+The first column gives the bias of each method's $a$ estimate over 150 simulations.
+The second column gives the precision of each method's $a$ estimate over 150 simulations.}
+
+    \label{fig:biasa_10}
 
@@
     \includegraphics[width=0.5\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/compare_crofton.pdf}
-    \caption{A comparison of this papers implementation of the Crofton Method with the results given in \cite{Crofton1971a}.
-Three of the 6 stations fit by \cite{Crofton1971a} are shown here and all three show that the two implementations give very similar results.}
+    \caption{A comparison of this papers implementation (solid line, circles) of the Crofton Method with the results given in \cite{Crofton1971a} (dashed line, diamonds).
+Figure A compares the predicted number of hosts in a population pre-mortality ($N_p$).
+Figure B compares the predicted parasite aggregation pre-mortality ($k_p$).
+Figure C compares the $\chi^2$ statistic for each implementation.
+Three of the 6 stations fit by \citeauthor{Crofton1971a} are shown here and all show that our implementation gives very similar results to those given by \citeauthor{Crofton1971a}.}
     \label{fig:crof_test}

### 2015-05-07 code/pihm_methods.py

@@ -307,6 +307,7 @@ class PIHM:
         guess : list
             Guess for a and b of the survival function
 
+
         Returns
         -------
         : tuple
@@ -314,6 +315,12 @@ class PIHM:
 
         Notes
         -----
+        When dealing with small sample sizes and trying to fit the full
+        distribution (i.e., not using the Crofton Method), the convergence of
+        the likelihood method will often fail or multiple likelihood peaks
+        might exist.  One way around this is to use the approach of Ferguson
+        and fix k and then just try to estimate
+
 
         """
 
@@ -466,6 +473,7 @@ class PIHM:
 
             red_nll = comp.nll(self.data, mod.nbinom(self.mup, self.kp))
 
+        # Approximately chi-squared...though this is a large sample size approx
         chi_sq = 2 * (-full_nll - (-red_nll))
         prob = chisqprob(chi_sq, 2)
 
@@ -501,6 +509,14 @@ def pihm_pmf(x, mup, kp, a, b, max_sum=5000):
     kern = lambda x: surv_prob(x, a, b) * \
                                     mod.nbinom.pmf(x, mup, kp)
     return(kern(x) / sum(kern(np.arange(0, max_sum))))
+
+def likefxn_def(mu, k, a, b, x):
+    """ Likelihood fxn for all parameters for pihm pmf
+    Returns negative log-likelihood
+
+    """
+
+    return -np.sum(np.log(pihm_pmf(x, mu, k, a, b)))
 
 
 
@@ -619,6 +635,81 @@ def extract_simulation_results(sim_results, keys, method_name, param,
 
         else:
             raise KeyError("Don't recognize parameter: should be a, b, or ld50")
+
+    return samp_sizes, biases, precisions
+
+def extract_full_likelihood_simulation(sim_results, keys, param,
+                                alpha=0.05):
+    """
+
+    Method to extract parameters from the simulation dicationary from the full
+    likelihood simulation
+
+    Parameters
+    ----------
+    sim_results : dict
+        A dictionary with analysis results
+
+    keys : list
+        First item is a value for mup, second item is a tuple specify the
+        (a, b) pair, third iterm is kp.
+
+    param: str
+        either "ld50" or "p".  If the param is p, then the type I error is
+        extracted. If the parameter is ld50 than the bias and precision for the
+        ld50 estimates are extracted.
+
+    alpha : float
[…]

### 2015-05-11 docs/parasite_mortatlity_manuscript.tex
Added Figure 3 simulations and Saras commentts

@@
 
-\title{Comparing methods for estimating parasite-induced host mortality from distributional data: implementation and limitations}
+\title{Methods for estimating parasite-induced mortality from intensity data and their limitations}
 \author{Mark Wilber, Sara Weinstein, and Cherie Briggs}
@@
 regulates host and parasite populations, making predictions about disease
-transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies(), bovine TB(), and
-rinderpest(),  is typically quantified based on the presence or absence of
+transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies (Coyne et al. 1989) [get citation], bovine TB \citep{Cox2005}, and
+rinderpest \citep{Tille1991}, is typically quantified based on the presence or absence of
 disease, and does not account for the number of infectious agents present.
-Although sufficient for many bacterial and viral agents that reproduce within a
-host, for macroparasites, hosts cannot be simply categorized as infected and
-uninfected because pathology is linked to the intensity of infection \citep{AndersonandMay1979}.  Helminths exhibiting this intensity dependent pathology have
-significant impacts on human health (), domestic livestock economics (),
-wildlife survival ().  While it is generally assumed that some fraction of
+This method is sufficient for many bacterial and viral agents that reproduce within a
+host, however for macroparasites pathology is linked to the intensity of infection and hosts cannot be simply categorized as infected and
+uninfected \citep{AndersonandMay1979,Lafferty2002}.  Helminths exhibiting this intensity dependent pathology have
+significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013},
+wildlife survival \citep{Kirk2003, Logiudice2003}.  While it is generally assumed that some fraction of
 wild host populations must succumb to parasitic infections, it is notoriously
 difficult to actually quantify parasite-induced host mortality (PIHM) in wild
-animal populations \citep{Lafferty2002, McCallum2000a}.
+animal populations \citep{McCallum2000a}.
 
-Ideally, parasite-induced host mortality would be
+Ideally, parasite-induced host mortality is
 quantified by experimentally infecting and tracking individual hosts in the
@@
 rarely feasible \citep{McCallum2000a}. Data on parasite intensity is much easier to collect and has
-often been used to identify the presence PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and quantify the
-relationship between infection intensity and host mortality \citep{Adjei1986} [others?].
+often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and quantify the
+relationship between infection intensity and host mortality \citep{Adjei1986}.
 
@@
 observed parasite distribution in the host population to the distribution
-predicted in the absence of parasite-induced host mortality. This method
-(“Crofton Method”) assumes that, prior to host mortality, parasites are
-distributed in the host population following a negative binomial distribution;
-however, as intensity dependent pathology removes heavily infected hosts from
-the population, the tail of the distribution is truncated. Mortality is assumed
-to not occur in hosts with low intensity infections, thus by iteratively
+predicted in the absence of parasite-induced host mortality.
+We briefly introduce the Crofton Method here and provide a more detailed explanation of its implementation in \emph{Supplementary Material (SI)} 1.
+This method
+assumes that, prior to host mortality, infection intensity in the host population follows a negative binomial distribution and the tail of the distribution is truncated as intensity dependent pathology removes the most heavily infected hosts.
+Assuming mortality occurs only in these heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
 fitting a negative binomial distribution to hosts with lower and lower parasite
-loads, and comparing the tail end of this predicted distribution to the
-observed parasite data, one could determine both whether PIHM was occurring and
-the parasite distribution in the host population prior to parasite induced
-mortality. We give a thorough description and implementation of the Crofton
-Method in \emph{Supplementary Information (SI)} 1 and discuss the validity of
-its assumptions in the \emph{Discussion}.
+loads, and comparing the tail ends of these predicted distributions to the
+observed parasite data.
 
-The Crofton Method may be able to detect the presence of PIHM however,
-quantifying the relationship between infection intensity and host survival
-probability is more complicated. \cite{Adjei1986} suggested that these values
-could be quantified by using the Crofton Method to first estimate the pre-
-mortality parasite distribution and then, using those parameters, calculate the
-probability of host survival with increasing parasite load. To do this,
-\cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI 2} for a technical description of the Adjei Method).
-These methods appeared to provide an estimate the parasite intensity at which
+The Crofton Method may be able to detect the presence of PIHM however, but it does not quantify the relationship between infection intensity and host survival
+probability. \cite{Adjei1986} suggested that this relationship
+could be calculated by first using the Crofton Method to estimate the pre-mortality parasite distribution and then using those distribution to calculate the
+probability of host survival with increasing parasite intensity. To do this,
+\cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI} 2 for a technical description of the Adjei Method).
+These methods appeared to provide an estimate for the parasite intensity at which
 50\% of hosts exhibit PIHM ($LD_{50}$), as well as the unmeasurable fraction of the
-population that was lost (SI 2). However, to implement this method the observed data must be
+population that was lost (\emph{SI} 2). However, to implement this method the observed data must be
 modified to fit the GLM framework and subjectively binned when mean
@@
 only be identified by visually examining plots and, with no clear decision
-rule; it can be difficult to determine the significance of PIHM across
+rule, it can be difficult to determine the significance of PIHM across
 different host-parasite systems. The survival function produced by the Adjei
 Method offers one solution; however, this method requires manipulating the
-original data and has never been thoroughly tested.
+original data and has never been tested.
 
-Estimating parasite impacts on host population using intensity data should only
-be done if methods exist to permit unbiased and accurate estimates of PIHM. In
+Intensity data should only
+be used to estimate parasite impacts on host populations if unbiased and accurate methods exist. In
 this study, we first propose a novel method for calculating PIHM. We
@@
 apply both methods to real datasets previously used in PIHM analyses and
-compare results. Finally, we discuss the limitations of inferring PIHM
-from distributional data and whether any method for inferring PIHM has a place
+compare the results. Finally, we discuss the limitations of inferring PIHM
+from intensity data and whether any method for inferring PIHM has a place
 in quantitative parasitology.
@@
 require binning or data alteration, reduces the number of
-parameters to be estimated, and allows the significance of PIHM to be determined
-using standard statistical techniques.
+parameters to be estimated, and  uses standard statistical techniques to determine PIHM significance. We provide Python code for implementing the Likelihood Method in \emph{SI} 4.
 
 As with all previously proposed methods for estimating PIHM, the
-Likelihood Method first assumes that the pre-mortality distribution of parasites across hosts follows a
-negative binomial distribution $g(x; \mu_p, k_p)$, where $\mu_p$ is the mean parasite intensity in hosts before mortality and $k_p$ is the parasite aggregation before mortality (smaller $k_p$ leads to more aggregation).
-Previous methods have also required a parameter $N_p$ specifying the total number of hosts before mortality \citep{Crofton1971a,Adjei1986}, but this is not a necessary parameter in the Likelihood Method.
+Likelihood Method first assumes that prior to mortality the parasite distribution is described by a negative binomial $g(x; \mu_p, k_p)$, where $\mu_p$ and $k_p$ are the mean parasite intensity and aggregation before mortality, respectively (smaller $k_p$ indicates more aggregation).
+Previous methods required calculating the total number of hosts before mortality ($N_p$) \citep{Crofton1971a,Adjei1986}, however this parameter is not needed in the Likelihood Method.
 
-The second assumption of the Likelihood Method is that the host survival function, the function specifying the probability of a host surviving with $x$ parasites, takes the form of a logistic curve given by
+The Likelihood Method then assumes that the host survival function, which specifies the probability of a host surviving with $x$ parasites, follows the logistic curve given by
+
 \begin{equation}
-    h(x ; a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+    h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
     \label{eq:logistic}
 \end{equation}
-For this equation, the parasite intensity at which 50\% of host experience PIHM ($LD_{50}$) can be calculated by $\exp{a / b}$.
-With these two explicit assumptions, the Likelihood Method estimates four parameters: $\mu_p$, $k_p$, $a$, and $b$.
-
-To estimate these parameters, we first define a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.
+With these two explicit assumptions, the Likelihood Method estimates four parameters: $\mu_p$, $k_p$, $a$, and $b$ by  first defining a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.
 Using standard rules of conditional probability this distribution can be written as
@@
 Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset.
-Alternatively, one could apply the Crofton Method to estimate $\mu_p$ and $k_p$ and then find the maximum likelihood estimates of $a$ and $b$ and the corresponding $LD_{50}$.
+The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.
 
 To estimate the significance of PIHM in a host-parasite system, a
-likelihood ratio test can be used in which the full model is given by equation
+likelihood ratio test is used in which the full model is given by equation
 \ref{eq:dist} and the reduced model is given by a negative binomial
 distribution.  If PIHM is not significant in the system, the resulting likelihood
-ratio statistic should approximately follow a $\chi^2$ distribution with degrees of freedom equal to 2.  We provide the code for implementing this Likelihood Method in SI 3.
+ratio statistic should approximately follow a $\chi^2$ distribution with two degrees of freedom.
 
@@
 
-To test the ability of the Adjei and the Likelihood Methods to identify whether or not PIHM was occurring in a system, we randomly generated data using the following procedure.
-First, we drew $N_p$ randomly infected hosts from a
-negative binomial distribution with parameters $\mu_p$ and $k_p$.
-This represented the dataset observed before mortality.
-Second, we chose values of $a$ and $b$ and calculated the probability of survival
-for all $N_p$ hosts using equation \ref{eq:logistic}.  Third, we drew $N_p$ random numbers from a uniform distribution
-between 0 and 1 and if host survival probability was less than this random
-number, the host experienced parasite-induced mortality.  The surviving
-hosts comprised the dataset that would be obtained in the field, after PIHM.
+We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters.
+First, we created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
+negative binomial distribution with parameters $\mu_p$ and $k_p$. Second, we chose values of $a$ and $b$ and calculated the probability of survival
+for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, for each host, we drew a random number from a uniform distribution
+between 0 and 1 and if the calculated host survival probability was less than this random
+number, the host experienced parasite-induced mortality.
+The parasite distribution in these simulated surviving hosts is equivalent to the observed parasite distribution in a wild host population that has undergone parasite-induced host mortality.
 
-Using both the pre-mortality and post-mortality simulated datasets,  we assumed
-that the values of $N_p$, $\mu_p$, and $k_p$ were known and tested the ability
-of both methods to correctly determine whether or not PIHM was occurring.
-While this scenario is unrealistic because the parameters $N_p$, $\mu_p$, and
-$k_p$ are always unknown, we implemented this scenario as a baseline to
-establish the efficacy of the methods independent of the estimates of $N_p$,
-$\mu_p$ and $k_p$.  If a method could not correctly predict whether or not PIHM
-was occurring under these idealistic conditions, we considered this strong
-evidence of the unreliability of this method.  For the Adjei Method, $N_p$,
+We used these simulated pre-mortality and post-mortality datasets to test the ability
+of both methods to correctly determine whether or not PIHM was occurring when the parameters $N_p$, $\mu_p$ and $k_p$ were known.
+  For the Adjei Method, $N_p$,
 $\mu_p$, and $k_p$ are estimated using the Crofton Method, while $\mu_p$ and
-$k_p$ in the Likelihood Method can be estimated jointly with $a$ and $b$ or via
-the Crofton Method (SI X).
+$k_p$ in the Likelihood Method can be estimated jointly with $a$ and $b$ or via the Crofton Method.
+Although the parameters $N_p$, $\mu_p$, and $k_p$ are always unknown in real systems, a method that fails under these ideal simulation conditions will certainly also fail using less ideal, empirical data.
 
-We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:surv}).
+We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:question1}A).
 For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.
@@
 aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
-combinations we simulated 150 datasets and tested the probability of each method incorrectly identifying PIHM in the pre-mortality dataset (Type I error) and correctly identifying PIHM in the post-mortality dataset (power).
+combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).
 For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.
@@
 combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
-precision \citep{Walther2005} for these estimates over varying pre-mortality host population sizes  $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
+precision \citep{Walther2005} for these estimates over pre-mortality host population sizes of $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
 10000].
@@
 
-In the above simulations, we tested the power of the Likelihood Method in the ideal scenario when the pre-mortality parameters were known (or estimated independently via the Crofton Method).
-In this simulation, test the ability of the Likelihood Method to correctly identify PIHM and estimate $LD_50$ when the pre-mortality parameters are not known.
-The simulations in the previous sections showed that the Likelihood method was highly effective at identifying PIHM and the $LD_{50}$ when $\mu_p = 10$ and $k_p = 1$, and as an illustrative example of the limitations of the Likelihood Method, we only use these pre-mortality values when simulating the dataset.
+In the final simulation, we test the ability of the Likelihood Method to correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality parameters are unknown.
+The previous simulations showed that the Likelihood Method effectively identified PIHM when $\mu_p$ and $k_p$ were known with values of 10 and 1, respectively.
+In the best-case scenario where a host-parasite system has these these parameters, we test the power of the Likelihood Method to identify PIHM for gradual, moderate and steep survival functions when the pre-mortality parameters also needed to be estimated.
+We perform 500 simulations over a range of different samples sizes following the simulation procedure described above.
 
@@
 PIHM in 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
-given in \cite{Adjei1986} (Table \ref{table:pihm}). In the \cite{Crofton1971a} datasets, the host was
-the snail \emph{Gammarus pulex} which acts as the intermediate host for the
-acanthocephalan \emph{Polmorphus minutus}. In the \cite{Adjei1986} datasets,
-the hosts were two species of lizard fish \emph{Saurida tumbil} and
-\emph{Saurida undosquamis} that were infected by the cestode
-\emph{Callitetrarhynchus gracilis}.  Males and females of both fish species
-were considered separately.
+given in \cite{Adjei1986} (Table \ref{table:pihm}). \citeauthor{Crofton1971a} analyzed infection patterns in the snail \emph{Gammarus pulex} infected with the
+acanthocephalan \emph{Polmorphus minutus}. \citeauthor{Adjei1986} analyzed males and females of two species of lizard fish \emph{Saurida tumbil} and
+\emph{S. undosquamis} that were infected by the cestode
+\emph{Callitetrarhynchus gracilis}.
 
-In both studies, the authors reported PIHM in some of the datasets and we test whether the Adjei
+In both earlier studies, the authors reported PIHM in some of the datasets and we test whether the Adjei
 Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
@@
 applied the Adjei Method and the Likelihood Method to determine whether or not
-PIHM was significant for these species and compared our results to those given by the authors.  All fitting to data was done with the code provided in SI 3.
+PIHM was significant for these species and compared our results to those given by the authors.  All fitting to data was done with the code provided in \emph{SI} 4.
 
@@
 
+The power of the Adjei Method to detect PIHM in a
+system was close to unity for larger sample sizes and tended to
+decrease as sample size decreased (Figure \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).  The Likelihood Method had a power close to
+unity for all parameter combinations and sample sizes considered.  With gradual
+survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (Fig. \ref{fig:question1}B, \emph{SI} 3 Figs 1-3).
+
 The Adjei Method showed highly inflated Type I error rates (i.e. falsely detected
 PIHM) for all parameter combinations that we
-considered (Figure \ref{fig:typeI10}; SI2 Figs 1, 2).  This method also showed the unintuitive pattern of Type I error
+considered (Fig. \ref{fig:question1}A; \emph{SI} 3 Figs 1-3).  This method also showed the unintuitive pattern of Type I error
 rate decreasing as sample size decreased.  This pattern was due to the issue of
-binning discussed in the \emph{Introduction} and \emph{SI X}. For small samples sizes, the
+binning discussed in the \emph{Introduction} and \emph{SI} 2. For small samples sizes, the
 applicability of the Adjei Method is compromised without binning the observed
@@
 error rate at or near the pre-set level of 0.05 for all parameter combinations
-and sample sizes considered.
-
-The ability of the Adjei Method to detect PIHM given that it was occurring in a
-system (i.e. the power) was close to 1 for larger sample sizes and tended to
-decrease as sample size decreased (Figure \ref{fig:typeI10}; SI2 Figs 1, 2).  The Likelihood Method had a power close to
-one for all parameter combinations and sample sizes considered.  With gradual
-survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (solid black lines; Figure \ref{fig:typeI10}, SI2 Figs 1, 2).
-
-% TODO: Need to figure out what to do with the mu = 100 that are being dropped
-% because the sample size is too small!
+and sample sizes considered (Fig. \ref{fig:question1}A; \emph{SI} 3 Figs 1-3).
 
@@
 The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$
-for all combinations of parameters examined in this study (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  Even for
-small sample sizes (< 500 hosts), the Likelihood Method's estimate of $LD_{50}$
+for all combinations of parameters examined in this study (Fig. \ref{fig:question2}, \emph{SI} 3 Figs 4-6).  Even for
+the smallest sample sizes we considered, the Likelihood Method's estimate of $LD_{50}$
 was largely unbiased, with small biases occurring for host survival functions
@@
 (increasing coefficient of variation) as sample size decreased for all
-parameter combinations we examined (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).
+parameter combinations we examined (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6).
 
-The Adjei Method
-always produced biased estimates of the $LD_{50}$ across all parameter combinations (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  For $\mu_p = 10$, the $LD_{50}$
-estimates from the Adjei Method were largely unbiased for large samples sizes,
-but showed increasing bias as sample size decreased, particularly for steep
-survival functions.  As $\mu_p$ increased, the Adjei method
+The Adjei Method produced biased estimates of the $LD_{50}$ across nearly all parameter combinations (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6)).  For $\mu_p = 10$, the $LD_{50}$
+estimates from the Adjei Method were largely unbiased for large samples sizes, but as $\mu_p$ increased, the Adjei Method
 produced biased estimates of $LD_{50}$ across all sample sizes, with bias
-increasing as sample size decreased (SI2 Fig 3, 4). The $LD_{50}$ estimates from the Adjei
-Method showed large decreases in precision occurring
-for the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).
+increasing as sample size decreased (Figure \ref{fig:question2}, SI2 Fig 4-6). The $LD_{50}$ estimates from the Adjei
+Method showed large decreases in precision with the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:question2}, SI2 Fig 4-6).
 
-% Set sample size to be small. (50 or 100) and then vary k and mu, Make some sort of heat map to show this?
+In terms of the host survival function, the Likelihood Method gave unbiased estimates of survival function parameters when sample sizes were large, however as sample size decreased these estimates became severely biased (Fig.
+\ref{fig:question2}, \emph{SI} Fig 7 - 9) The Adjei Method produced
+biased estimates of the host survival function across all sample sizes, with the bias
+consistently being larger when the survival function was steeper and $\mu_p$ was larger (Fig \ref{fig:question2}, \emph{SI} 3 Figs 7-9).
 
-In terms of the host survival function, the Likelihood Method gave
-asymptotically unbiased estimates of $a$ as sample size increased for
-all parameter combinations considered (Figure \ref{fig:biasa}, SI2 Fig 5, 6).  However, as sample size decreased, the
-Likelihood Method tended to produce severely biased estimates of $a$.
-This was generally more pronounced for steeper survival functions and more
-aggregated pre-mortality distributions (e.g. Figure \ref{fig:biasa}).  The Adjei Method produced
-biased estimates of $a$ across all sample sizes, with the bias
-consistently being larger when the survival function was steeper. The bias of
-the Adjei Method's estimate of $a$ also increased as $\mu_p$ increased (Figure \ref{fig:biasa}, SI2 Fig 5, 6).
+\subsection*{Detecting PIHM with unknown pre-mortality parameters}
+
+When all pre-mortality parameters were jointly estimated, the Likelihood Method had a power of greater than 0.8 when the survival function was moderate and steep for host sample sizes of 424 and 83 respectively (Figure \ref{fig:real_power}).
+When the host survival function was gradual, the Likelihood Method never had a power greater than 0.8 for any post-mortality samples sizes we considered.
 
@@
 of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$).
-
 The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results that the Adjei Method has a high Type I error rate.
@@
 
-Quantifying the impact of parasitism on wild host populations is critical in
-both disease modeling and wildlife management. Ideally the relationship between
+Quantifying the impact of parasitism on wild host populations is critical for managing wildlife populations and understanding parasite transmission. Ideally the relationship between
 infection intensity and host survival would be measured experimentally, but for
@@
 an alternative, likelihood-based, method to estimate host survival and the
-$LD_{50}$ from observed parasite intensity data.  At a theoretical level this
+$LD_{50}$ from observed parasite intensity data.  This
 method is a significant improvement over the previous methods because it requires fewer parameters,
@@
 
-Although superior to the Adjei Method, the Likelihood Method may still not
-always be applicable to real data.  Our simulations showed when the when pre-
-mortality parameters were not set a priori, the Likelihood Method needed at
-least X samples to have 80\% power and for steep survival functions and even
-more as the survival function became more gradual. While these sample sizes are
-reasonable for hosts such as invertebrates or small fish, they are completely
+Although superior to the Adjei Method, the Likelihood Method is not universally applicable to real data.  Our simulations showed when the when pre-
+mortality parameters were estimated directly, the Likelihood Method needed at
+least 83-424 samples to have 80\% power and for steep to moderate survival functions and even
+more as the survival function became more gradual. While some of these sample sizes are reasonable for hosts such as invertebrates or small fish, even the smallest sample sizes are completely
 unfeasible for many vertebrates, particularly the species of conservation
 concern where addressing the impact of parasitism would be most important. An
-even larger sample size would be required to capture the full parasite
-distribution when parasites are highly aggregated, mean infection intensity is
-high, or parasite prevalence is low [same as aggregation], all of which are
-common in many parasitic helminths.  Low parasite-induced host mortality, as
-might be predicted in many definitive hosts, may also be very difficult to
-detect and require impossibly large sample sizes. And, even when sample size is
-sufficient, these methods can only detect PIHM when the host survival curve is
-non-linear \citep{Lanciani1989}. Most host-parasite models assume a linear
-relationship between survival and infection intensity
-\citep{AndersonandMay1978,McCallum2000a}, however non-linear survival functions
-are not uncommon in empirical host-parasite systems \citep{Benesh2011}.  And,
+even larger sample size would be required to identify PIHM when parasites are highly aggregated, mean infection intensity is
+high, or parasite prevalence is low, all of which are
+common in many parasitic helminths.
+Moreover, our results are in agreement with previous work that has shown that as host-survival functions become progressively more linear, PIHM becomes all but impossible to detect \citep{Lanciani1989}.
+This result, however, does not preclude the use of this method as non-linear survival functions
+are not uncommon in empirical host-parasite systems \citep{Benesh2011}.  Finally,
 while linear functions make PIHM undetectable, at the other extreme, steep,
@@
 function. Give the interaction between all of these different factors, the
-Likelihood Method is probably limited to detecting PIHM and estimating
-$LD_{50}$ in systems where greater 100 hosts can be collected, parasites are
-common and only moderately aggregated, and substantial host mortality occurs at
-relatively low parasite intensity.
+Likelihood Method is probably limited to detecting PIHM in systems where greater than 100 hosts can be collected, parasites are
+common and only moderately aggregated, and substantial host mortality occurs at relatively low parasite intensity.
 
@@
 flexible and there is substantial empirical and theoretical evidence to support
-the assumption that, prior to any PIHM, parasite distributions can be fit by a negative binomial distribution \citep{Shaw1995,Shaw1998,Wilson2002}. Unfortunately, this
+the assumption that, prior to any PIHM, parasite distributions can be fit by a negative binomial distribution \citep{Shaw1995,Shaw1998,Wilson2002}.
+
+Unfortunately, this
 flexibility in the distribution may also reduce our ability to detect PIHM. If
@@
 that systems where these methods are applicable may be more the exception than
-the rule.  Finally, even when truncation of the negative binomial distribution is detected, it may be caused by other
+the rule.  Furthermore, even when truncation of the negative binomial distribution is detected, it may be caused by other
 processes such as within host density dependence, age dependent variation in host
@@
 
-Given these numerous caveats, is the is there a place in parasitology for methods that
-estimate PIHM from distributional data?  We are in agreement with
+Given these numerous caveats, is there a place in parasitology for methods that
+estimate PIHM from intensity data?  We are in agreement with
 \cite{Lester1984} that, at the very least, methods for estimating PIHM can
@@
 users should critically evaluate whether they think they have a large enough
-sample sizes and an appropriate host survival function/post-mortality distribution for the methods developed
+sample size and an appropriate host survival function/post-mortality distribution for the methods developed
 in this paper to be applicable.  Even if they are applicable, inferring PIHM
-from distributional data is no substitute for field or laboratory experiments
-and/or in depth understanding of the natural history of the host-parasite
+from distributional data is no substitute for field experiments
+and an in depth understanding of the natural history of the host-parasite
 system under consideration.
@@
 
-We thank X for their useful comments. This work was supported by X.
+TODO
 
@@
 
+\label{fig:question1}
+
 \end{figure}
@@
 
-    \caption{Bias (\textbf{A}) and precision (\textbf{B}) (coefficient of variation) for the Likelihood Method and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
+    \caption{Bias and precision (coefficient of variation) for the Likelihood Method and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
 The pre-mortality parameters for this simulation were $\mu_p = 50$ and $k_p = 0.5$.
@@
 
+\begin{figure}
+    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure3_for_manuscript.pdf}
+
+    \caption{The power of the Likelihood Method to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
+The curves were generated from 500 simulations for 10 pre-mortality population sizes, $N_p$.}
+
+    \label{fig:real_power}
+
+\end{figure}
+
 \end{document}

### 2015-05-11 code/analysis_for_manuscript_full_likelihood.py

@@ -21,7 +21,7 @@ ks = [1]
 mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], [])}
 
 Nps = np.round(np.logspace(1.8, 3.5, 10, base=10)).astype(np.int)#[50, 100, 200, 300, 400, 500] #[300, 500, 1000, 2000, 5000, 7500, 10000]
-N_samp = 150
+N_samp = 300
 
 scenario_1_results = {}
 
@@ -68,4 +68,9 @@ for mup in mu_ld50.iterkeys():
 
         scenario_1_results[mup][ld50][kp][Np] = (results_like, samp_sizes)
 
-pd.to_pickle(scenario_1_results, "../results/full_likelihood_analysis_results.pkl")
+
+# This saves the likelihood results with different means
+#pd.to_pickle(scenario_1_results, "../results/full_likelihood_analysis_results.pkl")
+
+# This saves the likelihood results with different means
+pd.to_pickle(scenario_1_results, "../results/full_likelihood_analysis_results_just_10.pkl")

### 2015-05-12 docs/parasite_mortatlity_manuscript.tex
Added in Saras comments and sent draft for lab meeting

@@
 
-\title{Methods for estimating parasite-induced mortality from intensity data and their limitations}
+\title{Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations}
 \author{Mark Wilber, Sara Weinstein, and Cherie Briggs}
@@
 
-\section*{Abstract}
+\abstract{
 
-TODO
+Parasites can often have significant impacts on host populations through parasite-induced host mortality.  Detecting and quantifying the magnitude of parasite-
+induced mortality, particularly for macroparasites where pathology is
+linked to to the intensity of infection, can provide important insight into
+parasite transmission and host population management. This is, however,
+notoriously difficult in practice as the only data often available are measures of parasite intensities. In this study we provide a novel method for detecting and
+quantifying parasite-induced mortality in wildlife populations from intensity
+data and, using simulations, we show that our method is more reliable for
+detecting and quantifying parasite-induced host mortality than the methods that are
+currently available.  We also show that this method provides quantitative
+assessments of parasite-induced mortality in empirical data that are consist
+with previously published qualitative assessments. However, we stress that this
+method, along with all methods for estimating parasite-induced mortality in
+host populations, have a number critical assumptions that limit their
+applicability in the real world.  We conclude that methods for estimating and
+detecting parasite-induced mortality from intensity data should only be used as an
+exploratory tool for informing more rigorous studies of parasite-induced host
+mortality.}
 
@@
 observed parasite distribution in the host population to the distribution
-predicted in the absence of parasite-induced host mortality.
+predicted in the absence of parasite-induced mortality.
 We briefly introduce the Crofton Method here and provide a more detailed explanation of its implementation in \emph{Supplementary Material (SI)} 1.
@@
 fitting a negative binomial distribution to hosts with lower and lower parasite
-loads, and comparing the tail ends of these predicted distributions to the
-observed parasite data.
+loads, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data.
 
-The Crofton Method may be able to detect the presence of PIHM however, but it does not quantify the relationship between infection intensity and host survival
+The Crofton Method may be able to detect the presence of PIHM, but it does not quantify the relationship between infection intensity and host survival
 probability. \cite{Adjei1986} suggested that this relationship
-could be calculated by first using the Crofton Method to estimate the pre-mortality parasite distribution and then using those distribution to calculate the
+could be calculated by first using the Crofton Method to estimate the pre-mortality parasite distribution and then using this distribution to calculate the
 probability of host survival with increasing parasite intensity. To do this,
 \cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI} 2 for a technical description of the Adjei Method).
-These methods appeared to provide an estimate for the parasite intensity at which
+This method appeared to provide an estimate for the parasite intensity at which
 50\% of hosts exhibit PIHM ($LD_{50}$), as well as the unmeasurable fraction of the
@@
 rule, it can be difficult to determine the significance of PIHM across
-different host-parasite systems. The survival function produced by the Adjei
+different host-parasite systems. The survival function given by the Adjei
 Method offers one solution; however, this method requires manipulating the
@@
 be used to estimate parasite impacts on host populations if unbiased and accurate methods exist. In
-this study, we first propose a novel method for calculating PIHM. We
+this study, we first propose a novel method for detecting and quantifying PIHM. We
 next use simulations to compare our method with the previous Adjei
-Method to test the ability of both methods to (1) detect occurrence of PIHM and (2) estimate the
+Method to test the ability of both to (1) detect occurrence of PIHM and (2) estimate the
 lethal parasite load ($LD_{50}$) and the associated survival function.  We then
@@
 The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.
+All parameters are defined in Table \ref{tab:params}.
 
@@
 
-We used these simulated pre-mortality and post-mortality datasets to test the ability
-of both methods to correctly determine whether or not PIHM was occurring when the parameters $N_p$, $\mu_p$ and $k_p$ were known.
-  For the Adjei Method, $N_p$,
-$\mu_p$, and $k_p$ are estimated using the Crofton Method, while $\mu_p$ and
-$k_p$ in the Likelihood Method can be estimated jointly with $a$ and $b$ or via the Crofton Method.
-Although the parameters $N_p$, $\mu_p$, and $k_p$ are always unknown in real systems, a method that fails under these ideal simulation conditions will certainly also fail using less ideal, empirical data.
+We used these simulated pre-mortality and post-mortality datasets to test the
+ability of both methods to correctly determine whether or not PIHM was
+occurring when the parameters $N_p$, $\mu_p$ and $k_p$ were known.  Although
+the parameters $N_p$, $\mu_p$, and $k_p$ are always unknown in real systems, a
+method that fails under these ideal simulation conditions will certainly also
+fail using less ideal, empirical data. In practice, for the Adjei Method, $N_p$, $\mu_p$,
+and $k_p$ are estimated using the Crofton Method \citep{Adjei1986}, while $\mu_p$ and $k_p$ in
+the Likelihood Method can be estimated jointly with $a$ and $b$ or via the
+Crofton Method.
 
@@
 
-In the final simulation, we test the ability of the Likelihood Method to correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality parameters are unknown.
-The previous simulations showed that the Likelihood Method effectively identified PIHM when $\mu_p$ and $k_p$ were known with values of 10 and 1, respectively.
-In the best-case scenario where a host-parasite system has these these parameters, we test the power of the Likelihood Method to identify PIHM for gradual, moderate and steep survival functions when the pre-mortality parameters also needed to be estimated.
-We perform 500 simulations over a range of different samples sizes following the simulation procedure described above.
+In the final simulation, we tested the ability of the Likelihood Method to
+correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality
+parameters are unknown. The previous simulations showed that the Likelihood
+Method effectively identified PIHM when $\mu_p$ and $k_p$ were known with
+values of 10 and 1, respectively.  As a best-case scenario, we simulated host-
+parasite systems with these pre-mortality parameters and tested the power of
+the Likelihood Method to identify PIHM for gradual, moderate and steep survival
+functions when the pre-mortality parameters $\mu_p$ and $k_p$ also needed to be estimated.  We
+perform 500 simulations over a range of different samples sizes following the
+simulation procedure described above.
 
@@
 
-In both earlier studies, the authors reported PIHM in some of the datasets and we test whether the Adjei
+In both earlier studies, the authors reported PIHM in some of the datasets and we tested whether the Adjei
 Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
@@
 system was close to unity for larger sample sizes and tended to
-decrease as sample size decreased (Figure \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).  The Likelihood Method had a power close to
+decrease as sample size decreased (Figure \ref{fig:question1}C; \emph{SI} 3 Figs 1-3).  The Likelihood Method had a power close to
 unity for all parameter combinations and sample sizes considered.  With gradual
-survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (Fig. \ref{fig:question1}B, \emph{SI} 3 Figs 1-3).
+survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (Fig. \ref{fig:question1}C, \emph{SI} 3 Figs 1-3).
 
@@
 PIHM) for all parameter combinations that we
-considered (Fig. \ref{fig:question1}A; \emph{SI} 3 Figs 1-3).  This method also showed the unintuitive pattern of Type I error
+considered (Fig. \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).  This method also showed the unintuitive pattern of Type I error
 rate decreasing as sample size decreased.  This pattern was due to the issue of
@@
 error rate at or near the pre-set level of 0.05 for all parameter combinations
-and sample sizes considered (Fig. \ref{fig:question1}A; \emph{SI} 3 Figs 1-3).
+and sample sizes considered (Fig. \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).
 
@@
 
-The Adjei Method produced biased estimates of the $LD_{50}$ across nearly all parameter combinations (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6)).  For $\mu_p = 10$, the $LD_{50}$
+The Adjei Method produced biased estimates of the $LD_{50}$ across nearly all parameter combinations (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6).  For $\mu_p = 10$, the $LD_{50}$
 estimates from the Adjei Method were largely unbiased for large samples sizes, but as $\mu_p$ increased, the Adjei Method
 produced biased estimates of $LD_{50}$ across all sample sizes, with bias
-increasing as sample size decreased (Figure \ref{fig:question2}, SI2 Fig 4-6). The $LD_{50}$ estimates from the Adjei
-Method showed large decreases in precision with the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:question2}, SI2 Fig 4-6).
+increasing as sample size decreased (Figure \ref{fig:question2}, \emph{SI} 3 Fig 4-6). The $LD_{50}$ estimates from the Adjei
+Method showed large decreases in precision with the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:question2}, \emph{SI} 3 Fig 4-6).
 
@@
 
-When all pre-mortality parameters were jointly estimated, the Likelihood Method had a power of greater than 0.8 when the survival function was moderate and steep for host sample sizes of 424 and 83 respectively (Figure \ref{fig:real_power}).
-When the host survival function was gradual, the Likelihood Method never had a power greater than 0.8 for any post-mortality samples sizes we considered.
+When all parameters were jointly estimated, the Likelihood Method showed highly
+context-dependent results when detecting PIHM in the best-case scenario of
+$\mu_p = 10$ and $k_p = 1$. The Likelihood Method's power of detecting PIHM was
+greater than 0.8 when when host sample sizes were 424 and 83  for survival
+functions that were moderate and steep, respectively (Fig
+\ref{fig:real_power}).  When the host survival function was gradual, the
+Likelihood Method never had a power greater than 0.8 for any post-mortality
+samples sizes we considered (Fig \ref{fig:real_power}).
 
@@
 
-Of the 10 datasets we considered, the previous authors visually detected PIHM
+Of the 10 datasets we considered, the previous authors qualitatively detected PIHM
 in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
@@
 \begin{table}
-
+    \centering
     \caption{Definition of parameters and functions used in the main text}
@@
     $x$     & Number of parasites in a given host \\
-    $g(x; \mu_p, k_p)$ & Pre-mortality negative binomial distribution parasites distribution \\
+    $g(x; \mu_p, k_p)$ & Pre-mortality negative binomial parasite distribution \\
     $a$ & Parameter of the logistic host survival function \\
@@
     $h(x; a, b)$ & The host survival function \\
-    $LD_{50}$ & $\exp(a / b)$, Parasite intensity at which 50\% of hosts die \\
+    $LD_{50}$ & $\exp(a / b)$, parasite intensity at which 50\% of hosts die \\
 
@@
     \caption{Comparison of the PIHM predictions of previously used host-parasite datasets to those given by the Adjei Method and the Likelihood Method.
-The first column specifies the identity of the dataset, the second column specifies whether or not the author indicated that PIHM was occurring in the system, the third column indicates whether or not the Likelihood Method with pre-mortality parameters estimated from the Crofton Method detects significant PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
-If a method detected significant PIHM the predicted $LD_{50}$ is given in parentheses}
+The first column specifies the identity of the dataset, the second column specifies whether or not the authors indicated that PIHM was occurring in the system based on a qualitative assessment, the third column indicates whether or not the Likelihood Method with pre-mortality parameters estimated from the Crofton Method detects significant PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
+If a method detected significant PIHM the predicted $LD_{50}$ is given in parentheses.}
 
@@
 The linear and immediate survival functions represent two potential extremes that we do not include in the simulations.
-For each of these survival functions and the parameter combinations described in the main text, we tested the Type I Error and Power of the Likelihood Method and Adjei Method.
-B) Gives the Type I Error of each method over a range of pre-mortality sample sizes with a pre-mortality mean parasite intensity ($\mu_p$) of 50 and pre-mortality parasite aggregation ($k_p$) at 0.5.
+For each of these survival functions and the parameter combinations described in the main text, we tested the Type I error and power of the Likelihood Method and Adjei Method.
+B) Gives the Type I error of each method over a range of pre-mortality sample sizes with a pre-mortality mean parasite intensity ($\mu_p$) of 50 and pre-mortality parasite aggregation ($k_p$) at 0.5.
 The red line shows the pre-set significance level of 0.05.
 C) Gives the Power of each method for detecting PIHM over a range of post-mortality sample sizes for $\mu_p = 50$ and $k_p = 0.5$.
-In general, the Likelihood Method has higher power and lower Type I Error than the Adjei Method.
-See the SI 3 Fig 1 - 3 for Type I Error and Power results for all parameter combinations.}
+In general, the Likelihood Method has higher power and lower Type I error than the Adjei Method.
+See the \emph{SI} 3 Fig 1 - 3 for Type I Error and power results for all parameter combinations.}
 
@@
     \caption{Bias and precision (coefficient of variation) for the Likelihood Method and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
+As the coefficient of variation increases, precision decreases.
 The pre-mortality parameters for this simulation were $\mu_p = 50$ and $k_p = 0.5$.
 The figure shows the simulations for three different host survival functions (gradual, moderate, and steep), each with the same $LD_{50}$.
-Bias and precision results of $LD_{50}$ and $a$ for all other parameter combinations can be found in SI 3 Fig 4 - 9.}
+Bias and precision results of $LD_{50}$ and $a$ for all other parameter combinations can be found in \emph{SI} 3 Fig 4 - 9.}
 

### 2015-05-12 docs/supplementary_material.tex
Added Figure 3 simulations and Saras commentts
Added in Saras comments and sent draft for lab meeting

@@
 with $n$ hosts where each host has some parasite intensity 0 to $p_{max}$. Starting
-with the full dataset, guess a vector of pre-mortality parameters $(N_p, \mu_p, k_p)$ and given
-these parameters calculate the predicted number of hosts with $0, 1, 2, \dots
-p_{max}$.  Compare the expected number of hosts with $0, 1, 2, \dots  p_{max}$ parasites
+with the full dataset, guess a vector of pre-mortality parameters $(N_p, \mu_p, k_p)$ where $N_p$ is the total number of hosts before mortality, $\mu_p$ is the parasite intensity before mortality, and $k_p$ is the parasite aggregation before mortality.
+Given
+these parameters, use a negative binomial distribution to calculate the predicted number of hosts with $0, 1, 2, \dots
+p_{max}$ parasites.  Compare the expected number of hosts with $0, 1, 2, \dots  p_{max}$ parasites
 to the observed number hosts with $0, 1, 2, \dots p_{max}$ parasites and calculate
@@
 Second, choose a truncation value ($t_1$) such that $t_1 <
-p_{max}$. Truncate the data such that $\text{data}_\text{truncated} <= t$ and repeat the above
+p_{max}$. Truncate the data such that $\text{data}_\text{truncated} <= t_1$ and repeat the above
 iterative procedure to calculate another set of parameters $(N_{p2}, \mu_{p2}, k_{p2})$
 that minimizes the $\chi^2$-squared statistic on the truncated data. Choose a new
-truncated value $t_2 < t_1$ and repeat the first two steps. Continue to truncated
+truncated value $t_2 < t_1$ and repeat the first two steps. Continue to truncate
 the dataset until it only contains hosts with 0, 1, and 2 parasites (or 3 bins).
-As the method attempts to estimate three parameters, at least 3 classes are
-needed if all 3 parameters are to be identifiable \citep{Royce1990}.
+Because the method attempts to estimate three parameters, at least 3 classes are
+needed for all 3 parameters to be identifiable \citep{Royce1990}.
+Parasite-induced host mortality is traditionally identified by then plotting your truncation values $t_i$ against your different values of $N_{po}$ and looking for a distinct kink in the resulting plot.
 
-We provide an implementation and unit tests of the Crofton Method Supplementary Information 4.
+We provide an implementation and unit tests of the Crofton Method in  Supplementary Information 4.
 Figure \ref{fig:crof_test} visually shows that our implementation of the Crofton Method agrees with results previously published by \cite{Crofton1971a}.
@@
 calculates the expected number of hosts with a given parasite load $x$ by using
-the equation $g(x ; \mu_p, k_p) * N_p$, where $g(x ; \cdot)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
+the equation $g(x ; \mu_p, k_p) * N_p$, where $g(x ; \mu_p, k_p)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
 with $x$ parasites are paired as a single data point and the method then assumes that
@@
 
-\section*{Appendix 3: Additional Figure}
+\section*{Appendix 3: Additional Figures}
 
@@
 
-% These can be in the supplementary in later drafts!
+\singlespacing
+\bibliographystyle{/Users/mqwilber/Dropbox/Documents/Bibformats/ecology_letters.bst}
+\bibliography{/Users/mqwilber/Dropbox/Documents/Bibfiles/Projects_and_Permits-parasite_host_mortality}
 
@@
 
-\singlespacing
-\bibliographystyle{/Users/mqwilber/Dropbox/Documents/Bibformats/ecology_letters.bst}
-\bibliography{/Users/mqwilber/Dropbox/Documents/Bibfiles/Projects_and_Permits-parasite_host_mortality}
-
 \end{document}

### 2015-05-13 docs/parasite_mortality_manuscript.tex
Added citations to draft and included kevins comments in directory, but not in manuscript yet

--- 2015-05-12 docs/parasite_mortatlity_manuscript.tex
@@
 
-\abstract{
+\section*{Abstract}
 
@@
 exploratory tool for informing more rigorous studies of parasite-induced host
-mortality.}
+mortality.
 
@@
 regulates host and parasite populations, making predictions about disease
-transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies (Coyne et al. 1989) [get citation], bovine TB \citep{Cox2005}, and
+transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies \citep{Coyne1989}, bovine TB \citep{Cox2005}, and
 rinderpest \citep{Tille1991}, is typically quantified based on the presence or absence of

### 2015-05-15 code/pihm_methods.py
Trying out piecewise survival funciton

@@ -9,6 +9,7 @@ import pymc as pm
 import matplotlib.pyplot as plt
 from scipy.stats import chisqprob
 
+
 """
 Module contains the updated functions and classes for assessing
 parasite-induced host mortality (PIHM)
@@ -329,9 +330,11 @@ class PIHM:
             if None in list(self.get_premort_params())[1:]:
                 raise TypeError("mup, kp must be preset")
 
-            params = opt.fmin(likefxn2, guess, args=(self.data, self.mup,
-                        self.kp), disp=disp)
-            out_params = [self.mup, self.kp] + list(params)
+            cons = lambda x: x[1] - x[0]
+            params = opt.fmin_cobyla(likefxn2, guess, cons, args=(self.data,
+                self.mup, self.kp), consargs=(self.data, self.mup, self.kp),
+            maxfun=3)
+            out_params = [self.mup, self.kp] + list(params['x'])
 
         else:
 
@@ -505,10 +508,10 @@ def pihm_pmf(x, mup, kp, a, b, max_sum=5000):
 
     """
 
-
     kern = lambda x: surv_prob(x, a, b) * \
                                     mod.nbinom.pmf(x, mup, kp)
     return(kern(x) / sum(kern(np.arange(0, max_sum))))
+
 
 def likefxn_def(mu, k, a, b, x):
     """ Likelihood fxn for all parameters for pihm pmf
@@ -517,7 +520,6 @@ def likefxn_def(mu, k, a, b, x):
     """
 
     return -np.sum(np.log(pihm_pmf(x, mu, k, a, b)))
-
 
 
 def likefxn1(params, x):
@@ -685,7 +687,7 @@ def extract_full_likelihood_simulation(sim_results, keys, param,
         res = sims[Np]
 
         p_vals, ab_vals = zip(*res[0])
-        a_vals, b_vals  = zip(*ab_vals)
+        a_vals, b_vals = zip(*ab_vals)
 
         N_vals = res[1]
 
@@ -706,7 +708,6 @@ def extract_full_likelihood_simulation(sim_results, keys, param,
             # Power
             precisions.append(np.sum(np.array(powerp) < alpha) / len(powerp))
             biases.append(np.nan)
-
 
         else:
             raise KeyError("Don't recognize parameter: should be p or ld50")
@@ -731,16 +732,42 @@ def split_data(bin_edges):
 
 def surv_prob(x, a, b):
     """
-    Calculate survival probability given a parasite load x
[…]

### 2015-05-24 docs/parasite_mortality_manuscript.tex
Adding in Cherie's comments

@@
 Parasites can often have significant impacts on host populations through parasite-induced host mortality.  Detecting and quantifying the magnitude of parasite-
-induced mortality, particularly for macroparasites where pathology is
+induced mortality, particularly for macroparasites in which pathology is
 linked to to the intensity of infection, can provide important insight into
-parasite transmission and host population management. This is, however,
+parasite dynamics and host population management. This is, however,
 notoriously difficult in practice as the only data often available are measures of parasite intensities. In this study we provide a novel method for detecting and
@@
 currently available.  We also show that this method provides quantitative
-assessments of parasite-induced mortality in empirical data that are consist
-with previously published qualitative assessments. However, we stress that this
+estimates of parasite-induced mortality in empirical data that are consist
+with previously published qualitative estimates. However, we stress that this
 method, along with all methods for estimating parasite-induced mortality in
-host populations, have a number critical assumptions that limit their
+host populations for intensity data alone, has a number critical assumptions that limit their
 applicability in the real world.  We conclude that methods for estimating and
-detecting parasite-induced mortality from intensity data should only be used as an
+detecting parasite-induced mortality from intensity data should be used only as an
 exploratory tool for informing more rigorous studies of parasite-induced host
@@
 regulates host and parasite populations, making predictions about disease
-transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies \citep{Coyne1989}, bovine TB \citep{Cox2005}, and
+transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of microparasite pathogens, such as rabies \citep{Coyne1989}, bovine TB \citep{Cox2005}, and
 rinderpest \citep{Tille1991}, is typically quantified based on the presence or absence of
@@
 This method is sufficient for many bacterial and viral agents that reproduce within a
-host, however for macroparasites pathology is linked to the intensity of infection and hosts cannot be simply categorized as infected and
+host, however for macroparasites, pathology is linked to the intensity of infection and hosts cannot be simply categorized as infected and
 uninfected \citep{AndersonandMay1979,Lafferty2002}.  Helminths exhibiting this intensity dependent pathology have
-significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013},
+significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013}, and
 wildlife survival \citep{Kirk2003, Logiudice2003}.  While it is generally assumed that some fraction of
@@
 rarely feasible \citep{McCallum2000a}. Data on parasite intensity is much easier to collect and has
-often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and quantify the
+often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and to quantify the
 relationship between infection intensity and host mortality \citep{Adjei1986}.
@@
 assumes that, prior to host mortality, infection intensity in the host population follows a negative binomial distribution and the tail of the distribution is truncated as intensity dependent pathology removes the most heavily infected hosts.
-Assuming mortality occurs only in these heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
+Assuming mortality occurs only in heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
 fitting a negative binomial distribution to hosts with lower and lower parasite
-loads, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data.
+loads, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data. [FIGURE]
 
@@
 \cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI} 2 for a technical description of the Adjei Method).
-This method appeared to provide an estimate for the parasite intensity at which
-50\% of hosts exhibit PIHM ($LD_{50}$), as well as the unmeasurable fraction of the
-population that was lost (\emph{SI} 2). However, to implement this method the observed data must be
+\citeauthor{Adjei1986} suggested that this method could provide an estimate for the parasite intensity at which a host has a 50\% chance of suffering parasite-induced mortalith ($LD_{50}$).
+However, to implement this method the observed data must be
 modified to fit the GLM framework and subjectively binned when mean
-infection intensity is high or sample sizes are small.
+infection intensity is high or sample sizes are small (see SI 2 for details).
 
@@
 Method offers one solution; however, this method requires manipulating the
-original data and has never been tested.
+original data and its accuracy has never been validated.
 
-Intensity data should only
-be used to estimate parasite impacts on host populations if unbiased and accurate methods exist. In
+Intensity data should
+be used to estimate parasite impacts on host populations only if unbiased and accurate methods exist. In
 this study, we first propose a novel method for detecting and quantifying PIHM. We
@@
 require binning or data alteration, reduces the number of
-parameters to be estimated, and  uses standard statistical techniques to determine PIHM significance. We provide Python code for implementing the Likelihood Method in \emph{SI} 4.
+parameters to be estimated, is highly generalizable, and uses standard statistical techniques to determine PIHM significance.
+%We provide Python code for implementing the Likelihood Method in \emph{SI} 4.
+The Likelihood Method first assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \phi)$, which specifies the probability of a host having $x$ parasites when it is observed.
+$\phi$ is a vector of parameters that described the shape and scale of this distribution.
 
-As with all previously proposed methods for estimating PIHM, the
-Likelihood Method first assumes that prior to mortality the parasite distribution is described by a negative binomial $g(x; \mu_p, k_p)$, where $\mu_p$ and $k_p$ are the mean parasite intensity and aggregation before mortality, respectively (smaller $k_p$ indicates more aggregation).
-Previous methods required calculating the total number of hosts before mortality ($N_p$) \citep{Crofton1971a,Adjei1986}, however this parameter is not needed in the Likelihood Method.
+The Likelihood Method then assumes that the probability of a host surviving for
+$t$ units given it has some parasite intensity $x$ is described by the function
+$h(\text{survival} ; x, t, \theta)$ where $\theta$ is a vector of parameters of
+the survival function. For most parasite-intensity datasets, the observer has
+no knowledge of how long a host has been infected with $x$.  One way to account
+for this would be to integrate out time such that we are left with the function
+$h(\text{survival} ; x, \theta)$ specifying the probability of survival given
+$x$ parasites and some additional parameters $\theta$.  While eliminating the
+functions dependence on time is far from ideal, this is an inherent limitation
+of attempting inference on cross-sectional data (jargon).
 
-The Likelihood Method then assumes that the host survival function, which specifies the probability of a host surviving with $x$ parasites, follows the logistic curve given by
-
-\begin{equation}
-    h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
-    \label{eq:logistic}
-\end{equation}
-With these two explicit assumptions, the Likelihood Method estimates four parameters: $\mu_p$, $k_p$, $a$, and $b$ by  first defining a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.
+With these two assumptions, we can define a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.
 Using standard rules of conditional probability this distribution can be written as
@@
 
-One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$.
+One can see that $P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \theta)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \phi)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \theta)  * g(x; \phi)$.
 Therefore equation \ref{eq:concept} can be written as
@@
 \begin{equation}
-    P(x | \text{survival}) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
+    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \theta)  * g(x; \phi)}{\sum_{x=0}^{\infty} h(\text{survival}; x, \theta)  * g(x; \phi)}
     \label{eq:dist}
@@
 
-Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset.
+% \begin{equation}
+%     h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+%     \label{eq:logistic}
+% \end{equation}
+
+Using this probability distribution, one can then find the parameters $\theta$
+and $\phi$ that maximize the likelihood of an observed host-parasite dataset.
+To estimate the significance of PIHM in a host-parasite system, a likelihood
+ratio test can be used in which the full model is given by equation
+\ref{eq:dist} and the reduced model is given by the pre-mortality distribution
+$g(x; \phi)$.  If PIHM is not significant in the system, the resulting
+likelihood ratio statistic should approximately follow a $\chi^2$ distribution
+with two degrees of freedom.
+
+Equation \ref{eq:dist} could be parameterized in many different ways depending
+on the parasite system of interest. In this study, we will follow the precedent
+set by all previous methods for estimating PIHM and assume that the pre-
+mortality parasite distribution $g(x; \phi)$ follows a negative binomial
+distribution with the parameters mean parasite intensity ($\mu_p$) and
+aggregation ($k_p$) before mortality, respectively (smaller $k_p$ indicates
+more aggregation).   The negative binomial distribution is not just a
+phenomenological assumption and can arise as the equilibrium parasite
+distribution under a variety of different biological assumptions
+\citep{Calabrese2011} [MORE]. However, it is also an incredibly flexible
+distribution that fits many host-parasite systems regardless of whether the
+underlying mechanisms lead to an exact negative binomial distribution
+\citep{Shaw1998}.
+
+Choosing a function for $h(\text{survival}; x, \theta)$
+
  The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.
 All parameters are defined in Table \ref{tab:params}.
-
-To estimate the significance of PIHM in a host-parasite system, a
-likelihood ratio test is used in which the full model is given by equation
-\ref{eq:dist} and the reduced model is given by a negative binomial
-distribution.  If PIHM is not significant in the system, the resulting likelihood
-ratio statistic should approximately follow a $\chi^2$ distribution with two degrees of freedom.
 

### 2015-05-26 docs/parasite_mortality_manuscript.tex
Finished added in Cherie's comments on the main text

@@
 
-Here we propose a novel, likelihood-based method (henceforth Likelihood Method) that does not
-require binning or data alteration, reduces the number of
-parameters to be estimated, is highly generalizable, and uses standard statistical techniques to determine PIHM significance.
-%We provide Python code for implementing the Likelihood Method in \emph{SI} 4.
-The Likelihood Method first assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \phi)$, which specifies the probability of a host having $x$ parasites when it is observed.
-$\phi$ is a vector of parameters that described the shape and scale of this distribution.
+Here we propose a novel, likelihood-based method (henceforth Likelihood Method)
+that does not require binning or data alteration, reduces the number of
+parameters to be estimated, is highly generalizable, and uses standard
+statistical techniques to determine PIHM significance.  The Likelihood Method
+begins with the same assumptions as the Adjei Method: namely that infection of
+a host, parasite-induced mortality of a host, an the sampling of a host
+population occur at distinct time intervals during a hosts life. As discussed by
+\citeauthor{Adjei1986}, this is not necessarily unrealistic as studies have
+shown that infection and host mortality are often age and/or body size dependent [citations].
 
-The Likelihood Method then assumes that the probability of a host surviving for
-$t$ units given it has some parasite intensity $x$ is described by the function
-$h(\text{survival} ; x, t, \theta)$ where $\theta$ is a vector of parameters of
-the survival function. For most parasite-intensity datasets, the observer has
-no knowledge of how long a host has been infected with $x$.  One way to account
-for this would be to integrate out time such that we are left with the function
-$h(\text{survival} ; x, \theta)$ specifying the probability of survival given
-$x$ parasites and some additional parameters $\theta$.  While eliminating the
-functions dependence on time is far from ideal, this is an inherent limitation
-of attempting inference on cross-sectional data (jargon).
+The Likelihood Method then assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \boldsymbol{\phi})$, which specifies the probability of a host having $x$ parasites when it is observed.
+$\boldsymbol{\phi}$ is a vector of parameters that describes the shape this distribution.
+
+The method then assumes that the probability of a host surviving with $x$ parasites from infection until sampling is given by $h(\text{survival} ; x, \boldsymbol{\theta})$ where $\boldsymbol{\theta}$ specifies any additional parameters needed to define the host survival function.
 
@@
 
-One can see that $P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \theta)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \phi)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \theta)  * g(x; \phi)$.
+$P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \boldsymbol{\theta})$, $P(x)$ is the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \phi)$.
 Therefore equation \ref{eq:concept} can be written as
@@
 \begin{equation}
-    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \theta)  * g(x; \phi)}{\sum_{x=0}^{\infty} h(\text{survival}; x, \theta)  * g(x; \phi)}
+    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}{\sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}
     \label{eq:dist}
@@
 
-Using this probability distribution, one can then find the parameters $\theta$
-and $\phi$ that maximize the likelihood of an observed host-parasite dataset.
+Using this probability distribution, one can then find the parameters $\boldsymbol{\theta}$
+and $\boldsymbol{\phi}$ that maximize the likelihood of an observed host-parasite dataset.
 To estimate the significance of PIHM in a host-parasite system, a likelihood
@@
 \ref{eq:dist} and the reduced model is given by the pre-mortality distribution
-$g(x; \phi)$.  If PIHM is not significant in the system, the resulting
+$g(x; \boldsymbol{\phi})$.  If PIHM is not significant in the system, the resulting
 likelihood ratio statistic should approximately follow a $\chi^2$ distribution
-with two degrees of freedom.
+with degree of freedom equal to the number of parameters in the full model with parasite-induced mortality minus the number of parameters in the reduced model without parasite-induced mortality [citation].
 
 Equation \ref{eq:dist} could be parameterized in many different ways depending
-on the parasite system of interest. In this study, we will follow the precedent
-set by all previous methods for estimating PIHM and assume that the pre-
-mortality parasite distribution $g(x; \phi)$ follows a negative binomial
+on the parasite system of interest. In this study, we adopt the typical assumption that the pre-
+mortality parasite distribution $g(x; \boldsymbol{\phi})$ follows a negative binomial
 distribution with the parameters mean parasite intensity ($\mu_p$) and
 aggregation ($k_p$) before mortality, respectively (smaller $k_p$ indicates
-more aggregation).   The negative binomial distribution is not just a
-phenomenological assumption and can arise as the equilibrium parasite
-distribution under a variety of different biological assumptions
-\citep{Calabrese2011} [MORE]. However, it is also an incredibly flexible
+more aggregation) \citep{Crofton1971a,AndersonandMay1978,Adjei1986}.   The negative binomial distribution can arise as the equilibrium parasite
+distribution under a variety of different biological and statistical assumptions
+\citep{Kendall1948a, Boswell1970, Calabrese2011}. However, it is also an incredibly flexible
 distribution that fits many host-parasite systems regardless of whether the
@@
 
-Choosing a function for $h(\text{survival}; x, \theta)$
+Choosing an appropriate function for $h(\text{survival}; x, \boldsymbol{\theta})$ will depend on the system under consideration.  Many theoretical models of
+parasite-induced host mortality assume that the parasite-induced death rate of
+hosts is a linear function of parasite intensity
+\citep{AndersonandMay1978,Dobson1992,Barbour2000}.
+It has been previously noted that parasite-induced mortality can be nearly impossible to detect from intensity data when the host survival function is a linear function of parasite intensity as the post-mortality distribution will be of a similar form as the pre-mortality distribution \citep{Lanciani1989}.
+For example, \cite{Barbour2000} showed that both simple host parasite models without parasite-induced host mortality and those with linear parasite-induced host mortality produced Poisson distributions at equilibrium.
+More complex, non-linear survival functions were needed to produce over-dispersed of under-dispersed post-mortality distributions.
+This is not unrealistic as empirical evidence has shown that non-linear host-survival functions are not uncommon
+\citep{Benesh2011} [phrasing].
 
- The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.
-All parameters are defined in Table \ref{tab:params}.
+As one of the goals of this study is to compare this new
+Likelihood Method to the previously proposed Adjei Method, we adopt the host-
+survival function used in their study and assume host-survival is non-linear and follows a logistic function given by
+
+\begin{equation}
+    h(\text{survival}; x, a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+    \label{eq:logistic}
+\end{equation}
+where $b / 4$ determines the maximum rate of decline of
+host survival probability with increasing parasite load, analogous to the pathogenicity parameter $\alpha$ in traditional macroparasite models \citep{AndersonandMay1978}.
+When $b$ is held constant, for every one unit increase in $a$ the parasite intensity at which 99\% of hosts survive increases by $ 1/ b$.
+The equation $\exp(a
+/ b)$ can also be used to calculate the parasite $LD_{50}$, here defined as the
+infection intensity at a host has a 50\% dying.
+This function is commonly used in toxicology and survival analysis and has the useful properties of being bounded between 0 and 1 and being differentiable for all $x$.
+That being said, it is phenomenological and there is little justification to use it rather than it tends to survival data.
+However, given that a goal of these analyses is to
+compare this method's results the those given by the Adjei Method it is natural
+to adopt the same host-survival function to facilitate comparison.  When
+applying the likelihood method to other systems, other more mechanistic host-survival functions can be used in place of equation \ref{eq:logistic}.
 
@@
 We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters.
-First, we created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
-negative binomial distribution with parameters $\mu_p$ and $k_p$. Second, we chose values of $a$ and $b$ and calculated the probability of survival
+Consistent with the assumptions of the model that parasite infection, mortality, and sampling occur at distinct life stages of the host, we first created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
+negative binomial distribution with parameters $\mu_p$ and $k_p$.
+This is equivalent to the period of hosts becoming infected with parasites given in \cite{Adjei1986}.
+In the Adjei Method and Crofton Method, $N_p$ is a necessary parameter that is defined as the number of hosts in the population before parasite-induced mortality.
+A more appropriate way to define this parameter is the number of hosts that would have been sampled had parasite-induced host mortality not occurred.
+This parameter is not necessary using the Likelihood Method because unlike the Adjei Method and Crofton Method which estimate parasite-induced mortality using absolute numbers of hosts, the Likelihood Method estimates parasite-induced mortality using probabilities.
+However, to compare the results of the Likelihood Method with the Adjei Method, we specified a value for $N_p$ for all simulations.
+
+Second, we chose values of $a$ and $b$ for the host survival function and calculated the probability of survival
 for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, for each host, we drew a random number from a uniform distribution
@@
 number, the host experienced parasite-induced mortality.
-The parasite distribution in these simulated surviving hosts is equivalent to the observed parasite distribution in a wild host population that has undergone parasite-induced host mortality.
+This was the period in which hosts died due to infection.
+The parasite distribution in these simulated surviving hosts represented the parasite distribution in a wild host population that has undergone parasite-induced host mortality.
 
@@
 
-We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:question1}A).
+We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had gradual, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:question1}A).
 For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.
@@
 $N_p$ is not technically the sample size on which the methods are being
-tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
+tested for the post-mortality data because PIHM reduces $N_p$ for each simulated
 dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.\\
 
-\noindent
-\emph{Question 2: Can we estimate fatal parasite intensity and the host survival function?}
-
-To compare the ability of the Adjei Method and the Likelihood Method to recover the $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above.
-For each parameter
-combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
-precision \citep{Walther2005} for these estimates over pre-mortality host population sizes of $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
-10000].
-We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.
-Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.
-
-\subsection*{Efficacy of the Likelihood Method with unknown pre-mortality parameters}
-
-In the final simulation, we tested the ability of the Likelihood Method to
+In the next simulation, we tested the ability of only the Likelihood Method to
 correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality
-parameters are unknown. The previous simulations showed that the Likelihood
-Method effectively identified PIHM when $\mu_p$ and $k_p$ were known with
-values of 10 and 1, respectively.  As a best-case scenario, we simulated host-
-parasite systems with these pre-mortality parameters and tested the power of
+parameters are unknown.  As a best-case scenario, we simulated host-
+parasite systems with $\mu_p = 10$ and $k = 1, 0.5, \text{and} 0.1$ [re-run for 0.5 and 0.1] as it is easier to detect PIHM from small samples sizes when mean parasite intensity is low. We then used
 the Likelihood Method to identify PIHM for gradual, moderate and steep survival
@@
 simulation procedure described above.
+
+\noindent
+\emph{Question 2: Can we estimate properties of the host survival function?}
+
+In the previous section we compared the ability of the Adjei Method and the Likelihood Method to correctly identify whether or not PIHM was a occurring in a system (i.e.
+a yes or no answer).
+In this section we compare the ability of the Adjei Method and the Likelihood Method to estimate properties of the survival function such as the parameters $a$, $b$ and $LD_{50}$.
+To do this, we used the same simulation procedure and parameter combinations described above.
+For each parameter
+combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
+precision \citep{Walther2005} for these estimates.
+Because estimating properties of the host survival function requires more information than simply detecting PIHM, we used larger values of $N_p$ for this simulation ($N_p$ = [300, 500, 1000, 2000, 5000, 7500,
+10000]).
+We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.
+Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.
 
@@
 $\mu_p = 10$ and $k_p = 1$. The Likelihood Method's power of detecting PIHM was
-greater than 0.8 when when host sample sizes were 424 and 83  for survival
+greater than 0.8 when host sample sizes exceeded 424 and 83  for survival
 functions that were moderate and steep, respectively (Fig
@@
 
-Of the 10 datasets we considered, the previous authors qualitatively detected PIHM
-in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
+The previous authors qualitatively detected PIHM
+in 7 of the 10 datasets considered(Table \ref{table:pihm}).  The Likelihood Method parameterized
 from the pre-mortality parameters of the Crofton Method detected significant
@@
 
-Quantifying the impact of parasitism on wild host populations is critical for managing wildlife populations and understanding parasite transmission. Ideally the relationship between
+Quantifying the impact of parasitism on wild host populations is critical for managing wildlife populations and understanding parasite-host dynamics. Ideally the relationship between
 infection intensity and host survival would be measured experimentally, but for
@@
 function at realistic sample sizes.  These simulations demonstrate that
-the Likelihood Method is more powerful and precise than the previously propose Adjei Method.
+the Likelihood Method is more powerful and precise than the previously proposed Adjei Method.
 
@@
 mortality parameters were estimated directly, the Likelihood Method needed at
-least 83-424 samples to have 80\% power and for steep to moderate survival functions and even
-more as the survival function became more gradual. While some of these sample sizes are reasonable for hosts such as invertebrates or small fish, even the smallest sample sizes are completely
+least 83-424 samples to have 80\% power for steep to moderate survival functions, an even larger sample size as the survival function became more gradual.
+While some of these sample sizes are reasonable for hosts such as invertebrates or small fish, even the smallest sample sizes are completely
 unfeasible for many vertebrates, particularly the species of conservation
@@
 exploration.  However, we stress that these methods should only be used as an
-exploratory tool when assessing the role of PIHM in a system and potential
+exploratory tool when assessing the role of PIHM in a system, and potential
 users should critically evaluate whether they think they have a large enough
@@
     \caption{Definition of parameters and functions used in the main text}
-    \begin{tabular}{c c}
+    \begin{tabular}{c p{12cm}}
     \hline
@@
     $g(x; \mu_p, k_p)$ & Pre-mortality negative binomial parasite distribution \\
-    $a$ & Parameter of the logistic host survival function \\
-    $b$ & Parameter of the logistic host survival function \\
-    $h(x; a, b)$ & The host survival function \\
-    $LD_{50}$ & $\exp(a / b)$, parasite intensity at which 50\% of hosts die \\
+
+    $h(\text{survival}; x, a, b)$ & The probability of host survival given a parasite load $x$ and logistic parameters $a$ and $b$\\
+
+    $b / 4$ & The maximum rate of decline in host survival probability with increasing parasite load  \\
+    $a$ & When $b$ is held constant a one unit increase in $a$ leads to a $1 / b$ increase in the parasite intensity at which 99\% of hosts survive\\
+    $LD_{50}$ & $\exp(a / b)$, parasite intensity at which a host has a 50\% chance of dying \\
 

### 2015-06-03 docs/parasite_mortality_manuscript.tex
Incorporated all draft3 comments and added new figures.

--- 2015-05-13 docs/parasite_mortality_manuscript.tex
@@
 
-\maketitle
+% \begingroup
+\setlength{\parindent}{0in}
+\renewcommand{\arraystretch}{2}
+% \renewcommand{\tabcolsep}{0}
+%\renewcommand{\tabrowsep}{1}
+\begin{tabular}{p{3cm} p{10cm}}
 
-\section*{Abstract}
+\large\textbf{} & \vspace\baselineskip \\
 
-Parasites can often have significant impacts on host populations through parasite-induced host mortality.  Detecting and quantifying the magnitude of parasite-
-induced mortality, particularly for macroparasites where pathology is
-linked to to the intensity of infection, can provide important insight into
-parasite transmission and host population management. This is, however,
-notoriously difficult in practice as the only data often available are measures of parasite intensities. In this study we provide a novel method for detecting and
-quantifying parasite-induced mortality in wildlife populations from intensity
-data and, using simulations, we show that our method is more reliable for
-detecting and quantifying parasite-induced host mortality than the methods that are
-currently available.  We also show that this method provides quantitative
-assessments of parasite-induced mortality in empirical data that are consist
-with previously published qualitative assessments. However, we stress that this
-method, along with all methods for estimating parasite-induced mortality in
-host populations, have a number critical assumptions that limit their
-applicability in the real world.  We conclude that methods for estimating and
-detecting parasite-induced mortality from intensity data should only be used as an
-exploratory tool for informing more rigorous studies of parasite-induced host
-mortality.
+\textbf{Title:} & Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations \\
+
+\textbf{Running title:} & Methods for quantifying parasite-induced host mortality  \\
+
+\textbf{Authors:} & Mark Q. Wilber, Sara B. Weinstein and Cheryl J. Briggs \\
+
+\textbf{Statement of Authorship:} & MW and SW designed research, performed
+analyses and wrote the manuscript, CB designed research and contributed to revisions. \\
+
+\textbf{Affiliations:} &
+1) Department of Ecology, Evolution and Marine Biology, University of California,
+Santa Barbara, Santa Barbara California \\
+
+\textbf{Email addresses:} &
+Mark Wilber: mark.wilber@lifesci.ucsb.edu \newline
+Sara Weinstein: sara.weinstein@lifesci.ucsb.edu \newline
+Cheryl Briggs: cherie.briggs@lifesci.ucsb.edu \\
+
+\textbf{Keywords:} & aggregation, negative binomial distribution, parasite intensity, Crofton Method, Likelihood Method
+\\
+
+\textbf{Corresponding \newline author:} &
+Mark Wilber \newline
+University of California, Santa Barbara \newline
+Department of Ecology, Evolution, and Marine Biology \newline
+2111 Noble Hall \newline
+Santa Barbara, CA 93106 \newline
+505-321-5381 \newline
+mark.wilber@lifesci.ucsb.edu
+
+\end{tabular}
+
+% \renewcommand{\arraystretch}{1}
+% \begin{tabular}{l l}
+
+% \textbf{Words - Abstract:} & 293 \\
+% \textbf{Words - Main Text:} & 4988 \\
+% \textbf{References:} & 50 \\
+% \textbf{Figures:} & 5 \\
+% \textbf{Tables:} & 1
+
+% \end{tabular}
+
+\newpage
+
+\abstract{
+
+Parasites can significantly impact animal populations by changing host
+behavior, reproduction and survival. Detecting and quantifying these impacts is
+critical for understanding disease dynamics and managing wild animal
+populations. However, for wild hosts infected with macroparasites, it is
+notoriously difficult to quantify the fatal parasite load and number of animals
+that have died due to disease. When ethical or logistical constraints prohibit
+experimental determination of these values, examination of parasite intensity
+and distribution data may offer an alternative solution. In this study we
+introduce a novel method for using intensity data to detect and quantify
+parasite-induced mortality in wildlife populations. Using simulations, we show
+that this method is more reliable than previously proposed methods while
+providing quantitative estimates of parasite-induced mortality from empirical
+data that are consistent with previously published qualitative estimates.
+However, we stress that this method, and all techniques that estimate parasite-
+induced mortality from intensity data alone, have several critical assumptions
+that limit their applicability in the real world. Due to these limitations,
+these methods should only be used as an exploratory tool to inform more
+rigorous studies of parasite-induced host mortality.
+
+\textbf{Keywords}: parasite aggregation, negative binomial distribution
+
+}
 
@@
 
-Infectious agents can have major impacts on animal populations through changing
+Infectious agents can impact on animal populations through changing
 population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
@@
 regulates host and parasite populations, making predictions about disease
-transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies \citep{Coyne1989}, bovine TB \citep{Cox2005}, and
-rinderpest \citep{Tille1991}, is typically quantified based on the presence or absence of
-disease, and does not account for the number of infectious agents present.
-This method is sufficient for many bacterial and viral agents that reproduce within a
-host, however for macroparasites pathology is linked to the intensity of infection and hosts cannot be simply categorized as infected and
-uninfected \citep{AndersonandMay1979,Lafferty2002}.  Helminths exhibiting this intensity dependent pathology have
-significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013},
-wildlife survival \citep{Kirk2003, Logiudice2003}.  While it is generally assumed that some fraction of
-wild host populations must succumb to parasitic infections, it is notoriously
-difficult to actually quantify parasite-induced host mortality (PIHM) in wild
-animal populations \citep{McCallum2000a}.
+transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of microparasite pathogens, such as rabies \citep{Coyne1989}, bovine tuberculosis \citep{Cox2005}, and
+rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, and host survival is not a function of the number of infectious agents present.
+In contrast, for macroparasites, pathology increases with parasite burden and host survival probability must be treated as a function of infection intensity.
+Helminths exhibiting this intensity-dependent pathology have significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013}, and wildlife survival \citep{Kirk2003,Logiudice2003}.
+While it is generally assumed that some fraction of wild host populations succumb to parasitic infection, it is notoriously difficult to actually quantify parasite-induced host mortality (PIHM) in wild animal populations because it is difficult to observe the dead or dying hosts most impacted by parasitism \citep{McCallum2000a}.
 
@@
 wild population; however, for logistical and ethical reasons this method is
-rarely feasible \citep{McCallum2000a}. Data on parasite intensity is much easier to collect and has
-often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and quantify the
+rarely feasible \citep{McCallum2000a}. Snapshot data of parasite intensities across multiple hosts is much easier to collect and has
+often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and to quantify the
 relationship between infection intensity and host mortality \citep{Adjei1986}.
 
-\cite{Crofton1971a} first proposed that PIHM could be identified by comparing the
-observed parasite distribution in the host population to the distribution
-predicted in the absence of parasite-induced mortality.
-We briefly introduce the Crofton Method here and provide a more detailed explanation of its implementation in \emph{Supplementary Material (SI)} 1.
-This method
+\cite{Crofton1971a} first proposed that PIHM could be identified from parasite intensity data by comparing the
+observed parasite distribution in sampled hosts to the distribution
+predicted in the absence of parasite-induced mortality. This method
 assumes that, prior to host mortality, infection intensity in the host population follows a negative binomial distribution and the tail of the distribution is truncated as intensity dependent pathology removes the most heavily infected hosts.
-Assuming mortality occurs only in these heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
-fitting a negative binomial distribution to hosts with lower and lower parasite
-loads, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data.
+Assuming mortality occurs only in heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
+fitting a negative binomial distribution to hosts with lower and lower parasite intensities, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data (Figure \ref{fig:crofton}, see \emph{Supplementary Material (SI)} 1 for additional detail).
 
-The Crofton Method may be able to detect the presence of PIHM, but it does not quantify the relationship between infection intensity and host survival
-probability. \cite{Adjei1986} suggested that this relationship
-could be calculated by first using the Crofton Method to estimate the pre-mortality parasite distribution and then using this distribution to calculate the
-probability of host survival with increasing parasite intensity. To do this,
-\cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI} 2 for a technical description of the Adjei Method).
-This method appeared to provide an estimate for the parasite intensity at which
-50\% of hosts exhibit PIHM ($LD_{50}$), as well as the unmeasurable fraction of the
-population that was lost (\emph{SI} 2). However, to implement this method the observed data must be
-modified to fit the GLM framework and subjectively binned when mean
-infection intensity is high or sample sizes are small.
+While the Crofton Method detects the presence of PIHM, it makes no attempt to
+quantify the relationship between infection intensity and host survival
+probability; an important relationship for determining the potential impact of
+a parasite on host dynamics \citep{AndersonandMay1978s}. \cite{Adjei1986}
+suggested that this relationship could be calculated by first using the Crofton
+Method to estimate the pre-mortality parasite distribution and then using this
+distribution to calculate the probability of host survival with increasing
+parasite intensity. To do this, \cite{Adjei1986} modeled host survival as a
+logistic function and then used a generalized linear model (GLM) to estimate
+the parameters of the host survival function (see \emph{SI} 2 for a technical
+description of the Adjei Method). \citeauthor{Adjei1986} noted that this
+survival function could provide useful information such as the parasite
+intensity at which a host has a greater than 50\% chance of suffering parasite-
+induced mortality. However, if mean infection intensity is high or sample sizes
+are small the observed intensity data must be subjectively binned into
+intensity ranges in order to fit the GLM framework. Furthermore, for the Adjei
+Method to work, any observed intensity values greater than predicted values
+must be modified and set equal to the predict values (see \emph{SI} 2 for
+details); a questionable act of data manipulation. These manipulations may introduce bias, reduce the precision and
+limit the power of this method to detect and quantify parasite-induced host
+mortality.
 
@@
 methods (particularly the Crofton Method) are still discussed among
-parasitologists and are the primary techniques for examining population level
-impacts of parasitism using parasite intensity data. In these methods, PIHM can
-only be identified by visually examining plots and, with no clear decision
-rule, it can be difficult to determine the significance of PIHM across
-different host-parasite systems. The survival function given by the Adjei
-Method offers one solution; however, this method requires manipulating the
-original data and has never been tested.
+parasitologists and are the primary techniques for examining population-level
+impacts of parasitism using parasite intensity data.
+In these methods, PIHM can only be identified by visually examining plots of the pre-mortality parameters predicted by Crofton Method and determining whether they show a ``kink'' over a range of truncation values \citep[Figure \ref{fig:crofton}B;][]{Lester1984,Ferguson2011}.
+This qualitative criteria makes it difficult to compare PIHM between studies and a more rigorous and quantitative method is needed to both detect and quantify host mortality.
+The survival function given by the Adjei
+Method may be used to do this; however, it requires manipulating the
+original data and its accuracy remains untested.
 
-Intensity data should only
-be used to estimate parasite impacts on host populations if unbiased and accurate methods exist. In
-this study, we first propose a novel method for detecting and quantifying PIHM. We
-next use simulations to compare our method with the previous Adjei
-Method to test the ability of both to (1) detect occurrence of PIHM and (2) estimate the
-lethal parasite load ($LD_{50}$) and the associated survival function.  We then
+In this study, we propose a novel method for detecting and quantifying PIHM that ameliorates many of the aforementioned deficiencies of the previous methods.
+Our method does not require data alteration, is highly generalizable, and uses standard statistical techniques to quantitatively determine whether PIHM is occurring in a system.
+We
+use simulations to compare our method with the Adjei
+Method to test the ability of both to (1) detect occurrence of PIHM and (2) estimate the host survival function.  We then
 apply both methods to real datasets previously used in PIHM analyses and
 compare the results. Finally, we discuss the limitations of inferring PIHM
-from intensity data and whether any method for inferring PIHM has a place
-in quantitative parasitology.
+from intensity data and the place of any of these methods in modern quantitative parasitology.
 
@@
 
-Here we propose a novel, likelihood-based method (henceforth Likelihood Method) that does not
-require binning or data alteration, reduces the number of
-parameters to be estimated, and  uses standard statistical techniques to determine PIHM significance. We provide Python code for implementing the Likelihood Method in \emph{SI} 4.
+Our method (henceforth the Likelihood Method) begins with the same assumptions as the Adjei Method: namely that infection of
+a host, parasite-induced mortality of a host, and the sampling of a host
+population occur at distinct time intervals during a host's life. As discussed by
+\citeauthor{Adjei1986}, this is not necessarily an unrealistic as some parasite infections are host age dependent with mortality occurring soon after infection \citep{Schotthoefer2003,Johnson2008}.
+[CHAT WITH SARA ABOUT THIS]
 
-As with all previously proposed methods for estimating PIHM, the
-Likelihood Method first assumes that prior to mortality the parasite distribution is described by a negative binomial $g(x; \mu_p, k_p)$, where $\mu_p$ and $k_p$ are the mean parasite intensity and aggregation before mortality, respectively (smaller $k_p$ indicates more aggregation).
-Previous methods required calculating the total number of hosts before mortality ($N_p$) \citep{Crofton1971a,Adjei1986}, however this parameter is not needed in the Likelihood Method.
+The Likelihood Method then assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \boldsymbol{\phi})$, which specifies the probability of a host having $x$ parasites when it is observed.
+$\boldsymbol{\phi}$ is a vector of parameters that describes the shape of this distribution.
+The method then assumes that the probability of a host surviving with $x$ parasites from infection until sampling is given by the host survival function $h(\text{survival} ; x, \boldsymbol{\theta})$ where $\boldsymbol{\theta}$ specifies any additional parameters needed to define the host survival function.
 
-The Likelihood Method then assumes that the host survival function, which specifies the probability of a host surviving with $x$ parasites, follows the logistic curve given by
-
-\begin{equation}
-    h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
-    \label{eq:logistic}
-\end{equation}
-With these two explicit assumptions, the Likelihood Method estimates four parameters: $\mu_p$, $k_p$, $a$, and $b$ by  first defining a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.
+With these two assumptions, we can define a distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival, $P(x | \text{survival})$.
 Using standard rules of conditional probability this distribution can be written as
@@
 
-One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$.
-Therefore equation \ref{eq:concept} can be written as
+$P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \boldsymbol{\theta})$, $P(x)$ is the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})$.
+Therefore, equation \ref{eq:concept} can be written as
 
 \begin{equation}
-    P(x | \text{survival}) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
+    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}{\sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}
     \label{eq:dist}
@@
 
-Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset.
-The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.
-All parameters are defined in Table \ref{tab:params}.
+Using this probability distribution, one can then find the parameters $\boldsymbol{\theta}$
+and $\boldsymbol{\phi}$ that maximize the likelihood of an observed host-parasite dataset.
+To estimate the significance of PIHM in a host-parasite system, a likelihood
+ratio test can be used in which the full model is given by equation
+\ref{eq:dist} and the reduced model is given by the pre-mortality distribution
+$g(x; \boldsymbol{\phi})$.  If PIHM is not significant in the system, the resulting
+likelihood ratio statistic should approximately follow a $\chi^2$ distribution
+with degrees of freedom equal to the number of parameters in the full model with parasite-induced mortality minus the number of parameters in the reduced model without parasite-induced mortality \citep{Bolker2008}.
 
-To estimate the significance of PIHM in a host-parasite system, a
-likelihood ratio test is used in which the full model is given by equation
-\ref{eq:dist} and the reduced model is given by a negative binomial
-distribution.  If PIHM is not significant in the system, the resulting likelihood
-ratio statistic should approximately follow a $\chi^2$ distribution with two degrees of freedom.
+The parameterization of equation \ref{eq:dist} depends on the parasite system of interest.
+Here, we assume that the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ follows a negative binomial
+distribution with the parameters mean parasite intensity ($\mu_p$) and
+aggregation ($k_p$) before mortality (smaller $k_p$ indicates
+more aggregation) \citep{Crofton1971a,AndersonandMay1978,Adjei1986}.
+A variety of different biological and statistical assumptions can result in an equilibrium parasite distribution that follows a negative binomial distribution
+\citep{Kendall1948a, Boswell1970, Calabrese2011}.
+Furthermore, the negative binomial distribution is an incredibly flexible distribution that fits many host-parasite systems even when the underlying mechanisms determining the empirical distribution are unknown \citep{Shaw1998}.
+
+The function for $h(\text{survival}; x,
+\boldsymbol{\theta})$ is also system specific.  Many
+theoretical models of parasite-induced host mortality assume that the parasite-induced death rate of hosts is a linear function of parasite intensity
+\citep{AndersonandMay1978,Dobson1992,Barbour2000}.
+In systems where there is truly a linear relationship between infection intensity and survival probability it will be nearly impossible to use intensity data to detect parasite-induced host mortality \citep{Lanciani1989}.
+However,  there is substantial empirical evidence that host-parasite systems do not universally exhibit linear host-survival functions \citep{Benesh2011}, suggesting a linear form for $h(\text{survival} ; x, \boldsymbol{\theta}$) is not always the most reasonable.
+
+To compare the Likelihood Method and the previously proposed Adjei Method, we adopt the host-survival function used in the earlier study and assume it is non-linear and follows a logistic function given by
+
+\begin{equation}
+    h(\text{survival}; x, a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+    \label{eq:logistic}
+\end{equation}
+
+Generally, a larger $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases, with the maximum rate of decline having a value of $b / 4$ (\emph{SI} 2).
+$b$ is in many ways analogous to the the pathogenicity parameter in classic macroparasite models which gives that rate at which host death rate changes with increasing parasite intensity \citep{AndersonandMay1978,Isham1995}.
+When $b$ is held constant, a larger $a$ allows for hosts to tolerate larger parasite intensities before experiencing parasite-induced mortality.
+More specifically, for every one unit increase in $a$ the log parasite intensity at which any percent of hosts survive (e.g.
+99\% of hosts survive) increases by $ 1 / b$ (\emph{SI} 2).
+
+The equation $\exp(a
+/ b)$ can also be used to calculate the parasite $LD_{50}$, here defined as the
+infection intensity above which a host has greater than 50\% probability of dying.
+Equation \ref{eq:logistic} is commonly used in toxicology and has the useful properties of being bounded between 0 and 1 and being differentiable for all $x$ \citep{Collet2002}.
+That being said, it is phenomenological and there is little theoretical justification to use it rather than it tends to fit survival data.
+However, given that a goal of these analyses is to
+compare the Likelihood Method's results to the Adjei Method, it is natural
+to adopt the same host-survival function to facilitate comparison.  When
+applying the Likelihood Method to other systems more mechanistic host-survival functions can be used in place of equation \ref{eq:logistic}.
 
@@
 We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters.
-First, we created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
-negative binomial distribution with parameters $\mu_p$ and $k_p$. Second, we chose values of $a$ and $b$ and calculated the probability of survival
-for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, for each host, we drew a random number from a uniform distribution
+Consistent with the model assumption that parasite infection, host mortality, and population sampling are temporally separate events, we first created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
+negative binomial distribution with parameters $\mu_p$ and $k_p$.
+This represents a host population that has become infected but not yet experienced parasite-induced mortality \citep{Adjei1986}.
+In the Adjei Method and Crofton Method, $N_p$ is a necessary parameter defined as the number of hosts in the population before parasite-induced mortality.
+More accurately, $N_p$  is the number of hosts that would have been sampled had parasite-induced host mortality not occurred.
+This parameter is not necessary when using the Likelihood Method because, unlike the Adjei Method and Crofton Method which estimate parasite-induced mortality using absolute numbers of hosts, the Likelihood Method estimates parasite-induced mortality using probabilities.
+However, to compare the results of the Likelihood Method with the Adjei Method, we specified a value for $N_p$ for all simulations.
+
+We next chose values of $a$ and $b$ for the host survival function and calculated the probability of survival
+for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, to simulate the period in which hosts died due to infection, for each host we drew a random number from a uniform distribution
 between 0 and 1 and if the calculated host survival probability was less than this random
-number, the host experienced parasite-induced mortality.
-The parasite distribution in these simulated surviving hosts is equivalent to the observed parasite distribution in a wild host population that has undergone parasite-induced host mortality.
+number, the host experienced parasite-induced mortality.  The surviving individuals represent the post-mortality hosts that would be sampled in the field.
 
-We used these simulated pre-mortality and post-mortality datasets to test the
+We then used these simulated pre-mortality and post-mortality datasets to test the
 ability of both methods to correctly determine whether or not PIHM was
@@
 the parameters $N_p$, $\mu_p$, and $k_p$ are always unknown in real systems, a
-method that fails under these ideal simulation conditions will certainly also
-fail using less ideal, empirical data. In practice, for the Adjei Method, $N_p$, $\mu_p$,
+method that fails under these ideal simulation conditions with known parameters will certainly also fail when these values must be estimated from empirical data.
+In practice, for the Adjei Method, $N_p$, $\mu_p$,
 and $k_p$ are estimated using the Crofton Method \citep{Adjei1986}, while $\mu_p$ and $k_p$ in
@@
 
-We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:question1}A).
+We compared the two methods using three different mean parasite intensity values ($\mu_p$ = 10, 50, 100) and three different host survival functions (gradual, moderate, and steep decreases in the host survival with increasing parasite intensity, Figure \ref{fig:question1}A).
 For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.
 We examined each $\mu_p$-survival function pair at  three levels of parasite
-aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
+aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these 27 parameter
 combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).
 For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.
-We tested each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500].
-$N_p$ is not technically the sample size on which the methods are being
-tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
-dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.\\
+We also examined the impact of sample size by simulating each parameter for pre-mortality sample sizes of $N_p$ = [50, 100, 200, 300, 400, 500].
+Wild host populations were assumed to be sampled after PIHM has occurred, thus we calculated the sample size in the power simulations as the average number of surviving hosts over all 150 simulations for each parameter combination.
+
+We then tested the ability of the Likelihood Method to correctly identify PIHM
+under the more realistic condition of unknown pre-mortality parameters. Based
+on the first set of simulations, we excluded the Adjei Method and only examined
+the power of the Likelihood Method under ``best-case'' scenario parameter
+values, setting $\mu_p = 10$ and $k = 1$ because PIHM is most detectable when
+parasites are less clumped and mean intensity is low. We examined the impact of survival function shape and
+sample size on the Likelihood Method's ability to identify PIHM when the pre-mortality parameters $\mu_p$ and $k_p$ and the survival function parameters $a$
+and $b$ needed to be estimated.  We performed 500 simulations over a range of
+different samples sizes for gradual, moderate, and steep survival functions, following the simulation procedure described above. \\
 
 \noindent
-\emph{Question 2: Can we estimate fatal parasite intensity and the host survival function?}
+\emph{Question 2: Can we estimate properties of the host survival function?}
 
-To compare the ability of the Adjei Method and the Likelihood Method to recover the $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above.
-For each parameter
-combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
-precision \citep{Walther2005} for these estimates over pre-mortality host population sizes of $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
-10000].
-We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.
-Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.
-
-\subsection*{Efficacy of the Likelihood Method with unknown pre-mortality parameters}
-
-In the final simulation, we tested the ability of the Likelihood Method to
-correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality
-parameters are unknown. The previous simulations showed that the Likelihood
-Method effectively identified PIHM when $\mu_p$ and $k_p$ were known with
-values of 10 and 1, respectively.  As a best-case scenario, we simulated host-
-parasite systems with these pre-mortality parameters and tested the power of
-the Likelihood Method to identify PIHM for gradual, moderate and steep survival
-functions when the pre-mortality parameters $\mu_p$ and $k_p$ also needed to be estimated.  We
-perform 500 simulations over a range of different samples sizes following the
-simulation procedure described above.
+In the previous section we compared the ability of the Adjei Method and the Likelihood Method to provide a ``yes'' or ``no'' answer for whether or not PIHM was occurring in a system.
+In this section we compare the ability of the Adjei Method and the Likelihood Method to estimate properties of the survival function such as the parameters $a$, $b$ and $LD_{50}$.
+Using the same simulation procedure and parameter combinations described above, we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
+precision for these estimates \citep{Walther2005}.
+Because estimating properties of the host survival function requires more information than simply detecting PIHM, we used larger values of $N_p$ for this simulation ($N_p$ = [300, 500, 1000, 2000, 5000, 7500,
+10000]).
+We used the average number of surviving hosts for each set of 150 simulated datasets as our measure of sample size.
+Although both $a$ and $b$ are necessary to estimate $LD_{50}$, the two parameters showed similar patterns of bias and precision so we only show the results for $a$.
 
@@
 We tested the ability of the Adjei Method and the Likelihood Method to identify
-PIHM in 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
+PIHM in six host-parasite datasets given in \cite{Crofton1971a} and four datasets
 given in \cite{Adjei1986} (Table \ref{table:pihm}). \citeauthor{Crofton1971a} analyzed infection patterns in the snail \emph{Gammarus pulex} infected with the
@@
 In both earlier studies, the authors reported PIHM in some of the datasets and we tested whether the Adjei
-Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
-\cite{Crofton1971a}, we truncated the data at 4 parasites, applied the Crofton
+Method and/or the Likelihood Method also predicted PIHM. For the six datasets from
+\cite{Crofton1971a}, we used the general conclusions of the author and truncated the data at four parasites, applied the Crofton
 Method to estimate the pre-mortality distribution, and then ran the Likelihood
@@
 \cite{Adjei1986} datasets, we followed the same procedure as the authors and
-first truncated the data at 2 parasites and then fit the Crofton Method for the
+first truncated the data at two parasites and then fit the Crofton Method for the
 female fish of both species.  Then, following the original authors' methods, we parameterized the male pre-mortality
@@
 applied the Adjei Method and the Likelihood Method to determine whether or not
-PIHM was significant for these species and compared our results to those given by the authors.  All fitting to data was done with the code provided in \emph{SI} 4.
+PIHM was significant for these species and compared our results to those given by the authors.  All code for the analyses is provided in \emph{SI} 4.
 
@@
 system was close to unity for larger sample sizes and tended to
-decrease as sample size decreased (Figure \ref{fig:question1}C; \emph{SI} 3 Figs 1-3).  The Likelihood Method had a power close to
+decrease as sample size decreased for all survival functions (Figure \ref{fig:question1}C; \emph{SI} 3 Figs 1-3).  The Likelihood Method had a power close to
 unity for all parameter combinations and sample sizes considered.  With gradual
@@
 PIHM) for all parameter combinations that we
-considered (Fig. \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).  This method also showed the unintuitive pattern of Type I error
-rate decreasing as sample size decreased.  This pattern was due to the issue of
-binning discussed in the \emph{Introduction} and \emph{SI} 2. For small samples sizes, the
-applicability of the Adjei Method is compromised without binning the observed
-data in some way.  In contrast, the Likelihood Method showed a Type I
+considered (Fig. \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).  This method also showed the unintuitive pattern of decreasing Type I error
+rate with decreasing sample size.
+This occurred because, at small samples sizes, intensity data must be binned before the Adjei Method can be used (\emph{SI} 2).
+In contrast, the Likelihood Method showed a Type I
 error rate at or near the pre-set level of 0.05 for all parameter combinations
 and sample sizes considered (Fig. \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).
+
+When all parameters were jointly estimated, the Likelihood Method showed highly
+context-dependent results even when detecting PIHM under the best-case scenario of
+$\mu_p = 10$ and $k_p = 1$.
+For steep survival curves, PIHM could be detected with a power of greater than 0.8 from a sample of less than 100 hosts (Fig \ref{fig:real_power}).
+However, for moderate survival functions over 400 hosts had to be sampled to achieve the same power and for gradual survival functions, no tested sample size ever achieved a power greater than 0.8 (Fig \ref{fig:real_power}).
 
@@
 the smallest sample sizes we considered, the Likelihood Method's estimate of $LD_{50}$
-was largely unbiased, with small biases occurring for host survival functions
-that were gradual. The precision of the Likelihood Method's $LD_{50}$ estimates decreased
+was largely unbiased, with small biases occurring for gradual host survival functions. The precision of the Likelihood Method's $LD_{50}$ estimates decreased
 (increasing coefficient of variation) as sample size decreased for all
@@
 
-The Adjei Method produced biased estimates of the $LD_{50}$ across nearly all parameter combinations (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6).  For $\mu_p = 10$, the $LD_{50}$
+The Adjei Method produced biased estimates of the $LD_{50}$ across nearly all parameter combinations, tending to underestimate the true value of the parameter (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6).
+For $\mu_p = 10$, the $LD_{50}$
 estimates from the Adjei Method were largely unbiased for large samples sizes, but as $\mu_p$ increased, the Adjei Method
@@
 increasing as sample size decreased (Figure \ref{fig:question2}, \emph{SI} 3 Fig 4-6). The $LD_{50}$ estimates from the Adjei
-Method showed large decreases in precision with the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:question2}, \emph{SI} 3 Fig 4-6).
+Method also showed large decreases in precision with the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:question2}, \emph{SI} 3 Fig 4-6).
 
-In terms of the host survival function, the Likelihood Method gave unbiased estimates of survival function parameters when sample sizes were large, however as sample size decreased these estimates became severely biased (Fig.
+In terms of the host survival function, the Likelihood Method gave unbiased estimates of survival function parameter $a$ when sample sizes were large, however as sample size decreased these estimates became severely biased (Fig.
 \ref{fig:question2}, \emph{SI} Fig 7 - 9) The Adjei Method produced
@@
 
-\subsection*{Detecting PIHM with unknown pre-mortality parameters}
-
-When all parameters were jointly estimated, the Likelihood Method showed highly
-context-dependent results when detecting PIHM in the best-case scenario of
-$\mu_p = 10$ and $k_p = 1$. The Likelihood Method's power of detecting PIHM was
-greater than 0.8 when when host sample sizes were 424 and 83  for survival
-functions that were moderate and steep, respectively (Fig
-\ref{fig:real_power}).  When the host survival function was gradual, the
-Likelihood Method never had a power greater than 0.8 for any post-mortality
-samples sizes we considered (Fig \ref{fig:real_power}).
-
 \subsection*{Application to real data}
 
-Of the 10 datasets we considered, the previous authors qualitatively detected PIHM
-in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
+The previous authors qualitatively detected PIHM
+in 7 of the 10 datasets considered (Table \ref{table:pihm}).  The Likelihood Method parameterized
 from the pre-mortality parameters of the Crofton Method detected significant
@@
 of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$).
-The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results that the Adjei Method has a high Type I error rate.
+The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results showing that the Adjei Method has a high Type I error rate.
 
@@
 
-Quantifying the impact of parasitism on wild host populations is critical for managing wildlife populations and understanding parasite transmission. Ideally the relationship between
-infection intensity and host survival would be measured experimentally, but for
-logistical and ethical reasons, this is often impossible \citep{McCallum2000a}.
-Looking for evidence of mortality in parasite distribution data requires the
-least amount of information, but is notoriously difficult to implement. The
-methodological flaws in the Adjei Method limit its utility, so here we propose
-an alternative, likelihood-based, method to estimate host survival and the
-$LD_{50}$ from observed parasite intensity data.  This
-method is a significant improvement over the previous methods because it requires fewer parameters,
-provides a statistical decision rule for identifying PIHM and does not require
-any data manipulation.
+Our likelihood-based method to estimate parasite-induced host mortality from
+observed parasite intensity data is a significant improvement over the previous
+methods.  In simulations, it had greater power for detecting PIHM over a wider
+range of parameter values and also exhibited fewer false detection events (Type
+I errors) in both simulations and when applied to published datasets previously
+used in PIHM analyses. The Likelihood Method was also generally less biased and
+more precise when quantifying parasite-induced mortality via the host survival
+function for the parameters we considered.  The superior performance of the
+Likelihood Method over the Adjei Method can be attributed to its fewer
+parameters, its lack of unnecessary data alteration, and its applicability
+across a variety of different parameter combinations. In short, the Likelihood
+Method is a better method for detecting and quantifying PIHM than the
+previously proposed Adjei Method.
 
-Using simulated data, we found that the Likelihood Method always out performed the Adjei Method. For simply detecting the presence of PIHM, the Likelihood
-Method was both more powerful and had fewer false detection events (Type I
-errors).  When both methods were applied to published datasets previously used
-in PIHM analyses, the Adjei Method tended to detect PIHM where it had not previously been
-reported, consistent with the high Type I error rate observed in our
-simulations. The Likelihood Method was also more precise and less
-biased in calculations of both the parasite $LD_{50}$ and host survival curve over the parameter values we considered.
-However, while only the Likelihood Method produced precise and unbiased $LD_{50}$
-estimates, neither method could provide unbiased estimates of the host survival
-function at realistic sample sizes.  These simulations demonstrate that
-the Likelihood Method is more powerful and precise than the previously propose Adjei Method.
-
-Although superior to the Adjei Method, the Likelihood Method is not universally applicable to real data.  Our simulations showed when the when pre-
-mortality parameters were estimated directly, the Likelihood Method needed at
-least 83-424 samples to have 80\% power and for steep to moderate survival functions and even
-more as the survival function became more gradual. While some of these sample sizes are reasonable for hosts such as invertebrates or small fish, even the smallest sample sizes are completely
+Although superior to the Adjei Method, the Likelihood Method still cannot be applied to all real datasets.
+For host-parasite systems where host mortality occurs as a steep, non-linear function of parasite intensity only 80 hosts must be sampled to have an 80\% power in detecting PIHM.
+However, as the maximum slope of the survival function decreases and the function becomes, hundreds, or possibly thousands of hosts would have to be sampled to achieve the same result.
+This is consistent with previous studies which illustrate the difficulty of detecting PIHM from linear host survival functions \citep{Lanciani1989}.
+While it may be feasible to sample several hundred invertebrates or small fish, even the smallest sample sizes are completely
 unfeasible for many vertebrates, particularly the species of conservation
@@
 high, or parasite prevalence is low, all of which are
-common in many parasitic helminths.
-Moreover, our results are in agreement with previous work that has shown that as host-survival functions become progressively more linear, PIHM becomes all but impossible to detect \citep{Lanciani1989}.
-This result, however, does not preclude the use of this method as non-linear survival functions
-are not uncommon in empirical host-parasite systems \citep{Benesh2011}.  Finally,
-while linear functions make PIHM undetectable, at the other extreme, steep,
+common in many parasitic helminths. Moreover, while linear functions make PIHM undetectable, at the other extreme, steep,
 non-linear survival curves produce severely biased estimates of the survival
@@
 Given these numerous caveats, is there a place in parasitology for methods that
-estimate PIHM from intensity data?  We are in agreement with
-\cite{Lester1984} that, at the very least, methods for estimating PIHM can
+estimate PIHM from intensity data?  As suggested by
+\cite{Lester1984} these methods for estimating PIHM can
 provide preliminary insight into whether or not PIHM is worth further
 exploration.  However, we stress that these methods should only be used as an
-exploratory tool when assessing the role of PIHM in a system and potential
+exploratory tool when assessing the role of PIHM in a system, and potential
 users should critically evaluate whether they think they have a large enough
@@
 
-TODO
+We thank Hamish McCallum for discussions on the inadequacy of estimating PIHM from intensity data and Kevin Lafferty for manuscript comments.
+We also thank the Briggs Lab Group and Theoretical Ecology group at UCSB for helpful feedback.
+This work was supported by X and X.
 
@@
     $k_p$   & Pre-mortality parasite aggregation \\
-    $N_p$   & Pre-mortality host population size \\
+    $N_p$   & Pre-mortality host sample size \\
     $x$     & Number of parasites in a given host \\
     $g(x; \mu_p, k_p)$ & Pre-mortality negative binomial parasite distribution \\
-    $a$ & Parameter of the logistic host survival function \\
-    $b$ & Parameter of the logistic host survival function \\
-    $h(x; a, b)$ & The host survival function \\
-    $LD_{50}$ & $\exp(a / b)$, parasite intensity at which 50\% of hosts die \\
+
+    $h(\text{survival}; x, a, b)$ & The probability of host survival given a parasite load $x$ and logistic parameters $a$ and $b$\\
+
+    $b / 4$ & The maximum rate of decline in host survival probability with increasing parasite load  \\
+    $a$ & When $b$ is held constant a one unit increase in $a$ leads to a $1 / b$ increase in the parasite intensity at which some percent of hosts survive\\
+    $LD_{50}$ & $\exp(a / b)$, parasite intensity above which a host has a greater than 50\% chance of dying \\
 
@@
 
+\newpage
+
+\begin{figure}
+
+    \begin{tikzpicture}
+        \node at (0, 0) (hist) {\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/concept_crofton_method}};
+
+        % %\node at (2, -2) (trunc) {Truncated tail due to PIHM};
+        % %\draw[->] (trunc.south)--(-1, -3);
+
+        % % \foreach \x in {2,...,4}
+        % %     \draw[ultra thick, dashed, opacity=(1 / \x)] (-3.5 + \x - 1, -3.68)--(-3.5 + \x - 1, 4.75) node[midway, rotate=90, fill=white,opacity=1] {$\text{iter}_{\x}$: Pre $\neq$ Post};
+
+        % \draw[ultra thick, dashed, opacity=0.2] (1.5, -3.68)--(1.5, 4.75) node[midway, rotate=90, fill=white,opacity=1] (ncm) {Pre $\neq$ Post};
+
+        % \draw[ultra thick, dashed] (-2, -3.68)--(-2, 4.75) node[midway, rotate=90, fill=white, opacity=1] (cm) {Pre = Post};
+
+        % \draw[->] (ncm.north)--(cm.south) node[midway, above=0.1cm, align=center, fill=white] {\tiny{Change truncation}\\\tiny{until pre = post}\\\tiny{to estimate pre-mortality}\\\tiny{distribution}};
+
+        % \draw[->] (cm.north)--(-3, 0.55);
+
+        % %\draw[<-, ultra thick, opacity=0.5] (-4, 3)--(-3.5, 3);
+        % %\draw[->, ultra thick] (-3.5, 0)--(-3, 0);
+
+        % %\node at (-3, 0) {Pre};
+
+    \end{tikzpicture}
+
+    \caption{A schematic representation of the iterative approach of the Crofton Method.
+\textbf{A)} The light gray shows the pre-mortality distribution that the Crofton Method is trying to estimate from the dark grey post-mortality distribution.
+The Crofton Method proceeds by truncating the post-mortality data at different levels ($t_i$, $i= 0, \dots, 5$) and finding the parameters pre-mortality host population size ($N_p$), pre-mortality mean parasite intensity ($\mu_p$), and pre-mortality parasite aggregation ($k_p$) that best fit the truncated data.
+\textbf{B)} The parameter $N_p$ is then plotted against the truncation level $t_i$ to determine if a ``kink'' occurs in the parameter values  \citep{Lester1984}.
+This ``kink'' indicates that PIHM is occurring in the system.
+In the above example, PIHM is occurring in the system as visualized by the distinct ``kink'' at $t_3$.}
+    \label{fig:crofton}
+
+\end{figure}
+
+\newpage
+
 \begin{figure}
@@
 
-    \draw[->] (0, 4.2)--(-1.5, 3.1);
-    \draw[->] (0, 4.2)--(1.5, 3.1);
+    \draw[->] (0, 4.2)--(-1.5, 3.2);
+    \draw[->] (0, 4.2)--(1.5, 3.2);
 
     \node at (-3.2, 4.9) {A.};
-    \node at (-6.5, 2.7) {B.};
-    \node at (1.4, 2.7) {C.};
+    \node at (-6.7, 2.9) {B.};
+    \node at (1.15, 2.9) {C.};
 
@@
 The linear and immediate survival functions represent two potential extremes that we do not include in the simulations.
-For each of these survival functions and the parameter combinations described in the main text, we tested the Type I error and power of the Likelihood Method and Adjei Method.
+For each of these survival functions and the parameter combinations described in the main text, we tested the Type I error and power of the Likelihood (Like.) Method and Adjei Method.
 B) Gives the Type I error of each method over a range of pre-mortality sample sizes with a pre-mortality mean parasite intensity ($\mu_p$) of 50 and pre-mortality parasite aggregation ($k_p$) at 0.5.
 The red line shows the pre-set significance level of 0.05.
-C) Gives the Power of each method for detecting PIHM over a range of post-mortality sample sizes for $\mu_p = 50$ and $k_p = 0.5$.
+C) Gives the power of each method for detecting PIHM over a range of post-mortality sample sizes for $\mu_p = 50$ and $k_p = 0.5$.
 In general, the Likelihood Method has higher power and lower Type I error than the Adjei Method.
@@
 
-    \caption{Bias and precision (coefficient of variation) for the Likelihood Method and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
+    \caption{Bias and precision (coefficient of variation) for the Likelihood Method (Like.) and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
 As the coefficient of variation increases, precision decreases.
@@
 
-    \caption{The power of the Likelihood Method to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
-The curves were generated from 500 simulations for 10 pre-mortality population sizes, $N_p$.}
+    \caption{The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
+The curves were generated from 500 simulations for 10 pre-mortality sample sizes, $N_p$.}
 

### 2015-06-03 docs/supplementary_material.tex
Incorporated all draft3 comments and added new figures.

@@
 \usepackage{lipsum}
+\usepackage{hyperref}
 
@@
 
-\section*{Appendix 1: Implementation of the Crofton Method}
+\section*{SI 1: Implementation of the Crofton Method}
 
@@
 with $n$ hosts where each host has some parasite intensity 0 to $p_{max}$. Starting
-with the full dataset, guess a vector of pre-mortality parameters $(N_p, \mu_p, k_p)$ where $N_p$ is the total number of hosts before mortality, $\mu_p$ is the parasite intensity before mortality, and $k_p$ is the parasite aggregation before mortality.
+with the full dataset, guess a vector of pre-mortality parameters $(N_{p1}, \mu_{p1}, k_{p1})$ where $N_{p1}$ is the total number of hosts before mortality, $\mu_{p1}$ is the parasite intensity before mortality, and $k_{p1}$ is the parasite aggregation before mortality.
 Given
@@
 to the observed number hosts with $0, 1, 2, \dots p_{max}$ parasites and calculate
-the $\chi^2$-squared statistic associated with your observed and predicted vectors. In reality, one often has to bin the parasite intensity data because all
-parasite intensities are not represented in the dataset. Continue to guess $(N_p,
-mu_p, k_p)$ vectors until a set of parameters is found that minimizes the $\chi^2$-
+the $\chi^2$-squared statistic associated with the observed and predicted vectors. In reality, one often has to bin the parasite intensity data because all
+parasite intensities are not represented in the dataset. Continue to guess $(N_{p1},
+\mu_{p1}, k_{p1})$ vectors until a set of parameters is found that minimizes the $\chi^2$-
 squared statistic.
 
-Second, choose a truncation value ($t_1$) such that $t_1 <
-p_{max}$. Truncate the data such that $\text{data}_\text{truncated} <= t_1$ and repeat the above
+Second, choose a truncation value ($t_2$) such that $t_2 <
+p_{max}$. Truncate the data such that $\text{data}_\text{truncated} <= t_2$ and repeat the above
 iterative procedure to calculate another set of parameters $(N_{p2}, \mu_{p2}, k_{p2})$
 that minimizes the $\chi^2$-squared statistic on the truncated data. Choose a new
-truncated value $t_2 < t_1$ and repeat the first two steps. Continue to truncate
+truncated value $t_3 < t_2$ and repeat the first two steps. Continue to truncate
 the dataset until it only contains hosts with 0, 1, and 2 parasites (or 3 bins).
@@
 needed for all 3 parameters to be identifiable \citep{Royce1990}.
-Parasite-induced host mortality is traditionally identified by then plotting your truncation values $t_i$ against your different values of $N_{po}$ and looking for a distinct kink in the resulting plot.
+
+Once the iterative procedure has been completed, parasite-induced host mortality is traditionally identified by plotting the different truncation values $t_{i}$ against the different values of $N_{pi}$ and looking for a distinct ``kink'' in the resulting plot.
+Once the ``kink'' as occurred, the values of $N_{pi}$ will typically remain close to constant as $t_i$ is decreased further.
+The ``true'' pre-mortality parameters $N_{pt}$, $\mu_{pt}$, and $k_{pt}$ are taken to be at the point where the ``kink'' occurs.
 
@@
 
-\section*{Appendix 2: Implementation of the Adjei Method}
+\section*{SI 2: Implementation of the Adjei Method}
 
@@
 step is to estimate the parameters of the pre-mortality host-parasite
-distribution using the Crofton Method.  The three parameters estimated are the
+distribution using the Crofton Method (see \emph{SI} 1).  The three parameters estimated are the
 total number of hosts before mortality $N_p$,  the mean number of parasites per
@@
 When $k_p$ is small, parasites are highly aggregated among hosts and when
-$k_p$ is large parasites are more evenly spread out \citep{Wilson2002}.
-The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][ and above]{Royce1990,Lester1984} and we provide a tested implementation of the method in SI 4.
+$k_p$ is large parasites are more evenly distributed across hosts \citep{Wilson2002}.
+The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][ and in \emph{SI} 4]{Royce1990,Lester1984} and we provide a tested implementation of the method in SI 4.
 
-The second step of the Adjei Method is to estimate the shape of the host survival function. \cite{Adjei1986} assume that the host survival function follows the logistic form
+The second step of the Adjei Method is to make the assumption that infection, host mortality, and sampling occur in that order and are temporally \cite{Adjei1986}.
+Next, the \citeauthor{Adjei1986} assume that the host survival function follows the logistic form
 
@@
 survival as parasite intensity increases. The value $\exp(a / b)$ is referred
-to as the $LD_{50}$, which gives the parasite intensity at which 50\% of hosts
-experience mortality.
+to as the $LD_{50}$. Individuals with loads higher than this will have a greater than 50\% chance of death.
 
-To estimate this function, the Adjei Method first
+By taking the first and second derivatives of equation \ref{eq:logistic}, one can easily find that the maximum rate of decline in host survival probability with increasing parasite intensity occurs at the $LD_{50}$ and has a value of $b / 4$.
+This is in many ways analogous to the parasite pathogenicity parameter $\alpha$ given in classic macroparasite models, which specifies the slope of the linear relationship between between host death rate and parasite intensity \citep{AndersonandMay1978, Isham1995}.
+The parameter $a$ is easily interpreted by holding $b$ constant and looking at how a one unit change in $a$ affects the log parasite intensity at which some percentage $p$ of hosts experience mortality.
+Letting $a_1$ and $a_2$ be two different values of $a$ and $x_1$ and $x_2$ be two different parasite intensities, a bit of rearranging of equation \ref{eq:logistic} gives.
+
+\begin{align*}
+    \log{\frac{p}{1 - p}} &= a_1 - b \log{x_1}\\
+    - \log{\frac{p}{1 - p}} &= a_1 - b \log{x_2}\\
+    0 &= a_1 - a_2 - b\log{x_1} + b\log{x_2}\\
+    a_2 - a_1 &= b(\log{x_2} - \log{x_1})
+\end{align*}
+If $a_2 - a_1 = 1$, than the change in log parasite intensity at which $p$ percentage of hosts survive is $1 / b$.
+
+To estimate the parameters in equation \ref{eq:logistic}, the Adjei Method first
 calculates the expected number of hosts with a given parasite load $x$ by using
@@
 
-% \section*{Appendix 3: Implementation of the Likelihood Method}
-
-% As described in the main text, the probability mass function for the likelihood method is given by
-
-% \begin{equation}
-%     P(x | \text{survival}) = \dfrac{P(\text{survival} | x) * P(x)}{P(\text{survival})}
-%     \label{eq:concept}
-% \end{equation}
-
-% One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$.
-Therefore equation \ref{eq:concept} can be written as
-
-% \begin{equation}
-%     P(x | \text{survival}) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
-%     \label{eq:dist}
-% \end{equation}
-
-% Using this equation, we seek to find estimates for $\mu_p$, $k_p$, $a$, and $b$
-% that maximize the likelihood of equation \ref{eq:dist} given a host-parasite
-% dataset. The first approach is use a standard maximization algorithm to try to
-% find all four parameters.  However, for small number of hosts and/or a gradual
-% survival function, the gradient of the likelihood will not contain much
-% information (check this) and the maximization will likely fail or give
-% unrealistic results. An alternative approach would be to parameterize $\mu_p$
-% and $k_p$ using the Crofton Method and then use equation \ref{eq:dist} to find the values for $a$ and $b$ that maximize the likelihood of a host-parasite dataset.
-
-\section*{Appendix 3: Additional Figures}
+\section*{SI 3: Additional Figures}
 
@@
 
-\section*{Appendix 4: Code and unit tests for estimating parasite-induced host mortality}
+\section*{SI 4: Code and unit tests for estimating parasite-induced host mortality}
 
-Python code, unit tests, and a help file for the Crofton Method, the Adjei Method and the Likelihood Method can be found in \verb*#pihm_methods.py#, \verb*#test_pihm_methods.py#, and \verb*#help_file.txt#, respectively.
+Python code, unit tests, and a help file for the Crofton Method, the Adjei Method and the Likelihood Method can be found at \url{https://github.com/mqwilber/parasite_mortality}
 

### 2015-06-03 code/pihm_methods.py

@@ -73,7 +73,8 @@ class PIHM:
         return mod.nbinom.fit_mle(self.data, k_array=k_array)
 
 
-    def crofton_method(self, crof_bin_edges, guess=None):
+    def crofton_method(self, crof_bin_edges, guess=None, bounded=False,
+        upper=[2, 2, 5]):
         """
         Crofton's method for estimating the pre-mortality parameters N_p
         (population size before mortality),
@@ -85,8 +86,6 @@ class PIHM:
 
         Parameters
         ----------
-        data : array
-            Parasite data across hosts
         crof_bin_edges : list
             List specifying upper and lower bin boundaries. Upper right most bin
             boundary is inclusive; all other upper bounds are exclusive.  Example:
@@ -98,6 +97,12 @@ class PIHM:
             If tuple, should be a guess for mu and k of the negative binomial.
             If None, uses fitted mu and k for a guess.
 
+        bounded : bool
+            If True, uses a bounded minimizer.  Else uses an unbounded
+        upper : list
+            A list of multipliers/upper bounds. The first multiplies the N upper
+            bound, the second item multiplies the mu upper bound, and the third
+            item specifies the k upper bound.
         Returns
         -------
         : Np, mup, kp
@@ -126,10 +131,24 @@ class PIHM:
             # Minimizing chi-squared
             return np.sum((obs - pred)**2 / pred)
 
-        full_out = opt.fmin(opt_fxn, np.array([N_guess, mu_guess, k_guess]),
-                                        disp=0)
+        if not bounded:
+
+            full_out = opt.fmin(opt_fxn, np.array([N_guess, mu_guess, k_guess]),
+                                            disp=0)
+
+            opt_params = full_out
+
+        else:
+
+            bounds = [(0, upper[0]*N_guess), (0, upper[1]*mu_guess),
+                                (0.001, upper[2])]
+            full_out = opt.fmin_l_bfgs_b(opt_fxn,
+                            np.array([N_guess, mu_guess, k_guess]),
+                            bounds=bounds, approx_grad=True, disp=0)
+            opt_params = full_out[0]
+
+
         # Calculate predicted in bins
-        opt_params = full_out
         # pred = [opt_params[0] * np.sum(mod.nbinom.pmf(s, opt_params[1],
         #     opt_params[2])) for s in splits]
 
@@ -330,7 +349,7 @@ class PIHM:
                 raise TypeError("mup, kp must be preset")
 
[…]

### 2015-06-08 docs/parasite_mortality_manuscript.tex
Added in Saras comment on 6-2 and added to TODO list

@@
 \usepackage{tikz}
+\usepackage{hyperref}
 
@@
 
-\textbf{Keywords:} & aggregation, negative binomial distribution, parasite intensity, Crofton Method, Likelihood Method
-\\
-
 \textbf{Corresponding \newline author:} &
@@
 introduce a novel method for using intensity data to detect and quantify
-parasite-induced mortality in wildlife populations. Using simulations, we show
+parasite-induced mortality in wildlife populations. We show
 that this method is more reliable than previously proposed methods while
@@
 data that are consistent with previously published qualitative estimates.
-However, we stress that this method, and all techniques that estimate parasite-
-induced mortality from intensity data alone, have several critical assumptions
+However, we stress that this method, and all techniques that estimate parasite-induced mortality from intensity data alone, have several critical assumptions
 that limit their applicability in the real world. Due to these limitations,
@@
 
-\textbf{Keywords}: parasite aggregation, negative binomial distribution
+\textbf{Keywords}: parasite aggregation, negative binomial distribution, Crofton Method, host survival function, $LD_{50}$
 
@@
 \linenumbers
-\section*{Introduction}
+\section{Introduction}
 
-Infectious agents can impact on animal populations through changing
+Infectious agents can impact animal populations by changing
 population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
@@
 regulates host and parasite populations, making predictions about disease
-transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of microparasite pathogens, such as rabies \citep{Coyne1989}, bovine tuberculosis \citep{Cox2005}, and
-rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, and host survival is not a function of the number of infectious agents present.
+transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies \citep{Coyne1989}, bovine tuberculosis \citep{Cox2005}, and
+rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, and host survival is not a function of the number of infectious agents present within in the host.
 In contrast, for macroparasites, pathology increases with parasite burden and host survival probability must be treated as a function of infection intensity.
@@
 quantify the relationship between infection intensity and host survival
-probability; an important relationship for determining the potential impact of
-a parasite on host dynamics \citep{AndersonandMay1978s}. \cite{Adjei1986}
+probability. \cite{Adjei1986}
 suggested that this relationship could be calculated by first using the Crofton
@@
 the parameters of the host survival function (see \emph{SI} 2 for a technical
-description of the Adjei Method). \citeauthor{Adjei1986} noted that this
-survival function could provide useful information such as the parasite
-intensity at which a host has a greater than 50\% chance of suffering parasite-
-induced mortality. However, if mean infection intensity is high or sample sizes
+description of the Adjei Method).  This
+survival function can provide useful information such as the parasite
+intensity at which a host has a greater than 50\% chance of suffering parasite-induced mortality. However, if mean infection intensity is high or sample sizes
 are small the observed intensity data must be subjectively binned into
@@
 
-\section*{Methods}
+\section{Methods}
 
-\subsection*{A novel, likelihood-based method for estimating PIHM}
+\subsection{A novel, likelihood-based method for estimating PIHM}
 
-Our method (henceforth the Likelihood Method) begins with the same assumptions as the Adjei Method: namely that infection of
-a host, parasite-induced mortality of a host, and the sampling of a host
-population occur at distinct time intervals during a host's life. As discussed by
-\citeauthor{Adjei1986}, this is not necessarily an unrealistic as some parasite infections are host age dependent with mortality occurring soon after infection \citep{Schotthoefer2003,Johnson2008}.
-[CHAT WITH SARA ABOUT THIS]
+Our method (henceforth the Likelihood Method) begins with the same assumptions as the Adjei Method: namely that infection has occurred and hosts with fatal parasite loads have died prior to the population sampling.
+As discussed by
+\citeauthor{Adjei1986}, this is not necessarily an unrealistic as some parasite infections occur primarily in younger hosts with parasite-induced mortality occurring soon after infection \citep[e.g.][]{Schotthoefer2003,Johnson2008}.
 
@@
 $\boldsymbol{\phi}$ is a vector of parameters that describes the shape of this distribution.
-The method then assumes that the probability of a host surviving with $x$ parasites from infection until sampling is given by the host survival function $h(\text{survival} ; x, \boldsymbol{\theta})$ where $\boldsymbol{\theta}$ specifies any additional parameters needed to define the host survival function.
+The probability of a host surviving with $x$ parasites from infection until sampling is given by the host survival function $h(\text{survival} ; x, \boldsymbol{\theta})$ where $\boldsymbol{\theta}$ specifies any additional parameters needed to define the host survival function.
 
@@
 Here, we assume that the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ follows a negative binomial
-distribution with the parameters mean parasite intensity ($\mu_p$) and
-aggregation ($k_p$) before mortality (smaller $k_p$ indicates
-more aggregation) \citep{Crofton1971a,AndersonandMay1978,Adjei1986}.
+distribution with parameters mean parasite intensity ($\mu_p$) and
+aggregation ($k_p$) before mortality \citep{Crofton1971a,AndersonandMay1978,Adjei1986}.
 A variety of different biological and statistical assumptions can result in an equilibrium parasite distribution that follows a negative binomial distribution
@@
 
-To compare the Likelihood Method and the previously proposed Adjei Method, we adopt the host-survival function used in the earlier study and assume it is non-linear and follows a logistic function given by
+To compare the Likelihood Method and the previously proposed Adjei Method, we adopt the non-linear, logistic host-survival function used in the earlier study given by
 
@@
 Generally, a larger $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases, with the maximum rate of decline having a value of $b / 4$ (\emph{SI} 2).
-$b$ is in many ways analogous to the the pathogenicity parameter in classic macroparasite models which gives that rate at which host death rate changes with increasing parasite intensity \citep{AndersonandMay1978,Isham1995}.
+$b$ is in many ways analogous to the the pathogenicity parameter in classic macroparasite models which gives the rate at which host death rate increases with increasing parasite intensity \citep{AndersonandMay1978,Isham1995}.
 When $b$ is held constant, a larger $a$ allows for hosts to tolerate larger parasite intensities before experiencing parasite-induced mortality.
@@
 
-\subsection*{Evaluating the Adjei and Likelihood Methods}
+\subsection{Evaluating the Adjei and Likelihood Methods}
 
@@
 In the previous section we compared the ability of the Adjei Method and the Likelihood Method to provide a ``yes'' or ``no'' answer for whether or not PIHM was occurring in a system.
-In this section we compare the ability of the Adjei Method and the Likelihood Method to estimate properties of the survival function such as the parameters $a$, $b$ and $LD_{50}$.
+In this section we compared the ability of the Adjei Method and the Likelihood Method to estimate properties of the survival function such as the parameters $a$, $b$ and $LD_{50}$.
 Using the same simulation procedure and parameter combinations described above, we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
@@
 
-\subsection*{Application to real data}
+\subsection{Application to real data}
 
@@
 
-\section*{Results}
+\section{Results}
 
-\subsection*{Question 1: Detecting presence of PIHM}
+\subsection{Question 1: Detecting presence of PIHM}
 
@@
 
-\subsection*{Question 2: Estimating the $LD_{50}$ and survival function}
+\subsection{Question 2: Estimating the $LD_{50}$ and survival function}
 
@@
 
-\section*{Discussion}
+\section{Discussion}
 
@@
 For host-parasite systems where host mortality occurs as a steep, non-linear function of parasite intensity only 80 hosts must be sampled to have an 80\% power in detecting PIHM.
-However, as the maximum slope of the survival function decreases and the function becomes, hundreds, or possibly thousands of hosts would have to be sampled to achieve the same result.
+However, as the maximum slope of the survival function decreases and the function becomes somewhat linear, hundreds, or possibly thousands of hosts would have to be sampled to achieve the same result.
 This is consistent with previous studies which illustrate the difficulty of detecting PIHM from linear host survival functions \citep{Lanciani1989}.
@@
 
-    $b / 4$ & The maximum rate of decline in host survival probability with increasing parasite load  \\
-    $a$ & When $b$ is held constant a one unit increase in $a$ leads to a $1 / b$ increase in the parasite intensity at which some percent of hosts survive\\
+    $b / 4$ & The maximum rate of decline in host survival probability with increasing parasite intensity  \\
+    $a$ & When $b$ is held constant a one unit increase in $a$ leads to a $1 / b$ increase in the log parasite intensity at which some percent of hosts survive\\
     $LD_{50}$ & $\exp(a / b)$, parasite intensity above which a host has a greater than 50\% chance of dying \\
@@
 \textbf{A)} The light gray shows the pre-mortality distribution that the Crofton Method is trying to estimate from the dark grey post-mortality distribution.
-The Crofton Method proceeds by truncating the post-mortality data at different levels ($t_i$, $i= 0, \dots, 5$) and finding the parameters pre-mortality host population size ($N_p$), pre-mortality mean parasite intensity ($\mu_p$), and pre-mortality parasite aggregation ($k_p$) that best fit the truncated data.
+The Crofton Method proceeds by truncating the post-mortality data at different levels ($t_i$, e.g.
+$i= 0, \dots, 5$) and finding the parameters pre-mortality host population size ($N_p$), pre-mortality mean parasite intensity ($\mu_p$), and pre-mortality parasite aggregation ($k_p$) that best fit the truncated data.
 \textbf{B)} The parameter $N_p$ is then plotted against the truncation level $t_i$ to determine if a ``kink'' occurs in the parameter values  \citep{Lester1984}.

### 2015-06-08 docs/parasite_mortality_manuscript.tex
Reformatted references for IJP

@@
 1) Department of Ecology, Evolution and Marine Biology, University of California,
-Santa Barbara, Santa Barbara California \\
+Santa Barbara, Santa Barbara California, United States of America \\
 
@@
 
-\textbf{Keywords}: parasite aggregation, negative binomial distribution, Crofton Method, host survival function, $LD_{50}$
+\textbf{Keywords}: parasite aggregation, negative binomial distribution, Crofton Method, host survival function, lethal dose
 
 }
+
+\bigskip
+
+\noindent
+\textbf{Highlights}
+\begin{itemize}
+    \item We develop and test a novel method for detecting and quantifying parasite-induced host mortality from parasite intensity data.
+    \item Our method outperforms previous methods for detecting parasite-induced host mortality.
+    \item Our method correctly quantifies parasite-induced host mortality in empirical data where mortality has been qualitatively reported
+    \item  All current methods for detecting parasite-induced host mortality have fundamental and untestable assumptions that limit their applicability to many datasets.
+\end{itemize}
 
@@
 \begin{equation}
-    h(\text{survival}; x, a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+    h(\text{survival}; x, a, b) = \dfrac{\exp{(a - b \log(x))}}{1 + \exp{(a - b \log(x))}}
     \label{eq:logistic}
@@
 processes such as within host density dependence, age dependent variation in host
-resistance and/or heterogeneous infection rates \citep{McCallum2000a,Anderson1982a,Rousset1996}.  This means that in the event
+resistance and/or heterogeneous infection rates \citep{Anderson1982a,Rousset1996, McCallum2000a}.  This means that in the event
 that PIHM is detected, it may actually not be the result of PIHM.
@@
 \singlespacing
-\bibliographystyle{/Users/mqwilber/Dropbox/Documents/Bibformats/ecology_letters.bst}
-\bibliography{/Users/mqwilber/Dropbox/Documents/Bibfiles/Projects_and_Permits-parasite_host_mortality}
+\bibliographystyle{int_j_para.bst}
+\bibliography{parasite_host_mortality}
 
@@
 The Crofton Method proceeds by truncating the post-mortality data at different levels ($t_i$, e.g.
-$i= 0, \dots, 5$) and finding the parameters pre-mortality host population size ($N_p$), pre-mortality mean parasite intensity ($\mu_p$), and pre-mortality parasite aggregation ($k_p$) that best fit the truncated data.
+$i= 0, \dots, 5$) and finding the pre-mortality host population size ($N_p$), pre-mortality mean parasite intensity ($\mu_p$), and pre-mortality parasite aggregation ($k_p$) that best fit the truncated data.
 \textbf{B)} The parameter $N_p$ is then plotted against the truncation level $t_i$ to determine if a ``kink'' occurs in the parameter values  \citep{Lester1984}.

### 2015-06-09 docs/parasite_mortality_manuscript.tex
Cleaning up code directory for manuscript submission

@@
 505-321-5381 \newline
-mark.wilber@lifesci.ucsb.edu
+mark.wilber@lifesci.ucsb.edu \newline
 
 \end{tabular}
+
+\textbf{Note}: Supplementary data associated with this article
 
@@
 introduce a novel method for using intensity data to detect and quantify
-parasite-induced mortality in wildlife populations. We show
+parasite-induced mortality in wildlife populations. We use simulations to show
 that this method is more reliable than previously proposed methods while
@@
     \item Our method correctly quantifies parasite-induced host mortality in empirical data where mortality has been qualitatively reported
-    \item  All current methods for detecting parasite-induced host mortality have fundamental and untestable assumptions that limit their applicability to many datasets.
+    \item  All current methods for detecting parasite-induced host mortality have fundamental and untestable assumptions that limit their applicability in the real world.
 \end{itemize}
@@
 Infectious agents can impact animal populations by changing
-population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
+population dynamics and stability \citep{Dobson1992,Tompkins2002}, altering predator-prey interactions \citep{Joly2004}, and
 even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
@@
 transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies \citep{Coyne1989}, bovine tuberculosis \citep{Cox2005}, and
-rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, and host survival is not a function of the number of infectious agents present within in the host.
+rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, such that host survival is not a function of the number of infectious agents present within the host.
 In contrast, for macroparasites, pathology increases with parasite burden and host survival probability must be treated as a function of infection intensity.
@@
 quantify the relationship between infection intensity and host survival
-probability. \cite{Adjei1986}
+probability; information that is necessary for estimating parasite impacts on host populations \citep{AndersonandMay1978,Tompkins2002}. \cite{Adjei1986}
 suggested that this relationship could be calculated by first using the Crofton
@@
 the parameters of the host survival function (see \emph{SI} 2 for a technical
-description of the Adjei Method).  This
-survival function can provide useful information such as the parasite
-intensity at which a host has a greater than 50\% chance of suffering parasite-induced mortality. However, if mean infection intensity is high or sample sizes
+description of the Adjei Method).  Although this method can predict the host survival function, it has several technical drawbacks. When mean infection intensity is high or sample sizes
 are small the observed intensity data must be subjectively binned into
@@
 impacts of parasitism using parasite intensity data.
-In these methods, PIHM can only be identified by visually examining plots of the pre-mortality parameters predicted by Crofton Method and determining whether they show a ``kink'' over a range of truncation values \citep[Figure \ref{fig:crofton}B;][]{Lester1984,Ferguson2011}.
+In these methods, PIHM can only be identified by visually examining plots of the pre-mortality parameters predicted by the Crofton Method and determining whether they show a ``kink'' over a range of truncation values \citep[Figure \ref{fig:crofton}B;][]{Lester1984,Ferguson2011}.
 This qualitative criteria makes it difficult to compare PIHM between studies and a more rigorous and quantitative method is needed to both detect and quantify host mortality.
@@
 compare the results. Finally, we discuss the limitations of inferring PIHM
-from intensity data and the place of any of these methods in modern quantitative parasitology.
+from intensity data and how these methods fit in modern quantitative parasitology.
 
@@
 Here, we assume that the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ follows a negative binomial
-distribution with parameters mean parasite intensity ($\mu_p$) and
-aggregation ($k_p$) before mortality \citep{Crofton1971a,AndersonandMay1978,Adjei1986}.
+distribution with two parameters mean parasite intensity ($\mu_p$) and
+aggregation ($k_p$, where smaller $k_p$ indicates a more aggregated parasite population) before mortality \citep{Crofton1971a,AndersonandMay1978,Adjei1986}.
 A variety of different biological and statistical assumptions can result in an equilibrium parasite distribution that follows a negative binomial distribution
@@
 In systems where there is truly a linear relationship between infection intensity and survival probability it will be nearly impossible to use intensity data to detect parasite-induced host mortality \citep{Lanciani1989}.
-However,  there is substantial empirical evidence that host-parasite systems do not universally exhibit linear host-survival functions \citep{Benesh2011}, suggesting a linear form for $h(\text{survival} ; x, \boldsymbol{\theta}$) is not always the most reasonable.
+However, some systems do exhibit non-linear host survival functions \citep{Benesh2011}, in which case these methods would be applicable.
 
@@
 Generally, a larger $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases, with the maximum rate of decline having a value of $b / 4$ (\emph{SI} 2).
-$b$ is in many ways analogous to the the pathogenicity parameter in classic macroparasite models which gives the rate at which host death rate increases with increasing parasite intensity \citep{AndersonandMay1978,Isham1995}.
+$b$ is in many ways analogous to the the pathogenicity parameter ($\alpha$) in classic macroparasite models that gives the parasite intensity dependent host death rate.
+\citep{AndersonandMay1978,Isham1995}.
 When $b$ is held constant, a larger $a$ allows for hosts to tolerate larger parasite intensities before experiencing parasite-induced mortality.
@@
 Equation \ref{eq:logistic} is commonly used in toxicology and has the useful properties of being bounded between 0 and 1 and being differentiable for all $x$ \citep{Collet2002}.
-That being said, it is phenomenological and there is little theoretical justification to use it rather than it tends to fit survival data.
+That being said, it is phenomenological and is used simply because it tends to fit survival data.
 However, given that a goal of these analyses is to
@@
 first truncated the data at two parasites and then fit the Crofton Method for the
-female fish of both species.  Then, following the original authors' methods, we parameterized the male pre-mortality
+female fish of both species.  Then, following the \citeauthor{Adjei1986}'s methods, we parameterized the male pre-mortality
 distributions for each species with the results from the females.  Finally, we
@@
 
-The Adjei Method showed highly inflated Type I error rates (i.e. falsely detected
+The Adjei Method had highly inflated Type I error rates (i.e. falsely detected
 PIHM) for all parameter combinations that we
@@
 \ref{fig:question2}, \emph{SI} Fig 7 - 9) The Adjei Method produced
-biased estimates of the host survival function across all sample sizes, with the bias
-consistently being larger when the survival function was steeper and $\mu_p$ was larger (Fig \ref{fig:question2}, \emph{SI} 3 Figs 7-9).
+biased estimates of the host survival function across all sample sizes, with consistently greater bias for steeper survival functions and higher mean parasite loads.
+(Fig \ref{fig:question2}, \emph{SI} 3 Figs 7-9).
 
@@
 
-We thank Hamish McCallum for discussions on the inadequacy of estimating PIHM from intensity data and Kevin Lafferty for manuscript comments.
-We also thank the Briggs Lab Group and Theoretical Ecology group at UCSB for helpful feedback.
-This work was supported by X and X.
+We thank the Briggs Lab group and Theoretical Ecology group at UCSB for helpful feedback.
+This work was supported by National Science Foundation Graduate Research Fellowship (Grant No.
+DGE 1144085) and X.
+
+% We thank Hamish McCallum for discussions on the inadequacy of estimating PIHM from intensity data and Kevin Lafferty for manuscript comments.
 

### 2015-06-09 docs/supplementary_material.tex
Cleaning up code directory for manuscript submission

@@
 $k_p$ is large parasites are more evenly distributed across hosts \citep{Wilson2002}.
-The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][ and in \emph{SI} 4]{Royce1990,Lester1984} and we provide a tested implementation of the method in SI 4.
+The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][ and in \emph{SI} 1]{Royce1990,Lester1984} and we provide a tested implementation of the method in \emph{SI} 4.
 
 The second step of the Adjei Method is to make the assumption that infection, host mortality, and sampling occur in that order and are temporally \cite{Adjei1986}.
-Next, the \citeauthor{Adjei1986} assume that the host survival function follows the logistic form
+Next, \citeauthor{Adjei1986} assume that the host survival function follows the logistic form
 
 \begin{equation}
-    h(x | a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+    h(\text{survival}; x, a, b) = \dfrac{\exp{(a - b \log(x))}}{1 + \exp{(a - b \log(x))}}
     \label{eq:logistic}
@@
     \includegraphics[width=0.5\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/compare_crofton.pdf}
-    \caption{A comparison of this papers implementation (solid line, circles) of the Crofton Method with the results given in \cite{Crofton1971a} (dashed line, diamonds).
+    \caption{A comparison of this paper's implementation (solid line, circles) of the Crofton Method with the results given in \cite{Crofton1971a} (dashed line, diamonds).
 Figure A compares the predicted number of hosts in a population pre-mortality ($N_p$).

### 2015-06-09 code/analysis_for_manuscript.py

@@ -36,20 +36,6 @@ mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], []),
 Nps = [300, 500, 1000, 2000, 5000, 7500, 10000]
 N_samp = 150
 
-# ks = [1] # 0.5, .1]
-
-# mu_ld50 = {50: ([(5, -1.4), (10, -2.8), (20, -5.6)], [])}
-
-# # {100: ([(5, -1.15), (10, -2.3), (20, -4.6)], [])}
-
-# # {10: ([(5, -2.5), (10, -5), (20, -10)], []),
-# #             100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
-# #             500: ([(10, -1.8), (20, -3.6), (60, -10.8)], [])}
-
-# Nps = [300] # 500, 1000, 2000, 5000, 7500, 10000]
-# N_samp = 10
-
-
 
 scenario_1_results = {}
 

### 2015-06-09 code/analysis_for_manuscript_compare_crofton_method.py

@@ -1,6 +1,5 @@
 import numpy as np
-import crofton_method as cm
-reload(cm)
+import pihm_methods as pm
 
 """
 Description
@@ -71,12 +70,15 @@ crof_st1_compare_ss = []
 for i in trunc:
 
     bin_edges = list(np.arange(0, i + 1)) + [i + 0.9]
-    our_meth = cm.crofton_method(st1_raw, bin_edges)
-    N, mu, k = our_meth[0]
+    pm_obj = pm.PIHM(st1_raw)
+
+
+
+    N, mu, k = pm_obj.crofton_method(bin_edges)
     our_st1_ks.append(k)
     our_st1_N.append(N)
 
-    our_pred = cm.get_pred(N, mu, k, np.arange(0, 10))[0]
+    our_pred = pm_obj.obs_pred(np.arange(0, 10))['predicted']
     crof_ss = chi_sq(st1_pred[i], st1_obs)
     our_ss = chi_sq(our_pred, st1_obs)
     crof_st1_compare_ss.append((crof_ss, our_ss))
@@ -92,12 +94,13 @@ crof_st2_compare_ss = []
 for i in trunc:
 
     bin_edges = list(np.arange(0, i + 1)) + [i + 0.9]
-    our_meth = cm.crofton_method(st2_raw, bin_edges)
-    N, mu, k = our_meth[0]
+    pm_obj = pm.PIHM(st2_raw)
+
+    N, mu, k = pm_obj.crofton_method(bin_edges)
     our_st2_ks.append(k)
     our_st2_N.append(N)
 
-    our_pred = cm.get_pred(N, mu, k, np.arange(0, 10))[0]
+    our_pred = pm_obj.obs_pred(np.arange(0, 10))['predicted']
     crof_ss = chi_sq(st2_pred[i], st2_obs)
     our_ss = chi_sq(our_pred, st2_obs)
     crof_st2_compare_ss.append((crof_ss, our_ss))
@@ -115,12 +118,13 @@ crof_st3_compare_ss = []
 for i in trunc:
 
     bin_edges = list(np.arange(0, i + 1)) + [i + 0.9]
-    our_meth = cm.crofton_method(st3_raw, bin_edges)
-    N, mu, k = our_meth[0]
+    pm_obj = pm.PIHM(st3_raw)
+
+    N, mu, k = pm_obj.crofton_method(bin_edges)
     our_st3_ks.append(k)
     our_st3_N.append(N)
 
[…]

### 2015-06-09 code/analysis_for_manuscript_full_likelihood.py

@@ -7,10 +7,15 @@ import pandas as pd
 Description
 -----------
 
+The analysis looks at the Power of the Likelihood Method for detecting PIHM
+when all the parameters (a, b, mu_, and k_p) must be esitmated from parasite
+intensity data.  Just looks at the best case scenario when k_p=1 and mu_p=10.
+
+A thorough description of this analysis can be found in
+../doc/parasite_mortality_manuscript.tex
 
 
 Author : Mark Wilber
-
 
 """
 

### 2015-06-09 code/analysis_for_manuscript_test_data.py

@@ -1,5 +1,4 @@
 import pihm_methods as pm
-reload(pm)
 import numpy as np
 import pandas as pd
 
@@ -13,6 +12,8 @@ A couple different tests
 
 1. Run Crofton Method and then test for PIHM (Adjei and Likelihood)
 2. For Likelihood, fit with all parameters
+
+These results are given in Table 2 in the manuscript
 
 """
 

### 2015-06-09 code/analysis_for_manuscript_typeIandII.py

@@ -26,7 +26,6 @@ Author : Mark Wilber
 """
 
 
-# # Parameters
 # Parameters
 ks = [1, 0.5, .1]
 
@@ -34,7 +33,7 @@ mu_ld50 = {10: ([(5, -2.5), (10, -5), (20, -10)], []),
             100: ([(5, -1.15), (10, -2.3), (20, -4.6)], []),
             50: ([(5, -1.4), (10, -2.8), (20, -5.6)], [])}
 
-Nps = [50, 100, 200, 300, 400, 500] #[300, 500, 1000, 2000, 5000, 7500, 10000]
+Nps = [50, 100, 200, 300, 400, 500]
 N_samp = 150
 
 scenario_1_results = {}

### 2015-06-09 code/test_pihm_methods.py

@@ -11,6 +11,13 @@ import scipy.stats as stats
 import matplotlib.pyplot as plt
 import macroeco.models as mod
 
+"""
+Dsecription
+------------
+Unittests for methods given in pihm_methods
+
+"""
+
 class TestPIHM(TestCase):
 
     def test_crofton_method(self):
@@ -133,7 +140,8 @@ class TestPIHM(TestCase):
 
         assert_array_almost_equal(crof_st3_ks, our_st3_ks, decimal=1)
 
-        # Close to equal but not quite
+        # Close to equal but not quite.  Which is ok.  We don't expect them
+        # to be exactly equal
         # Uncomment and run nosetests to see
 
         # assert_array_equal(crof_st3_N, our_st3_N)

### 2015-06-09 code/pihm_methods.py
Updated comments in pihm_methods

@@ -11,7 +11,8 @@ from scipy.stats import chisqprob
 
 """
 Module contains the updated functions and classes for assessing
-parasite-induced host mortality (PIHM)
+parasite-induced host mortality (PIHM).  For examples of use see the
+analysis*.py files in this same directory.
 
 """
 
@@ -20,10 +21,22 @@ class PIHM:
     Class PIHM (parasite-induced host morality) provides an interface for
     analyzing the effect of parasite-induced host mortality on a given dataset
 
-    Description
-    ------------
-    The class PIHM has two primary functions: 1) To test whether an observed
-    host-parasite distribution is experiencing parasite induced host morality.
+    There are 5 key paramters:
+
+    Np : Pre-mortality host population/sample size
+    mup : Premortality mean parasite intensity
+    kp : Premortality parasite aggregation.
+    a : Parameter of the host survival function.
+    b : Parameter of the host survival function.
+
+    See ../doc/parasite_mortality_manuscript.tex for a more thorough Description
+    of these parameters.
+
+    Examples
+    --------
+
+    For examples of use see the analysis*.py files in this same directory.
+
 
     """
 
@@ -40,6 +53,13 @@ class PIHM:
     def set_premort_params(self, Np, mup, kp):
         """
         Set the premortality parameters
+
+        Parameters
+        ----------
+        Np : Pre-mortality host population/sample size
+        mup : Premortality mean parasite intensity
+        kp : Premortality parasite aggregation
+
         """
 
         self.Np, self.mup, self.kp = Np, mup, kp
@@ -47,6 +67,15 @@ class PIHM:
     def set_all_params(self, Np, mup, kp, a, b):
         """
         Set the premortality parameters
+
+        Parameters
+        ----------
+
+        Np : Pre-mortality host population/sample size
+        mup : Premortality mean parasite intensity
+        kp : Premortality parasite aggregation.
+        a : Parameter of the host survival function.
+        b : Parameter of the host survival function.
         """
 
         self.Np, self.mup, self.kp, self.a, self.b = Np, mup, kp, a, b
@@ -67,7 +96,13 @@ class PIHM:
 
     def fit_nbd(self, k_array=np.linspace(0.01, 20, 100)):
         """
-        Fit a NBD to the parasite data
+        Fit a negative binomial distribution to the parasite data
+
+        Parameters
+        ----------
+        k_array : array-like
+            Array of values over which to fit
+
[…]

### 2015-06-09 docs/parasite_mortality_manuscript.tex
Adding highlights and moving materials

@@
 }
-
-\bigskip
-
-\noindent
-\textbf{Highlights}
-\begin{itemize}
-    \item We develop and test a novel method for detecting and quantifying parasite-induced host mortality from parasite intensity data.
-    \item Our method outperforms previous methods for detecting parasite-induced host mortality.
-    \item Our method correctly quantifies parasite-induced host mortality in empirical data where mortality has been qualitatively reported
-    \item  All current methods for detecting parasite-induced host mortality have fundamental and untestable assumptions that limit their applicability in the real world.
-\end{itemize}
 
@@
 
-        % %\node at (2, -2) (trunc) {Truncated tail due to PIHM};
-        % %\draw[->] (trunc.south)--(-1, -3);
-
-        % % \foreach \x in {2,...,4}
-        % %     \draw[ultra thick, dashed, opacity=(1 / \x)] (-3.5 + \x - 1, -3.68)--(-3.5 + \x - 1, 4.75) node[midway, rotate=90, fill=white,opacity=1] {$\text{iter}_{\x}$: Pre $\neq$ Post};
-
-        % \draw[ultra thick, dashed, opacity=0.2] (1.5, -3.68)--(1.5, 4.75) node[midway, rotate=90, fill=white,opacity=1] (ncm) {Pre $\neq$ Post};
-
-        % \draw[ultra thick, dashed] (-2, -3.68)--(-2, 4.75) node[midway, rotate=90, fill=white, opacity=1] (cm) {Pre = Post};
-
-        % \draw[->] (ncm.north)--(cm.south) node[midway, above=0.1cm, align=center, fill=white] {\tiny{Change truncation}\\\tiny{until pre = post}\\\tiny{to estimate pre-mortality}\\\tiny{distribution}};
-
-        % \draw[->] (cm.north)--(-3, 0.55);
-
-        % %\draw[<-, ultra thick, opacity=0.5] (-4, 3)--(-3.5, 3);
-        % %\draw[->, ultra thick] (-3.5, 0)--(-3, 0);
-
-        % %\node at (-3, 0) {Pre};
-
     \end{tikzpicture}

### 2015-06-09 code/pihm_methods.py

@@ -5,8 +5,6 @@ import macroeco.compare as comp
 import scipy.optimize as opt
 import pandas as pd
 import statsmodels.api as sm
-import pymc as pm
-import matplotlib.pyplot as plt
 from scipy.stats import chisqprob
 
 """

### 2015-06-12 docs/figures_tables/figure2.tex
Moved figures out of main text and into separate file for journal formatting

--- 2015-06-09 docs/parasite_mortality_manuscript.tex
@@
 
-% Format section and subsection headers
-\makeatletter
-\renewcommand{\section}{\@startsection
-{section}%                   % the name
-{1}%                         % the level
-{0mm}%                       % the indent
-{-\baselineskip}%            % the before skip
-{0.5\baselineskip}%          % the after skip
-{\normalfont\bf\large}} % the style
-
-\renewcommand{\subsection}{\@startsection
-{subsection}%                   % the name
-{2}%                         % the level
-{0mm}%                       % the indent
-{-\baselineskip}%            % the before skip
-{0.5\baselineskip}%          % the after skip
-{\normalfont\bf}} % the style
-\makeatother
-
 % Other
@@
 
-% Bibliography
-\usepackage[numbers, compress]{natbib} % Bibliography - APA
-\bibpunct{(}{)}{;}{a}{}{,}
-
-% Format the Bibliography appropriately
-% increase \bibhang to take care of the numbers
-\setlength{\bibhang}{0pc}
-\makeatletter
-% patch \@lbibitem to print the current number before the authors
-\patchcmd{\@lbibitem}
-  {]}
-  {] \theNAT@ctr. \newline }
-  {}{}
-\makeatother
-
-%%%%%% FRONT MATTER %%%%%%%%%
-
-\title{Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations}
-\author{Mark Wilber, Sara Weinstein, and Cherie Briggs}
-
 \begin{document}
-
-% \begingroup
-\setlength{\parindent}{0in}
-\renewcommand{\arraystretch}{2}
-% \renewcommand{\tabcolsep}{0}
-%\renewcommand{\tabrowsep}{1}
-\begin{tabular}{p{3cm} p{10cm}}
-
-\large\textbf{} & \vspace\baselineskip \\
-
-\textbf{Title:} & Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations \\
-
-\textbf{Running title:} & Methods for quantifying parasite-induced host mortality  \\
-
-\textbf{Authors:} & Mark Q. Wilber, Sara B. Weinstein and Cheryl J. Briggs \\
-
-\textbf{Statement of Authorship:} & MW and SW designed research, performed
-analyses and wrote the manuscript, CB designed research and contributed to revisions. \\
-
-\textbf{Affiliations:} &
-1) Department of Ecology, Evolution and Marine Biology, University of California,
-Santa Barbara, Santa Barbara California, United States of America \\
-
-\textbf{Email addresses:} &
-Mark Wilber: mark.wilber@lifesci.ucsb.edu \newline
-Sara Weinstein: sara.weinstein@lifesci.ucsb.edu \newline
-Cheryl Briggs: cherie.briggs@lifesci.ucsb.edu \\
-
-\textbf{Corresponding \newline author:} &
-Mark Wilber \newline
-University of California, Santa Barbara \newline
-Department of Ecology, Evolution, and Marine Biology \newline
-2111 Noble Hall \newline
-Santa Barbara, CA 93106 \newline
-505-321-5381 \newline
-mark.wilber@lifesci.ucsb.edu \newline
-
-\end{tabular}
-
-\textbf{Note}: Supplementary data associated with this article
-
-% \renewcommand{\arraystretch}{1}
-% \begin{tabular}{l l}
-
-% \textbf{Words - Abstract:} & 293 \\
-% \textbf{Words - Main Text:} & 4988 \\
-% \textbf{References:} & 50 \\
-% \textbf{Figures:} & 5 \\
-% \textbf{Tables:} & 1
-
-% \end{tabular}
-
-\newpage
-
-\abstract{
-
-Parasites can significantly impact animal populations by changing host
-behavior, reproduction and survival. Detecting and quantifying these impacts is
-critical for understanding disease dynamics and managing wild animal
-populations. However, for wild hosts infected with macroparasites, it is
-notoriously difficult to quantify the fatal parasite load and number of animals
-that have died due to disease. When ethical or logistical constraints prohibit
-experimental determination of these values, examination of parasite intensity
-and distribution data may offer an alternative solution. In this study we
-introduce a novel method for using intensity data to detect and quantify
-parasite-induced mortality in wildlife populations. We use simulations to show
-that this method is more reliable than previously proposed methods while
-providing quantitative estimates of parasite-induced mortality from empirical
-data that are consistent with previously published qualitative estimates.
-However, we stress that this method, and all techniques that estimate parasite-induced mortality from intensity data alone, have several critical assumptions
-that limit their applicability in the real world. Due to these limitations,
-these methods should only be used as an exploratory tool to inform more
-rigorous studies of parasite-induced host mortality.
-
-\textbf{Keywords}: parasite aggregation, negative binomial distribution, Crofton Method, host survival function, lethal dose
-
-}
-
-\doublespacing
-
-\linenumbers
-\section{Introduction}
-
-Infectious agents can impact animal populations by changing
-population dynamics and stability \citep{Dobson1992,Tompkins2002}, altering predator-prey interactions \citep{Joly2004}, and
-even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
-of these infectious agents in wildlife is critical to understanding what
-regulates host and parasite populations, making predictions about disease
-transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies \citep{Coyne1989}, bovine tuberculosis \citep{Cox2005}, and
-rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, such that host survival is not a function of the number of infectious agents present within the host.
-In contrast, for macroparasites, pathology increases with parasite burden and host survival probability must be treated as a function of infection intensity.
-Helminths exhibiting this intensity-dependent pathology have significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013}, and wildlife survival \citep{Kirk2003,Logiudice2003}.
-While it is generally assumed that some fraction of wild host populations succumb to parasitic infection, it is notoriously difficult to actually quantify parasite-induced host mortality (PIHM) in wild animal populations because it is difficult to observe the dead or dying hosts most impacted by parasitism \citep{McCallum2000a}.
-
-Ideally, parasite-induced host mortality is
-quantified by experimentally infecting and tracking individual hosts in the
-wild population; however, for logistical and ethical reasons this method is
-rarely feasible \citep{McCallum2000a}. Snapshot data of parasite intensities across multiple hosts is much easier to collect and has
-often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and to quantify the
-relationship between infection intensity and host mortality \citep{Adjei1986}.
-
-\cite{Crofton1971a} first proposed that PIHM could be identified from parasite intensity data by comparing the
-observed parasite distribution in sampled hosts to the distribution
-predicted in the absence of parasite-induced mortality. This method
-assumes that, prior to host mortality, infection intensity in the host population follows a negative binomial distribution and the tail of the distribution is truncated as intensity dependent pathology removes the most heavily infected hosts.
-Assuming mortality occurs only in heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
-fitting a negative binomial distribution to hosts with lower and lower parasite intensities, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data (Figure \ref{fig:crofton}, see \emph{Supplementary Material (SI)} 1 for additional detail).
-
-While the Crofton Method detects the presence of PIHM, it makes no attempt to
-quantify the relationship between infection intensity and host survival
-probability; information that is necessary for estimating parasite impacts on host populations \citep{AndersonandMay1978,Tompkins2002}. \cite{Adjei1986}
-suggested that this relationship could be calculated by first using the Crofton
-Method to estimate the pre-mortality parasite distribution and then using this
-distribution to calculate the probability of host survival with increasing
-parasite intensity. To do this, \cite{Adjei1986} modeled host survival as a
-logistic function and then used a generalized linear model (GLM) to estimate
-the parameters of the host survival function (see \emph{SI} 2 for a technical
-description of the Adjei Method).  Although this method can predict the host survival function, it has several technical drawbacks. When mean infection intensity is high or sample sizes
-are small the observed intensity data must be subjectively binned into
-intensity ranges in order to fit the GLM framework. Furthermore, for the Adjei
-Method to work, any observed intensity values greater than predicted values
-must be modified and set equal to the predict values (see \emph{SI} 2 for
-details); a questionable act of data manipulation. These manipulations may introduce bias, reduce the precision and
-limit the power of this method to detect and quantify parasite-induced host
-mortality.
-
-After 30 years, and despite clear limitations \citep{McCallum2000a}, these
-methods (particularly the Crofton Method) are still discussed among
-parasitologists and are the primary techniques for examining population-level
-impacts of parasitism using parasite intensity data.
-In these methods, PIHM can only be identified by visually examining plots of the pre-mortality parameters predicted by the Crofton Method and determining whether they show a ``kink'' over a range of truncation values \citep[Figure \ref{fig:crofton}B;][]{Lester1984,Ferguson2011}.
-This qualitative criteria makes it difficult to compare PIHM between studies and a more rigorous and quantitative method is needed to both detect and quantify host mortality.
-The survival function given by the Adjei
-Method may be used to do this; however, it requires manipulating the
-original data and its accuracy remains untested.
-
-In this study, we propose a novel method for detecting and quantifying PIHM that ameliorates many of the aforementioned deficiencies of the previous methods.
-Our method does not require data alteration, is highly generalizable, and uses standard statistical techniques to quantitatively determine whether PIHM is occurring in a system.
-We
-use simulations to compare our method with the Adjei
-Method to test the ability of both to (1) detect occurrence of PIHM and (2) estimate the host survival function.  We then
-apply both methods to real datasets previously used in PIHM analyses and
-compare the results. Finally, we discuss the limitations of inferring PIHM
-from intensity data and how these methods fit in modern quantitative parasitology.
-
-\section{Methods}
-
-\subsection{A novel, likelihood-based method for estimating PIHM}
-
-Our method (henceforth the Likelihood Method) begins with the same assumptions as the Adjei Method: namely that infection has occurred and hosts with fatal parasite loads have died prior to the population sampling.
-As discussed by
-\citeauthor{Adjei1986}, this is not necessarily an unrealistic as some parasite infections occur primarily in younger hosts with parasite-induced mortality occurring soon after infection \citep[e.g.][]{Schotthoefer2003,Johnson2008}.
-
-The Likelihood Method then assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \boldsymbol{\phi})$, which specifies the probability of a host having $x$ parasites when it is observed.
-$\boldsymbol{\phi}$ is a vector of parameters that describes the shape of this distribution.
-The probability of a host surviving with $x$ parasites from infection until sampling is given by the host survival function $h(\text{survival} ; x, \boldsymbol{\theta})$ where $\boldsymbol{\theta}$ specifies any additional parameters needed to define the host survival function.
-
-With these two assumptions, we can define a distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival, $P(x | \text{survival})$.
-Using standard rules of conditional probability this distribution can be written as
-
-\begin{equation}
-    P(x | \text{survival}) = \dfrac{P(\text{survival} | x) * P(x)}{P(\text{survival})}
-    \label{eq:concept}
-\end{equation}
-
-$P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \boldsymbol{\theta})$, $P(x)$ is the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})$.
-Therefore, equation \ref{eq:concept} can be written as
-
-\begin{equation}
-    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}{\sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}
-    \label{eq:dist}
-\end{equation}
-
-Using this probability distribution, one can then find the parameters $\boldsymbol{\theta}$
-and $\boldsymbol{\phi}$ that maximize the likelihood of an observed host-parasite dataset.
-To estimate the significance of PIHM in a host-parasite system, a likelihood
-ratio test can be used in which the full model is given by equation
-\ref{eq:dist} and the reduced model is given by the pre-mortality distribution
-$g(x; \boldsymbol{\phi})$.  If PIHM is not significant in the system, the resulting
-likelihood ratio statistic should approximately follow a $\chi^2$ distribution
-with degrees of freedom equal to the number of parameters in the full model with parasite-induced mortality minus the number of parameters in the reduced model without parasite-induced mortality \citep{Bolker2008}.
-
-The parameterization of equation \ref{eq:dist} depends on the parasite system of interest.
-Here, we assume that the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ follows a negative binomial
-distribution with two parameters mean parasite intensity ($\mu_p$) and
-aggregation ($k_p$, where smaller $k_p$ indicates a more aggregated parasite population) before mortality \citep{Crofton1971a,AndersonandMay1978,Adjei1986}.
-A variety of different biological and statistical assumptions can result in an equilibrium parasite distribution that follows a negative binomial distribution
-\citep{Kendall1948a, Boswell1970, Calabrese2011}.
-Furthermore, the negative binomial distribution is an incredibly flexible distribution that fits many host-parasite systems even when the underlying mechanisms determining the empirical distribution are unknown \citep{Shaw1998}.
-
-The function for $h(\text{survival}; x,
-\boldsymbol{\theta})$ is also system specific.  Many
-theoretical models of parasite-induced host mortality assume that the parasite-induced death rate of hosts is a linear function of parasite intensity
-\citep{AndersonandMay1978,Dobson1992,Barbour2000}.
-In systems where there is truly a linear relationship between infection intensity and survival probability it will be nearly impossible to use intensity data to detect parasite-induced host mortality \citep{Lanciani1989}.
-However, some systems do exhibit non-linear host survival functions \citep{Benesh2011}, in which case these methods would be applicable.
-
-To compare the Likelihood Method and the previously proposed Adjei Method, we adopt the non-linear, logistic host-survival function used in the earlier study given by
-
-\begin{equation}
-    h(\text{survival}; x, a, b) = \dfrac{\exp{(a - b \log(x))}}{1 + \exp{(a - b \log(x))}}
-    \label{eq:logistic}
-\end{equation}
-
-Generally, a larger $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases, with the maximum rate of decline having a value of $b / 4$ (\emph{SI} 2).
-$b$ is in many ways analogous to the the pathogenicity parameter ($\alpha$) in classic macroparasite models that gives the parasite intensity dependent host death rate.
-\citep{AndersonandMay1978,Isham1995}.
-When $b$ is held constant, a larger $a$ allows for hosts to tolerate larger parasite intensities before experiencing parasite-induced mortality.
-More specifically, for every one unit increase in $a$ the log parasite intensity at which any percent of hosts survive (e.g.
-99\% of hosts survive) increases by $ 1 / b$ (\emph{SI} 2).
-
-The equation $\exp(a
-/ b)$ can also be used to calculate the parasite $LD_{50}$, here defined as the
-infection intensity above which a host has greater than 50\% probability of dying.
-Equation \ref{eq:logistic} is commonly used in toxicology and has the useful properties of being bounded between 0 and 1 and being differentiable for all $x$ \citep{Collet2002}.
-That being said, it is phenomenological and is used simply because it tends to fit survival data.
-However, given that a goal of these analyses is to
-compare the Likelihood Method's results to the Adjei Method, it is natural
-to adopt the same host-survival function to facilitate comparison.  When
-applying the Likelihood Method to other systems more mechanistic host-survival functions can be used in place of equation \ref{eq:logistic}.
-
-\subsection{Evaluating the Adjei and Likelihood Methods}
-
-\emph{Question 1: Can we detect PIHM?}
-
-We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters.
-Consistent with the model assumption that parasite infection, host mortality, and population sampling are temporally separate events, we first created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
-negative binomial distribution with parameters $\mu_p$ and $k_p$.
-This represents a host population that has become infected but not yet experienced parasite-induced mortality \citep{Adjei1986}.
-In the Adjei Method and Crofton Method, $N_p$ is a necessary parameter defined as the number of hosts in the population before parasite-induced mortality.
-More accurately, $N_p$  is the number of hosts that would have been sampled had parasite-induced host mortality not occurred.
-This parameter is not necessary when using the Likelihood Method because, unlike the Adjei Method and Crofton Method which estimate parasite-induced mortality using absolute numbers of hosts, the Likelihood Method estimates parasite-induced mortality using probabilities.
-However, to compare the results of the Likelihood Method with the Adjei Method, we specified a value for $N_p$ for all simulations.
-
-We next chose values of $a$ and $b$ for the host survival function and calculated the probability of survival
-for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, to simulate the period in which hosts died due to infection, for each host we drew a random number from a uniform distribution
-between 0 and 1 and if the calculated host survival probability was less than this random
-number, the host experienced parasite-induced mortality.  The surviving individuals represent the post-mortality hosts that would be sampled in the field.
-
-We then used these simulated pre-mortality and post-mortality datasets to test the
-ability of both methods to correctly determine whether or not PIHM was
-occurring when the parameters $N_p$, $\mu_p$ and $k_p$ were known.  Although
-the parameters $N_p$, $\mu_p$, and $k_p$ are always unknown in real systems, a
-method that fails under these ideal simulation conditions with known parameters will certainly also fail when these values must be estimated from empirical data.
-In practice, for the Adjei Method, $N_p$, $\mu_p$,
-and $k_p$ are estimated using the Crofton Method \citep{Adjei1986}, while $\mu_p$ and $k_p$ in
-the Likelihood Method can be estimated jointly with $a$ and $b$ or via the
-Crofton Method.
-
-We compared the two methods using three different mean parasite intensity values ($\mu_p$ = 10, 50, 100) and three different host survival functions (gradual, moderate, and steep decreases in the host survival with increasing parasite intensity, Figure \ref{fig:question1}A).
-For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.
-We examined each $\mu_p$-survival function pair at  three levels of parasite
-aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these 27 parameter
-combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).
-For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.
-We also examined the impact of sample size by simulating each parameter for pre-mortality sample sizes of $N_p$ = [50, 100, 200, 300, 400, 500].
-Wild host populations were assumed to be sampled after PIHM has occurred, thus we calculated the sample size in the power simulations as the average number of surviving hosts over all 150 simulations for each parameter combination.
-
-We then tested the ability of the Likelihood Method to correctly identify PIHM
-under the more realistic condition of unknown pre-mortality parameters. Based
-on the first set of simulations, we excluded the Adjei Method and only examined
-the power of the Likelihood Method under ``best-case'' scenario parameter
-values, setting $\mu_p = 10$ and $k = 1$ because PIHM is most detectable when
-parasites are less clumped and mean intensity is low. We examined the impact of survival function shape and
-sample size on the Likelihood Method's ability to identify PIHM when the pre-mortality parameters $\mu_p$ and $k_p$ and the survival function parameters $a$
-and $b$ needed to be estimated.  We performed 500 simulations over a range of
-different samples sizes for gradual, moderate, and steep survival functions, following the simulation procedure described above. \\
-
-\noindent
-\emph{Question 2: Can we estimate properties of the host survival function?}
-
-In the previous section we compared the ability of the Adjei Method and the Likelihood Method to provide a ``yes'' or ``no'' answer for whether or not PIHM was occurring in a system.
-In this section we compared the ability of the Adjei Method and the Likelihood Method to estimate properties of the survival function such as the parameters $a$, $b$ and $LD_{50}$.
-Using the same simulation procedure and parameter combinations described above, we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
-precision for these estimates \citep{Walther2005}.
-Because estimating properties of the host survival function requires more information than simply detecting PIHM, we used larger values of $N_p$ for this simulation ($N_p$ = [300, 500, 1000, 2000, 5000, 7500,
-10000]).
-We used the average number of surviving hosts for each set of 150 simulated datasets as our measure of sample size.
-Although both $a$ and $b$ are necessary to estimate $LD_{50}$, the two parameters showed similar patterns of bias and precision so we only show the results for $a$.
-
-\subsection{Application to real data}
-
-We tested the ability of the Adjei Method and the Likelihood Method to identify
-PIHM in six host-parasite datasets given in \cite{Crofton1971a} and four datasets
-given in \cite{Adjei1986} (Table \ref{table:pihm}). \citeauthor{Crofton1971a} analyzed infection patterns in the snail \emph{Gammarus pulex} infected with the
-acanthocephalan \emph{Polmorphus minutus}. \citeauthor{Adjei1986} analyzed males and females of two species of lizard fish \emph{Saurida tumbil} and
-\emph{S. undosquamis} that were infected by the cestode
-\emph{Callitetrarhynchus gracilis}.
-
-In both earlier studies, the authors reported PIHM in some of the datasets and we tested whether the Adjei
-Method and/or the Likelihood Method also predicted PIHM. For the six datasets from
-\cite{Crofton1971a}, we used the general conclusions of the author and truncated the data at four parasites, applied the Crofton
-Method to estimate the pre-mortality distribution, and then ran the Likelihood
-Method and Adjei Method using these pre-mortality parameters.  For the
-\cite{Adjei1986} datasets, we followed the same procedure as the authors and
-first truncated the data at two parasites and then fit the Crofton Method for the
-female fish of both species.  Then, following the \citeauthor{Adjei1986}'s methods, we parameterized the male pre-mortality
-distributions for each species with the results from the females.  Finally, we
-applied the Adjei Method and the Likelihood Method to determine whether or not
-PIHM was significant for these species and compared our results to those given by the authors.  All code for the analyses is provided in \emph{SI} 4.
-
-\section{Results}
-
-\subsection{Question 1: Detecting presence of PIHM}
-
-The power of the Adjei Method to detect PIHM in a
-system was close to unity for larger sample sizes and tended to
-decrease as sample size decreased for all survival functions (Figure \ref{fig:question1}C; \emph{SI} 3 Figs 1-3).  The Likelihood Method had a power close to
-unity for all parameter combinations and sample sizes considered.  With gradual
-survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (Fig. \ref{fig:question1}C, \emph{SI} 3 Figs 1-3).
-
-The Adjei Method had highly inflated Type I error rates (i.e. falsely detected
-PIHM) for all parameter combinations that we
-considered (Fig. \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).  This method also showed the unintuitive pattern of decreasing Type I error
-rate with decreasing sample size.
-This occurred because, at small samples sizes, intensity data must be binned before the Adjei Method can be used (\emph{SI} 2).
-In contrast, the Likelihood Method showed a Type I
-error rate at or near the pre-set level of 0.05 for all parameter combinations
-and sample sizes considered (Fig. \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).
-
-When all parameters were jointly estimated, the Likelihood Method showed highly
-context-dependent results even when detecting PIHM under the best-case scenario of
-$\mu_p = 10$ and $k_p = 1$.
-For steep survival curves, PIHM could be detected with a power of greater than 0.8 from a sample of less than 100 hosts (Fig \ref{fig:real_power}).
-However, for moderate survival functions over 400 hosts had to be sampled to achieve the same power and for gradual survival functions, no tested sample size ever achieved a power greater than 0.8 (Fig \ref{fig:real_power}).
-
-\subsection{Question 2: Estimating the $LD_{50}$ and survival function}
-
-The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$
-for all combinations of parameters examined in this study (Fig. \ref{fig:question2}, \emph{SI} 3 Figs 4-6).  Even for
-the smallest sample sizes we considered, the Likelihood Method's estimate of $LD_{50}$
-was largely unbiased, with small biases occurring for gradual host survival functions. The precision of the Likelihood Method's $LD_{50}$ estimates decreased
-(increasing coefficient of variation) as sample size decreased for all
-parameter combinations we examined (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6).
-
-The Adjei Method produced biased estimates of the $LD_{50}$ across nearly all parameter combinations, tending to underestimate the true value of the parameter (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6).
-For $\mu_p = 10$, the $LD_{50}$
-estimates from the Adjei Method were largely unbiased for large samples sizes, but as $\mu_p$ increased, the Adjei Method
-produced biased estimates of $LD_{50}$ across all sample sizes, with bias
-increasing as sample size decreased (Figure \ref{fig:question2}, \emph{SI} 3 Fig 4-6). The $LD_{50}$ estimates from the Adjei
-Method also showed large decreases in precision with the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:question2}, \emph{SI} 3 Fig 4-6).
-
-In terms of the host survival function, the Likelihood Method gave unbiased estimates of survival function parameter $a$ when sample sizes were large, however as sample size decreased these estimates became severely biased (Fig.
-\ref{fig:question2}, \emph{SI} Fig 7 - 9) The Adjei Method produced
-biased estimates of the host survival function across all sample sizes, with consistently greater bias for steeper survival functions and higher mean parasite loads.
-(Fig \ref{fig:question2}, \emph{SI} 3 Figs 7-9).
-
-\subsection{Application to real data}
-
-The previous authors qualitatively detected PIHM
-in 7 of the 10 datasets considered (Table \ref{table:pihm}).  The Likelihood Method parameterized
-from the pre-mortality parameters of the Crofton Method detected significant
-PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
-dataset in which the Likelihood Method did not detect a significant effect of PIHM was the Adjei dataset
-for female \emph{S. tumbil}.  For this dataset there was a marginally significant effect
-of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$).
-The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results showing that the Adjei Method has a high Type I error rate.
-
-\section{Discussion}
-
-Our likelihood-based method to estimate parasite-induced host mortality from
-observed parasite intensity data is a significant improvement over the previous
-methods.  In simulations, it had greater power for detecting PIHM over a wider
-range of parameter values and also exhibited fewer false detection events (Type
-I errors) in both simulations and when applied to published datasets previously
-used in PIHM analyses. The Likelihood Method was also generally less biased and
-more precise when quantifying parasite-induced mortality via the host survival
-function for the parameters we considered.  The superior performance of the
-Likelihood Method over the Adjei Method can be attributed to its fewer
-parameters, its lack of unnecessary data alteration, and its applicability
-across a variety of different parameter combinations. In short, the Likelihood
-Method is a better method for detecting and quantifying PIHM than the
-previously proposed Adjei Method.
-
-Although superior to the Adjei Method, the Likelihood Method still cannot be applied to all real datasets.
-For host-parasite systems where host mortality occurs as a steep, non-linear function of parasite intensity only 80 hosts must be sampled to have an 80\% power in detecting PIHM.
-However, as the maximum slope of the survival function decreases and the function becomes somewhat linear, hundreds, or possibly thousands of hosts would have to be sampled to achieve the same result.
-This is consistent with previous studies which illustrate the difficulty of detecting PIHM from linear host survival functions \citep{Lanciani1989}.
-While it may be feasible to sample several hundred invertebrates or small fish, even the smallest sample sizes are completely
-unfeasible for many vertebrates, particularly the species of conservation
-concern where addressing the impact of parasitism would be most important. An
-even larger sample size would be required to identify PIHM when parasites are highly aggregated, mean infection intensity is
-high, or parasite prevalence is low, all of which are
-common in many parasitic helminths. Moreover, while linear functions make PIHM undetectable, at the other extreme, steep,
-non-linear survival curves produce severely biased estimates of the survival
-function. Give the interaction between all of these different factors, the
-Likelihood Method is probably limited to detecting PIHM in systems where greater than 100 hosts can be collected, parasites are
-common and only moderately aggregated, and substantial host mortality occurs at relatively low parasite intensity.
-
-While we have improved on the existing methods for quantifying
-PIHM from parasite intensity data, all such methods require several
-fundamental, and potentially problematic assumptions.
-Nearly all current methods derive from \cite{Crofton1971a} \citep[but see][]{Ferguson2011} and assume that, prior to any PIHM, parasites are distributed in the host population following a
-negative binomial distribution. But, it is fundamentally impossible to know
-what the pre-mortality parasite distribution was in a wild host population and
-it is widely recognized that different processes can lead to a variety of
-parasite distributions in hosts \citep{Anderson1982a, Duerr2003}. However, the negative binomial is extremely
-flexible and there is substantial empirical and theoretical evidence to support
-the assumption that, prior to any PIHM, parasite distributions can be fit by a negative binomial distribution \citep{Shaw1995,Shaw1998,Wilson2002}.
-
-Unfortunately, this
-flexibility in the distribution may also reduce our ability to detect PIHM. If
-a negative binomial can be fit to the observed post-mortality parasite
-distribution then, regardless of how lethal the parasite was, it will be
-impossible to detect PIHM because there is no need for a more complex model.
-Most observed parasite distributions are well fit by the negative binomial distribution \citep{Shaw1998}, suggesting
-that systems where these methods are applicable may be more the exception than
-the rule.  Furthermore, even when truncation of the negative binomial distribution is detected, it may be caused by other
-processes such as within host density dependence, age dependent variation in host
-resistance and/or heterogeneous infection rates \citep{Anderson1982a,Rousset1996, McCallum2000a}.  This means that in the event
-that PIHM is detected, it may actually not be the result of PIHM.
-
-Given these numerous caveats, is there a place in parasitology for methods that
-estimate PIHM from intensity data?  As suggested by
-\cite{Lester1984} these methods for estimating PIHM can
-provide preliminary insight into whether or not PIHM is worth further
-exploration.  However, we stress that these methods should only be used as an
-exploratory tool when assessing the role of PIHM in a system, and potential
-users should critically evaluate whether they think they have a large enough
-sample size and an appropriate host survival function/post-mortality distribution for the methods developed
-in this paper to be applicable.  Even if they are applicable, inferring PIHM
-from distributional data is no substitute for field experiments
-and an in depth understanding of the natural history of the host-parasite
-system under consideration.
-
-\section*{Acknowledgments}
-
-We thank the Briggs Lab group and Theoretical Ecology group at UCSB for helpful feedback.
-This work was supported by National Science Foundation Graduate Research Fellowship (Grant No.
-DGE 1144085) and X.
-
-% We thank Hamish McCallum for discussions on the inadequacy of estimating PIHM from intensity data and Kevin Lafferty for manuscript comments.
-
-\singlespacing
-\bibliographystyle{int_j_para.bst}
-\bibliography{parasite_host_mortality}
-
-\newpage
-
-\renewcommand{\arraystretch}{1.2}
-
-\begin{table}
-    \centering
-    \caption{Definition of parameters and functions used in the main text}
-    \begin{tabular}{c p{12cm}}
-    \hline
-    Parameter & Definition \\
-    \hline\hline
-    $\mu_p$ & Pre-mortality mean parasite intensity \\
-    $k_p$   & Pre-mortality parasite aggregation \\
-    $N_p$   & Pre-mortality host sample size \\
-    $x$     & Number of parasites in a given host \\
-    $g(x; \mu_p, k_p)$ & Pre-mortality negative binomial parasite distribution \\
-
-    $h(\text{survival}; x, a, b)$ & The probability of host survival given a parasite load $x$ and logistic parameters $a$ and $b$\\
-
-    $b / 4$ & The maximum rate of decline in host survival probability with increasing parasite intensity  \\
-    $a$ & When $b$ is held constant a one unit increase in $a$ leads to a $1 / b$ increase in the log parasite intensity at which some percent of hosts survive\\
-    $LD_{50}$ & $\exp(a / b)$, parasite intensity above which a host has a greater than 50\% chance of dying \\
-
-    \end{tabular}
-    \label{tab:params}
-\end{table}
-
-\renewcommand{\arraystretch}{1.2}
-\begin{sidewaystable}
-
-    \caption{Comparison of the PIHM predictions of previously used host-parasite datasets to those given by the Adjei Method and the Likelihood Method.
-The first column specifies the identity of the dataset, the second column specifies whether or not the authors indicated that PIHM was occurring in the system based on a qualitative assessment, the third column indicates whether or not the Likelihood Method with pre-mortality parameters estimated from the Crofton Method detects significant PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
-If a method detected significant PIHM the predicted $LD_{50}$ is given in parentheses.}
-
-    \centering
-    \begin{tabular}{l  p{3cm} p{3cm} l}
-
-    \hline\hline
-    Data Set (sample size) & Author detected PIHM? & Likelihood Method?  & Adjei Method? \\
-
-    \hline
-    Crofton, Station 1 ($n=538$) & Yes & Yes (7.27) & Yes (9.33) \\
-    Crofton, Station 2 ($n=507$) & Yes & Yes (6.92) &  Yes (14.95)\\
-    Crofton, Station 3 ($n=633$) & Yes & Yes (5.93) &  Yes (5.98) \\
-    Crofton, Station 4 ($n=486$) & No & No &  Yes (7.99) \\
-    Crofton, Station 5 ($n=276$) & No & No & Yes (10.58) \\
-    Crofton, Station 6 ($n=191$) & No & No & No \\
-    Adjei, \emph{S. tumbil} female ($n=446$) & Yes (5.7) & No & Yes (6.37) \\
-    Adjei, \emph{S. tumbil} male ($n=452$) & Yes (3.4) & Yes (3.42) & Yes (3.66)  \\
-    Adjei, \emph{S. undosquamis} female ($n=2573$) & Yes (3.2) & Yes (3.04) & Yes (3.11) \\
-    Adjei, \emph{S. undosquamis} male ($n=2440$) & Yes (1.8) & Yes (1.83) & Yes (1.78) \\
-
-    \end{tabular}
-    \label{table:pihm}
-
-\end{sidewaystable}
-
-\newpage
-
-\begin{figure}
-
-    \begin{tikzpicture}
-        \node at (0, 0) (hist) {\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/concept_crofton_method}};
-
-    \end{tikzpicture}
-
-    \caption{A schematic representation of the iterative approach of the Crofton Method.
-\textbf{A)} The light gray shows the pre-mortality distribution that the Crofton Method is trying to estimate from the dark grey post-mortality distribution.
-The Crofton Method proceeds by truncating the post-mortality data at different levels ($t_i$, e.g.
-$i= 0, \dots, 5$) and finding the pre-mortality host population size ($N_p$), pre-mortality mean parasite intensity ($\mu_p$), and pre-mortality parasite aggregation ($k_p$) that best fit the truncated data.
-\textbf{B)} The parameter $N_p$ is then plotted against the truncation level $t_i$ to determine if a ``kink'' occurs in the parameter values  \citep{Lester1984}.
-This ``kink'' indicates that PIHM is occurring in the system.
-In the above example, PIHM is occurring in the system as visualized by the distinct ``kink'' at $t_3$.}
-    \label{fig:crofton}
-
-\end{figure}
-
-\newpage
 
@@
 
-\caption{A) Five potential shapes for a host-survival functions.
-In the simulations we used a gradual survival function (dotted line), and moderate survival function (dashed line), and a steep survival function (solid line).
-The linear and immediate survival functions represent two potential extremes that we do not include in the simulations.
-For each of these survival functions and the parameter combinations described in the main text, we tested the Type I error and power of the Likelihood (Like.) Method and Adjei Method.
-B) Gives the Type I error of each method over a range of pre-mortality sample sizes with a pre-mortality mean parasite intensity ($\mu_p$) of 50 and pre-mortality parasite aggregation ($k_p$) at 0.5.
-The red line shows the pre-set significance level of 0.05.
-C) Gives the power of each method for detecting PIHM over a range of post-mortality sample sizes for $\mu_p = 50$ and $k_p = 0.5$.
-In general, the Likelihood Method has higher power and lower Type I error than the Adjei Method.
-See the \emph{SI} 3 Fig 1 - 3 for Type I Error and power results for all parameter combinations.}
-
-\label{fig:question1}
-
 \end{figure}
 
-\begin{figure}
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure2_for_manuscript.pdf}
+\centering
 
-    \caption{Bias and precision (coefficient of variation) for the Likelihood Method (Like.) and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
-As the coefficient of variation increases, precision decreases.
-The pre-mortality parameters for this simulation were $\mu_p = 50$ and $k_p = 0.5$.
-The figure shows the simulations for three different host survival functions (gradual, moderate, and steep), each with the same $LD_{50}$.
-Bias and precision results of $LD_{50}$ and $a$ for all other parameter combinations can be found in \emph{SI} 3 Fig 4 - 9.}
-
-    \label{fig:question2}
-
-\end{figure}
-
-\begin{figure}
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure3_for_manuscript.pdf}
-
-    \caption{The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
-The curves were generated from 500 simulations for 10 pre-mortality sample sizes, $N_p$.}
-
-    \label{fig:real_power}
-
-\end{figure}
+\textbf{Figure 1}
 

### 2015-06-12 docs/parasite_mortality_manuscript.tex
Moved figures out of main text and into separate file for journal formatting

@@
 % increase \bibhang to take care of the numbers
-\setlength{\bibhang}{0pc}
-\makeatletter
-% patch \@lbibitem to print the current number before the authors
-\patchcmd{\@lbibitem}
-  {]}
-  {] \theNAT@ctr. \newline }
-  {}{}
-\makeatother
+% \setlength{\bibhang}{0pc}
+% \makeatletter
+% % patch \@lbibitem to print the current number before the authors
+% \patchcmd{\@lbibitem}
+%   {]}
+%   {] \theNAT@ctr. \newline }
+%   {}{}
+% \makeatother
 
 %%%%%% FRONT MATTER %%%%%%%%%
-
-\title{Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations}
-\author{Mark Wilber, Sara Weinstein, and Cherie Briggs}
 
@@
 
-\section{Methods}
+\section{Materials and methods}
 
@@
 More specifically, for every one unit increase in $a$ the log parasite intensity at which any percent of hosts survive (e.g.
-99\% of hosts survive) increases by $ 1 / b$ (\emph{SI} 2).
+99\% of hosts survive) increases by $1 / b$ (\emph{SI} 2).
 
@@
 We thank the Briggs Lab group and Theoretical Ecology group at UCSB for helpful feedback.
-This work was supported by National Science Foundation Graduate Research Fellowship (Grant No.
-DGE 1144085) and X.
+MW was supported by National Science Foundation Graduate Research Fellowship (Grant No.
+DGE 1144085) and the UC Regents CB was supported by X.
 
@@
 
-\renewcommand{\arraystretch}{1.2}
+% \renewcommand{\arraystretch}{1.2}
 
-\begin{table}
-    \centering
-    \caption{Definition of parameters and functions used in the main text}
-    \begin{tabular}{c p{12cm}}
-    \hline
-    Parameter & Definition \\
-    \hline\hline
-    $\mu_p$ & Pre-mortality mean parasite intensity \\
-    $k_p$   & Pre-mortality parasite aggregation \\
-    $N_p$   & Pre-mortality host sample size \\
-    $x$     & Number of parasites in a given host \\
-    $g(x; \mu_p, k_p)$ & Pre-mortality negative binomial parasite distribution \\
+% \begin{table}
+%     \centering
+%     \caption{Definition of parameters and functions used in the main text}
+%     \begin{tabular}{c p{12cm}}
+%     \hline
+%     Parameter & Definition \\
+%     \hline\hline
+%     $\mu_p$ & Pre-mortality mean parasite intensity \\
+%     $k_p$   & Pre-mortality parasite aggregation \\
+%     $N_p$   & Pre-mortality host sample size \\
+%     $x$     & Number of parasites in a given host \\
+%     $g(x; \mu_p, k_p)$ & Pre-mortality negative binomial parasite distribution \\
 
-    $h(\text{survival}; x, a, b)$ & The probability of host survival given a parasite load $x$ and logistic parameters $a$ and $b$\\
+%     $h(\text{survival}; x, a, b)$ & The probability of host survival given a parasite load $x$ and logistic parameters $a$ and $b$\\
 
-    $b / 4$ & The maximum rate of decline in host survival probability with increasing parasite intensity  \\
-    $a$ & When $b$ is held constant a one unit increase in $a$ leads to a $1 / b$ increase in the log parasite intensity at which some percent of hosts survive\\
-    $LD_{50}$ & $\exp(a / b)$, parasite intensity above which a host has a greater than 50\% chance of dying \\
+%     $b / 4$ & The maximum rate of decline in host survival probability with increasing parasite intensity  \\
+%     $a$ & When $b$ is held constant a one unit increase in $a$ leads to a $1 / b$ increase in the log parasite intensity at which some percent of hosts survive\\
+%     $LD_{50}$ & $\exp(a / b)$, parasite intensity above which a host has a greater than 50\% chance of dying \\
 
-    \end{tabular}
-    \label{tab:params}
-\end{table}
+%     \end{tabular}
+%     \label{tab:params}
+% \end{table}
 
@@
 
-\newpage
-
 \begin{figure}
 
-    \begin{tikzpicture}
-        \node at (0, 0) (hist) {\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/concept_crofton_method}};
-
-    \end{tikzpicture}
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/concept_crofton_method}
+    \captionsetup{justification=raggedright, singlelinecheck=false}
 
@@
 
-\newpage
-
 \begin{figure}
 
-\begin{tikzpicture}
-
-    \node at (0, 0) (pic) {\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure1_partII_for_manuscript50}};
-
-    \node[above=0.1cm] at (pic.north) (concept) {\includegraphics[width=0.6\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure1_partI_for_manuscript}};
-
-    \draw[->] (0, 4.2)--(-1.5, 3.2);
-    \draw[->] (0, 4.2)--(1.5, 3.2);
-
-    \node at (-3.2, 4.9) {A.};
-    \node at (-6.7, 2.9) {B.};
-    \node at (1.15, 2.9) {C.};
-
-\end{tikzpicture}
+    \captionsetup{justification=raggedright, singlelinecheck=false}
 
@@
 \begin{figure}
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure2_for_manuscript.pdf}
+
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure2_for_manuscript.pdf}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
 
@@
 \begin{figure}
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure3_for_manuscript.pdf}
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure3_for_manuscript.pdf}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
 

### 2015-06-19 docs/parasite_mortality_manuscript.tex
Remade figures and added final comments before submission

@@
 transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies \citep{Coyne1989}, bovine tuberculosis \citep{Cox2005}, and
-rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, such that host survival is not a function of the number of infectious agents present within the host.
-In contrast, for macroparasites, pathology increases with parasite burden and host survival probability must be treated as a function of infection intensity.
+rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, such that host survival is not typically considered to be a function of the number of infectious agents present within the host.
+In contrast, models of macroparasites generally assume that pathology increases with parasite burden and host survival probability must be treated as a function of infection intensity.
 Helminths exhibiting this intensity-dependent pathology have significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013}, and wildlife survival \citep{Kirk2003,Logiudice2003}.
@@
 
-The Likelihood Method then assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \boldsymbol{\phi})$, which specifies the probability of a host having $x$ parasites when it is observed.
+The Likelihood Method then assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \boldsymbol{\phi})$, which specifies the probability of a host having $x$ parasites before mortality occurs.
 $\boldsymbol{\phi}$ is a vector of parameters that describes the shape of this distribution.
@@
 Generally, a larger $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases, with the maximum rate of decline having a value of $b / 4$ (\emph{SI} 2).
-$b$ is in many ways analogous to the the pathogenicity parameter ($\alpha$) in classic macroparasite models that gives the parasite intensity dependent host death rate.
+$b$ is in many ways analogous to the pathogenicity parameter ($\alpha$) in classic macroparasite models that gives the parasite intensity dependent host death rate.
 \citep{AndersonandMay1978,Isham1995}.
@@
 MW was supported by National Science Foundation Graduate Research Fellowship (Grant No.
-DGE 1144085) and the UC Regents CB was supported by X.
+DGE 1144085) and the UC Regents CB was supported by NIH grant 1R01GM109499 from the Ecology of Infectious Disease program.
 

### 2015-08-04 docs/response_to_reviewers.md
Beginning to respond to reviewers

## Response to Reviewers

### Reviewer 1

We thank Reviewer 1 for the helpful suggestions. We agree with Reviewer 1 that host-parasite distributions are dynamic and that it would be interesting to apply these models to

### Reviewer 2

We thank Reviewer 2 for their helpful feedback and address each of their points in turn.

1. The major critique provided by Reviewer 2 was that we were too negative about our results in the Discussion and the Abstract. We agree with this comment and have reworded the Abstract and the Discussion to reflect why we think .  That being said, we do still feel it is important to highlight the limitations of

### 2015-08-04 docs/parasite_mortality_manuscript.tex
Beginning to respond to reviewers

@@
 data that are consistent with previously published qualitative estimates.
-However, we stress that this method, and all techniques that estimate parasite-induced mortality from intensity data alone, have several critical assumptions
-that limit their applicability in the real world. Due to these limitations,
-these methods should only be used as an exploratory tool to inform more
-rigorous studies of parasite-induced host mortality.
+However, this method, and all techniques that estimate parasite-induced
+mortality from intensity data alone, have several important assumptions that
+must be scrutinized before applying them to real-world data. Given that these
+assumptions are met, our method is new exploratory tool that can help inform
+more rigorous studies of parasite-induced host mortality.
 
@@
 Generally, a larger $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases, with the maximum rate of decline having a value of $b / 4$ (\emph{SI} 2).
-$b$ is in many ways analogous to the pathogenicity parameter ($\alpha$) in classic macroparasite models that gives the parasite intensity dependent host death rate.
-\citep{AndersonandMay1978,Isham1995}.
+$b$ is in many ways analogous to the pathogenicity parameter ($\alpha$) in classic macroparasite models that gives the parasite intensity dependent host death rate \citep{AndersonandMay1978,Isham1995}.
 When $b$ is held constant, a larger $a$ allows for hosts to tolerate larger parasite intensities before experiencing parasite-induced mortality.
@@
 
-We thank the Briggs Lab group and Theoretical Ecology group at UCSB for helpful feedback.
+We thank the Briggs Lab group, Kuris/Lafferty Lab group, Theoretical Ecology group at UCSB, and two anonymous reviewers for helpful feedback.
 MW was supported by National Science Foundation Graduate Research Fellowship (Grant No.
-DGE 1144085) and the UC Regents CB was supported by NIH grant 1R01GM109499 from the Ecology of Infectious Disease program.
+DGE 1144085) and the UC Regents.
+CB was supported by NIH grant 1R01GM109499 from the Ecology of Infectious Disease program.
 

### 2015-08-04 docs/supplementary_material.tex
Beginning to respond to reviewers

@@
 
-The second step of the Adjei Method is to make the assumption that infection, host mortality, and sampling occur in that order and are temporally \cite{Adjei1986}.
+The second step of the Adjei Method is to make the assumption that infection, host mortality, and sampling occur in that order and are temporally separate \citep{Adjei1986}.
 Next, \citeauthor{Adjei1986} assume that the host survival function follows the logistic form
@@
     \log{\frac{p}{1 - p}} &= a_1 - b \log{x_1}\\
-    - \log{\frac{p}{1 - p}} &= a_1 - b \log{x_2}\\
+    - \log{\frac{p}{1 - p}} &= a_2 - b \log{x_2}\\
     0 &= a_1 - a_2 - b\log{x_1} + b\log{x_2}\\

### 2015-08-08 docs/response_to_reviewers.md
Incorporated reviewer comments

@@
 
-We thank Reviewer 1 for the helpful suggestions. We agree with Reviewer 1 that host-parasite distributions are dynamic and that it would be interesting to apply these models to
+We thank Reviewer 1 for the helpful suggestions. We agree with Reviewer 1 that
+host-parasite distributions are dynamic and that it would be interesting to
+apply the methods developed in this paper to a stochastic host-parasite model. However, we believe that this approach is unnecessary/problematic to include in the current manuscript for a few reasons.
+
+First, stochastic host-parasite models can range from relatively simple models
+with an analytical solutions at any given point in time (e.g. a simple death-
+immigration process) to complex models with clumped infections, host
+heterogeneity, and acquired immunity that can only be explored via simulation.
+Given the space alloted in this manuscript and the time alloted for these
+revisions, in would be difficult to provide a comprehensive treatment of how our
+method works across the numerous different assumptions that have been included
+in stochastic host-parasite models. We believe that this exploration would be
+better suited for a follow-up study.
+
+However, we can gain some insight into whether our model might be successful in
+identifying PIHM in a simple birth-death-immigration process in which the number
+of parasites increase by one, decrease by one, or stay the same in a given time
+step.  The stationary distribution for a BDI process in which birth rate < death
+rate and there is no parasite-induced host mortality is a negative binomial
+distribution with $k =$ immigration rate / birth rate. This result is consistent
+with our assumption that the pre-mortality distribution of parasites across
+hosts is negative binomial. If we were to include parasite-induced host
+mortality into this stochastic process, we know from extensive literature that
+the variance to mean ratio would decrease and the parasite distribution with
+PIHM would be less aggregated than the distribution without PIHM.  If the
+immigration and birth rate (or self-reinfection rate) was known for the system,
+our model could be applied in which the null model was a negative binomial with
+a fixed $k = $ immigration / birth and the alternative was model was defined by
+equation X in the manuscript.
+If the immigration and birth rate was not known, our method could be applied as given in the manuscript.
+The key criteria necessary for our method to work would be that the resulting distribution post mortality took a form significantly different than a negative binomial distribution.
+We have included an additional Appendix (Appendix X) in which we simulate a birth-death-immigration process with host mortality and test the applicability of our method.
+We have also added the following the the Discussion to address Reviewer 1's comments.
 
@@
 1.
-The major critique provided by Reviewer 2 was that we were too negative about our results in the Discussion and the Abstract.
-We agree with this comment and have reworded the Abstract and the Discussion to reflect why we think .
-That being said, we do still feel it is important to highlight the limitations of
+Normally I would be asking authors to insert caveats into conclusions when they over spin their work, but in your case I do think you are way too negative about your own work when you talk about the caveats in the discussion and your conclusion / statements in the abstract.
+
+We thank Reviewer 2 for this comment and have made changes in the Abstract and Discussion to cast our method in a more positive light.
+In particular, we changed the the following sentences in the abstract from
+
+*However, we stress that this method, and all techniques that estimate parasite- induced mortality from intensity data alone, have several critical assumptions that limit their applicability in the real world.
+Due to these limitations, these methods should only be used as an exploratory tool to inform more rigorous studies of parasite-induced host mortality.*
+
+to
+
+**However, this method, and all techniques that estimate parasite-induced
+mortality from intensity data alone, have several important assumptions that
+must be scrutinized before applying them to real-world data. Given that these
+assumptions are met, our method is new exploratory tool that can help inform
+more rigorous studies of parasite-induced host mortality.**
+
+2.
+Pg 8-9 - to me it could be made clearer why we should care about the Type I error - it is just stated on Pg 9 with no justification.
+I am not saying the authors are incorrect.
+I would get this justification in at the start of the section - ie lines 150~ Pg 8.
+In addition, I would insert here the "by estimating power"after "pre-mortality parameters" (Line 154) and then expand a bit.
+Just needs for me a couple of sentences setting up Type 1 and power and why done for the reader.
+The more detailed description that follows then flows.
+
+To address these comments, we added the following justification for using Type I error and power at the beginning the subsection "Evaluating the Adjei and Lieklihood Methods"
+
+**We used statistical power and Type I error to test the ability of the Adjei and the Likelihood Methods to correctly identify the presence of PIHM on simulated data with known pre-mortality parameters.
+The power of a method is the probability of correctly detecting PIHM given that it is occurring and the Type I error is the probability of incorrectly identifying PIHM given that it is not occurring.
+If a method has low Type I error we can be confident that when we detect PIHM it is actually occurring.
+If one method has higher power for detecting PIHM than another, we will need to sample fewer hosts to detect PIHM.**
+
+3.
+Pg 8 Line 169-175 - this may be V2 of your work and not for this paper, but I wonder what would be the impact of non-random mortality.
+I am thinking here of say where mortality is a combination of burden and some host factor - eg sex.
+As said this maybe beyond the scope of the paper, but could be mentioned in the discussion if agree?
+
+This is a great point by Reviewer 2.
+
+4.
+lines 201-202 - I assume that any confidence intervals around the average no of surviving hosts are nice and symmetrical - ie that we are happy to just take the mean and do not need to worry about some horrible non linear type distribution of surviving hosts.
+
+Reviewer 2 is correct in assuming that the distribution of surviving hosts was generally quite symmetrical and the standard deviation was small relative to the mean.
+We made this clear in the manuscript by adding the following line
+
+**The distribution of surviving hosts over the 150 simulations was generally symmetrical and the standard deviation was small compared to the mean (maximum coefficient of variation was approximately 0.06 across all parameter combinations), suggesting that the mean number of surviving hosts was an adequate summary statistic of the number of hosts sampled post-mortality.**
+
+5.
+Pg 13 - lines 295-302 - is it of interest that the variation your ML method observed in the LD50 seemed a lot less than the Adjei method?
+Given this is the same host, same parasite and similar habitats I would be a little concerned at the marked variation in LD50 in the Adjei method
+
+6. Pg 13-14 - lines 316-> - see comments above about being more positive about your work; Pg 15 - I think the conclusions need to be more positive
+
+We addressed these concerns in above section.
+
+7.
+Figure 2, 3, supplementary figures - distinction between ML and Adeji not always very clear.
+Would suggest (a) reordering the legend to ML grad, mode,steep then the Adjei equivalents; (b) having the lines in the legend box longer; (c) ideally some colour to distinguish between ML and Adjei, if not put into the figure legends themselves Adjei=gray, ML=black; (d) can the lines be any thicker on the graphs?
+
+We changed Figures 2, 3 and corresponding supplementary figures according the the suggestions by the reviewer.
+We reordered the legends, lengthened the lines in the legend box, colored the Adjei Lines green, and made all the lines thicker.
+We believe that this makes the figures much more readable.
+
+8. Figure 4 - I would recommend having vertical lines coming down from the steep and moderate to see what sample size is associated with 80% power
+
+Added the recommended vertical lines to the Figure.
+
+9. Figure S4 missing from my download
+
+Seems to be present in the version that we are looking at.
+
+10. Pg 2 lines 13-15 ref for sentence would be good
+
+Added a citation AndersonandMay1978
+
+11. Pg 5 line 84 - an unnecessary "an"
+
+Deleted the extra "an"
+
+12. Pg 12 - Figure 4 referred to in text before Figure 3 - suggest swapping order of figures
+
+Swapped the order of the figures

### 2015-08-08 docs/parasite_mortality_manuscript.tex
Incorporated reviewer comments

@@
 rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, such that host survival is not typically considered to be a function of the number of infectious agents present within the host.
-In contrast, models of macroparasites generally assume that pathology increases with parasite burden and host survival probability must be treated as a function of infection intensity.
+In contrast, models of macroparasites generally assume that pathology increases with parasite burden and host survival probability must be treated as a function of infection intensity \citep{AndersonandMay1978}.
 Helminths exhibiting this intensity-dependent pathology have significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013}, and wildlife survival \citep{Kirk2003,Logiudice2003}.
@@
 As discussed by
-\citeauthor{Adjei1986}, this is not necessarily an unrealistic as some parasite infections occur primarily in younger hosts with parasite-induced mortality occurring soon after infection \citep[e.g.][]{Schotthoefer2003,Johnson2008}.
+\citeauthor{Adjei1986}, this is not necessarily unrealistic as some parasite infections occur primarily in younger hosts with parasite-induced mortality occurring soon after infection \citep[e.g.][]{Schotthoefer2003,Johnson2008}.
 
@@
 
-We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters.
+We used statistical power and Type I error to test the ability of the Adjei Method and the Likelihood Method to correctly identify the presence of PIHM on simulated data with known pre-mortality parameters.
+The power of a method is the probability of correctly detecting PIHM given that it is occurring and the Type I error is the probability of incorrectly identifying PIHM given that it is not occurring.
+If a method has low Type I error we can be confident that when we detect PIHM it is actually occurring.
+If one method has higher power for detecting PIHM than another, we will need to sample fewer hosts to detect PIHM.
+
 Consistent with the model assumption that parasite infection, host mortality, and population sampling are temporally separate events, we first created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
@@
 Wild host populations were assumed to be sampled after PIHM has occurred, thus we calculated the sample size in the power simulations as the average number of surviving hosts over all 150 simulations for each parameter combination.
+The distribution of surviving hosts over the 150 simulations was generally symmetrical and the standard deviation was small compared to the mean (maximum coefficient of variation was approximately 0.06 across all parameter combinations), suggesting that the mean number of surviving hosts was an adequate summary statistic of the number of hosts sampled post-mortality.
 
@@
 PIHM from parasite intensity data, all such methods require several
-fundamental, and potentially problematic assumptions.
+fundamental assumptions.
 Nearly all current methods derive from \cite{Crofton1971a} \citep[but see][]{Ferguson2011} and assume that, prior to any PIHM, parasites are distributed in the host population following a
@@
 
-Unfortunately, this
-flexibility in the distribution may also reduce our ability to detect PIHM. If
+It is important to note that the flexibility of the negative binomial distribution may also reduce our ability to detect PIHM. If
 a negative binomial can be fit to the observed post-mortality parasite
@@
 impossible to detect PIHM because there is no need for a more complex model.
-Most observed parasite distributions are well fit by the negative binomial distribution \citep{Shaw1998}, suggesting
-that systems where these methods are applicable may be more the exception than
-the rule.  Furthermore, even when truncation of the negative binomial distribution is detected, it may be caused by other
+Many observed parasite distributions are well-fit by the negative binomial distribution \citep{Shaw1998}, suggesting
+that systems where these methods are applicable without any a prior knowledge may be uncommon.
+However, if one has a prior knowledge about some aspect of the pre-mortality distribution \citep[e.g.
+assumes/knows the value of $k_p$,][]{Ferguson2011}, then the Likelihood Method could be applicable even if the the post-mortality distribution was well-fit by a negative binomial.
+
+If one has evidence that the pre-mortality is not negative binomial, the generality of our method easily allows another distribution to be specified for $g(x, \boldsymbol{\phi})$.
+In particular, one could use the
+resulting stationary host-parasite distribution from a stochastic host-parasite
+model without parasite-induced host mortality \citep{Anderson1982a} to specify
+the form of $g(x, \boldsymbol{\phi})$ and then apply the techniques discussed
+in this paper to detect PIHM.  The general criteria necessary
+for the Likelihood Method to detect PIHM in a stochastic host-parasite process
+is that the stationary distribution of the process with
+mortality is significantly different than the stationary distribution without mortality.
+It is widely recognized that parasite-induced host mortality decreases the
+aggregation of host-parasite distributions relative to those without mortality
+\citep{Barbour2000}, suggesting that Likelihood Method could be applicable for
+detecting PIHM in host-parasite systems experiencing stochastic parasite births, deaths, and immigrations in addition to host mortality.
+
+If the Likelihood Method is applicable and the truncation of the negative binomial distribution is detected, one must be aware that the pattern may be caused by other
 processes such as within host density dependence, age dependent variation in host
@@
 that PIHM is detected, it may actually not be the result of PIHM.
+Moreover, if host mortality depends on parasite intensity and additional variables (e.g.
+host sex, host size), failure to identify these important confounding variables could significantly affect the ability of these methods to correctly identify PIHM.
+However, both of these issues -- inferring process from pattern and confounding variables -- are well-recognized limitations of most statistical inference and are addressed via judicious model specification and selection \citep{Seber2003}.
 
-Given these numerous caveats, is there a place in parasitology for methods that
-estimate PIHM from intensity data?  As suggested by
-\cite{Lester1984} these methods for estimating PIHM can
+As suggested by \cite{Lester1984} these methods for estimating PIHM can
 provide preliminary insight into whether or not PIHM is worth further
-exploration.  However, we stress that these methods should only be used as an
-exploratory tool when assessing the role of PIHM in a system, and potential
+exploration.  However, we stress that these methods are an
+exploratory tool for assessing the role of PIHM in a system, and potential
 users should critically evaluate whether they think they have a large enough
@@
 \begin{figure}
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure3_for_manuscript.pdf}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
+
+    \caption{The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
+The curves were generated from 500 simulations for 10 pre-mortality sample sizes, $N_p$.
+The vertical, dotted-dashed lines indicate the sample size at which the power the for the Likelihood Method with steep and moderate survival functions is 0.8 (75 hosts for steep functions and 402 for moderate functions).
+The Likelihood Method with a gradual survival function never has a power above 0.8.}
+
+    \label{fig:real_power}
+
+\end{figure}
+
+\begin{figure}
 
@@
 
-\begin{figure}
-    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure3_for_manuscript.pdf}
-
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
-The curves were generated from 500 simulations for 10 pre-mortality sample sizes, $N_p$.}
-
-    \label{fig:real_power}
-
-\end{figure}
-
 \end{document}

### 2015-08-08 code/pihm_methods.py

@@ -631,7 +631,7 @@ def likefxn2(params, x, mu, k):
 
 
 def extract_simulation_results(sim_results, keys, method_name, param,
-                                alpha=0.05):
+                                alpha=0.05, get_std=False):
     """
 
     Method to extract parameters from a simulation dictionary as generated
@@ -656,6 +656,10 @@ def extract_simulation_results(sim_results, keys, method_name, param,
 
     alpha : float
         The significance level.  Only used if param == "p"
+
+    get_std : bool
+        If True calculates the standard deviation of the post-mortality samples
+        rather than the mean.
 
     Returns
     : tuple
@@ -690,7 +694,10 @@ def extract_simulation_results(sim_results, keys, method_name, param,
 
         N_vals = res[2]
 
-        samp_sizes.append(np.mean(N_vals))
+        if not get_std:
+            samp_sizes.append(np.mean(N_vals))
+        else:
+            samp_sizes.append(np.std(N_vals, ddof=1))
 
         if param == "a":
 

### 2015-08-08 docs/parasite_mortality_manuscript.tex
Added sentence on Adjei variability

@@
 The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results showing that the Adjei Method has a high Type I error rate.
+Moreover, the Adjei Method estimates of the $LD_{50}$ were quite variable for the Crofton data, consistent with our simulation results that the Adjei Method $LD_{50}$ estimates could be imprecise for sample sizes of less than 1000 hosts (\emph{SI} Fig.
+4-6).
 
@@
 Although superior to the Adjei Method, the Likelihood Method still cannot be applied to all real datasets.
-For host-parasite systems where host mortality occurs as a steep, non-linear function of parasite intensity only 80 hosts must be sampled to have an 80\% power in detecting PIHM.
+For host-parasite systems where host mortality occurs as a steep, non-linear function of parasite intensity only 75 hosts must be sampled to have an 80\% power in detecting PIHM.
 However, as the maximum slope of the survival function decreases and the function becomes somewhat linear, hundreds, or possibly thousands of hosts would have to be sampled to achieve the same result.

### 2015-08-10 docs/figures_tables/figure2.tex
Remade figures and added final comments before submission
Remade figures and sent revised manuscript to coauthors for comments

@@
-% Document
-% Document
-\documentclass[12pt, a4paper]{article}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage{authblk}
-\usepackage{lipsum}
+\documentclass[tikz]{standalone}
+
+\usepackage{amsmath}
 \usepackage{tikz}
-\usepackage{hyperref}
-
-% Figure and table formating
-\usepackage{epsfig}
-\usepackage{tabu}
-\usepackage{rotating}
-\usepackage{pbox}
-\usepackage{framed, multicol}
-\usepackage[framemethod=TikZ]{mdframed}
-
-% Set up Frame
-\mdfdefinestyle{MyFrame}{%
-    linecolor=black,
-    skipabove=10pt,
-    outerlinewidth=2pt,
-    roundcorner=20pt,
-    innertopmargin=10pt,
-    innerbottommargin=10pt,%\baselineskip,
-    innerrightmargin=20pt,
-    innerleftmargin=20pt,
-    backgroundcolor=gray!50!white}
-
-\usepackage{float}
-\usepackage[left=1 in, right=1 in, top=1.25 in, bottom=1.25 in]{geometry}
-
-% Fonts - Mathtime
-%\usepackage{txfonts}
-\usepackage{amsmath} % Add amssymb if not using Mathtime
-
-% Text
-\setlength{\parindent}{0.5in}
-\frenchspacing  \tolerance = 800  \hyphenpenalty = 800
-
-\usepackage{lineno} % Line numbers
-\def\linenumberfont{\normalfont\footnotesize\ttfamily}
-\setlength\linenumbersep{0.1 in}
-
-\usepackage{setspace}
-
-% Other
-\usepackage{graphicx}
-\usepackage[singlelinecheck=false,font=small,labelfont=bf]{caption}
+\usetikzlibrary{matrix}
+\usetikzlibrary{backgrounds}
 
 \begin{document}
-
-\begin{figure}
 
@@
 
-    \node at (-3.2, 4.9) {A.};
-    \node at (-6.7, 2.9) {B.};
-    \node at (1.15, 2.9) {C.};
+    \node at (-2.5, 3.8) {A.};
+    \node at (-5, 2.2) {B.};
+    \node at (1, 2.2) {C.};
 
@@
 
-\end{figure}
-
-\centering
-
-\textbf{Figure 1}
-
 \end{document}

### 2015-08-10 docs/parasite_mortality_manuscript.tex
Remade figures and sent revised manuscript to coauthors for comments

@@
 must be scrutinized before applying them to real-world data. Given that these
-assumptions are met, our method is new exploratory tool that can help inform
+assumptions are met, our method is a new exploratory tool that can help inform
 more rigorous studies of parasite-induced host mortality.
@@
 Many observed parasite distributions are well-fit by the negative binomial distribution \citep{Shaw1998}, suggesting
-that systems where these methods are applicable without any a prior knowledge may be uncommon.
-However, if one has a prior knowledge about some aspect of the pre-mortality distribution \citep[e.g.
+that systems where these methods are applicable without any \emph{a priori} knowledge may be uncommon.
+However, if one has \emph{a priori} knowledge about some aspect of the pre-mortality distribution \citep[e.g.
 assumes/knows the value of $k_p$,][]{Ferguson2011}, then the Likelihood Method could be applicable even if the the post-mortality distribution was well-fit by a negative binomial.
@@
 If one has evidence that the pre-mortality is not negative binomial, the generality of our method easily allows another distribution to be specified for $g(x, \boldsymbol{\phi})$.
-In particular, one could use the
+For example, one could use the
 resulting stationary host-parasite distribution from a stochastic host-parasite
@@
 aggregation of host-parasite distributions relative to those without mortality
-\citep{Barbour2000}, suggesting that Likelihood Method could be applicable for
-detecting PIHM in host-parasite systems experiencing stochastic parasite births, deaths, and immigrations in addition to host mortality.
+\citep{Barbour2000}, suggesting that the Likelihood Method could be generally applicable to host-parasite systems that follow the assumptions of many stochastic host-parasite models.
+This is an intriguing area for further research.
 
-If the Likelihood Method is applicable and the truncation of the negative binomial distribution is detected, one must be aware that the pattern may be caused by other
+If the Likelihood Method is applicable and the truncation of the negative binomial distribution is detected, one must be aware that the truncation pattern may be caused by other
 processes such as within host density dependence, age dependent variation in host
@@
 In general, the Likelihood Method has higher power and lower Type I error than the Adjei Method.
-See the \emph{SI} 3 Fig 1 - 3 for Type I Error and power results for all parameter combinations.}
+See the \emph{SI} 3 Fig 1 - 3 for Type I error and power results for all parameter combinations.}
 
@@
 The curves were generated from 500 simulations for 10 pre-mortality sample sizes, $N_p$.
-The vertical, dotted-dashed lines indicate the sample size at which the power the for the Likelihood Method with steep and moderate survival functions is 0.8 (75 hosts for steep functions and 402 for moderate functions).
+The vertical, dotted-dashed lines indicate the sample size at which the power for the Likelihood Method with steep and moderate survival functions is 0.8 (75 hosts for steep functions and 408 for moderate functions).
 The Likelihood Method with a gradual survival function never has a power above 0.8.}

### 2015-08-13 docs/parasite_mortality_manuscript.tex
Added Saras revisions

@@
 transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies \citep{Coyne1989}, bovine tuberculosis \citep{Cox2005}, and
-rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, such that host survival is not typically considered to be a function of the number of infectious agents present within the host.
+rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, such that host survival is not generally considered to be a function of the number of infectious agents present within the host.
 In contrast, models of macroparasites generally assume that pathology increases with parasite burden and host survival probability must be treated as a function of infection intensity \citep{AndersonandMay1978}.
@@
 In these methods, PIHM can only be identified by visually examining plots of the pre-mortality parameters predicted by the Crofton Method and determining whether they show a ``kink'' over a range of truncation values \citep[Figure \ref{fig:crofton}B;][]{Lester1984,Ferguson2011}.
-This qualitative criteria makes it difficult to compare PIHM between studies and a more rigorous and quantitative method is needed to both detect and quantify host mortality.
+These qualitative criteria makes it difficult to compare PIHM between studies and a more rigorous and quantitative method is needed to both detect and quantify host mortality.
 The survival function given by the Adjei
@@
 the form of $g(x, \boldsymbol{\phi})$ and then apply the techniques discussed
-in this paper to detect PIHM.  The general criteria necessary
+in this paper to detect PIHM.  The general requirement
 for the Likelihood Method to detect PIHM in a stochastic host-parasite process

### 2015-08-15 docs/figures_tables/figure2_IJPara15_261.tex
Renamed figures with manuscript number

\documentclass[tikz]{standalone}

\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{matrix}
\usetikzlibrary{backgrounds}
\tikzset{>=latex}

\begin{document}

\begin{tikzpicture}

    \node at (0, 0) (pic) {\includegraphics{/Users/mqwilber/Repos/parasite_mortality/results/figure1_partII_for_manuscript50}};

    \node[above=1cm] at (pic.north) (concept) {\includegraphics{/Users/mqwilber/Repos/parasite_mortality/results/figure1_partI_for_manuscript}};

    \node[above=1.3cm, at=(concept.south)] (temp) {};
    \node[right=6cm, at=(pic.north)] (right) {};
    \node[left=6cm, at=(pic.north)] (left) {};
    \draw[->, ultra thick]  (temp)--(right);
    \draw[->, ultra thick] (temp)--(left);

    \node at (-2.5, 3.8) {A};
    \node at (-5.1, 2.2) {B};
    \node at (0.9, 2.2) {C};

\end{tikzpicture}

\end{document}

### 2015-08-15 docs/figures_tables/table1_IJPara15_261.tex
Renamed figures with manuscript number

--- 2015-05-11 docs/parasite_mortatlity_manuscript.tex
@@
-% Document
 % Document
@@
 \usepackage{authblk}
-\usepackage{lipsum}
-\usepackage{tikz}
+\usepackage{hyperref}
 
 % Figure and table formating
-\usepackage{epsfig}
 \usepackage{tabu}
@@
 \usepackage{pbox}
-\usepackage{framed, multicol}
-\usepackage[framemethod=TikZ]{mdframed}
-
-% Set up Frame
-\mdfdefinestyle{MyFrame}{%
-    linecolor=black,
-    skipabove=10pt,
-    outerlinewidth=2pt,
-    roundcorner=20pt,
-    innertopmargin=10pt,
-    innerbottommargin=10pt,%\baselineskip,
-    innerrightmargin=20pt,
-    innerleftmargin=20pt,
-    backgroundcolor=gray!50!white}
 
 \usepackage{float}
-\usepackage[left=1 in, right=1 in, top=1.25 in, bottom=1.25 in]{geometry}
+\usepackage[left=0.75 in, right=0.75 in, top=1.25 in, bottom=1.25 in]{geometry}
 
@@
 %\usepackage{txfonts}
-\usepackage{amsmath} % Add amssymb if not using Mathtime
 
-% Text
-\setlength{\parindent}{0.5in}
-\frenchspacing  \tolerance = 800  \hyphenpenalty = 800
-
-\usepackage{lineno} % Line numbers
-\def\linenumberfont{\normalfont\footnotesize\ttfamily}
-\setlength\linenumbersep{0.1 in}
-
-\usepackage{setspace}
-
-% Format section and subsection headers
-\makeatletter
-\renewcommand{\section}{\@startsection
-{section}%                   % the name
-{1}%                         % the level
-{0mm}%                       % the indent
-{-\baselineskip}%            % the before skip
-{0.5\baselineskip}%          % the after skip
-{\normalfont\bf\large}} % the style
-
-\renewcommand{\subsection}{\@startsection
-{subsection}%                   % the name
-{2}%                         % the level
-{0mm}%                       % the indent
-{-\baselineskip}%            % the before skip
-{0.5\baselineskip}%          % the after skip
-{\normalfont\bf}} % the style
-\makeatother
-
-% Other
-\usepackage{graphicx}
 \usepackage[singlelinecheck=false,font=small,labelfont=bf]{caption}
@@
 
-% Format the Bibliography appropriately
-% increase \bibhang to take care of the numbers
-\setlength{\bibhang}{0pc}
-\makeatletter
-% patch \@lbibitem to print the current number before the authors
-\patchcmd{\@lbibitem}
-  {]}
-  {] \theNAT@ctr. \newline }
-  {}{}
-\makeatother
-
 %%%%%% FRONT MATTER %%%%%%%%%
-
-\title{Methods for estimating parasite-induced mortality from intensity data and their limitations}
-\author{Mark Wilber, Sara Weinstein, and Cherie Briggs}
 
@@
 
-\maketitle
-
-\section*{Abstract}
-
-TODO
-
-\doublespacing
-
-\linenumbers
-\section*{Introduction}
-
-Infectious agents can have major impacts on animal populations through changing
-population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
-even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
-of these infectious agents in wildlife is critical to understanding what
-regulates host and parasite populations, making predictions about disease
-transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies (Coyne et al. 1989) [get citation], bovine TB \citep{Cox2005}, and
-rinderpest \citep{Tille1991}, is typically quantified based on the presence or absence of
-disease, and does not account for the number of infectious agents present.
-This method is sufficient for many bacterial and viral agents that reproduce within a
-host, however for macroparasites pathology is linked to the intensity of infection and hosts cannot be simply categorized as infected and
-uninfected \citep{AndersonandMay1979,Lafferty2002}.  Helminths exhibiting this intensity dependent pathology have
-significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013},
-wildlife survival \citep{Kirk2003, Logiudice2003}.  While it is generally assumed that some fraction of
-wild host populations must succumb to parasitic infections, it is notoriously
-difficult to actually quantify parasite-induced host mortality (PIHM) in wild
-animal populations \citep{McCallum2000a}.
-
-Ideally, parasite-induced host mortality is
-quantified by experimentally infecting and tracking individual hosts in the
-wild population; however, for logistical and ethical reasons this method is
-rarely feasible \citep{McCallum2000a}. Data on parasite intensity is much easier to collect and has
-often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and quantify the
-relationship between infection intensity and host mortality \citep{Adjei1986}.
-
-\cite{Crofton1971a} first proposed that PIHM could be identified by comparing the
-observed parasite distribution in the host population to the distribution
-predicted in the absence of parasite-induced host mortality.
-We briefly introduce the Crofton Method here and provide a more detailed explanation of its implementation in \emph{Supplementary Material (SI)} 1.
-This method
-assumes that, prior to host mortality, infection intensity in the host population follows a negative binomial distribution and the tail of the distribution is truncated as intensity dependent pathology removes the most heavily infected hosts.
-Assuming mortality occurs only in these heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
-fitting a negative binomial distribution to hosts with lower and lower parasite
-loads, and comparing the tail ends of these predicted distributions to the
-observed parasite data.
-
-The Crofton Method may be able to detect the presence of PIHM however, but it does not quantify the relationship between infection intensity and host survival
-probability. \cite{Adjei1986} suggested that this relationship
-could be calculated by first using the Crofton Method to estimate the pre-mortality parasite distribution and then using those distribution to calculate the
-probability of host survival with increasing parasite intensity. To do this,
-\cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI} 2 for a technical description of the Adjei Method).
-These methods appeared to provide an estimate for the parasite intensity at which
-50\% of hosts exhibit PIHM ($LD_{50}$), as well as the unmeasurable fraction of the
-population that was lost (\emph{SI} 2). However, to implement this method the observed data must be
-modified to fit the GLM framework and subjectively binned when mean
-infection intensity is high or sample sizes are small.
-
-After 30 years, and despite clear limitations \citep{McCallum2000a}, these
-methods (particularly the Crofton Method) are still discussed among
-parasitologists and are the primary techniques for examining population level
-impacts of parasitism using parasite intensity data. In these methods, PIHM can
-only be identified by visually examining plots and, with no clear decision
-rule, it can be difficult to determine the significance of PIHM across
-different host-parasite systems. The survival function produced by the Adjei
-Method offers one solution; however, this method requires manipulating the
-original data and has never been tested.
-
-Intensity data should only
-be used to estimate parasite impacts on host populations if unbiased and accurate methods exist. In
-this study, we first propose a novel method for calculating PIHM. We
-next use simulations to compare our method with the previous Adjei
-Method to test the ability of both methods to (1) detect occurrence of PIHM and (2) estimate the
-lethal parasite load ($LD_{50}$) and the associated survival function.  We then
-apply both methods to real datasets previously used in PIHM analyses and
-compare the results. Finally, we discuss the limitations of inferring PIHM
-from intensity data and whether any method for inferring PIHM has a place
-in quantitative parasitology.
-
-\section*{Methods}
-
-\subsection*{A novel, likelihood-based method for estimating PIHM}
-
-Here we propose a novel, likelihood-based method (henceforth Likelihood Method) that does not
-require binning or data alteration, reduces the number of
-parameters to be estimated, and  uses standard statistical techniques to determine PIHM significance. We provide Python code for implementing the Likelihood Method in \emph{SI} 4.
-
-As with all previously proposed methods for estimating PIHM, the
-Likelihood Method first assumes that prior to mortality the parasite distribution is described by a negative binomial $g(x; \mu_p, k_p)$, where $\mu_p$ and $k_p$ are the mean parasite intensity and aggregation before mortality, respectively (smaller $k_p$ indicates more aggregation).
-Previous methods required calculating the total number of hosts before mortality ($N_p$) \citep{Crofton1971a,Adjei1986}, however this parameter is not needed in the Likelihood Method.
-
-The Likelihood Method then assumes that the host survival function, which specifies the probability of a host surviving with $x$ parasites, follows the logistic curve given by
-
-\begin{equation}
-    h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
-    \label{eq:logistic}
-\end{equation}
-With these two explicit assumptions, the Likelihood Method estimates four parameters: $\mu_p$, $k_p$, $a$, and $b$ by  first defining a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.
-Using standard rules of conditional probability this distribution can be written as
-
-\begin{equation}
-    P(x | \text{survival}) = \dfrac{P(\text{survival} | x) * P(x)}{P(\text{survival})}
-    \label{eq:concept}
-\end{equation}
-
-One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$.
-Therefore equation \ref{eq:concept} can be written as
-
-\begin{equation}
-    P(x | \text{survival}) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
-    \label{eq:dist}
-\end{equation}
-
-Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset.
-The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.
-
-To estimate the significance of PIHM in a host-parasite system, a
-likelihood ratio test is used in which the full model is given by equation
-\ref{eq:dist} and the reduced model is given by a negative binomial
-distribution.  If PIHM is not significant in the system, the resulting likelihood
-ratio statistic should approximately follow a $\chi^2$ distribution with two degrees of freedom.
-
-\subsection*{Evaluating the Adjei and Likelihood Methods}
-
-\emph{Question 1: Can we detect PIHM?}
-
-We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters.
-First, we created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
-negative binomial distribution with parameters $\mu_p$ and $k_p$. Second, we chose values of $a$ and $b$ and calculated the probability of survival
-for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, for each host, we drew a random number from a uniform distribution
-between 0 and 1 and if the calculated host survival probability was less than this random
-number, the host experienced parasite-induced mortality.
-The parasite distribution in these simulated surviving hosts is equivalent to the observed parasite distribution in a wild host population that has undergone parasite-induced host mortality.
-
-We used these simulated pre-mortality and post-mortality datasets to test the ability
-of both methods to correctly determine whether or not PIHM was occurring when the parameters $N_p$, $\mu_p$ and $k_p$ were known.
-  For the Adjei Method, $N_p$,
-$\mu_p$, and $k_p$ are estimated using the Crofton Method, while $\mu_p$ and
-$k_p$ in the Likelihood Method can be estimated jointly with $a$ and $b$ or via the Crofton Method.
-Although the parameters $N_p$, $\mu_p$, and $k_p$ are always unknown in real systems, a method that fails under these ideal simulation conditions will certainly also fail using less ideal, empirical data.
-
-We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:question1}A).
-For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.
-We examined each $\mu_p$-survival function pair at  three levels of parasite
-aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
-combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).
-For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.
-We tested each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500].
-$N_p$ is not technically the sample size on which the methods are being
-tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
-dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.\\
-
-\noindent
-\emph{Question 2: Can we estimate fatal parasite intensity and the host survival function?}
-
-To compare the ability of the Adjei Method and the Likelihood Method to recover the $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above.
-For each parameter
-combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
-precision \citep{Walther2005} for these estimates over pre-mortality host population sizes of $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
-10000].
-We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.
-Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.
-
-\subsection*{Efficacy of the Likelihood Method with unknown pre-mortality parameters}
-
-In the final simulation, we test the ability of the Likelihood Method to correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality parameters are unknown.
-The previous simulations showed that the Likelihood Method effectively identified PIHM when $\mu_p$ and $k_p$ were known with values of 10 and 1, respectively.
-In the best-case scenario where a host-parasite system has these these parameters, we test the power of the Likelihood Method to identify PIHM for gradual, moderate and steep survival functions when the pre-mortality parameters also needed to be estimated.
-We perform 500 simulations over a range of different samples sizes following the simulation procedure described above.
-
-\subsection*{Application to real data}
-
-We tested the ability of the Adjei Method and the Likelihood Method to identify
-PIHM in 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
-given in \cite{Adjei1986} (Table \ref{table:pihm}). \citeauthor{Crofton1971a} analyzed infection patterns in the snail \emph{Gammarus pulex} infected with the
-acanthocephalan \emph{Polmorphus minutus}. \citeauthor{Adjei1986} analyzed males and females of two species of lizard fish \emph{Saurida tumbil} and
-\emph{S. undosquamis} that were infected by the cestode
-\emph{Callitetrarhynchus gracilis}.
-
-In both earlier studies, the authors reported PIHM in some of the datasets and we test whether the Adjei
-Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
-\cite{Crofton1971a}, we truncated the data at 4 parasites, applied the Crofton
-Method to estimate the pre-mortality distribution, and then ran the Likelihood
-Method and Adjei Method using these pre-mortality parameters.  For the
-\cite{Adjei1986} datasets, we followed the same procedure as the authors and
-first truncated the data at 2 parasites and then fit the Crofton Method for the
-female fish of both species.  Then, following the original authors' methods, we parameterized the male pre-mortality
-distributions for each species with the results from the females.  Finally, we
-applied the Adjei Method and the Likelihood Method to determine whether or not
-PIHM was significant for these species and compared our results to those given by the authors.  All fitting to data was done with the code provided in \emph{SI} 4.
-
-\section*{Results}
-
-\subsection*{Question 1: Detecting presence of PIHM}
-
-The power of the Adjei Method to detect PIHM in a
-system was close to unity for larger sample sizes and tended to
-decrease as sample size decreased (Figure \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).  The Likelihood Method had a power close to
-unity for all parameter combinations and sample sizes considered.  With gradual
-survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (Fig. \ref{fig:question1}B, \emph{SI} 3 Figs 1-3).
-
-The Adjei Method showed highly inflated Type I error rates (i.e. falsely detected
-PIHM) for all parameter combinations that we
-considered (Fig. \ref{fig:question1}A; \emph{SI} 3 Figs 1-3).  This method also showed the unintuitive pattern of Type I error
-rate decreasing as sample size decreased.  This pattern was due to the issue of
-binning discussed in the \emph{Introduction} and \emph{SI} 2. For small samples sizes, the
-applicability of the Adjei Method is compromised without binning the observed
-data in some way.  In contrast, the Likelihood Method showed a Type I
-error rate at or near the pre-set level of 0.05 for all parameter combinations
-and sample sizes considered (Fig. \ref{fig:question1}A; \emph{SI} 3 Figs 1-3).
-
-\subsection*{Question 2: Estimating the $LD_{50}$ and survival function}
-
-The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$
-for all combinations of parameters examined in this study (Fig. \ref{fig:question2}, \emph{SI} 3 Figs 4-6).  Even for
-the smallest sample sizes we considered, the Likelihood Method's estimate of $LD_{50}$
-was largely unbiased, with small biases occurring for host survival functions
-that were gradual. The precision of the Likelihood Method's $LD_{50}$ estimates decreased
-(increasing coefficient of variation) as sample size decreased for all
-parameter combinations we examined (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6).
-
-The Adjei Method produced biased estimates of the $LD_{50}$ across nearly all parameter combinations (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6)).  For $\mu_p = 10$, the $LD_{50}$
-estimates from the Adjei Method were largely unbiased for large samples sizes, but as $\mu_p$ increased, the Adjei Method
-produced biased estimates of $LD_{50}$ across all sample sizes, with bias
-increasing as sample size decreased (Figure \ref{fig:question2}, SI2 Fig 4-6). The $LD_{50}$ estimates from the Adjei
-Method showed large decreases in precision with the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:question2}, SI2 Fig 4-6).
-
-In terms of the host survival function, the Likelihood Method gave unbiased estimates of survival function parameters when sample sizes were large, however as sample size decreased these estimates became severely biased (Fig.
-\ref{fig:question2}, \emph{SI} Fig 7 - 9) The Adjei Method produced
-biased estimates of the host survival function across all sample sizes, with the bias
-consistently being larger when the survival function was steeper and $\mu_p$ was larger (Fig \ref{fig:question2}, \emph{SI} 3 Figs 7-9).
-
-\subsection*{Detecting PIHM with unknown pre-mortality parameters}
-
-When all pre-mortality parameters were jointly estimated, the Likelihood Method had a power of greater than 0.8 when the survival function was moderate and steep for host sample sizes of 424 and 83 respectively (Figure \ref{fig:real_power}).
-When the host survival function was gradual, the Likelihood Method never had a power greater than 0.8 for any post-mortality samples sizes we considered.
-
-\subsection*{Application to real data}
-
-Of the 10 datasets we considered, the previous authors visually detected PIHM
-in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
-from the pre-mortality parameters of the Crofton Method detected significant
-PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
-dataset in which the Likelihood Method did not detect a significant effect of PIHM was the Adjei dataset
-for female \emph{S. tumbil}.  For this dataset there was a marginally significant effect
-of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$).
-The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results that the Adjei Method has a high Type I error rate.
-
-\section*{Discussion}
-
-Quantifying the impact of parasitism on wild host populations is critical for managing wildlife populations and understanding parasite transmission. Ideally the relationship between
-infection intensity and host survival would be measured experimentally, but for
-logistical and ethical reasons, this is often impossible \citep{McCallum2000a}.
-Looking for evidence of mortality in parasite distribution data requires the
-least amount of information, but is notoriously difficult to implement. The
-methodological flaws in the Adjei Method limit its utility, so here we propose
-an alternative, likelihood-based, method to estimate host survival and the
-$LD_{50}$ from observed parasite intensity data.  This
-method is a significant improvement over the previous methods because it requires fewer parameters,
-provides a statistical decision rule for identifying PIHM and does not require
-any data manipulation.
-
-Using simulated data, we found that the Likelihood Method always out performed the Adjei Method. For simply detecting the presence of PIHM, the Likelihood
-Method was both more powerful and had fewer false detection events (Type I
-errors).  When both methods were applied to published datasets previously used
-in PIHM analyses, the Adjei Method tended to detect PIHM where it had not previously been
-reported, consistent with the high Type I error rate observed in our
-simulations. The Likelihood Method was also more precise and less
-biased in calculations of both the parasite $LD_{50}$ and host survival curve over the parameter values we considered.
-However, while only the Likelihood Method produced precise and unbiased $LD_{50}$
-estimates, neither method could provide unbiased estimates of the host survival
-function at realistic sample sizes.  These simulations demonstrate that
-the Likelihood Method is more powerful and precise than the previously propose Adjei Method.
-
-Although superior to the Adjei Method, the Likelihood Method is not universally applicable to real data.  Our simulations showed when the when pre-
-mortality parameters were estimated directly, the Likelihood Method needed at
-least 83-424 samples to have 80\% power and for steep to moderate survival functions and even
-more as the survival function became more gradual. While some of these sample sizes are reasonable for hosts such as invertebrates or small fish, even the smallest sample sizes are completely
-unfeasible for many vertebrates, particularly the species of conservation
-concern where addressing the impact of parasitism would be most important. An
-even larger sample size would be required to identify PIHM when parasites are highly aggregated, mean infection intensity is
-high, or parasite prevalence is low, all of which are
-common in many parasitic helminths.
-Moreover, our results are in agreement with previous work that has shown that as host-survival functions become progressively more linear, PIHM becomes all but impossible to detect \citep{Lanciani1989}.
-This result, however, does not preclude the use of this method as non-linear survival functions
-are not uncommon in empirical host-parasite systems \citep{Benesh2011}.  Finally,
-while linear functions make PIHM undetectable, at the other extreme, steep,
-non-linear survival curves produce severely biased estimates of the survival
-function. Give the interaction between all of these different factors, the
-Likelihood Method is probably limited to detecting PIHM in systems where greater than 100 hosts can be collected, parasites are
-common and only moderately aggregated, and substantial host mortality occurs at relatively low parasite intensity.
-
-While we have improved on the existing methods for quantifying
-PIHM from parasite intensity data, all such methods require several
-fundamental, and potentially problematic assumptions.
-Nearly all current methods derive from \cite{Crofton1971a} \citep[but see][]{Ferguson2011} and assume that, prior to any PIHM, parasites are distributed in the host population following a
-negative binomial distribution. But, it is fundamentally impossible to know
-what the pre-mortality parasite distribution was in a wild host population and
-it is widely recognized that different processes can lead to a variety of
-parasite distributions in hosts \citep{Anderson1982a, Duerr2003}. However, the negative binomial is extremely
-flexible and there is substantial empirical and theoretical evidence to support
-the assumption that, prior to any PIHM, parasite distributions can be fit by a negative binomial distribution \citep{Shaw1995,Shaw1998,Wilson2002}.
-
-Unfortunately, this
-flexibility in the distribution may also reduce our ability to detect PIHM. If
-a negative binomial can be fit to the observed post-mortality parasite
-distribution then, regardless of how lethal the parasite was, it will be
-impossible to detect PIHM because there is no need for a more complex model.
-Most observed parasite distributions are well fit by the negative binomial distribution \citep{Shaw1998}, suggesting
-that systems where these methods are applicable may be more the exception than
-the rule.  Furthermore, even when truncation of the negative binomial distribution is detected, it may be caused by other
-processes such as within host density dependence, age dependent variation in host
-resistance and/or heterogeneous infection rates \citep{McCallum2000a,Anderson1982a,Rousset1996}.  This means that in the event
-that PIHM is detected, it may actually not be the result of PIHM.
-
-Given these numerous caveats, is there a place in parasitology for methods that
-estimate PIHM from intensity data?  We are in agreement with
-\cite{Lester1984} that, at the very least, methods for estimating PIHM can
-provide preliminary insight into whether or not PIHM is worth further
-exploration.  However, we stress that these methods should only be used as an
-exploratory tool when assessing the role of PIHM in a system and potential
-users should critically evaluate whether they think they have a large enough
-sample size and an appropriate host survival function/post-mortality distribution for the methods developed
-in this paper to be applicable.  Even if they are applicable, inferring PIHM
-from distributional data is no substitute for field experiments
-and an in depth understanding of the natural history of the host-parasite
-system under consideration.
-
-\section*{Acknowledgments}
-
-TODO
-
-\singlespacing
-\bibliographystyle{/Users/mqwilber/Dropbox/Documents/Bibformats/ecology_letters.bst}
-\bibliography{/Users/mqwilber/Dropbox/Documents/Bibfiles/Projects_and_Permits-parasite_host_mortality}
-
-\newpage
-
-\renewcommand{\arraystretch}{1.2}
+\thispagestyle{empty}
 
@@
 
-    \caption{Definition of parameters and functions used in the main text}
-    \begin{tabular}{c c}
-    \hline
-    Parameter & Definition \\
-    \hline\hline
-    $\mu_p$ & Pre-mortality mean parasite intensity \\
-    $k_p$   & Pre-mortality parasite aggregation \\
-    $N_p$   & Pre-mortality host population size \\
-    $x$     & Number of parasites in a given host \\
-    $g(x; \mu_p, k_p)$ & Pre-mortality negative binomial distribution parasites distribution \\
-    $a$ & Parameter of the logistic host survival function \\
-    $b$ & Parameter of the logistic host survival function \\
-    $h(x; a, b)$ & The host survival function \\
-    $LD_{50}$ & $\exp(a / b)$, Parasite intensity at which 50\% of hosts die \\
-
-    \end{tabular}
-    \label{tab:params}
-\end{table}
-
-\renewcommand{\arraystretch}{1.2}
-\begin{sidewaystable}
-
     \caption{Comparison of the PIHM predictions of previously used host-parasite datasets to those given by the Adjei Method and the Likelihood Method.
-The first column specifies the identity of the dataset, the second column specifies whether or not the author indicated that PIHM was occurring in the system, the third column indicates whether or not the Likelihood Method with pre-mortality parameters estimated from the Crofton Method detects significant PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
-If a method detected significant PIHM the predicted $LD_{50}$ is given in parentheses}
+The first column specifies the identity of the dataset, the second column specifies whether or not the authors indicated that PIHM was occurring in the system based on a qualitative assessment, the third column indicates whether or not the Likelihood Method with pre-mortality parameters estimated from the Crofton Method detects significant PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
+If a method detected significant PIHM the predicted $LD_{50}$ is given in parentheses.}
 
@@
 
-\end{sidewaystable}
-
-\begin{figure}
-
-\begin{tikzpicture}
-
-    \node at (0, 0) (pic) {\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure1_partII_for_manuscript50}};
-
-    \node[above=0.1cm] at (pic.north) (concept) {\includegraphics[width=0.6\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure1_partI_for_manuscript}};
-
-    \draw[->] (0, 4.2)--(-1.5, 3.1);
-    \draw[->] (0, 4.2)--(1.5, 3.1);
-
-    \node at (-3.2, 4.9) {A.};
-    \node at (-6.5, 2.7) {B.};
-    \node at (1.4, 2.7) {C.};
-
-\end{tikzpicture}
-
-\caption{A) Five potential shapes for a host-survival functions.
-In the simulations we used a gradual survival function (dotted line), and moderate survival function (dashed line), and a steep survival function (solid line).
-The linear and immediate survival functions represent two potential extremes that we do not include in the simulations.
-For each of these survival functions and the parameter combinations described in the main text, we tested the Type I Error and Power of the Likelihood Method and Adjei Method.
-B) Gives the Type I Error of each method over a range of pre-mortality sample sizes with a pre-mortality mean parasite intensity ($\mu_p$) of 50 and pre-mortality parasite aggregation ($k_p$) at 0.5.
-The red line shows the pre-set significance level of 0.05.
-C) Gives the Power of each method for detecting PIHM over a range of post-mortality sample sizes for $\mu_p = 50$ and $k_p = 0.5$.
-In general, the Likelihood Method has higher power and lower Type I Error than the Adjei Method.
-See the SI 3 Fig 1 - 3 for Type I Error and Power results for all parameter combinations.}
-
-\label{fig:question1}
-
-\end{figure}
-
-\begin{figure}
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure2_for_manuscript.pdf}
-
-    \caption{Bias and precision (coefficient of variation) for the Likelihood Method and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
-The pre-mortality parameters for this simulation were $\mu_p = 50$ and $k_p = 0.5$.
-The figure shows the simulations for three different host survival functions (gradual, moderate, and steep), each with the same $LD_{50}$.
-Bias and precision results of $LD_{50}$ and $a$ for all other parameter combinations can be found in SI 3 Fig 4 - 9.}
-
-    \label{fig:question2}
-
-\end{figure}
-
-\begin{figure}
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure3_for_manuscript.pdf}
-
-    \caption{The power of the Likelihood Method to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
-The curves were generated from 500 simulations for 10 pre-mortality population sizes, $N_p$.}
-
-    \label{fig:real_power}
-
-\end{figure}
+\end{table}
 

### 2015-08-15 code/analysis_for_manuscript_full_likelihood.py
Fixed comment in file

@@ -8,7 +8,7 @@ Description
 -----------
 
 The analysis looks at the Power of the Likelihood Method for detecting PIHM
-when all the parameters (a, b, mu_, and k_p) must be esitmated from parasite
+when all the parameters (a, b, mu_p, and k_p) must be esitmated from parasite
 intensity data.  Just looks at the best case scenario when k_p=1 and mu_p=10.
 
 A thorough description of this analysis can be found in

### 2015-08-15 docs/parasite_mortality_manuscript.tex
Edited manuscript and supplementary material to meet journal specs

@@
 \usepackage{authblk}
-\usepackage{lipsum}
-\usepackage{tikz}
 \usepackage{hyperref}
@@
 % Figure and table formating
-\usepackage{epsfig}
 \usepackage{tabu}
@@
 \usepackage{pbox}
-\usepackage{framed, multicol}
-\usepackage[framemethod=TikZ]{mdframed}
-
-% Set up Frame
-\mdfdefinestyle{MyFrame}{%
-    linecolor=black,
-    skipabove=10pt,
-    outerlinewidth=2pt,
-    roundcorner=20pt,
-    innertopmargin=10pt,
-    innerbottommargin=10pt,%\baselineskip,
-    innerrightmargin=20pt,
-    innerleftmargin=20pt,
-    backgroundcolor=gray!50!white}
 
@@
 
-% Format the Bibliography appropriately
-% increase \bibhang to take care of the numbers
-% \setlength{\bibhang}{0pc}
-% \makeatletter
-% % patch \@lbibitem to print the current number before the authors
-% \patchcmd{\@lbibitem}
-%   {]}
-%   {] \theNAT@ctr. \newline }
-%   {}{}
-% \makeatother
-
 %%%%%% FRONT MATTER %%%%%%%%%
@@
 \begin{document}
+
+\linenumbers
 
@@
 
-\large\textbf{} & \vspace\baselineskip \\
-
 \textbf{Title:} & Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations \\
 
-\textbf{Running title:} & Methods for quantifying parasite-induced host mortality  \\
+\end{tabular}
+
+\begin{tabular}{p{3cm} p{10cm}}
 
@@
 
-\textbf{Statement of Authorship:} & MW and SW designed research, performed
-analyses and wrote the manuscript, CB designed research and contributed to revisions. \\
+\end{tabular}
+
+\begin{tabular}{p{3cm} p{10cm}}
 
@@
 
-\textbf{Email addresses:} &
-Mark Wilber: mark.wilber@lifesci.ucsb.edu \newline
-Sara Weinstein: sara.weinstein@lifesci.ucsb.edu \newline
-Cheryl Briggs: cherie.briggs@lifesci.ucsb.edu \\
+\end{tabular}
+
+\begin{tabular}{p{3cm} p{10cm}}
 
@@
 
+\doublespacing
+
 \abstract{
@@
 
-\doublespacing
-
-\linenumbers
 \section{Introduction}
@@
 Assuming mortality occurs only in heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
-fitting a negative binomial distribution to hosts with lower and lower parasite intensities, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data (Figure \ref{fig:crofton}, see \emph{Supplementary Material (SI)} 1 for additional detail).
+fitting a negative binomial distribution to hosts with lower and lower parasite intensities, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data (Fig.
+\ref{fig:crofton}, see Supplementary Data S1 for additional detail).
 
@@
 logistic function and then used a generalized linear model (GLM) to estimate
-the parameters of the host survival function (see \emph{SI} 2 for a technical
+the parameters of the host survival function (see Supplementary Data S2 for a technical
 description of the Adjei Method).  Although this method can predict the host survival function, it has several technical drawbacks. When mean infection intensity is high or sample sizes
@@
 Method to work, any observed intensity values greater than predicted values
-must be modified and set equal to the predict values (see \emph{SI} 2 for
+must be modified and set equal to the predict values (see Supplementary Data S2 for
 details); a questionable act of data manipulation. These manipulations may introduce bias, reduce the precision and
@@
 impacts of parasitism using parasite intensity data.
-In these methods, PIHM can only be identified by visually examining plots of the pre-mortality parameters predicted by the Crofton Method and determining whether they show a ``kink'' over a range of truncation values \citep[Figure \ref{fig:crofton}B;][]{Lester1984,Ferguson2011}.
+In these methods, PIHM can only be identified by visually examining plots of the pre-mortality parameters predicted by the Crofton Method and determining whether they show a ``kink'' over a range of truncation values \citep[Fig.
+\ref{fig:crofton}B;][]{Lester1984,Ferguson2011}.
 These qualitative criteria makes it difficult to compare PIHM between studies and a more rigorous and quantitative method is needed to both detect and quantify host mortality.
@@
 
-Generally, a larger $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases, with the maximum rate of decline having a value of $b / 4$ (\emph{SI} 2).
+Generally, a larger $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases, with the maximum rate of decline having a value of $b / 4$ (Supplementary Data S2).
 $b$ is in many ways analogous to the pathogenicity parameter ($\alpha$) in classic macroparasite models that gives the parasite intensity dependent host death rate \citep{AndersonandMay1978,Isham1995}.
@@
 More specifically, for every one unit increase in $a$ the log parasite intensity at which any percent of hosts survive (e.g.
-99\% of hosts survive) increases by $1 / b$ (\emph{SI} 2).
+99\% of hosts survive) increases by $1 / b$ (Supplementary Data S2).
 
@@
 
-We compared the two methods using three different mean parasite intensity values ($\mu_p$ = 10, 50, 100) and three different host survival functions (gradual, moderate, and steep decreases in the host survival with increasing parasite intensity, Figure \ref{fig:question1}A).
+We compared the two methods using three different mean parasite intensity values ($\mu_p$ = 10, 50, 100) and three different host survival functions (gradual, moderate, and steep decreases in the host survival with increasing parasite intensity, Fig.
+\ref{fig:question1}A).
 For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.
@@
 PIHM in six host-parasite datasets given in \cite{Crofton1971a} and four datasets
-given in \cite{Adjei1986} (Table \ref{table:pihm}). \citeauthor{Crofton1971a} analyzed infection patterns in the snail \emph{Gammarus pulex} infected with the
+given in \cite{Adjei1986} (Table 1). \citeauthor{Crofton1971a} analyzed infection patterns in the snail \emph{Gammarus pulex} infected with the
 acanthocephalan \emph{Polmorphus minutus}. \citeauthor{Adjei1986} analyzed males and females of two species of lizard fish \emph{Saurida tumbil} and
@@
 applied the Adjei Method and the Likelihood Method to determine whether or not
-PIHM was significant for these species and compared our results to those given by the authors.  All code for the analyses is provided in \emph{SI} 4.
+PIHM was significant for these species and compared our results to those given by the authors.  All code for the analyses is provided in Supplementary Data S3.
 
@@
 system was close to unity for larger sample sizes and tended to
-decrease as sample size decreased for all survival functions (Figure \ref{fig:question1}C; \emph{SI} 3 Figs 1-3).  The Likelihood Method had a power close to
+decrease as sample size decreased for all survival functions (Fig. \ref{fig:question1}C; Supplementary Fig. S1-S3).  The Likelihood Method had a power close to
 unity for all parameter combinations and sample sizes considered.  With gradual
-survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (Fig. \ref{fig:question1}C, \emph{SI} 3 Figs 1-3).
+survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (Fig. \ref{fig:question1}C, Supplementary Fig. S1-S3).
 
@@
 PIHM) for all parameter combinations that we
-considered (Fig. \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).  This method also showed the unintuitive pattern of decreasing Type I error
+considered (Fig. \ref{fig:question1}B; Supplementary Fig. S1-S3).  This method also showed the unintuitive pattern of decreasing Type I error
 rate with decreasing sample size.
-This occurred because, at small samples sizes, intensity data must be binned before the Adjei Method can be used (\emph{SI} 2).
+This occurred because, at small samples sizes, intensity data must be binned before the Adjei Method can be used (Supplementary Data S2).
 In contrast, the Likelihood Method showed a Type I
 error rate at or near the pre-set level of 0.05 for all parameter combinations
-and sample sizes considered (Fig. \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).
+and sample sizes considered (Fig. \ref{fig:question1}B; Supplementary Fig. S1-S3).
 
@@
 $\mu_p = 10$ and $k_p = 1$.
-For steep survival curves, PIHM could be detected with a power of greater than 0.8 from a sample of less than 100 hosts (Fig \ref{fig:real_power}).
-However, for moderate survival functions over 400 hosts had to be sampled to achieve the same power and for gradual survival functions, no tested sample size ever achieved a power greater than 0.8 (Fig \ref{fig:real_power}).
+For steep survival curves, PIHM could be detected with a power of greater than 0.8 from a sample of less than 100 hosts (Fig.
+\ref{fig:real_power}).
+However, for moderate survival functions over 400 hosts had to be sampled to achieve the same power and for gradual survival functions, no tested sample size ever achieved a power greater than 0.8 (Fig.
+\ref{fig:real_power}).
 
@@
 The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$
-for all combinations of parameters examined in this study (Fig. \ref{fig:question2}, \emph{SI} 3 Figs 4-6).  Even for
+for all combinations of parameters examined in this study (Fig. \ref{fig:question2}, Supplementary Fig. S4-S6).  Even for
 the smallest sample sizes we considered, the Likelihood Method's estimate of $LD_{50}$
@@
 (increasing coefficient of variation) as sample size decreased for all
-parameter combinations we examined (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6).
+parameter combinations we examined (Fig. \ref{fig:question2}, Supplementary Fig. S4-S6).
 
-The Adjei Method produced biased estimates of the $LD_{50}$ across nearly all parameter combinations, tending to underestimate the true value of the parameter (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6).
+The Adjei Method produced biased estimates of the $LD_{50}$ across nearly all parameter combinations, tending to underestimate the true value of the parameter (Fig.
+\ref{fig:question2}, Supplementary Fig.
+S4-S6).
 For $\mu_p = 10$, the $LD_{50}$
@@
 produced biased estimates of $LD_{50}$ across all sample sizes, with bias
-increasing as sample size decreased (Figure \ref{fig:question2}, \emph{SI} 3 Fig 4-6). The $LD_{50}$ estimates from the Adjei
-Method also showed large decreases in precision with the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:question2}, \emph{SI} 3 Fig 4-6).
+increasing as sample size decreased (Fig. \ref{fig:question2}, Supplementary Fig. S4-S6). The $LD_{50}$ estimates from the Adjei
+Method also showed large decreases in precision with the steepest survival function across all values of $\mu_p$ (Fig. \ref{fig:question2}, Supplementary Fig. S4-S6).
 
 In terms of the host survival function, the Likelihood Method gave unbiased estimates of survival function parameter $a$ when sample sizes were large, however as sample size decreased these estimates became severely biased (Fig.
-\ref{fig:question2}, \emph{SI} Fig 7 - 9) The Adjei Method produced
+\ref{fig:question2}, Supplementary Fig.
+S7 - S9) The Adjei Method produced
 biased estimates of the host survival function across all sample sizes, with consistently greater bias for steeper survival functions and higher mean parasite loads.
-(Fig \ref{fig:question2}, \emph{SI} 3 Figs 7-9).
+(Fig.
+\ref{fig:question2}, Supplementary Fig.
+S7-S9).
 
@@
 The previous authors qualitatively detected PIHM
-in 7 of the 10 datasets considered (Table \ref{table:pihm}).  The Likelihood Method parameterized
+in 7 of the 10 datasets considered (Table 1).  The Likelihood Method parameterized
 from the pre-mortality parameters of the Crofton Method detected significant
@@
 of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$).
-The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results showing that the Adjei Method has a high Type I error rate.
-Moreover, the Adjei Method estimates of the $LD_{50}$ were quite variable for the Crofton data, consistent with our simulation results that the Adjei Method $LD_{50}$ estimates could be imprecise for sample sizes of less than 1000 hosts (\emph{SI} Fig.
-4-6).
+The Adjei Method detected PIHM in 9 of the 10 datasets (Table 1), consistent with our simulation results showing that the Adjei Method has a high Type I error rate.
+Moreover, the Adjei Method estimates of the $LD_{50}$ were quite variable for the Crofton data, consistent with our simulation results that the Adjei Method $LD_{50}$ estimates could be imprecise for sample sizes of less than 1000 hosts (Supplementary Fig.
+S4-S6).
 
@@
 
-We thank the Briggs Lab group, Kuris/Lafferty Lab group, Theoretical Ecology group at UCSB, and two anonymous reviewers for helpful feedback.
+We thank the Briggs Lab group, Kuris/Lafferty Lab group, Theoretical Ecology group at University of California, Santa Barbara (United States), and two anonymous reviewers for helpful feedback.
 MW was supported by National Science Foundation Graduate Research Fellowship (Grant No.
-DGE 1144085) and the UC Regents.
-CB was supported by NIH grant 1R01GM109499 from the Ecology of Infectious Disease program.
+DGE 1144085) and the Uuniversity of California Regents (United States).
+CB was supported by National Institute of Health (United States) grant 1R01GM109499 from the Ecology of Infectious Disease program.
 
@@
 
-\newpage
+% \newpage
+
+% \newgeometry{left=1cm, right=1cm}
 
 % \renewcommand{\arraystretch}{1.2}
+% \begin{table}
 
-% \begin{table}
-%     \centering
-%     \caption{Definition of parameters and functions used in the main text}
-%     \begin{tabular}{c p{12cm}}
-%     \hline
-%     Parameter & Definition \\
-%     \hline\hline
-%     $\mu_p$ & Pre-mortality mean parasite intensity \\
-%     $k_p$   & Pre-mortality parasite aggregation \\
-%     $N_p$   & Pre-mortality host sample size \\
-%     $x$     & Number of parasites in a given host \\
-%     $g(x; \mu_p, k_p)$ & Pre-mortality negative binomial parasite distribution \\
-
-%     $h(\text{survival}; x, a, b)$ & The probability of host survival given a parasite load $x$ and logistic parameters $a$ and $b$\\
-
-%     $b / 4$ & The maximum rate of decline in host survival probability with increasing parasite intensity  \\
-%     $a$ & When $b$ is held constant a one unit increase in $a$ leads to a $1 / b$ increase in the log parasite intensity at which some percent of hosts survive\\
-%     $LD_{50}$ & $\exp(a / b)$, parasite intensity above which a host has a greater than 50\% chance of dying \\
-
-%     \end{tabular}
-%     \label{tab:params}
-% \end{table}
-
-\renewcommand{\arraystretch}{1.2}
-\begin{sidewaystable}
-
-    \caption{Comparison of the PIHM predictions of previously used host-parasite datasets to those given by the Adjei Method and the Likelihood Method.
+%     \caption{Comparison of the PIHM predictions of previously used host-parasite datasets to those given by the Adjei Method and the Likelihood Method.
 The first column specifies the identity of the dataset, the second column specifies whether or not the authors indicated that PIHM was occurring in the system based on a qualitative assessment, the third column indicates whether or not the Likelihood Method with pre-mortality parameters estimated from the Crofton Method detects significant PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
@@
 
-    \centering
-    \begin{tabular}{l  p{3cm} p{3cm} l}
+%     \centering
+%     \begin{tabular}{l  p{3cm} p{3cm} l}
 
-    \hline\hline
-    Data Set (sample size) & Author detected PIHM? & Likelihood Method?  & Adjei Method? \\
+%     \hline\hline
+%     Data Set (sample size) & Author detected PIHM? & Likelihood Method?  & Adjei Method? \\
 
-    \hline
-    Crofton, Station 1 ($n=538$) & Yes & Yes (7.27) & Yes (9.33) \\
-    Crofton, Station 2 ($n=507$) & Yes & Yes (6.92) &  Yes (14.95)\\
-    Crofton, Station 3 ($n=633$) & Yes & Yes (5.93) &  Yes (5.98) \\
-    Crofton, Station 4 ($n=486$) & No & No &  Yes (7.99) \\
-    Crofton, Station 5 ($n=276$) & No & No & Yes (10.58) \\
-    Crofton, Station 6 ($n=191$) & No & No & No \\
-    Adjei, \emph{S. tumbil} female ($n=446$) & Yes (5.7) & No & Yes (6.37) \\
-    Adjei, \emph{S. tumbil} male ($n=452$) & Yes (3.4) & Yes (3.42) & Yes (3.66)  \\
-    Adjei, \emph{S. undosquamis} female ($n=2573$) & Yes (3.2) & Yes (3.04) & Yes (3.11) \\
-    Adjei, \emph{S. undosquamis} male ($n=2440$) & Yes (1.8) & Yes (1.83) & Yes (1.78) \\
+%     \hline
+%     Crofton, Station 1 ($n=538$) & Yes & Yes (7.27) & Yes (9.33) \\
+%     Crofton, Station 2 ($n=507$) & Yes & Yes (6.92) &  Yes (14.95)\\
+%     Crofton, Station 3 ($n=633$) & Yes & Yes (5.93) &  Yes (5.98) \\
+%     Crofton, Station 4 ($n=486$) & No & No &  Yes (7.99) \\
+%     Crofton, Station 5 ($n=276$) & No & No & Yes (10.58) \\
+%     Crofton, Station 6 ($n=191$) & No & No & No \\
+%     Adjei, \emph{S. tumbil} female ($n=446$) & Yes (5.7) & No & Yes (6.37) \\
+%     Adjei, \emph{S. tumbil} male ($n=452$) & Yes (3.4) & Yes (3.42) & Yes (3.66)  \\
+%     Adjei, \emph{S. undosquamis} female ($n=2573$) & Yes (3.2) & Yes (3.04) & Yes (3.11) \\
+%     Adjei, \emph{S. undosquamis} male ($n=2440$) & Yes (1.8) & Yes (1.83) & Yes (1.78) \\
 
-    \end{tabular}
-    \label{table:pihm}
+%     \end{tabular}
+%     \label{table:pihm}
 
-\end{sidewaystable}
+% \end{table}
+
+\newpage
 
@@
 
-    \caption{A schematic representation of the iterative approach of the Crofton Method.
-\textbf{A)} The light gray shows the pre-mortality distribution that the Crofton Method is trying to estimate from the dark grey post-mortality distribution.
+    \caption{\doublespacing A schematic representation of the iterative approach of the Crofton Method.
+(A) The light gray shows the pre-mortality distribution that the Crofton Method is trying to estimate from the dark grey post-mortality distribution.
 The Crofton Method proceeds by truncating the post-mortality data at different levels ($t_i$, e.g.
 $i= 0, \dots, 5$) and finding the pre-mortality host population size ($N_p$), pre-mortality mean parasite intensity ($\mu_p$), and pre-mortality parasite aggregation ($k_p$) that best fit the truncated data.
-\textbf{B)} The parameter $N_p$ is then plotted against the truncation level $t_i$ to determine if a ``kink'' occurs in the parameter values  \citep{Lester1984}.
+(B) The parameter $N_p$ is then plotted against the truncation level $t_i$ to determine if a ``kink'' occurs in the parameter values  \citep{Lester1984}.
 This ``kink'' indicates that PIHM is occurring in the system.
-In the above example, PIHM is occurring in the system as visualized by the distinct ``kink'' at $t_3$.}
+In the above example, PIHM is occurring in the system as visualized by the distinct ``kink'' at $t_4$.}
     \label{fig:crofton}
@@
 
-    \caption{A) Five potential shapes for a host-survival functions.
+    \caption{\doublespacing The simulation results comparing the power and the Type I error of the Adjei Method and the Likelihood Method across a range of different sample sizes.
+(A) Five potential shapes for a host-survival functions.
 In the simulations we used a gradual survival function (dotted line), and moderate survival function (dashed line), and a steep survival function (solid line).
@@
 For each of these survival functions and the parameter combinations described in the main text, we tested the Type I error and power of the Likelihood (Like.) Method and Adjei Method.
-B) Gives the Type I error of each method over a range of pre-mortality sample sizes with a pre-mortality mean parasite intensity ($\mu_p$) of 50 and pre-mortality parasite aggregation ($k_p$) at 0.5.
+(B) Gives the Type I error of each method over a range of pre-mortality sample sizes with a pre-mortality mean parasite intensity ($\mu_p$) of 50 and pre-mortality parasite aggregation ($k_p$) at 0.5.
 The red line shows the pre-set significance level of 0.05.
-C) Gives the power of each method for detecting PIHM over a range of post-mortality sample sizes for $\mu_p = 50$ and $k_p = 0.5$.
+(C) Gives the power of each method for detecting PIHM over a range of post-mortality sample sizes for $\mu_p = 50$ and $k_p = 0.5$.
 In general, the Likelihood Method has higher power and lower Type I error than the Adjei Method.
-See the \emph{SI} 3 Fig 1 - 3 for Type I error and power results for all parameter combinations.}
+See the Supplementary Fig.
+S1-S3 for Type I error and power results for all parameter combinations.}
 
@@
 
-    \caption{The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
+    \caption{\doublespacing The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
 The curves were generated from 500 simulations for 10 pre-mortality sample sizes, $N_p$.
@@
 
-    \caption{Bias and precision (coefficient of variation) for the Likelihood Method (Like.) and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
+    \caption{\doublespacing Bias and precision (coefficient of variation) for the Likelihood Method (Like.) and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
 As the coefficient of variation increases, precision decreases.
@@
 The figure shows the simulations for three different host survival functions (gradual, moderate, and steep), each with the same $LD_{50}$.
-Bias and precision results of $LD_{50}$ and $a$ for all other parameter combinations can be found in \emph{SI} 3 Fig 4 - 9.}
+Bias and precision results of $LD_{50}$ and $a$ for all other parameter combinations can be found in Supplementary Fig.
+S4-S9.}
 
@@
 
+% SUPPLEMENTARY FIGURE LEGENDS
+
+\setcounter{figure}{0}
+\renewcommand\thefigure{S\arabic{figure}}
+
+\begin{figure}
+
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu10}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
+
+    \caption{\doublespacing The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$.
+The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
+The red line gives the the pre-set type I error rate of $\alpha = 0.05$.
+The second column gives the power of a given method to detect PIHM when it is actually occurring.
+}
+    \label{fig:typeI10}
+
+\end{figure}
+
+\begin{figure}
+
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu50}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
+
+    \caption{\doublespacing The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$.
+The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
+The red line gives the the pre-set type I error rate of $\alpha = 0.05$.
+The second column gives the power of a given method to detect PIHM when it is actually occurring.
+}
+    \label{fig:typeI50}
+
+\end{figure}
+
+\begin{figure}
+
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu100}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
+
+    \caption{\doublespacing The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$.
+The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
+The red line gives the the pre-set type I error rate of $\alpha = 0.05$.
+The second column gives the power of a given method to detect PIHM when it is actually occurring.
+}
+    \label{fig:typeI100}
+
+\end{figure}
+
+\begin{figure}
+
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu10}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
+
+    \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
+The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}
+
+    \label{fig:biasld50_10}
+
+\end{figure}
+
+\begin{figure}
+
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu50}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
+
+    \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
+The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}
+
+    \label{fig:biasld50_50}
+
+\end{figure}
+
+\begin{figure}
+
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu100}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
+
+    \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
+The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}
+
+    \label{fig:biasld50_100}
+
+\end{figure}
+
+\begin{figure}
+
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu10}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
+
+    \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+The first column gives the bias of each method's $a$ estimate over 150 simulations.
+The second column gives the precision of each method's $a$ estimate over 150 simulations.}
+
+    \label{fig:biasa_10}
+
+\end{figure}
+
+\begin{figure}
+
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu50}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
+
+    \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+The first column gives the bias of each method's $a$ estimate over 150 simulations.
+The second column gives the precision of each method's $a$ estimate over 150 simulations.}
+
+    \label{fig:biasa_50}
+
+\end{figure}
+
+\begin{figure}
+
+    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu100}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
+
+    \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+The first column gives the bias of each method's $a$ estimate over 150 simulations.
+The second column gives the precision of each method's $a$ estimate over 150 simulations.}
+
+    \label{fig:biasa_100}
+
+\end{figure}
+
+\begin{figure}
+    %\includegraphics[width=0.5\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/compare_crofton.pdf}
+
+    \captionsetup{justification=raggedright, singlelinecheck=false}
+    \caption{\doublespacing A comparison of this paper's implementation (solid lines, circles) of the Crofton Method with the results given in \cite{Crofton1971a} (dashed lines, diamonds).
+(A) compares the predicted number of hosts in a population pre-mortality ($N_p$).
+(B) compares the predicted parasite aggregation pre-mortality ($k_p$).
+(C) compares the $\chi^2$ statistic for each implementation.
+Three of the 6 stations fit by \citeauthor{Crofton1971a} are shown here and all show that our implementation gives very similar results to those given by \citeauthor{Crofton1971a}.\\\hspace{\textwidth}Crofton, H.D.
+1971.
+A quantitative approach to parasitism.
+Parasitology.
+62, 179–193.}
+    \label{fig:crof_test}
+
+\end{figure}
+
 \end{document}

### 2015-08-15 docs/supplementary_material.tex
Edited manuscript and supplementary material to meet journal specs

@@
 \noindent
-\Large{\textbf{Supporting Information}}
+\Large{\textbf{Supplementary Data}}
 
@@
 
-\section*{SI 1: Implementation of the Crofton Method}
+\section{Implementation of the Crofton Method}
 
@@
 We provide an implementation and unit tests of the Crofton Method in  Supplementary Information 4.
-Figure \ref{fig:crof_test} visually shows that our implementation of the Crofton Method agrees with results previously published by \cite{Crofton1971a}.
+Figure S\ref{fig:crof_test} visually shows that our implementation of the Crofton Method agrees with results previously published by \cite{Crofton1971a}.
 
-\section*{SI 2: Implementation of the Adjei Method}
+\section{Implementation of the Adjei Method}
 
@@
 subjectivity (i.e. which bins should you use?) to a method that already has
-serious potential issues.  In this analysis, we always assume the Adjei Method is not binning the data, though we provide code for applying the binning method in SI 4.
+serious potential issues.  In this analysis, we always assume the Adjei Method is not binning the data, though we provide code for applying the binning method in Supplementary Data S3.
 
-\section*{SI 3: Additional Figures}
-
-See Figures 1 - 9.
-
-\section*{SI 4: Code and unit tests for estimating parasite-induced host mortality}
+\section{Code and unit tests for estimating parasite-induced host mortality}
 
@@
 
-\begin{figure}
+% \renewcommand\thefigure{S\arabic{figure}}
 
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu10}
+% \begin{figure}
 
-    \caption{The type I error rate and the power of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$.
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu10}
+
+%     \caption{\doublespacing The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$.
 The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
@@
 }
-    \label{fig:typeI10}
+%     \label{fig:typeI10}
 
-\end{figure}
+% \end{figure}
 
-\begin{figure}
+% \begin{figure}
 
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu50}
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu50}
 
-    \caption{The type I error rate and the power of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$.
+%     \caption{\doublespacing The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$.
 The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
@@
 }
-    \label{fig:typeI50}
+%     \label{fig:typeI50}
 
-\end{figure}
+% \end{figure}
 
-\begin{figure}
+% \begin{figure}
 
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu100}
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu100}
 
-    \caption{The type I error rate and the power of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$.
+%     \caption{\doublespacing The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$.
 The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
@@
 }
-    \label{fig:typeI100}
+%     \label{fig:typeI100}
 
-\end{figure}
+% \end{figure}
 
-\begin{figure}
+% \begin{figure}
 
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu10}
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu10}
 
-    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+%     \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
 The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
@@
 
-    \label{fig:biasld50_10}
+%     \label{fig:biasld50_10}
 
-\end{figure}
+% \end{figure}
 
-\begin{figure}
+% \begin{figure}
 
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu50}
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu50}
 
-    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+%     \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
 The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
@@
 
-    \label{fig:biasld50_50}
+%     \label{fig:biasld50_50}
 
-\end{figure}
+% \end{figure}
 
-\begin{figure}
+% \begin{figure}
 
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu100}
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu100}
 
-    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+%     \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
 The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
@@
 
-    \label{fig:biasld50_100}
+%     \label{fig:biasld50_100}
 
-\end{figure}
+% \end{figure}
 
-\begin{figure}
+% \begin{figure}
 
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu10}
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu10}
 
-    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+%     \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
 The first column gives the bias of each method's $a$ estimate over 150 simulations.
@@
 
-    \label{fig:biasa_10}
+%     \label{fig:biasa_10}
 
-\end{figure}
+% \end{figure}
 
-\begin{figure}
+% \begin{figure}
 
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu50}
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu50}
 
-    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+%     \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
 The first column gives the bias of each method's $a$ estimate over 150 simulations.
@@
 
-    \label{fig:biasa_50}
+%     \label{fig:biasa_50}
 
-\end{figure}
+% \end{figure}
 
-\begin{figure}
+% \begin{figure}
 
-    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu100}
+%     \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu100}
 
-    \caption{The bias and the precision of the Likelihood Method (solid line) and the Adjei Method (dashed lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+%     \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
 The first column gives the bias of each method's $a$ estimate over 150 simulations.
@@
 
-    \label{fig:biasa_100}
+%     \label{fig:biasa_100}
 
-\end{figure}
+% \end{figure}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/compare_crofton.pdf}
-    \caption{A comparison of this paper's implementation (solid line, circles) of the Crofton Method with the results given in \cite{Crofton1971a} (dashed line, diamonds).
-Figure A compares the predicted number of hosts in a population pre-mortality ($N_p$).
-Figure B compares the predicted parasite aggregation pre-mortality ($k_p$).
-Figure C compares the $\chi^2$ statistic for each implementation.
-Three of the 6 stations fit by \citeauthor{Crofton1971a} are shown here and all show that our implementation gives very similar results to those given by \citeauthor{Crofton1971a}.}
-    \label{fig:crof_test}
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=0.5\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/compare_crofton.pdf}
+%     \caption{\doublespacing A comparison of this paper's implementation (solid lines, circles) of the Crofton Method with the results given in Crofton (1971) (dashed lines, diamonds).
+(A) compares the predicted number of hosts in a population pre-mortality ($N_p$).
+(B) compares the predicted parasite aggregation pre-mortality ($k_p$).
+(C) compares the $\chi^2$ statistic for each implementation.
+Three of the 6 stations fit by Crofton are shown here and all show that our implementation gives very similar results to those given by Crofton.\\\hspace{\textwidth}Crofton, H.D.
+1971.
+A quantitative approach to parasitism.
+Parasitology.
+62, 179–193.}
+%     \label{fig:crof_test}
 
-\end{figure}
+% \end{figure}
 

### 2015-08-15 docs/parasite_mortality_manuscript.tex
Made final figure and manuscript updates before revised submission

@@
 
-% \begingroup
-\setlength{\parindent}{0in}
-\renewcommand{\arraystretch}{2}
-% \renewcommand{\tabcolsep}{0}
-%\renewcommand{\tabrowsep}{1}
-\begin{tabular}{p{3cm} p{10cm}}
+\noindent
+\textbf{Title:} Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations
 
-\textbf{Title:} & Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations \\
+\bigskip
+\noindent
+\textbf{Authors:} Mark Q. Wilber, Sara B. Weinstein and Cheryl J. Briggs
 
-\end{tabular}
+\bigskip
+\noindent
+\textbf{Affiliations:} 1) Department of Ecology, Evolution and Marine Biology, University of California,
+Santa Barbara, Santa Barbara California, United States of America
 
-\begin{tabular}{p{3cm} p{10cm}}
+\bigskip
+\noindent
+\textbf{Corresponding author:}\\Mark Q. Wilber\\University of California, Santa Barbara\\
+Department of Ecology, Evolution, and Marine Biology\\
+2111 Noble Hall\\
+Santa Barbara, CA 93106\\
+mark.wilber@lifesci.ucsb.edu
 
-\textbf{Authors:} & Mark Q. Wilber, Sara B. Weinstein and Cheryl J. Briggs \\
-
-\end{tabular}
-
-\begin{tabular}{p{3cm} p{10cm}}
-
-\textbf{Affiliations:} &
-1) Department of Ecology, Evolution and Marine Biology, University of California,
-Santa Barbara, Santa Barbara California, United States of America \\
-
-\end{tabular}
-
-\begin{tabular}{p{3cm} p{10cm}}
-
-\textbf{Corresponding \newline author:} &
-Mark Wilber \newline
-University of California, Santa Barbara \newline
-Department of Ecology, Evolution, and Marine Biology \newline
-2111 Noble Hall \newline
-Santa Barbara, CA 93106 \newline
-505-321-5381 \newline
-mark.wilber@lifesci.ucsb.edu \newline
-
-\end{tabular}
-
+\bigskip\noindent
 \textbf{Note}: Supplementary data associated with this article
-
-% \renewcommand{\arraystretch}{1}
-% \begin{tabular}{l l}
-
-% \textbf{Words - Abstract:} & 293 \\
-% \textbf{Words - Main Text:} & 4988 \\
-% \textbf{References:} & 50 \\
-% \textbf{Figures:} & 5 \\
-% \textbf{Tables:} & 1
-
-% \end{tabular}
 
@@
 \ref{fig:question2}, Supplementary Fig.
-S7 - S9) The Adjei Method produced
+S7-S9) The Adjei Method produced
 biased estimates of the host survival function across all sample sizes, with consistently greater bias for steeper survival functions and higher mean parasite loads.

### 2015-08-15 docs/supplementary_material.tex
A few grammar changes.
Made final figure and manuscript updates before revised submission

@@
 
-We provide an implementation and unit tests of the Crofton Method in  Supplementary Information 4.
-Figure S\ref{fig:crof_test} visually shows that our implementation of the Crofton Method agrees with results previously published by \cite{Crofton1971a}.
+We provide an implementation and unit tests of the Crofton Method in  Supplementary Data S3.
+Figure S10 visually shows that our implementation of the Crofton Method agrees with results previously published by \cite{Crofton1971a}.
 
@@
 step is to estimate the parameters of the pre-mortality host-parasite
-distribution using the Crofton Method (see \emph{SI} 1).  The three parameters estimated are the
+distribution using the Crofton Method (see Supplementary Data S1).  The three parameters estimated are the
 total number of hosts before mortality $N_p$,  the mean number of parasites per
@@
 $k_p$ is large parasites are more evenly distributed across hosts \citep{Wilson2002}.
-The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][ and in \emph{SI} 1]{Royce1990,Lester1984} and we provide a tested implementation of the method in \emph{SI} 4.
+The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][ and in Supplementary Data S1]{Royce1990,Lester1984} and we provide a tested implementation of the method in Supplementary Data S3.
 
@@
 \end{align*}
-If $a_2 - a_1 = 1$, than the change in log parasite intensity at which $p$ percentage of hosts survive is $1 / b$.
+If $a_2 - a_1 = 1$, then the change in log parasite intensity at which $p$ percentage of hosts survive is $1 / b$.
 

### 2015-08-25 docs/parasite_mortality_manuscript.tex
Converting manuscript and supplementary to .doc

@@
 
-\abstract{
+\textbf{Abstract}
 
@@
 
-}
-
-\section{Introduction}
+\textbf{1 Introduction}
 
@@
 fitting a negative binomial distribution to hosts with lower and lower parasite intensities, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data (Fig.
-\ref{fig:crofton}, see Supplementary Data S1 for additional detail).
+1, see Supplementary Data S1 for additional detail).
 
@@
 In these methods, PIHM can only be identified by visually examining plots of the pre-mortality parameters predicted by the Crofton Method and determining whether they show a ``kink'' over a range of truncation values \citep[Fig.
-\ref{fig:crofton}B;][]{Lester1984,Ferguson2011}.
+1B;][]{Lester1984,Ferguson2011}.
 These qualitative criteria makes it difficult to compare PIHM between studies and a more rigorous and quantitative method is needed to both detect and quantify host mortality.
@@
 
-\section{Materials and methods}
+\textbf{2 Materials and methods}
 
-\subsection{A novel, likelihood-based method for estimating PIHM}
+\textbf{2.1 A novel, likelihood-based method for estimating PIHM}
 
@@
 
-The Likelihood Method then assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \boldsymbol{\phi})$, which specifies the probability of a host having $x$ parasites before mortality occurs.
-$\boldsymbol{\phi}$ is a vector of parameters that describes the shape of this distribution.
-The probability of a host surviving with $x$ parasites from infection until sampling is given by the host survival function $h(\text{survival} ; x, \boldsymbol{\theta})$ where $\boldsymbol{\theta}$ specifies any additional parameters needed to define the host survival function.
+The Likelihood Method then assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \phi)$, which specifies the probability of a host having $x$ parasites before mortality occurs.
+$\phi$ is a vector of parameters that describes the shape of this distribution.
+The probability of a host surviving with $x$ parasites from infection until sampling is given by the host survival function $h(\text{survival} ; x, \theta)$ where $\theta$ specifies any additional parameters needed to define the host survival function.
 
@@
 
-$P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \boldsymbol{\theta})$, $P(x)$ is the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})$.
-Therefore, equation \ref{eq:concept} can be written as
+$P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \theta)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \phi)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \theta)  * g(x; \phi)$.
+Therefore, equation 1 can be written as
 
 \begin{equation}
-    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}{\sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}
+    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \theta)  * g(x; \phi)}{\sum_{x=0}^{\infty} h(\text{survival}; x, \theta)  * g(x; \phi)}
     \label{eq:dist}
@@
 
-Using this probability distribution, one can then find the parameters $\boldsymbol{\theta}$
-and $\boldsymbol{\phi}$ that maximize the likelihood of an observed host-parasite dataset.
+Using this probability distribution, one can then find the parameters $\theta$
+and $\phi$ that maximize the likelihood of an observed host-parasite dataset.
 To estimate the significance of PIHM in a host-parasite system, a likelihood
 ratio test can be used in which the full model is given by equation
-\ref{eq:dist} and the reduced model is given by the pre-mortality distribution
-$g(x; \boldsymbol{\phi})$.  If PIHM is not significant in the system, the resulting
+2 and the reduced model is given by the pre-mortality distribution
+$g(x; \phi)$.  If PIHM is not significant in the system, the resulting
 likelihood ratio statistic should approximately follow a $\chi^2$ distribution
@@
 
-The parameterization of equation \ref{eq:dist} depends on the parasite system of interest.
-Here, we assume that the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ follows a negative binomial
+The parameterization of equation \ref{eq:dist} depends on the parasite system of interest.  Here, we assume that the pre-mortality parasite distribution $g(x; \phi)$ follows a negative binomial
 distribution with two parameters mean parasite intensity ($\mu_p$) and
@@
 The function for $h(\text{survival}; x,
-\boldsymbol{\theta})$ is also system specific.  Many
+\theta)$ is also system specific.  Many
 theoretical models of parasite-induced host mortality assume that the parasite-induced death rate of hosts is a linear function of parasite intensity
@@
 infection intensity above which a host has greater than 50\% probability of dying.
-Equation \ref{eq:logistic} is commonly used in toxicology and has the useful properties of being bounded between 0 and 1 and being differentiable for all $x$ \citep{Collet2002}.
+Equation 3 is commonly used in toxicology and has the useful properties of being bounded between 0 and 1 and being differentiable for all $x$ \citep{Collet2002}.
 That being said, it is phenomenological and is used simply because it tends to fit survival data.
@@
 to adopt the same host-survival function to facilitate comparison.  When
-applying the Likelihood Method to other systems more mechanistic host-survival functions can be used in place of equation \ref{eq:logistic}.
+applying the Likelihood Method to other systems more mechanistic host-survival functions can be used in place of equation 3.
 
-\subsection{Evaluating the Adjei and Likelihood Methods}
+\textbf{2.2 Evaluating the Adjei and Likelihood Methods}
 
@@
 We next chose values of $a$ and $b$ for the host survival function and calculated the probability of survival
-for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, to simulate the period in which hosts died due to infection, for each host we drew a random number from a uniform distribution
+for all $N_p$ hosts using equation 3.  Then, to simulate the period in which hosts died due to infection, for each host we drew a random number from a uniform distribution
 between 0 and 1 and if the calculated host survival probability was less than this random
@@
 We compared the two methods using three different mean parasite intensity values ($\mu_p$ = 10, 50, 100) and three different host survival functions (gradual, moderate, and steep decreases in the host survival with increasing parasite intensity, Fig.
-\ref{fig:question1}A).
+2A).
 For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.
@@
 
-\subsection{Application to real data}
+\textbf{2.3 Application to real data}
 
@@
 
-\section{Results}
+\textbf{3 Results}
 
-\subsection{Question 1: Detecting presence of PIHM}
+\textbf{3.1 Question 1: Detecting presence of PIHM}
 
@@
 system was close to unity for larger sample sizes and tended to
-decrease as sample size decreased for all survival functions (Fig. \ref{fig:question1}C; Supplementary Fig. S1-S3).  The Likelihood Method had a power close to
+decrease as sample size decreased for all survival functions (Fig. 2C; Supplementary Fig. S1-S3).  The Likelihood Method had a power close to
 unity for all parameter combinations and sample sizes considered.  With gradual
-survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (Fig. \ref{fig:question1}C, Supplementary Fig. S1-S3).
+survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (Fig. 2C, Supplementary Fig. S1-S3).
 
@@
 PIHM) for all parameter combinations that we
-considered (Fig. \ref{fig:question1}B; Supplementary Fig. S1-S3).  This method also showed the unintuitive pattern of decreasing Type I error
+considered (Fig. 2B; Supplementary Fig. S1-S3).  This method also showed the unintuitive pattern of decreasing Type I error
 rate with decreasing sample size.
@@
 error rate at or near the pre-set level of 0.05 for all parameter combinations
-and sample sizes considered (Fig. \ref{fig:question1}B; Supplementary Fig. S1-S3).
+and sample sizes considered (Fig. 2B; Supplementary Fig. S1-S3).
 
@@
 For steep survival curves, PIHM could be detected with a power of greater than 0.8 from a sample of less than 100 hosts (Fig.
-\ref{fig:real_power}).
+3).
 However, for moderate survival functions over 400 hosts had to be sampled to achieve the same power and for gradual survival functions, no tested sample size ever achieved a power greater than 0.8 (Fig.
-\ref{fig:real_power}).
+3).
 
-\subsection{Question 2: Estimating the $LD_{50}$ and survival function}
+\textbf{3.2 Question 2: Estimating the $LD_{50}$ and survival function}
 
 The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$
-for all combinations of parameters examined in this study (Fig. \ref{fig:question2}, Supplementary Fig. S4-S6).  Even for
+for all combinations of parameters examined in this study (Fig. 4, Supplementary Fig. S4-S6).  Even for
 the smallest sample sizes we considered, the Likelihood Method's estimate of $LD_{50}$
@@
 (increasing coefficient of variation) as sample size decreased for all
-parameter combinations we examined (Fig. \ref{fig:question2}, Supplementary Fig. S4-S6).
+parameter combinations we examined (Fig. 4, Supplementary Fig. S4-S6).
 
 The Adjei Method produced biased estimates of the $LD_{50}$ across nearly all parameter combinations, tending to underestimate the true value of the parameter (Fig.
-\ref{fig:question2}, Supplementary Fig.
+4, Supplementary Fig.
 S4-S6).
@@
 produced biased estimates of $LD_{50}$ across all sample sizes, with bias
-increasing as sample size decreased (Fig. \ref{fig:question2}, Supplementary Fig. S4-S6). The $LD_{50}$ estimates from the Adjei
-Method also showed large decreases in precision with the steepest survival function across all values of $\mu_p$ (Fig. \ref{fig:question2}, Supplementary Fig. S4-S6).
+increasing as sample size decreased (Fig. 4, Supplementary Fig. S4-S6). The $LD_{50}$ estimates from the Adjei
+Method also showed large decreases in precision with the steepest survival function across all values of $\mu_p$ (Fig. 4, Supplementary Fig. S4-S6).
 
 In terms of the host survival function, the Likelihood Method gave unbiased estimates of survival function parameter $a$ when sample sizes were large, however as sample size decreased these estimates became severely biased (Fig.
-\ref{fig:question2}, Supplementary Fig.
+4, Supplementary Fig.
 S7-S9) The Adjei Method produced
@@
 (Fig.
-\ref{fig:question2}, Supplementary Fig.
+4, Supplementary Fig.
 S7-S9).
 
-\subsection{Application to real data}
+\textbf{3.3 Application to real data}
 
@@
 
-\section{Discussion}
+\textbf{4 Discussion}
 
@@
 
-If one has evidence that the pre-mortality is not negative binomial, the generality of our method easily allows another distribution to be specified for $g(x, \boldsymbol{\phi})$.
-For example, one could use the
+If one has evidence that the pre-mortality is not negative binomial, the generality of our method easily allows another distribution to be specified for $g(x, \phi)$. For example, one could use the
 resulting stationary host-parasite distribution from a stochastic host-parasite
 model without parasite-induced host mortality \citep{Anderson1982a} to specify
-the form of $g(x, \boldsymbol{\phi})$ and then apply the techniques discussed
+the form of $g(x, \phi)$ and then apply the techniques discussed
 in this paper to detect PIHM.  The general requirement
@@
 
-\section*{Acknowledgments}
+\textbf{Acknowledgments}
 
@@
 MW was supported by National Science Foundation Graduate Research Fellowship (Grant No.
-DGE 1144085) and the Uuniversity of California Regents (United States).
+DGE 1144085) and the University of California Regents (United States).
 CB was supported by National Institute of Health (United States) grant 1R01GM109499 from the Ecology of Infectious Disease program.
@@
 
-\singlespacing
-\bibliographystyle{int_j_para.bst}
-\bibliography{parasite_host_mortality}
+% \singlespacing
+% \bibliographystyle{int_j_para.bst}
+% \bibliography{parasite_host_mortality}
 
@@
 
-% \newgeometry{left=1cm, right=1cm}
-
-% \renewcommand{\arraystretch}{1.2}
-% \begin{table}
-
-%     \caption{Comparison of the PIHM predictions of previously used host-parasite datasets to those given by the Adjei Method and the Likelihood Method.
-The first column specifies the identity of the dataset, the second column specifies whether or not the authors indicated that PIHM was occurring in the system based on a qualitative assessment, the third column indicates whether or not the Likelihood Method with pre-mortality parameters estimated from the Crofton Method detects significant PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
-If a method detected significant PIHM the predicted $LD_{50}$ is given in parentheses.}
-
-%     \centering
-%     \begin{tabular}{l  p{3cm} p{3cm} l}
-
-%     \hline\hline
-%     Data Set (sample size) & Author detected PIHM? & Likelihood Method?  & Adjei Method? \\
-
-%     \hline
-%     Crofton, Station 1 ($n=538$) & Yes & Yes (7.27) & Yes (9.33) \\
-%     Crofton, Station 2 ($n=507$) & Yes & Yes (6.92) &  Yes (14.95)\\
-%     Crofton, Station 3 ($n=633$) & Yes & Yes (5.93) &  Yes (5.98) \\
-%     Crofton, Station 4 ($n=486$) & No & No &  Yes (7.99) \\
-%     Crofton, Station 5 ($n=276$) & No & No & Yes (10.58) \\
-%     Crofton, Station 6 ($n=191$) & No & No & No \\
-%     Adjei, \emph{S. tumbil} female ($n=446$) & Yes (5.7) & No & Yes (6.37) \\
-%     Adjei, \emph{S. tumbil} male ($n=452$) & Yes (3.4) & Yes (3.42) & Yes (3.66)  \\
-%     Adjei, \emph{S. undosquamis} female ($n=2573$) & Yes (3.2) & Yes (3.04) & Yes (3.11) \\
-%     Adjei, \emph{S. undosquamis} male ($n=2440$) & Yes (1.8) & Yes (1.83) & Yes (1.78) \\
-
-%     \end{tabular}
-%     \label{table:pihm}
-
-% \end{table}
-
 \newpage
 
-\begin{figure}
+%\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/concept_crofton_method}
 
-    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/concept_crofton_method}
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{\doublespacing A schematic representation of the iterative approach of the Crofton Method.
+\textbf{Figure 1}: A schematic representation of the iterative approach of the Crofton Method.
 (A) The light gray shows the pre-mortality distribution that the Crofton Method is trying to estimate from the dark grey post-mortality distribution.
@@
 This ``kink'' indicates that PIHM is occurring in the system.
-In the above example, PIHM is occurring in the system as visualized by the distinct ``kink'' at $t_4$.}
-    \label{fig:crofton}
+In the above example, PIHM is occurring in the system as visualized by the distinct ``kink'' at $t_4$.
 
-\end{figure}
-
-\begin{figure}
-
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{\doublespacing The simulation results comparing the power and the Type I error of the Adjei Method and the Likelihood Method across a range of different sample sizes.
+\textbf{Figure 2}: The simulation results comparing the power and the Type I error of the Adjei Method and the Likelihood Method across a range of different sample sizes.
 (A) Five potential shapes for a host-survival functions.
@@
 See the Supplementary Fig.
-S1-S3 for Type I error and power results for all parameter combinations.}
+S1-S3 for Type I error and power results for all parameter combinations.
 
-\label{fig:question1}
+%\includegraphics[width=\textwidth]{/users/mqwilber/repos/parasite_mortality/results/figure3_for_manuscript.pdf}
 
-\end{figure}
-
-\begin{figure}
-    %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure3_for_manuscript.pdf}
-
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{\doublespacing The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
+\textbf{Figure 3}: The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
 The curves were generated from 500 simulations for 10 pre-mortality sample sizes, $N_p$.
 The vertical, dotted-dashed lines indicate the sample size at which the power for the Likelihood Method with steep and moderate survival functions is 0.8 (75 hosts for steep functions and 408 for moderate functions).
-The Likelihood Method with a gradual survival function never has a power above 0.8.}
-
-    \label{fig:real_power}
-
-\end{figure}
-
-\begin{figure}
+The Likelihood Method with a gradual survival function never has a power above 0.8.
 
@@
 
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{\doublespacing Bias and precision (coefficient of variation) for the Likelihood Method (Like.) and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
+\textbf{Figure 4}: Bias and precision (coefficient of variation) for the Likelihood Method (Like.) and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
 As the coefficient of variation increases, precision decreases.
@@
 Bias and precision results of $LD_{50}$ and $a$ for all other parameter combinations can be found in Supplementary Fig.
-S4-S9.}
-
-    \label{fig:question2}
-
-\end{figure}
-
-% SUPPLEMENTARY FIGURE LEGENDS
-
-\setcounter{figure}{0}
-\renewcommand\thefigure{S\arabic{figure}}
-
-\begin{figure}
+S4-S9.
 
@@
 
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{\doublespacing The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$.
+\textbf{Figure S1}: The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$.
 The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
@@
 The second column gives the power of a given method to detect PIHM when it is actually occurring.
-}
-    \label{fig:typeI10}
-
-\end{figure}
-
-\begin{figure}
 
@@
 
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{\doublespacing The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$.
+\textbf{Figure S2}: The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$.
 The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
@@
 The second column gives the power of a given method to detect PIHM when it is actually occurring.
-}
-    \label{fig:typeI50}
-
-\end{figure}
-
-\begin{figure}
 
@@
 
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{\doublespacing The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$.
+\textbf{Figure S3}: The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$.
 The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
@@
 The second column gives the power of a given method to detect PIHM when it is actually occurring.
-}
-    \label{fig:typeI100}
-
-\end{figure}
-
-\begin{figure}
 
@@
 
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+\textbf{Figure S4}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
 The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
-The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}
-
-    \label{fig:biasld50_10}
-
-\end{figure}
-
-\begin{figure}
+The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.
 
@@
 
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+\textbf{Figure S5}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
 The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
-The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}
-
-    \label{fig:biasld50_50}
-
-\end{figure}
-
-\begin{figure}
+The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.
 
@@
 
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+\textbf{Figure S6}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
 The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
-The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.}
-
-    \label{fig:biasld50_100}
-
-\end{figure}
-
-\begin{figure}
+The second column gives the precision of each method's $LD_{50}$ estimate over 150 simulations.
 
@@
 
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+\textbf{Figure S7}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
 The first column gives the bias of each method's $a$ estimate over 150 simulations.
-The second column gives the precision of each method's $a$ estimate over 150 simulations.}
-
-    \label{fig:biasa_10}
-
-\end{figure}
-
-\begin{figure}
+The second column gives the precision of each method's $a$ estimate over 150 simulations.
 
@@
 
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-
-    \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+\textbf{Figure S8}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
 The first column gives the bias of each method's $a$ estimate over 150 simulations.
-The second column gives the precision of each method's $a$ estimate over 150 simulations.}
-
-    \label{fig:biasa_50}
-
-\end{figure}
-
-\begin{figure}
+The second column gives the precision of each method's $a$ estimate over 150 simulations.
 
@@
 
-    \captionsetup{justification=raggedright, singlelinecheck=false}
+\textbf{Figure S9}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+The first column gives the bias of each method's $a$ estimate over 150 simulations.
+The second column gives the precision of each method's $a$ estimate over 150 simulations.
 
-    \caption{\doublespacing The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
-The first column gives the bias of each method's $a$ estimate over 150 simulations.
-The second column gives the precision of each method's $a$ estimate over 150 simulations.}
-
-    \label{fig:biasa_100}
-
-\end{figure}
-
-\begin{figure}
 %\includegraphics[width=0.5\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/compare_crofton.pdf}
 
-    \captionsetup{justification=raggedright, singlelinecheck=false}
-    \caption{\doublespacing A comparison of this paper's implementation (solid lines, circles) of the Crofton Method with the results given in \cite{Crofton1971a} (dashed lines, diamonds).
+\textbf{Figure S10}: A comparison of this paper's implementation (solid lines, circles) of the Crofton Method with the results given in \cite{Crofton1971a} (dashed lines, diamonds).
 (A) compares the predicted number of hosts in a population pre-mortality ($N_p$).
@@
 (C) compares the $\chi^2$ statistic for each implementation.
-Three of the 6 stations fit by \citeauthor{Crofton1971a} are shown here and all show that our implementation gives very similar results to those given by \citeauthor{Crofton1971a}.\\\hspace{\textwidth}Crofton, H.D.
-1971.
-A quantitative approach to parasitism.
-Parasitology.
-62, 179–193.}
-    \label{fig:crof_test}
+Three of the 6 stations fit by \citeauthor{Crofton1971a} are shown here and all show that our implementation gives very similar results to those given by \citeauthor{Crofton1971a}.
 
-\end{figure}
+\bigskip
+
+Crofton, H.D. 1971. A quantitative approach to parasitism. Parasitology. 62, 179–193.
+
+\textbf{References}
 

### 2015-08-25 docs/supplementary_material.tex
Converting manuscript and supplementary to .doc

@@
 \noindent
-\Large{\textbf{Supplementary Data}}
+\textbf{Supplementary Data}
 
-\normalsize
-
-\section{Implementation of the Crofton Method}
+\textbf{1 Implementation of the Crofton Method}
 
@@
 Second, choose a truncation value ($t_2$) such that $t_2 <
-p_{max}$. Truncate the data such that $\text{data}_\text{truncated} <= t_2$ and repeat the above
+p_{max}$. Truncate the data such that $\text{data}_\text{truncated} \leq t_2$ and repeat the above
 iterative procedure to calculate another set of parameters $(N_{p2}, \mu_{p2}, k_{p2})$
@@
 
-\section{Implementation of the Adjei Method}
+\textbf{2 Implementation of the Adjei Method}
 
@@
 
-By taking the first and second derivatives of equation \ref{eq:logistic}, one can easily find that the maximum rate of decline in host survival probability with increasing parasite intensity occurs at the $LD_{50}$ and has a value of $b / 4$.
+By taking the first and second derivatives of equation 1, one can easily find that the maximum rate of decline in host survival probability with increasing parasite intensity occurs at the $LD_{50}$ and has a value of $b / 4$.
 This is in many ways analogous to the parasite pathogenicity parameter $\alpha$ given in classic macroparasite models, which specifies the slope of the linear relationship between between host death rate and parasite intensity \citep{AndersonandMay1978, Isham1995}.
 The parameter $a$ is easily interpreted by holding $b$ constant and looking at how a one unit change in $a$ affects the log parasite intensity at which some percentage $p$ of hosts experience mortality.
-Letting $a_1$ and $a_2$ be two different values of $a$ and $x_1$ and $x_2$ be two different parasite intensities, a bit of rearranging of equation \ref{eq:logistic} gives.
+Letting $a_1$ and $a_2$ be two different values of $a$ and $x_1$ and $x_2$ be two different parasite intensities, a bit of rearranging of equation 1 gives
 
-\begin{align*}
-    \log{\frac{p}{1 - p}} &= a_1 - b \log{x_1}\\
-    - \log{\frac{p}{1 - p}} &= a_2 - b \log{x_2}\\
-    0 &= a_1 - a_2 - b\log{x_1} + b\log{x_2}\\
-    a_2 - a_1 &= b(\log{x_2} - \log{x_1})
-\end{align*}
+$$\log{\frac{p}{1 - p}} = a_1 - b \log{x_1}$$
+$$- \log{\frac{p}{1 - p}} = a_2 - b \log{x_2}$$
+$$0 = a_1 - a_2 - b\log{x_1} + b\log{x_2}$$
+$$a_2 - a_1 = b(\log{x_2} - \log{x_1})$$
+
 If $a_2 - a_1 = 1$, then the change in log parasite intensity at which $p$ percentage of hosts survive is $1 / b$.
 
-To estimate the parameters in equation \ref{eq:logistic}, the Adjei Method first
+To estimate the parameters in equation 1, the Adjei Method first
 calculates the expected number of hosts with a given parasite load $x$ by using
@@
 \citep{Adjei1986}.
-After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation \ref{eq:logistic}.
+After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation 1.
 This model provides estimates for parameters $a$, $b$ and $LD_{50}$.
@@
 
-\section{Code and unit tests for estimating parasite-induced host mortality}
+\textbf{3 Code and unit tests for estimating parasite-induced host mortality}
 
@@
 
-\singlespacing
-\bibliographystyle{int_j_para.bst}
-\bibliography{parasite_host_mortality}
+\textbf{References}
+
+% \singlespacing
+% \bibliographystyle{int_j_para.bst}
+% \bibliography{parasite_host_mortality}
 

### 2015-08-26 docs/figures_tables/table1_IJPara15_261.tex
Made word versions for all the documents

@@
 
+\textbf{Table 1}: Comparison of the PIHM predictions of previously used host-parasite datasets to those given by the Adjei Method and the Likelihood Method.
+The first column specifies the identity of the dataset, the second column specifies whether or not the authors indicated that PIHM was occurring in the system based on a qualitative assessment, the third column indicates whether or not the Likelihood Method with pre-mortality parameters estimated from the Crofton Method detects significant PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
+If a method detected significant PIHM the predicted $LD_{50}$ is given in parentheses.
+
 \begin{table}
-
-    \caption{Comparison of the PIHM predictions of previously used host-parasite datasets to those given by the Adjei Method and the Likelihood Method.
-The first column specifies the identity of the dataset, the second column specifies whether or not the authors indicated that PIHM was occurring in the system based on a qualitative assessment, the third column indicates whether or not the Likelihood Method with pre-mortality parameters estimated from the Crofton Method detects significant PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.
-If a method detected significant PIHM the predicted $LD_{50}$ is given in parentheses.}
 
@@
     \end{tabular}
-    \label{table:pihm}
 

### 2015-08-26 docs/parasite_mortality_manuscript.tex
Made word versions for all the documents

@@
 non-linear survival curves produce severely biased estimates of the survival
-function. Give the interaction between all of these different factors, the
+function. Given the interaction between all of these different factors, the
 Likelihood Method is probably limited to detecting PIHM in systems where greater than 100 hosts can be collected, parasites are

### 2015-09-04 docs/other_docs/manuscript_comments/parasite_mortality_manuscript_current.tex
Rearranged format of directories

--- 2015-05-12 docs/parasite_mortatlity_manuscript.tex
@@
 
-\abstract{
+\section*{Abstract}
 
 Parasites can often have significant impacts on host populations through parasite-induced host mortality.  Detecting and quantifying the magnitude of parasite-
-induced mortality, particularly for macroparasites where pathology is
+induced mortality, particularly for macroparasites in which pathology is
 linked to to the intensity of infection, can provide important insight into
-parasite transmission and host population management. This is, however,
+parasite dynamics and host population management. This is, however,
 notoriously difficult in practice as the only data often available are measures of parasite intensities. In this study we provide a novel method for detecting and
@@
 currently available.  We also show that this method provides quantitative
-assessments of parasite-induced mortality in empirical data that are consist
-with previously published qualitative assessments. However, we stress that this
+estimates of parasite-induced mortality in empirical data that are consist
+with previously published qualitative estimates. However, we stress that this
 method, along with all methods for estimating parasite-induced mortality in
-host populations, have a number critical assumptions that limit their
+host populations for intensity data alone, has a number critical assumptions that limit their
 applicability in the real world.  We conclude that methods for estimating and
-detecting parasite-induced mortality from intensity data should only be used as an
+detecting parasite-induced mortality from intensity data should be used only as an
 exploratory tool for informing more rigorous studies of parasite-induced host
-mortality.}
+mortality.
 
@@
 regulates host and parasite populations, making predictions about disease
-transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies (Coyne et al. 1989) [get citation], bovine TB \citep{Cox2005}, and
+transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of microparasite pathogens, such as rabies \citep{Coyne1989}, bovine TB \citep{Cox2005}, and
 rinderpest \citep{Tille1991}, is typically quantified based on the presence or absence of
@@
 
-Here we propose a novel, likelihood-based method (henceforth Likelihood Method) that does not
-require binning or data alteration, reduces the number of
-parameters to be estimated, and  uses standard statistical techniques to determine PIHM significance. We provide Python code for implementing the Likelihood Method in \emph{SI} 4.
+Here we propose a novel, likelihood-based method (henceforth Likelihood Method)
+that does not require binning or data alteration, reduces the number of
+parameters to be estimated, is highly generalizable, and uses standard
+statistical techniques to determine PIHM significance.  The Likelihood Method
+begins with the same assumptions as the Adjei Method: namely that infection of
+a host, parasite-induced mortality of a host, and the sampling of a host
+population occur at distinct time intervals during a host's life. As discussed by
+\citeauthor{Adjei1986}, this is not necessarily unrealistic as studies have
+shown that infection and host mortality are often age and/or body size dependent [citations].
 
-As with all previously proposed methods for estimating PIHM, the
-Likelihood Method first assumes that prior to mortality the parasite distribution is described by a negative binomial $g(x; \mu_p, k_p)$, where $\mu_p$ and $k_p$ are the mean parasite intensity and aggregation before mortality, respectively (smaller $k_p$ indicates more aggregation).
-Previous methods required calculating the total number of hosts before mortality ($N_p$) \citep{Crofton1971a,Adjei1986}, however this parameter is not needed in the Likelihood Method.
+The Likelihood Method then assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \boldsymbol{\phi})$, which specifies the probability of a host having $x$ parasites when it is observed.
+$\boldsymbol{\phi}$ is a vector of parameters that describes the shape of this distribution.
+The method then assumes that the probability of a host surviving with $x$ parasites from infection until sampling is given by $h(\text{survival} ; x, \boldsymbol{\theta})$ where $\boldsymbol{\theta}$ specifies any additional parameters needed to define the host survival function.
 
-The Likelihood Method then assumes that the host survival function, which specifies the probability of a host surviving with $x$ parasites, follows the logistic curve given by
-
-\begin{equation}
-    h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
-    \label{eq:logistic}
-\end{equation}
-With these two explicit assumptions, the Likelihood Method estimates four parameters: $\mu_p$, $k_p$, $a$, and $b$ by  first defining a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.
+With these two assumptions, we can define a distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival, $P(x | \text{survival})$.
 Using standard rules of conditional probability this distribution can be written as
@@
 
-One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$.
+$P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \boldsymbol{\theta})$, $P(x)$ is the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})$.
 Therefore equation \ref{eq:concept} can be written as
@@
 \begin{equation}
-    P(x | \text{survival}) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
+    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}{\sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}
     \label{eq:dist}
@@
 
-Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset.
-The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.
-All parameters are defined in Table \ref{tab:params}.
+% \begin{equation}
+%     h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+%     \label{eq:logistic}
+% \end{equation}
 
-To estimate the significance of PIHM in a host-parasite system, a
-likelihood ratio test is used in which the full model is given by equation
-\ref{eq:dist} and the reduced model is given by a negative binomial
-distribution.  If PIHM is not significant in the system, the resulting likelihood
-ratio statistic should approximately follow a $\chi^2$ distribution with two degrees of freedom.
+Using this probability distribution, one can then find the parameters $\boldsymbol{\theta}$
+and $\boldsymbol{\phi}$ that maximize the likelihood of an observed host-parasite dataset.
+To estimate the significance of PIHM in a host-parasite system, a likelihood
+ratio test can be used in which the full model is given by equation
+\ref{eq:dist} and the reduced model is given by the pre-mortality distribution
+$g(x; \boldsymbol{\phi})$.  If PIHM is not significant in the system, the resulting
+likelihood ratio statistic should approximately follow a $\chi^2$ distribution
+with degrees of freedom equal to the number of parameters in the full model with parasite-induced mortality minus the number of parameters in the reduced model without parasite-induced mortality [citation].
+
+Equation \ref{eq:dist} could be parameterized in many different ways depending
+on the parasite system of interest. In this study, we adopt the typical assumption that the pre-
+mortality parasite distribution $g(x; \boldsymbol{\phi})$ follows a negative binomial
+distribution with the parameters mean parasite intensity ($\mu_p$) and
+aggregation ($k_p$) before mortality (smaller $k_p$ indicates
+more aggregation) \citep{Crofton1971a,AndersonandMay1978,Adjei1986}.   The negative binomial distribution can arise as the equilibrium parasite
+distribution under a variety of different biological and statistical assumptions
+\citep{Kendall1948a, Boswell1970, Calabrese2011}. However, it is also an incredibly flexible
+distribution that fits many host-parasite systems regardless of whether the
+underlying mechanisms lead to an exact negative binomial distribution
+\citep{Shaw1998}.
+
+Choosing an appropriate function for $h(\text{survival}; x, \boldsymbol{\theta})$ will depend on the system under consideration.  Many theoretical models of
+parasite-induced host mortality assume that the parasite-induced death rate of
+hosts is a linear function of parasite intensity
+\citep{AndersonandMay1978,Dobson1992,Barbour2000}.
+It has been previously noted that parasite-induced mortality can be nearly impossible to detect from intensity data when the host survival function is a linear function of parasite intensity as the post-mortality distribution will be of a similar form as the pre-mortality distribution \citep{Lanciani1989}.
+That being said there is substantial empirical evidence for non-linear host-survival functions  \citep{Benesh2011}.
+
+As one of the goals of this study is to compare this new
+Likelihood Method to the previously proposed Adjei Method, we adopt the host-
+survival function used in their study and assume host-survival is non-linear and follows a logistic function given by
+
+\begin{equation}
+    h(\text{survival}; x, a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
+    \label{eq:logistic}
+\end{equation}
+where $b / 4$ determines the maximum rate of decline of
+host survival probability with increasing parasite load, analogous to the pathogenicity parameter $\alpha$ in traditional macroparasite models \citep{AndersonandMay1978}.
+When $b$ is held constant, for every one unit increase in $a$ the parasite intensity at which 99\% of hosts survive increases by $ 1/ b$.
+The equation $\exp(a
+/ b)$ can also be used to calculate the parasite $LD_{50}$, here defined as the
+infection intensity at a host has a 50\% probability of dying.
+This function is commonly used in toxicology and survival analysis and has the useful properties of being bounded between 0 and 1 and being differentiable for all $x$.
+That being said, it is phenomenological and there is little theoretical justification to use it rather than it tends to fit survival data.
+However, given that a goal of these analyses is to
+compare this method's results to the Adjei Method, to facilitate comparison it is natural
+to adopt the same host-survival function.  When
+applying the likelihood method to other systems, other, more mechanistic host-survival functions can be used in place of equation \ref{eq:logistic}.
 
@@
 We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters.
-First, we created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
-negative binomial distribution with parameters $\mu_p$ and $k_p$. Second, we chose values of $a$ and $b$ and calculated the probability of survival
+Consistent with the assumptions of the model that parasite infection, mortality, and sampling occur at distinct life stages of the host, we first created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
+negative binomial distribution with parameters $\mu_p$ and $k_p$.
+This is equivalent to the period of hosts becoming infected with parasites given in \cite{Adjei1986}.
+In the Adjei Method and Crofton Method, $N_p$ is a necessary parameter that is defined as the number of hosts in the population before parasite-induced mortality.
+A more appropriate way to define this parameter is the number of hosts that would have been sampled had parasite-induced host mortality not occurred.
+This parameter is not necessary when using the Likelihood Method because unlike the Adjei Method and Crofton Method which estimate parasite-induced mortality using absolute numbers of hosts, the Likelihood Method estimates parasite-induced mortality using probabilities.
+However, to compare the results of the Likelihood Method with the Adjei Method, we specified a value for $N_p$ for all simulations.
+
+Second, we chose values of $a$ and $b$ for the host survival function and calculated the probability of survival
 for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, for each host, we drew a random number from a uniform distribution
@@
 $N_p$ is not technically the sample size on which the methods are being
-tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
-dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.\\
+tested for the post-mortality data because PIHM reduces $N_p$ for each simulated
+dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.
 
-\noindent
-\emph{Question 2: Can we estimate fatal parasite intensity and the host survival function?}
-
-To compare the ability of the Adjei Method and the Likelihood Method to recover the $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above.
-For each parameter
-combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
-precision \citep{Walther2005} for these estimates over pre-mortality host population sizes of $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
-10000].
-We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.
-Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.
-
-\subsection*{Efficacy of the Likelihood Method with unknown pre-mortality parameters}
-
-In the final simulation, we tested the ability of the Likelihood Method to
+In the next simulation, we tested the ability of only the Likelihood Method to
 correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality
-parameters are unknown. The previous simulations showed that the Likelihood
-Method effectively identified PIHM when $\mu_p$ and $k_p$ were known with
-values of 10 and 1, respectively.  As a best-case scenario, we simulated host-
-parasite systems with these pre-mortality parameters and tested the power of
+parameters are unknown.  As a best-case scenario, we simulated host-
+parasite systems with $\mu_p = 10$ and $k = 1, 0.5, \text{and} 0.1$ [re-run for 0.5 and 0.1] as it is easier to detect PIHM from small samples sizes when mean parasite intensity is low. We then used
 the Likelihood Method to identify PIHM for gradual, moderate and steep survival
@@
 perform 500 simulations over a range of different samples sizes following the
-simulation procedure described above.
+simulation procedure described above. \\
+
+\noindent
+\emph{Question 2: Can we estimate properties of the host survival function?}
+
+In the previous section we compared the ability of the Adjei Method and the Likelihood Method to correctly identify whether or not PIHM was a occurring in a system (i.e.
+a yes or no answer).
+In this section we compare the ability of the Adjei Method and the Likelihood Method to estimate properties of the survival function such as the parameters $a$, $b$ and $LD_{50}$.
+To do this, we used the same simulation procedure and parameter combinations described above.
+For each parameter
+combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
+precision \citep{Walther2005} for these estimates.
+Because estimating properties of the host survival function requires more information than simply detecting PIHM, we used larger values of $N_p$ for this simulation ($N_p$ = [300, 500, 1000, 2000, 5000, 7500,
+10000]).
+We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.
+Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.
 

### 2017-09-12 docs/parasite_mortality_manuscript.tex
Formatted figures for dissertation

@@
 
-\textbf{Abstract}
+\section{Abstract}
 
@@
 
-\textbf{1 Introduction}
+\section{Introduction}
 
@@
 
-\textbf{2 Materials and methods}
+\section{Materials and methods}
 
-\textbf{2.1 A novel, likelihood-based method for estimating PIHM}
+\subsection{A novel, likelihood-based method for estimating PIHM}
 
@@
 
-\textbf{2.2 Evaluating the Adjei and Likelihood Methods}
+\subsection{Evaluating the Adjei and Likelihood Methods}
 
+\noindent
 \emph{Question 1: Can we detect PIHM?}
@@
 
-\textbf{2.3 Application to real data}
+\subsection{Application to real data}
 
@@
 
-\textbf{3 Results}
+\section{Results}
 
-\textbf{3.1 Question 1: Detecting presence of PIHM}
+\subsection{Question 1: Detecting presence of PIHM}
 
@@
 
-\textbf{3.2 Question 2: Estimating the $LD_{50}$ and survival function}
+\subsection{Question 2: Estimating the $LD_{50}$ and survival function}
 
@@
 
-\textbf{3.3 Application to real data}
+\subsection{Application to real data}
 
@@
 
-\textbf{4 Discussion}
+\section{Discussion}
 
@@
 
-\textbf{Acknowledgments}
+\section{Acknowledgments}
 
@@
 
-\newpage
+\clearpage
 
@@
 
-\textbf{Figure 1}: A schematic representation of the iterative approach of the Crofton Method.
+\begin{figure}
+    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/completed_projects/parasite_mortality/results/concept_crofton_method}
+    \caption{A schematic representation of the iterative approach of the Crofton Method.
 (A) The light gray shows the pre-mortality distribution that the Crofton Method is trying to estimate from the dark grey post-mortality distribution.
 The Crofton Method proceeds by truncating the post-mortality data at different levels ($t_i$, e.g.
-$i= 0, \dots, 5$) and finding the pre-mortality host population size ($N_p$), pre-mortality mean parasite intensity ($\mu_p$), and pre-mortality parasite aggregation ($k_p$) that best fit the truncated data.
+$i= 0, \dots, 6$) and finding the pre-mortality host population size ($N_p$), pre-mortality mean parasite intensity ($\mu_p$), and pre-mortality parasite aggregation ($k_p$) that best fit the truncated data.
 (B) The parameter $N_p$ is then plotted against the truncation level $t_i$ to determine if a ``kink'' occurs in the parameter values  \citep{Lester1984}.
 This ``kink'' indicates that PIHM is occurring in the system.
-In the above example, PIHM is occurring in the system as visualized by the distinct ``kink'' at $t_4$.
+In the above example, PIHM is occurring in the system as visualized by the distinct ``kink'' at $t_4$.}
+\end{figure}
 
-\textbf{Figure 2}: The simulation results comparing the power and the Type I error of the Adjei Method and the Likelihood Method across a range of different sample sizes.
+\begin{figure}
+\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/completed_projects/parasite_mortality/docs/figures_tables/figure2_IJPara15_261.pdf}
+\caption{The simulation results comparing the power and the Type I error of the Adjei Method and the Likelihood Method across a range of different sample sizes.
 (A) Five potential shapes for a host-survival functions.
@@
 See the Supplementary Fig.
-S1-S3 for Type I error and power results for all parameter combinations.
+S1-S3 for Type I error and power results for all parameter combinations.}
+\end{figure}
 
@@
 
-\textbf{Figure 3}: The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
+\begin{figure}
+    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/completed_projects/parasite_mortality/docs/figures_tables/figure3_IJPara15_261.pdf}
+    \caption{The power of the Likelihood Method (Like.) to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated.
 The curves were generated from 500 simulations for 10 pre-mortality sample sizes, $N_p$.
 The vertical, dotted-dashed lines indicate the sample size at which the power for the Likelihood Method with steep and moderate survival functions is 0.8 (75 hosts for steep functions and 408 for moderate functions).
-The Likelihood Method with a gradual survival function never has a power above 0.8.
+The Likelihood Method with a gradual survival function never has a power above 0.8.}
+\end{figure}
 
@@
 
-\textbf{Figure 4}: Bias and precision (coefficient of variation) for the Likelihood Method (Like.) and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
+\begin{figure}
+\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/completed_projects/parasite_mortality/docs/figures_tables/figure4_IJPara15_261.pdf}
+\caption{Bias and precision (coefficient of variation) for the Likelihood Method (Like.) and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes.
 As the coefficient of variation increases, precision decreases.
@@
 Bias and precision results of $LD_{50}$ and $a$ for all other parameter combinations can be found in Supplementary Fig.
-S4-S9.
+S4-S9.}
+\end{figure}
 
@@
 
-\textbf{Figure S1}: The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$.
+% \textbf{Figure S1}: The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$.
 The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
@@
 
-%\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu50}
+% %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu50}
 
-\textbf{Figure S2}: The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$.
+% \textbf{Figure S2}: The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$.
 The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
@@
 
-%\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu100}
+% %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/typeIpower_figure_for_mu100}
 
-\textbf{Figure S3}: The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$.
+% \textbf{Figure S3}: The type I error rate and the power of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$.
 The first column gives the type I error rate of each method for falsely detecting PIHM when none is present.
@@
 
-%\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu10}
+% %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu10}
 
-\textbf{Figure S4}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+% \textbf{Figure S4}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
 The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
@@
 
-%\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu50}
+% %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu50}
 
-\textbf{Figure S5}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+% \textbf{Figure S5}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
 The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
@@
 
-%\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu100}
+% %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_ld50_mu100}
 
-\textbf{Figure S6}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
+% \textbf{Figure S6}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating $LD_{50}$.
 The first column gives the bias of each method's $LD_{50}$ estimate over 150 simulations.
@@
 
-%\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu10}
+% %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu10}
 
-\textbf{Figure S7}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+% \textbf{Figure S7}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 10$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
 The first column gives the bias of each method's $a$ estimate over 150 simulations.
@@
 
-%\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu50}
+% %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu50}
 
-\textbf{Figure S8}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+% \textbf{Figure S8}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 50$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
 The first column gives the bias of each method's $a$ estimate over 150 simulations.
@@
 
-%\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu100}
+% %\includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/bais_prec_figure_for_a_mu100}
 
-\textbf{Figure S9}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
+% \textbf{Figure S9}: The bias and the precision of the Likelihood Method (black lines) and the Adjei Method (green lines) when $\mu_p = 100$ for various shapes of the host survival function and levels of aggregation $k_p$ when estimating the $a$ parameter of the host survival function.
 The first column gives the bias of each method's $a$ estimate over 150 simulations.
@@
 
-%\includegraphics[width=0.5\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/compare_crofton.pdf}
+% %\includegraphics[width=0.5\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/compare_crofton.pdf}
 
-\textbf{Figure S10}: A comparison of this paper's implementation (solid lines, circles) of the Crofton Method with the results given in \cite{Crofton1971a} (dashed lines, diamonds).
+% \textbf{Figure S10}: A comparison of this paper's implementation (solid lines, circles) of the Crofton Method with the results given in \cite{Crofton1971a} (dashed lines, diamonds).
 (A) compares the predicted number of hosts in a population pre-mortality ($N_p$).
